### 2023-11-10 paper.qmd
first commit

---
title: "A Short Demo Article: Regression Models for Count Data in R"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Achim Zeileis
    affiliations:
      - name: Universität Innsbruck
        department: Department of Statistics
        address: Universitätsstr. 15
        city: Innsbruck
        country: Austria
        postal-code: 6020
      - Journal of Statistical Software
    orcid: 0000-0003-0918-3766
    email: Achim.Zeileis@R-project.org
    url: https://www.zeileis.org/
  - name: Second  Author
    affiliations:
      - Plus Affiliation
abstract: |
  This short article illustrates how to write a manuscript for the
  _Journal of Statistical Software_ (JSS) using its {{< latex >}} style files.
  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
  it is recommended to keep the {{< latex >}} code as simple as possible,
  i.e., avoid inclusion of packages/commands that are not necessary.
  For outlining the typical structure of a JSS article some brief text snippets are employed that have been inspired by @ZeileisKleiberJackman2008, discussing count data regression in [R]{.proglang}. Editorial comments and instructions are marked by vertical bars.

keywords: [JSS, style guide, comma-separated, not capitalized, R]
keywords-formatted: [JSS, style guide, comma-separated, not capitalized, "[R]{.proglang}"]

bibliography: bibliography.bib  
---

## Introduction: Count data regression in R {#sec-intro}

::: callout
The introduction is in principle "as usual". However, it should usually embed both the implemented *methods* and the *software* into the respective relevant literature. For the latter both competing and complementary software should be discussed (within the same software environment and beyond), bringing out relative (dis)advantages. All software mentioned should be properly `@cited`'d. (See also [Using BibTeX] for more details on {{< bibtex >}}.)

For writing about software JSS requires authors to use the markup `[]{.proglang}` (programming languages and large programmable systems), `[]{.pkg}` (software packages), back ticks like \`code\` for code (functions, commands, arguments, etc.).

If there is such markup in (sub)section titles (as above), a plain text version has to be provided in the {{< latex >}} command as well. Below we also illustrate how abbrevations should be introduced and citation commands can be employed. See the {{< latex >}} code for more details.
:::

Modeling count variables is a common task in economics and the social sciences. The classical Poisson regression model for count data is often of limited use in these disciplines because empirical count data sets typically exhibit overdispersion and/or an excess number of zeros. The former issue can be addressed by extending the plain Poisson regression model in various directions: e.g., using sandwich covariances or estimating an additional dispersion parameter (in a so-called quasi-Poisson model). Another more formal way is to use a negative binomial (NB) regression. All of these models belong to the family of generalized linear models (GLMs). However, although these models typically can capture overdispersion rather well, they are in many applications not sufficient for modeling excess zeros. Since @Mullahy1986 there is increased interest in zero-augmented models that address this issue by a second model component capturing zero counts. An overview of count data models in econometrics, including hurdle and zero-inflated models, is provided in @CameronTrivedi2013.

In [R]{.proglang} @R, GLMs are provided by the model fitting functions [glm]{.fct} in the [stats]{.pkg} package and [glm.nb]{.fct} in the [MASS]{.pkg} package [@VenablesRipley2002] along with associated methods for diagnostics and inference. The manuscript that this document is based on [@ZeileisKleiberJackman2008] then introduced hurdle and zero-inflated count models in the functions [hurdle]{.fct} and [zeroinfl]{.fct} in the [pcsl]{.pkg} package @Jackman2015. Of course, much more software could be discussed here, including (but not limited to) generalized additive models for count data as available in the [R]{.proglang} packages [mgcv]{.pkg} @Wood2006, [gamlss]{.pkg} @StasinopoulosRigby2007, or [VGAM]{.pkg} @Yee2009.

## Models and software {#sec-models}

The basic Poisson regression model for count data is a special case of the GLM framework @McCullaghNelder1989. It describes the dependence of a count response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution $y_i \sim \mathrm{Pois}(\mu_i)$. The dependence of the conditional mean $E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a log link and a linear predictor

$$
\log(\mu_i) \quad = \quad x_i^\top \beta,
$$ {#eq-mean}

where the regression coefficients $\beta$ are estimated by maximum likelihood (ML) using the iterative weighted least squares (IWLS) algorithm.

::: callout
TODO: Note that around the equation above there should be no spaces (avoided in the {{< latex >}} code by `%` lines) so that "normal" spacing is used and not a new paragraph started.
:::

[R]{.proglang} provides a very flexible implementation of the general GLM framework in the function [glm]{.fct} @ChambersHastie1992 in the [stats]{.pkg} package. Its most important arguments are

```r
glm(formula, data, subset, na.action, weights, offset,
  family = gaussian, start = NULL, control = glm.control(…),
  model = TRUE, y = TRUE, x = FALSE, …)
```

where `formula` plus `data` is the now standard way of specifying regression relationships in [R]{.proglang}/[S]{.proglang} introduced in @ChambersHastie1992. The remaining arguments in the first line (`subset`, `na.action`, `weights`, and `offset`) are also standard for setting up formula-based regression models in [R]{.proglang}/[S]{.proglang}. The arguments in the second line control aspects specific to GLMs while the arguments in the last line specify which components are returned in the fitted model object (of class [glm]{.class} which inherits from [lm]{.class}). For further arguments to [glm]{.fct} (including alternative specifications of starting values) see `?glm`. For estimating a Poisson model `family = poisson` has to be specified.

::: callout
As the synopsis above is a code listing that is not meant to be executed, one can use either the dedicated `{Code}` environment or a simple `{verbatim}` environment for this. Again, spaces before and after should be avoided.

Finally, there might be a reference to a `{table}` such as @tbl-overview. Usually, these are placed at the top of the page (`[t!]`), centered (`\centering`), with a caption below the table, column headers and captions in sentence style, and if possible avoiding vertical lines.
:::

| Type           | Distribution | Method   | Description                                                                                                                                                                                  |
|---------|---------|---------|-----------------------------------------------|
| GLM            | Poisson      | ML       | Poisson regression: classical GLM, estimated by maximum likelihood (ML)                                                                                                                      |
|                |              | Quasi    | "Quasi-Poisson regression'': same mean function, estimated by quasi-ML (QML) or equivalently generalized estimating equations (GEE), inference adjustment via estimated dispersion parameter |
|                |              | Adjusted | "Adjusted Poisson regression'': same mean function, estimated by QML/GEE, inference adjustment via sandwich covariances                                                                      |
|                | NB           | ML       | NB regression: extended GLM, estimated by ML including additional shape parameter                                                                                                            |
| Zero-augmented | Poisson      | ML       | Zero-inflated Poisson (ZIP), hurdle Poisson                                                                                                                                                  |
|                | NB           | ML       | Zero-inflated NB (ZINB), hurdle NB                                                                                                                                                           |
: Overview of various count regression models. The table is usually placed at the top of the page (`[t!]`), centered (`centering`), has a caption below the table, column headers and captions are in sentence style, and if possible vertical lines should be avoided. {#tbl-overview}

## Illustrations {#sec-illustrations}

For a simple illustration of basic Poisson and NB count regression the
`quine` data from the [MASS]{.pkg} package is used. This provides the number
of `Days` that children were absent from school in Australia in a
particular year, along with several covariates that can be employed as regressors. The data can be loaded by

```{R}
#| prompt: true
data(mtcars)
```
and a basic frequency distribution of the response variable is displayed in
@fig-quine.

:::{.callout}
For code input and output, the style files provide dedicated environments.
Either the "agnostic" `{CodeInput}` and `{CodeOutput}` can be used
or, equivalently, the environments `{Sinput}` and `{Soutput}` as
produced by [Sweave]{.fct} or [knitr]{.pkg} when using the `render_sweave()`
hook. Please make sure that all code is properly spaced, e.g., using
`y = a + b * x` and _not_ `y=a+b*x`. Moreover, code input should use "the usual" command prompt in the respective software system. For
[R]{.proglang} code, the prompt `R> ` should be used with `+  ` as
the continuation prompt. Generally, comments within the code chunks should be
avoided -- and made in the regular {{< latex >}} text instead. Finally, empty lines before and after code input/output should be avoided (see above).
:::

## Summary and discussion {#sec-summary}

:::{.callout}

As usual…

:::

## Computational details {.unnumbered}

:::{.callout}

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualizations,
maps, tables, …) that are not cited in the main text can be credited here.

:::

The results in this paper were obtained using [R]{.proglang}~3.4.1 with the
[MASS]{.pkg}~7.3.47 package. [R]{.proglang} itself and all packages used are available from the Comprehensive [R]{.proglang} Archive Network (CRAN) at
[https://CRAN.R-project.org/].

## Acknowledgments {.unnumbered}

:::{.callout}

All acknowledgments (note the AE spelling) should be collected in this
unnumbered section before the references. It may contain the usual information
about funding and feedback from colleagues/reviewers/etc. Furthermore,
information such as relative contributions of the authors may be added here
(if any).

:::

## References {.unnumbered}

:::{#refs}

:::

{{< pagebreak >}}

## More technical details {#sec-techdetails .unnumbered}

:::{.callout}

Appendices can be included after the bibliography (with a page break). Each
section within the appendix should have a proper section title (rather than
just _Appendix_).

For more technical style details, please check out JSS's style FAQ at
[https://www.jstatsoft.org/pages/view/style#frequently-asked-questions]
which includes the following topics:

- Title vs. sentence case.
- Graphics formatting.
- Naming conventions.
- Turning JSS manuscripts into [R]{.proglang} package vignettes.
- Trouble shooting.
- Many other potentially helpful details…

:::

## Using BibTeX {#sec-bibtex .unnumbered}

:::{.callout}

References need to be provided in a {{< bibtex >}} file (`.bib`). All
references should be made with `@cite` syntax. This commands yield different
formats of author-year citations and allow to include additional details (e.g.,pages, chapters, \dots) in brackets. In case you are not familiar with these
commands see the JSS style FAQ for details.

Cleaning up {{< bibtex >}} files is a somewhat tedious task -- especially
when acquiring the entries automatically from mixed online sources. However,
it is important that informations are complete and presented in a consistent
style to avoid confusions. JSS requires the following format.

- item JSS-specific markup (`\proglang`, `\pkg`, `\code`) should be used in the references.
- item Titles should be in title case.
- item Journal titles should not be abbreviated and in title case.
- item DOIs should be included where available.
- item Software should be properly cited as well. For [R]{.proglang} packages `citation("pkgname")` typically provides a good starting point.

:::

### 2023-11-10 paper.tex
first commit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  article]{jss}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{orcidlink,thumbpdf,lmodern}

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}{}
\makeatother
\makeatletter
\makeatother
\makeatletter
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Short Demo Article: Regression Models for Count Data in R},
  pdfauthor={Achim Zeileis; Second Author},
  pdfkeywords={JSS, style guide, comma-separated, not capitalized, R},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% Author information
\author{Achim Zeileis~\orcidlink{0000-0003-0918-3766}\\Universität
Innsbruck \And Second Author\\Plus Affiliation}
\Plainauthor{Achim Zeileis, Second Author} %% comma-separated

\title{A Short Demo Article: Regression Models for Count Data in R}
\Plaintitle{A Short Demo Article: Regression Models for Count Data in
R} %% without formatting

%% an abstract and keywords
\Abstract{This short article illustrates how to write a manuscript for
the \emph{Journal of Statistical Software} (JSS) using its {\LaTeX}
style files. Generally, we ask to follow JSS's style guide and FAQs
precisely. Also, it is recommended to keep the {\LaTeX} code as simple
as possible, i.e., avoid inclusion of packages/commands that are not
necessary. For outlining the typical structure of a JSS article some
brief text snippets are employed that have been inspired by
\citet{ZeileisKleiberJackman2008}, discussing count data regression in
\proglang{R}. Editorial comments and instructions are marked by vertical
bars.}

%% at least one keyword must be supplied
\Keywords{JSS, style guide, comma-separated, not
capitalized, \proglang{R}}
\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}
%% \setcounter{page}{1}
%% \Pages{1--xx}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Achim Zeileis\\
Department of Statistics\\
Universitätsstr. 15\\
Innsbruck Austria\\
E-mail: \email{Achim.Zeileis@R-project.org}\\
URL: \url{https://www.zeileis.org/}\\
\\~
Second Author\\
\\~

}

\begin{document}
\maketitle
\section{Introduction: Count data regression in R}\label{sec-intro}

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

The introduction is in principle ``as usual''. However, it should
usually embed both the implemented \emph{methods} and the
\emph{software} into the respective relevant literature. For the latter
both competing and complementary software should be discussed (within
the same software environment and beyond), bringing out relative
(dis)advantages. All software mentioned should be properly
\texttt{@cited}'d.~(See also \hyperref[sec-bibtex]{Using BibTeX} for
more details on \textsc{Bib}{\TeX}.)

For writing about software JSS requires authors to use the markup
\texttt{{[}{]}\{.proglang\}} (programming languages and large
programmable systems), \texttt{{[}{]}\{.pkg\}} (software packages), back
ticks like `code` for code (functions, commands, arguments, etc.).

If there is such markup in (sub)section titles (as above), a plain text
version has to be provided in the {\LaTeX} command as well. Below we
also illustrate how abbrevations should be introduced and citation
commands can be employed. See the {\LaTeX} code for more details.

\end{tcolorbox}

Modeling count variables is a common task in economics and the social
sciences. The classical Poisson regression model for count data is often
of limited use in these disciplines because empirical count data sets
typically exhibit overdispersion and/or an excess number of zeros. The
former issue can be addressed by extending the plain Poisson regression
model in various directions: e.g., using sandwich covariances or
estimating an additional dispersion parameter (in a so-called
quasi-Poisson model). Another more formal way is to use a negative
binomial (NB) regression. All of these models belong to the family of
generalized linear models (GLMs). However, although these models
typically can capture overdispersion rather well, they are in many
applications not sufficient for modeling excess zeros. Since
\citet{Mullahy1986} there is increased interest in zero-augmented models
that address this issue by a second model component capturing zero
counts. An overview of count data models in econometrics, including
hurdle and zero-inflated models, is provided in
\citet{CameronTrivedi2013}.

In \proglang{R} \citet{R}, GLMs are provided by the model fitting
functions \fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the
\pkg{MASS} package \citep{VenablesRipley2002} along with associated
methods for diagnostics and inference. The manuscript that this document
is based on \citep{ZeileisKleiberJackman2008} then introduced hurdle and
zero-inflated count models in the functions \fct{hurdle} and
\fct{zeroinfl} in the \pkg{pcsl} package \citet{Jackman2015}. Of course,
much more software could be discussed here, including (but not limited
to) generalized additive models for count data as available in the
\proglang{R} packages \pkg{mgcv} \citet{Wood2006}, \pkg{gamlss}
\citet{StasinopoulosRigby2007}, or \pkg{VGAM} \citet{Yee2009}.

\section{Models and software}\label{sec-models}

The basic Poisson regression model for count data is a special case of
the GLM framework \citet{McCullaghNelder1989}. It describes the
dependence of a count response variable \(y_i\) (\(i = 1, \dots, n\)) by
assuming a Poisson distribution \(y_i \sim \mathrm{Pois}(\mu_i)\). The
dependence of the conditional mean \(E[y_i \, | \, x_i] = \mu_i\) on the
regressors \(x_i\) is then specified via a log link and a linear
predictor

\begin{equation}\phantomsection\label{eq-mean}{
\log(\mu_i) \quad = \quad x_i^\top \beta,
}\end{equation}

where the regression coefficients \(\beta\) are estimated by maximum
likelihood (ML) using the iterative weighted least squares (IWLS)
algorithm.

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

TODO: Note that around the equation above there should be no spaces
(avoided in the {\LaTeX} code by \texttt{\%} lines) so that ``normal''
spacing is used and not a new paragraph started.

\end{tcolorbox}

\proglang{R} provides a very flexible implementation of the general GLM
framework in the function \fct{glm} \citet{ChambersHastie1992} in the
\pkg{stats} package. Its most important arguments are

\begin{verbatim}
glm(formula, data, subset, na.action, weights, offset,
  family = gaussian, start = NULL, control = glm.control(…),
  model = TRUE, y = TRUE, x = FALSE, …)
\end{verbatim}

where \texttt{formula} plus \texttt{data} is the now standard way of
specifying regression relationships in \proglang{R}/\proglang{S}
introduced in \citet{ChambersHastie1992}. The remaining arguments in the
first line (\texttt{subset}, \texttt{na.action}, \texttt{weights}, and
\texttt{offset}) are also standard for setting up formula-based
regression models in \proglang{R}/\proglang{S}. The arguments in the
second line control aspects specific to GLMs while the arguments in the
last line specify which components are returned in the fitted model
object (of class \class{glm} which inherits from \class{lm}). For
further arguments to \fct{glm} (including alternative specifications of
starting values) see \texttt{?glm}. For estimating a Poisson model
\texttt{family\ =\ poisson} has to be specified.

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

As the synopsis above is a code listing that is not meant to be
executed, one can use either the dedicated \texttt{\{Code\}} environment
or a simple \texttt{\{verbatim\}} environment for this. Again, spaces
before and after should be avoided.

Finally, there might be a reference to a \texttt{\{table\}} such as
Table~\ref{tbl-overview}. Usually, these are placed at the top of the
page (\texttt{{[}t!{]}}), centered (\texttt{\textbackslash{}centering}),
with a caption below the table, column headers and captions in sentence
style, and if possible avoiding vertical lines.

\end{tcolorbox}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.6351}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Distribution
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Distribution
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
GLM & Poisson & ML & Poisson regression: classical GLM, estimated by
maximum likelihood (ML) \\
& & Quasi & ``Quasi-Poisson regression'\,': same mean function,
estimated by quasi-ML (QML) or equivalently generalized estimating
equations (GEE), inference adjustment via estimated dispersion
parameter \\
& & Adjusted & ``Adjusted Poisson regression'\,': same mean function,
estimated by QML/GEE, inference adjustment via sandwich covariances \\
& NB & ML & NB regression: extended GLM, estimated by ML including
additional shape parameter \\
Zero-augmented & Poisson & ML & Zero-inflated Poisson (ZIP), hurdle
Poisson \\
& NB & ML & Zero-inflated NB (ZINB), hurdle NB \\
\caption{Overview of various count regression models. The table is
usually placed at the top of the page (\texttt{{[}t!{]}}), centered
(\texttt{centering}), has a caption below the table, column headers and
captions are in sentence style, and if possible vertical lines should be
avoided.}\label{tbl-overview}\tabularnewline
\end{longtable}

\section{Illustrations}\label{sec-illustrations}

For a simple illustration of basic Poisson and NB count regression the
\texttt{quine} data from the \pkg{MASS} package is used. This provides
the number of \texttt{Days} that children were absent from school in
Australia in a particular year, along with several covariates that can
be employed as regressors. The data can be loaded by

\begin{verbatim}
R> data(mtcars)
\end{verbatim}

and a basic frequency distribution of the response variable is displayed
in \textbf{?@fig-quine}.

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

For code input and output, the style files provide dedicated
environments. Either the ``agnostic'' \texttt{\{CodeInput\}} and
\texttt{\{CodeOutput\}} can be used or, equivalently, the environments
\texttt{\{Sinput\}} and \texttt{\{Soutput\}} as produced by \fct{Sweave}
or \pkg{knitr} when using the \texttt{render\_sweave()} hook. Please
make sure that all code is properly spaced, e.g., using
\texttt{y\ =\ a\ +\ b\ *\ x} and \emph{not} \texttt{y=a+b*x}. Moreover,
code input should use ``the usual'' command prompt in the respective
software system. For \proglang{R} code, the prompt
\texttt{R\textgreater{}} should be used with \texttt{+} as the
continuation prompt. Generally, comments within the code chunks should
be avoided -- and made in the regular {\LaTeX} text instead. Finally,
empty lines before and after code input/output should be avoided (see
above).

\end{tcolorbox}

\section{Summary and discussion}\label{sec-summary}

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

As usual\ldots{}

\end{tcolorbox}

\section*{Computational details}\label{computational-details}
\addcontentsline{toc}{section}{Computational details}

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be
included in an unnumbered section. Also, auxiliary packages (say, for
visualizations, maps, tables, \ldots) that are not cited in the main
text can be credited here.

\end{tcolorbox}

The results in this paper were obtained using
\proglang{R}\textasciitilde3.4.1 with the
\pkg{MASS}\textasciitilde7.3.47 package. \proglang{R} itself and all
packages used are available from the Comprehensive \proglang{R} Archive
Network (CRAN) at {[}https://CRAN.R-project.org/{]}.

\section*{Acknowledgments}\label{acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

All acknowledgments (note the AE spelling) should be collected in this
unnumbered section before the references. It may contain the usual
information about funding and feedback from colleagues/reviewers/etc.
Furthermore, information such as relative contributions of the authors
may be added here (if any).

\end{tcolorbox}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\renewcommand{\bibsection}{}
\bibliography{bibliography.bib}

\newpage{}

\section*{More technical details}\label{sec-techdetails}
\addcontentsline{toc}{section}{More technical details}

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

Appendices can be included after the bibliography (with a page break).
Each section within the appendix should have a proper section title
(rather than just \emph{Appendix}).

For more technical style details, please check out JSS's style FAQ at
{[}https://www.jstatsoft.org/pages/view/style\#frequently-asked-questions{]}
which includes the following topics:

\begin{itemize}
\tightlist
\item
  Title vs.~sentence case.
\item
  Graphics formatting.
\item
  Naming conventions.
\item
  Turning JSS manuscripts into \proglang{R} package vignettes.
\item
  Trouble shooting.
\item
  Many other potentially helpful details\ldots{}
\end{itemize}

\end{tcolorbox}

\section*{Using BibTeX}\label{sec-bibtex}
\addcontentsline{toc}{section}{Using BibTeX}

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]

References need to be provided in a \textsc{Bib}{\TeX} file
(\texttt{.bib}). All references should be made with \texttt{@cite}
syntax. This commands yield different formats of author-year citations
and allow to include additional details (e.g.,pages, chapters, \dots) in
brackets. In case you are not familiar with these commands see the JSS
style FAQ for details.

Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task --
especially when acquiring the entries automatically from mixed online
sources. However, it is important that informations are complete and
presented in a consistent style to avoid confusions. JSS requires the
following format.

\begin{itemize}
\tightlist
\item
  item JSS-specific markup (\texttt{\textbackslash{}proglang},
  \texttt{\textbackslash{}pkg}, \texttt{\textbackslash{}code}) should be
  used in the references.
\item
  item Titles should be in title case.
\item
  item Journal titles should not be abbreviated and in title case.
\item
  item DOIs should be included where available.
\item
  item Software should be properly cited as well. For \proglang{R}
  packages \texttt{citation("pkgname")} typically provides a good
  starting point.
\end{itemize}

\end{tcolorbox}

\end{document}

### 2023-11-10 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
 \makeatother
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacityback=0, bottomrule=.15mm, left=2mm, arc=.35mm, rightrule=.15mm, colback=white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, breakable, rightrule=.15mm, colback=white, toprule=.15mm, leftrule=.75mm, opacityback=0, bottomrule=.15mm, left=2mm]
 

### 2023-11-11 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 \makeatother
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, breakable, rightrule=.15mm, colback=white, toprule=.15mm, leftrule=.75mm, opacityback=0, bottomrule=.15mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, breakable, bottomrule=.15mm, rightrule=.15mm, colback=white, left=2mm, arc=.35mm, opacityback=0, leftrule=.75mm]
 

### 2023-11-11 paper.qmd
Update paper.qmd

@@
 ---
-title: "A Short Demo Article: Regression Models for Count Data in R"
+title: "Reproducible development environments with rix"
 format:
@@
 author:
-  - name: Achim Zeileis
+  - name: Bruno Rodrigues 
     affiliations:
-      - name: Universität Innsbruck
+      - name: Ministry of higher education and research, Luxembourg 
         department: Department of Statistics
-        address: Universitätsstr. 15
-        city: Innsbruck
-        country: Austria
-        postal-code: 6020
+        address: 18, Montée de la Pétrusse
+        city: Luxembourg 
+        country: Luxembourg 
+        postal-code: 2327
       - Journal of Statistical Software
-    orcid: 0000-0003-0918-3766
-    email: Achim.Zeileis@R-project.org
-    url: https://www.zeileis.org/
-  - name: Second  Author
+    orcid: 0000-0002-3211-3689
+    email: bruno@brodrigues.co
+    url: https://www.brodrigues.co 
+  - name: Philipp Baumann
     affiliations:

### 2023-11-11 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 \makeatother
@@
 \hypersetup{
-  pdftitle={A Short Demo Article: Regression Models for Count Data in R},
-  pdfauthor={Achim Zeileis; Second Author},
+  pdftitle={Reproducible development environments with rix},
+  pdfauthor={Bruno Rodrigues; Philipp Baumann},
   pdfkeywords={JSS, style guide, comma-separated, not capitalized, R},
@@
 %% Author information
-\author{Achim Zeileis~\orcidlink{0000-0003-0918-3766}\\Universität
-Innsbruck \And Second Author\\Plus Affiliation}
-\Plainauthor{Achim Zeileis, Second Author} %% comma-separated
+\author{Bruno Rodrigues~\orcidlink{0000-0002-3211-3689}\\Ministry of
+higher education and research, Luxembourg \And Philipp Baumann\\Plus
+Affiliation}
+\Plainauthor{Bruno Rodrigues, Philipp Baumann} %% comma-separated
 
-\title{A Short Demo Article: Regression Models for Count Data in R}
-\Plaintitle{A Short Demo Article: Regression Models for Count Data in
-R} %% without formatting
+\title{Reproducible development environments with rix}
+\Plaintitle{Reproducible development environments with
+rix} %% without formatting
 
@@
 \Address{
-Achim Zeileis\\
+Bruno Rodrigues\\
 Department of Statistics\\
-Universitätsstr. 15\\
-Innsbruck Austria\\
-E-mail: \email{Achim.Zeileis@R-project.org}\\
-URL: \url{https://www.zeileis.org/}\\
+18, Montée de la Pétrusse\\
+Luxembourg Luxembourg\\
+E-mail: \email{bruno@brodrigues.co}\\
+URL: \url{https://www.brodrigues.co}\\
 \\~
-Second Author\\
+Philipp Baumann\\
 \\~
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, breakable, bottomrule=.15mm, rightrule=.15mm, colback=white, left=2mm, arc=.35mm, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, arc=.35mm, colback=white, bottomrule=.15mm, left=2mm, toprule=.15mm, leftrule=.75mm, opacityback=0, breakable]
 

### 2023-12-10 paper.qmd
started intro

--- 2023-11-10 paper.qmd
@@
 author:
-  - name: Achim Zeileis
+  - name: Bruno Rodrigues 
     affiliations:
-      - name: Universität Innsbruck
+      - name: Ministry of Research and Higher education, Luxembourg 
         department: Department of Statistics
-        address: Universitätsstr. 15
-        city: Innsbruck
-        country: Austria
-        postal-code: 6020
+        address: 18, Montée de la Pétrusse
+        city: Luxembourg 
+        country: Luxembourg 
+        postal-code: 2327
       - Journal of Statistical Software
-    orcid: 0000-0003-0918-3766
-    email: Achim.Zeileis@R-project.org
-    url: https://www.zeileis.org/
-  - name: Second  Author
+    orcid: 0000-0002-3211-3689
+    email: bruno@brodrigues.co
+    url: https://www.brodrigues.co 
+  - name: Philipp Baumann
     affiliations:
@@
 abstract: |
-  This short article illustrates how to write a manuscript for the
-  _Journal of Statistical Software_ (JSS) using its {{< latex >}} style files.
-  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
-  it is recommended to keep the {{< latex >}} code as simple as possible,
-  i.e., avoid inclusion of packages/commands that are not necessary.
-  For outlining the typical structure of a JSS article some brief text snippets are employed that have been inspired by @ZeileisKleiberJackman2008, discussing count data regression in [R]{.proglang}.
-Editorial comments and instructions are marked by vertical bars.
+  In order to create an analysis that is easily reproducible,
+  it is not enough to write clean code and document it well. One
+  must also make sure to list all the dependencies of the analysis
+  clearly and ideally provide an easy way to install said dependencies.
+  There are several tools that can be used to list dependencies and to
+  make them easily installable by someone that wishes to reproduce a study,
+  such as Docker, a containerization solution. This paper will present
+  the Nix package manager, and an R package called {rix} that lowers
+  Nix's learning curve for users of the R programming language.
 
-keywords: [JSS, style guide, comma-separated, not capitalized, R]
-keywords-formatted: [JSS, style guide, comma-separated, not capitalized, "[R]{.proglang}"]
+keywords: [reproducibility, R, Nix]
+keywords-formatted: [reproducibility, "[R]{.proglang}", "[Nix]{.proglang}"]
 
@@
 
-## Introduction: Count data regression in R {#sec-intro}
+## Introduction: Reproducibility is also about software {#sec-intro}
 
@@
 
-Modeling count variables is a common task in economics and the social sciences.
-The classical Poisson regression model for count data is often of limited use in these disciplines because empirical count data sets typically exhibit overdispersion and/or an excess number of zeros.
-The former issue can be addressed by extending the plain Poisson regression model in various directions: e.g., using sandwich covariances or estimating an additional dispersion parameter (in a so-called quasi-Poisson model).
-Another more formal way is to use a negative binomial (NB) regression.
-All of these models belong to the family of generalized linear models (GLMs).
-However, although these models typically can capture overdispersion rather well, they are in many applications not sufficient for modeling excess zeros.
-Since @Mullahy1986 there is increased interest in zero-augmented models that address this issue by a second model component capturing zero counts.
-An overview of count data models in econometrics, including hurdle and zero-inflated models, is provided in @CameronTrivedi2013.
+@peng2011 introduced the idea of reproducibility being on a continuum: on one of
+the ends of this continuum, we only have access to the paper describing the
+studies, which is not reproducible at all. Then, if to this paper we make the
+original source code of the analysis that was written to compute the results of
+the study available, reproducibility is improved, albeit only by a little.
+Adding the original data to this package improves reproducibility yet again. Finally,
+if to all this we add what Roger Peng named the *linked and executable code and data*,
+we reach the gold standard of full replication.
 
-In [R]{.proglang} @R, GLMs are provided by the model fitting functions [glm]{.fct} in the [stats]{.pkg} package and [glm.nb]{.fct} in the [MASS]{.pkg} package [@VenablesRipley2002] along with associated methods for diagnostics and inference.
-The manuscript that this document is based on [@ZeileisKleiberJackman2008] then introduced hurdle and zero-inflated count models in the functions [hurdle]{.fct} and [zeroinfl]{.fct} in the [pcsl]{.pkg} package @Jackman2015.
-Of course, much more software could be discussed here, including (but not limited to) generalized additive models for count data as available in the [R]{.proglang} packages [mgcv]{.pkg} @Wood2006, [gamlss]{.pkg} @StasinopoulosRigby2007, or [VGAM]{.pkg} @Yee2009.
+What is this *linked and executable code and data*? Another way to name this
+crucial piece of the reproducibility puzzle is *computational environment*. The
+computational environment is all the software required to actually run the
+analysis. Here too, we can speak of a continuum. One could simply name and list
+the software used: for example, the [R]{.proglang} programming language.
+Sometimes, authors have the courtesy to also state the version of [R]{.proglang}
+used. Some authors go further, and also list the packages used, and ideally with
+their versions as well. Authors rarely state the operating system on which the
+analysis was done, even though it has been shown that running the same analysis
+with the same software but on different operating systems could lead to different
+results, as described in @neupane2019. Authors also only very rarely provide
+instructions to install the required tools and software in order to reproduce
+their studies.
+
+There are exceptions of course, a great example of a paper that provides everything
+needed to reproduce its results is @mcdermott2021. The author of this paper set up
+an accompagnying Github repository to the paper^[https://github.com/grantmcdermott/skeptic-priors]
+containing all the instructions to install the required software and then run the
+analysis. If we take a closer look at this repository, we will notice that several
+tools were used to capture the compatutational environment and make it available to
+other researchers:
+
+- First, the version of [R]{.proglang} was stated
 

### 2023-12-10 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, enhanced, breakable, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 \makeatother
@@
   pdfauthor={Bruno Rodrigues; Philipp Baumann},
-  pdfkeywords={JSS, style guide, comma-separated, not capitalized, R},
+  pdfkeywords={reproducibility, R, Nix},
   colorlinks=true,
@@
 \author{Bruno Rodrigues~\orcidlink{0000-0002-3211-3689}\\Ministry of
-higher education and research, Luxembourg \And Philipp Baumann\\Plus
+Research and Higher education, Luxembourg \And Philipp Baumann\\Plus
 Affiliation}
@@
 %% an abstract and keywords
-\Abstract{This short article illustrates how to write a manuscript for
-the \emph{Journal of Statistical Software} (JSS) using its {\LaTeX}
-style files. Generally, we ask to follow JSS's style guide and FAQs
-precisely. Also, it is recommended to keep the {\LaTeX} code as simple
-as possible, i.e., avoid inclusion of packages/commands that are not
-necessary. For outlining the typical structure of a JSS article some
-brief text snippets are employed that have been inspired by
-\citet{ZeileisKleiberJackman2008}, discussing count data regression in
-\proglang{R}. Editorial comments and instructions are marked by vertical
-bars.}
+\Abstract{In order to create an analysis that is easily reproducible, it
+is not enough to write clean code and document it well. One must also
+make sure to list all the dependencies of the analysis clearly and
+ideally provide an easy way to install said dependencies. There are
+several tools that can be used to list dependencies and to make them
+easily installable by someone that wishes to reproduce a study, such as
+Docker, a containerization solution. This paper will present the Nix
+package manager, and an R package called \{rix\} that lowers Nix's
+learning curve for users of the R programming language.}
 
 %% at least one keyword must be supplied
-\Keywords{JSS, style guide, comma-separated, not
-capitalized, \proglang{R}}
-\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}
+\Keywords{reproducibility, \proglang{R}, \proglang{Nix}}
+\Plainkeywords{reproducibility, R, Nix}
 
@@
 \maketitle
-\section{Introduction: Count data regression in R}\label{sec-intro}
+\section{Introduction: Reproducibility is also about
+software}\label{sec-intro}
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, arc=.35mm, colback=white, bottomrule=.15mm, left=2mm, toprule=.15mm, leftrule=.75mm, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colback=white, rightrule=.15mm, opacityback=0, leftrule=.75mm]
 
@@
 
-Modeling count variables is a common task in economics and the social
-sciences. The classical Poisson regression model for count data is often
-of limited use in these disciplines because empirical count data sets
-typically exhibit overdispersion and/or an excess number of zeros. The
-former issue can be addressed by extending the plain Poisson regression
-model in various directions: e.g., using sandwich covariances or
-estimating an additional dispersion parameter (in a so-called
-quasi-Poisson model). Another more formal way is to use a negative
-binomial (NB) regression. All of these models belong to the family of
-generalized linear models (GLMs). However, although these models
-typically can capture overdispersion rather well, they are in many
-applications not sufficient for modeling excess zeros. Since
-\citet{Mullahy1986} there is increased interest in zero-augmented models
-that address this issue by a second model component capturing zero
-counts. An overview of count data models in econometrics, including
-hurdle and zero-inflated models, is provided in
-\citet{CameronTrivedi2013}.
+\citet{peng2011} introduced the idea of reproducibility being on a
+continuum: on one of the ends of this continuum, we only have access to
+the paper describing the studies, which is not reproducible at all.
+Then, if to this paper we make the original source code of the analysis
+that was written to compute the results of the study available,
+reproducibility is improved, albeit only by a little. Adding the
+original data to this package improves reproducibility yet again.
+Finally, if to all this we add what Roger Peng named the \emph{linked
+and executable code and data}, we reach the gold standard of full
+replication.
 
-In \proglang{R} \citet{R}, GLMs are provided by the model fitting
-functions \fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the
-\pkg{MASS} package \citep{VenablesRipley2002} along with associated
-methods for diagnostics and inference. The manuscript that this document
-is based on \citep{ZeileisKleiberJackman2008} then introduced hurdle and
-zero-inflated count models in the functions \fct{hurdle} and
-\fct{zeroinfl} in the \pkg{pcsl} package \citet{Jackman2015}. Of course,
-much more software could be discussed here, including (but not limited
-to) generalized additive models for count data as available in the
-\proglang{R} packages \pkg{mgcv} \citet{Wood2006}, \pkg{gamlss}
-\citet{StasinopoulosRigby2007}, or \pkg{VGAM} \citet{Yee2009}.
+What is this \emph{linked and executable code and data}? Another way to
+name this crucial piece of the reproducibility puzzle is
+\emph{computational environment}. The computational environment is all
+the software required to actually run the analysis. Here too, we can
+speak of a continuum. One could simply name and list the software used:
+for example, the \proglang{R} programming language. Sometimes, authors
+have the courtesy to also state the version of \proglang{R} used. Some
+authors go further, and also list the packages used, and ideally with
+their versions as well. Authors rarely state the operating system on
+which the analysis was done, even though it has been shown that running
+the same analysis with the same software but on different operating
+systems could lead to different results, as described in
+\citet{neupane2019}. Authors also only very rarely provide instructions
+to install the required tools and software in order to reproduce their
+studies.
+
+There are exceptions of course, a great example of a paper that provides
+everything needed to reproduce its results is \citet{mcdermott2021}. The
+author of this paper set up an accompagnying Github repository to the
+paper\footnote{https://github.com/grantmcdermott/skeptic-priors}
+containing all the instructions to install the required software and
+then run the analysis. If we take a closer look at this repository, we
+will notice that several tools were used to capture the compatutational
+environment and make it available to other researchers:
+
+\begin{itemize}
+\tightlist
+\item
+  First, the version of \proglang{R} was stated
+\end{itemize}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, arc=.35mm, colback=white, bottomrule=.15mm, left=2mm, toprule=.15mm, leftrule=.75mm, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colback=white, rightrule=.15mm, opacityback=0, leftrule=.75mm]
 

### 2023-12-12 paper.qmd
more intro

@@
 the ends of this continuum, we only have access to the paper describing the
-studies, which is not reproducible at all. Then, if to this paper we make the
-original source code of the analysis that was written to compute the results of
-the study available, reproducibility is improved, albeit only by a little.
-Adding the original data to this package improves reproducibility yet again. Finally,
-if to all this we add what Roger Peng named the *linked and executable code and data*,
-we reach the gold standard of full replication.
+studies, which is not reproducible at all. Then, if in addition, to this paper
+we make the original source code of the analysis that was written to compute the
+results of the study available, reproducibility is improved, albeit only by a
+little. Adding the original data improves reproducibility yet again. Finally, if
+to all this we add what Roger Peng named the *linked and executable code and
+data*, we reach the gold standard of full replication.
 
@@
 
-- First, the version of [R]{.proglang} was stated
+- The version of [R]{.proglang} was stated;
+- Packages and their versions were listed and saved into an `renv.lock` file;
+- A `Makefile` was used to run the whole analysis and compile the paper;
+- A `Dockerfile` was used to provide the complete computational environment, including the right version of [R]{.proglang} and run the whole analysis easily.
 
-## Models and software {#sec-models}
+As we go down this list, we get closer to the gold standard of a perfectly reproducible
+study.
 
-The basic Poisson regression model for count data is a special case of the GLM framework @McCullaghNelder1989.
-It describes the dependence of a count response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution $y_i \sim \mathrm{Pois}(\mu_i)$.
-The dependence of the conditional mean $E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a log link and a linear predictor
+However, reaching this gold standard is quite costly: one needs to learn a tool
+to deal with package versions, then a build automation tool such as
+[make]{.proglang} and ideally [Docker]{.proglang} should be added to the list.
+[Docker]{.proglang} is a tool that makes it possible to run arbitrary code in a
+completely controlled and isolated environment as a way to capture the complete
+underlying system libraries.
 
-$$
-\log(\mu_i) \quad = \quad x_i^\top \beta,
-$$ {#eq-mean}
-
-where the regression coefficients $\beta$ are estimated by maximum likelihood (ML) using the iterative weighted least squares (IWLS) algorithm.
-
-::: callout
-TODO: Note that around the equation above there should be no spaces (avoided in the {{< latex >}} code by `%` lines) so that "normal" spacing is used and not a new paragraph started.
-:::
+### Prior art {#subsec-prior}
 

### 2023-12-12 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, enhanced, breakable, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 \makeatother
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colback=white, rightrule=.15mm, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, arc=.35mm, rightrule=.15mm, bottomrule=.15mm, colback=white, toprule=.15mm]
 
@@
 the paper describing the studies, which is not reproducible at all.
-Then, if to this paper we make the original source code of the analysis
-that was written to compute the results of the study available,
-reproducibility is improved, albeit only by a little. Adding the
-original data to this package improves reproducibility yet again.
-Finally, if to all this we add what Roger Peng named the \emph{linked
-and executable code and data}, we reach the gold standard of full
-replication.
+Then, if in addition, to this paper we make the original source code of
+the analysis that was written to compute the results of the study
+available, reproducibility is improved, albeit only by a little. Adding
+the original data improves reproducibility yet again. Finally, if to all
+this we add what Roger Peng named the \emph{linked and executable code
+and data}, we reach the gold standard of full replication.
 
@@
 \item
-  First, the version of \proglang{R} was stated
+  The version of \proglang{R} was stated;
+\item
+  Packages and their versions were listed and saved into an
+  \texttt{renv.lock} file;
+\item
+  A \texttt{Makefile} was used to run the whole analysis and compile the
+  paper;
+\item
+  A \texttt{Dockerfile} was used to provide the complete computational
+  environment, including the right version of \proglang{R} and run the
+  whole analysis easily.
 \end{itemize}
 
-\section{Models and software}\label{sec-models}
+As we go down this list, we get closer to the gold standard of a
+perfectly reproducible study.
 
-The basic Poisson regression model for count data is a special case of
-the GLM framework \citet{McCullaghNelder1989}. It describes the
-dependence of a count response variable \(y_i\) (\(i = 1, \dots, n\)) by
-assuming a Poisson distribution \(y_i \sim \mathrm{Pois}(\mu_i)\). The
-dependence of the conditional mean \(E[y_i \, | \, x_i] = \mu_i\) on the
-regressors \(x_i\) is then specified via a log link and a linear
-predictor
+However, reaching this gold standard is quite costly: one needs to learn
+a tool to deal with package versions, then a build automation tool such
+as \proglang{make} and ideally \proglang{Docker} should be added to the
+list. \proglang{Docker} is a tool that makes it possible to run
+arbitrary code in a completely controlled and isolated environment as a
+way to capture the complete underlying system libraries.
 
-\begin{equation}\phantomsection\label{eq-mean}{
-\log(\mu_i) \quad = \quad x_i^\top \beta,
-}\end{equation}
-
-where the regression coefficients \(\beta\) are estimated by maximum
-likelihood (ML) using the iterative weighted least squares (IWLS)
-algorithm.
-
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colback=white, rightrule=.15mm, opacityback=0, leftrule=.75mm]
-
-TODO: Note that around the equation above there should be no spaces
-(avoided in the {\LaTeX} code by \texttt{\%} lines) so that ``normal''
-spacing is used and not a new paragraph started.
-
-\end{tcolorbox}
+\subsection{Prior art}\label{subsec-prior}
 

### 2023-12-13 paper.qmd
review of solutions to restore packages

@@
   make them easily installable by someone that wishes to reproduce a study,
-  such as Docker, a containerization solution. This paper will present
-  the Nix package manager, and an R package called {rix} that lowers
-  Nix's learning curve for users of the R programming language.
+  such as [Docker].{proglang}, a containerization solution. This paper will present
+  the Nix package manager, and an [R].{proglang} package called [rix]{.pkg} that lowers
+  Nix's learning curve for users of the [R].{proglang} programming language.
 
@@
 ## Introduction: Reproducibility is also about software {#sec-intro}
-
-::: callout
-The introduction is in principle "as usual".
-However, it should usually embed both the implemented *methods* and the *software* into the respective relevant literature.
-For the latter both competing and complementary software should be discussed (within the same software environment and beyond), bringing out relative (dis)advantages.
-All software mentioned should be properly `@cited`'d.
-(See also [Using BibTeX] for more details on {{< bibtex >}}.)
-
-For writing about software JSS requires authors to use the markup `[]{.proglang}` (programming languages and large programmable systems), `[]{.pkg}` (software packages), back ticks like \`code\` for code (functions, commands, arguments, etc.).
-
-If there is such markup in (sub)section titles (as above), a plain text version has to be provided in the {{< latex >}} command as well.
-Below we also illustrate how abbrevations should be introduced and citation commands can be employed.
-See the {{< latex >}} code for more details.
-:::
 
@@
 
+Tools can be used at each of these steps to reach the gold standard of full
+replication. Let's first consider the task of listing the software used.
+[R]{.proglang} provides the `sessionInfo()` function whose output can be saved
+into a file. Below is an example output of `sessionInfo()`:
+
+```{r, eval = FALSE}
+sessionInfo()
+```
+
+```
+R version 4.3.2 (2023-10-31)
+Platform: aarch64-unknown-linux-gnu (64-bit)
+Running under: Ubuntu 22.04.3 LTS
+
+Matrix products: default
+BLAS:   /usr/lib/aarch64-linux-gnu/openblas-pthread/libblas.so.3 
+LAPACK: /usr/lib/aarch64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0
+
+locale:
+ [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+ [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
+ [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
+ [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
+ [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+[11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C
+
+time zone: Etc/UTC
+tzcode source: system (glibc)
+
+attached base packages:
+[1] stats     graphics  grDevices utils     datasets  methods  
+[7] base
+
+other attached packages:
+[1] nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163
+
+loaded via a namespace (and not attached):
+[1] compiler_4.3.2 Matrix_1.6-1.1 tools_4.3.2    splines_4.3.2 
+[5] grid_4.3.2     lattice_0.21-9
+```
+
+If an author provides this information, other people trying to reproduce the
+study (or the author him- or herself in the future) can read this file
+and see which version of [R]{.proglang} was used, and which packages (and their
+versions) were used as well. However, others would still need to install the
+correct software themselves. An alternative to this is instead to use the [renv]{.pkg}
+package which generates a so-called `renv.lock` file which also lists
+[R]{.proglang} and package versions. Here is an example of such an `renv.lock` file:
+
+```
+{
+"R": {
+  "Version": "4.2.2",
+  "Repositories": [
+  {
+   "Name": "CRAN",
+   "URL": "https://packagemanager.rstudio.com/all/latest"
+  }
+  ]
+},
+"Packages": {
+  "MASS": {
+    "Package": "MASS",
+    "Version": "7.3-58.1",
+    "Source": "Repository",
+    "Repository": "CRAN",
+    "Hash": "762e1804143a332333c054759f89a706",
+    "Requirements": []
+  },
+  "Matrix": {
+    "Package": "Matrix",
+    "Version": "1.5-1",
+    "Source": "Repository",
+    "Repository": "CRAN",
+    "Hash": "539dc0c0c05636812f1080f473d2c177",
+    "Requirements": [
+      "lattice"
+    ]
+
+    ***and many more packages***
+```
+
+This file lists every package alongside their versions and the repository from
+which they were downloaded. Generating this file only requires one to run the
+`renv::init()` function. Someone else can then restore the same package library
+by running `renv::restore()`. The exact same packages get installed in an
+isolated, project-specific, library which doesn't interfere with the other,
+main, library of the user. [renv]{.pkg} does not restore the [R]{.proglang}
+itself though, so installing the right version of R needs to be handled
+separately. Before continuing, it should be noted that other packages exist
+which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
+by @simonsohn2023 which makes it rather easy to install packages as they were
+on CRAN at a given date. For example, the code snippet below install the
+[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:
+
+```
+groundhog.library("
+    library(purrr)
+    library(ggplot2)",
+    "2017-10-04",
+    tolerate.R.version = "4.2.2")
+```
+
+These packages also get installed in a project-specfic library so there is no
+interferences between these packages and other versions of the same packages
+that one might use for other projects. Because [groundhog]{.pkg} does not
+install [R]{.proglang} itself, users should either install the required version
+themselves, or they should use the `tolerate.R.version` argument as shown in the
+example above. Otherwise, [groundhog]{.pkg} would not continue with the
+installation of the packages. Another such package, developed by @chan2023 is
+[rang]{.pkg}, which also installs packages as they were on a given date.
+Yet another way to install packages as they were on a give date is to use
+the Posit Package Manager, which provides snapshots of CRAN. For example,
+to install the required packages for an analysis as they were on the 30th of June
+2023, one could add the following line to the `.Rprofile` file:
+
+```
+options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"))
+```
+
+The `.Rprofile` file is a file that gets read by R when starting a new session, which
+means that every call to the `install.packages()` function will now install the packages
+from this snapshotted mirror.
+
 There are exceptions of course, a great example of a paper that provides everything
@@
 
-As we go down this list, we get closer to the gold standard of a perfectly reproducible
-study.
+As we go down this list, we get closer to the gold standard of a perfectly
+reproducible study.
 
@@
 completely controlled and isolated environment as a way to capture the complete
-underlying system libraries.
+underlying system libraries. Simply put, a `Dockerfile` provides a complete
+description of the computational environment as well as the required instructions
+to build the complete project. One can then use [Docker]{.proglang} to build
+a so-called *image* which contains all the required software and instructions
+to build the project. From this image, it is then possible to run a *container*
+which actually runs the analysis.
 

### 2023-12-13 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, frame hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 \makeatother
@@
 easily installable by someone that wishes to reproduce a study, such as
-Docker, a containerization solution. This paper will present the Nix
-package manager, and an R package called \{rix\} that lowers Nix's
-learning curve for users of the R programming language.}
+{[}Docker{]}.\{proglang\}, a containerization solution. This paper will
+present the Nix package manager, and an {[}R{]}.\{proglang\} package
+called \pkg{rix} that lowers Nix's learning curve for users of the
+{[}R{]}.\{proglang\} programming language.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, arc=.35mm, rightrule=.15mm, bottomrule=.15mm, colback=white, toprule=.15mm]
-
-The introduction is in principle ``as usual''. However, it should
-usually embed both the implemented \emph{methods} and the
-\emph{software} into the respective relevant literature. For the latter
-both competing and complementary software should be discussed (within
-the same software environment and beyond), bringing out relative
-(dis)advantages. All software mentioned should be properly
-\texttt{@cited}'d.~(See also \hyperref[sec-bibtex]{Using BibTeX} for
-more details on \textsc{Bib}{\TeX}.)
-
-For writing about software JSS requires authors to use the markup
-\texttt{{[}{]}\{.proglang\}} (programming languages and large
-programmable systems), \texttt{{[}{]}\{.pkg\}} (software packages), back
-ticks like `code` for code (functions, commands, arguments, etc.).
-
-If there is such markup in (sub)section titles (as above), a plain text
-version has to be provided in the {\LaTeX} command as well. Below we
-also illustrate how abbrevations should be introduced and citation
-commands can be employed. See the {\LaTeX} code for more details.
-
-\end{tcolorbox}
-
 \citet{peng2011} introduced the idea of reproducibility being on a
@@
 studies.
+
+Tools can be used at each of these steps to reach the gold standard of
+full replication. Let's first consider the task of listing the software
+used. \proglang{R} provides the \texttt{sessionInfo()} function whose
+output can be saved into a file. Below is an example output of
+\texttt{sessionInfo()}:
+
+\begin{verbatim}
+sessionInfo()
+\end{verbatim}
+
+\begin{verbatim}
+R version 4.3.2 (2023-10-31)
+Platform: aarch64-unknown-linux-gnu (64-bit)
+Running under: Ubuntu 22.04.3 LTS
+
+Matrix products: default
+BLAS:   /usr/lib/aarch64-linux-gnu/openblas-pthread/libblas.so.3 
+LAPACK: /usr/lib/aarch64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0
+
+locale:
+ [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+ [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
+ [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
+ [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
+ [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+[11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C
+
+time zone: Etc/UTC
+tzcode source: system (glibc)
+
+attached base packages:
+[1] stats     graphics  grDevices utils     datasets  methods  
+[7] base
+
+other attached packages:
+[1] nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163
+
+loaded via a namespace (and not attached):
+[1] compiler_4.3.2 Matrix_1.6-1.1 tools_4.3.2    splines_4.3.2 
+[5] grid_4.3.2     lattice_0.21-9
+\end{verbatim}
+
+If an author provides this information, other people trying to reproduce
+the study (or the author him- or herself in the future) can read this
+file and see which version of \proglang{R} was used, and which packages
+(and their versions) were used as well. However, others would still need
+to install the correct software themselves. An alternative to this is
+instead to use the \pkg{renv} package which generates a so-called
+\texttt{renv.lock} file which also lists \proglang{R} and package
+versions. Here is an example of such an \texttt{renv.lock} file:
+
+\begin{verbatim}
+{
+"R": {
+  "Version": "4.2.2",
+  "Repositories": [
+  {
+   "Name": "CRAN",
+   "URL": "https://packagemanager.rstudio.com/all/latest"
+  }
+  ]
+},
+"Packages": {
+  "MASS": {
+    "Package": "MASS",
+    "Version": "7.3-58.1",
+    "Source": "Repository",
+    "Repository": "CRAN",
+    "Hash": "762e1804143a332333c054759f89a706",
+    "Requirements": []
+  },
+  "Matrix": {
+    "Package": "Matrix",
+    "Version": "1.5-1",
+    "Source": "Repository",
+    "Repository": "CRAN",
+    "Hash": "539dc0c0c05636812f1080f473d2c177",
+    "Requirements": [
+      "lattice"
+    ]
+
+    ***and many more packages***
+\end{verbatim}
+
+This file lists every package alongside their versions and the
+repository from which they were downloaded. Generating this file only
+requires one to run the \texttt{renv::init()} function. Someone else can
+then restore the same package library by running
+\texttt{renv::restore()}. The exact same packages get installed in an
+isolated, project-specific, library which doesn't interfere with the
+other, main, library of the user. \pkg{renv} does not restore the
+\proglang{R} itself though, so installing the right version of R needs
+to be handled separately. Before continuing, it should be noted that
+other packages exist which provide similar functionality to \pkg{renv}:
+there is \pkg{groundhog} by \citet{simonsohn2023} which makes it rather
+easy to install packages as they were on CRAN at a given date. For
+example, the code snippet below install the \pkg{purrr} and
+\pkg{ggplot2} packages as they were on April 4th, 2017:
+
+\begin{verbatim}
+groundhog.library("
+    library(purrr)
+    library(ggplot2)",
+    "2017-10-04",
+    tolerate.R.version = "4.2.2")
+\end{verbatim}
+
+These packages also get installed in a project-specfic library so there
+is no interferences between these packages and other versions of the
+same packages that one might use for other projects. Because
+\pkg{groundhog} does not install \proglang{R} itself, users should
+either install the required version themselves, or they should use the
+\texttt{tolerate.R.version} argument as shown in the example above.
+Otherwise, \pkg{groundhog} would not continue with the installation of
+the packages. Another such package, developed by \citet{chan2023} is
+\pkg{rang}, which also installs packages as they were on a given date.
+Yet another way to install packages as they were on a give date is to
+use the Posit Package Manager, which provides snapshots of CRAN. For
+example, to install the required packages for an analysis as they were
+on the 30th of June 2023, one could add the following line to the
+\texttt{.Rprofile} file:
+
+\begin{verbatim}
+options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"))
+\end{verbatim}
+
+The \texttt{.Rprofile} file is a file that gets read by R when starting
+a new session, which means that every call to the
+\texttt{install.packages()} function will now install the packages from
+this snapshotted mirror.
 
@@
 arbitrary code in a completely controlled and isolated environment as a
-way to capture the complete underlying system libraries.
+way to capture the complete underlying system libraries. Simply put, a
+\texttt{Dockerfile} provides a complete description of the computational
+environment as well as the required instructions to build the complete
+project. One can then use \proglang{Docker} to build a so-called
+\emph{image} which contains all the required software and instructions
+to build the project. From this image, it is then possible to run a
+\emph{container} which actually runs the analysis.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, arc=.35mm, rightrule=.15mm, bottomrule=.15mm, colback=white, toprule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, leftrule=.75mm, bottomrule=.15mm, left=2mm, breakable, toprule=.15mm, colback=white, opacityback=0, rightrule=.15mm]
 

### 2023-12-13 paper.qmd
started talking about rig

@@
 
+The next step in reaching the gold standard of reproducibility would be to not only
+install the right packages used for the analysis, but also the right version of
+[R]{.proglang}. Of course, it would be possible to install the right version manually,
+but here too, there are tools that simplify the process such as @rlib2023.
+
 There are exceptions of course, a great example of a paper that provides everything

### 2023-12-13 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, frame hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, interior hidden, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 \makeatother
@@
 
+The next step in reaching the gold standard of reproducibility would be
+to not only install the right packages used for the analysis, but also
+the right version of \proglang{R}. Of course, it would be possible to
+install the right version manually, but here too, there are tools that
+simplify the process such as \citet{rlib2023}.
+
 There are exceptions of course, a great example of a paper that provides
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, leftrule=.75mm, bottomrule=.15mm, left=2mm, breakable, toprule=.15mm, colback=white, opacityback=0, rightrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, leftrule=.75mm, colback=white, rightrule=.15mm, left=2mm, opacityback=0, breakable, bottomrule=.15mm]
 

### 2023-12-14 paper.qmd
first draft of intro

@@
   make them easily installable by someone that wishes to reproduce a study,
-  such as [Docker].{proglang}, a containerization solution. This paper will present
-  the Nix package manager, and an [R].{proglang} package called [rix]{.pkg} that lowers
-  Nix's learning curve for users of the [R].{proglang} programming language.
+  such as [Docker]{.proglang}, a containerization solution. This paper will present
+  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that lowers
+  Nix's learning curve for users of the [R]{.proglang} programming language.
 
@@
 BLAS:   /usr/lib/aarch64-linux-gnu/openblas-pthread/libblas.so.3 
-LAPACK: /usr/lib/aarch64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0
+LAPACK: /usr/lib/aarch64-linux-gnu/openblas-pthread/libopenblasp[...]
 
@@
 ```
-options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"))
+options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/
+                               __linux__/jammy/2023-06-30"))
 ```
 
-The `.Rprofile` file is a file that gets read by R when starting a new session, which
-means that every call to the `install.packages()` function will now install the packages
-from this snapshotted mirror.
+The `.Rprofile` file gets read by R when starting a new session, which means
+that every call to the `install.packages()` function will now install the
+packages from this snapshotted mirror.
 
-The next step in reaching the gold standard of reproducibility would be to not only
-install the right packages used for the analysis, but also the right version of
-[R]{.proglang}. Of course, it would be possible to install the right version manually,
-but here too, there are tools that simplify the process such as @rlib2023.
+The next step in reaching the gold standard of reproducibility would be to not
+only install the right packages used for the analysis, but also the right
+version of [R]{.proglang}. Of course, it would be possible to install the right
+version manually, but here too, there are tools that simplify the process such
+as [rig]{.proglang} by the R infrastructure team [-@rlib2023]. One could thus
+install the right version of [R]{.proglang} using [rig]{.proglang} allows to
+easily install different versions of [R]{.proglang}, one could use to install
+the right version to reproduce an analysis, and then use one of the listed
+packages above to install the right library of of [R]{.proglang} packages.
+However, this involves many manual steps and is thus error prone. It is also
+time-consuming to do so, especially if one wants to reproduce many studies.
 
-There are exceptions of course, a great example of a paper that provides everything
-needed to reproduce its results is @mcdermott2021. The author of this paper set up
-an accompagnying Github repository to the paper^[https://github.com/grantmcdermott/skeptic-priors]
-containing all the instructions to install the required software and then run the
-analysis. If we take a closer look at this repository, we will notice that several
-tools were used to capture the compatutational environment and make it available to
-other researchers:
+The final step towards the gold standard of reproducibility is to package the
+right version of [R]{.proglang} and [R]{.proglang} packages inside a
+[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
+[Docker]{.proglang} it is possible to package some *data product* with its
+dependencies into an image. From this image, containers can be run to reproduce
+the data product with its exact dependencies. A statistical analysis, from the
+simplest to the most complex one, can be seen as such a data product that comes
+with many software dependencies. *Dockerizing* an analysis consists in first
+building an image: in the build step, the dependencies of the analysis have to
+be installed and this can be achieved using the tools mentioned above. The
+scripts to run the analysis and the data are also added to the image at build
+time. Instances of that image, called containers, can then be executed, which
+give access to the exact development environment originally used for the study.
+It is also possible to script the building of the analysis itself, such that in
+one single call to the `docker run` command, a complete analysis is reproduced.
 
-- The version of [R]{.proglang} was stated;
+The Rocker project initiated by @boettiger2017 provides many pre-built
+[Docker]{.proglang} to the [R]{.proglang} community of users. These images
+can be used as bases to build other images containing statistical analyses
+more easily than starting from a bare-bones image.
+
+An optional step towards the gold standard is to use a build automation tool
+such as [Make]{.proglang} to run the whole analysis when the container is
+executed. Build automation tools make it easier to run arbitrary code in a
+series of well-defined steps. [R]{.proglang} programmers can use the
+[targets]{.pkg} by @landau2021 as a build automation tool.
+
+@mcdermott2021 is an example of a scientific study that reached the gold
+standard of reproducibility. The author of this paper set up an accompagnying
+Github repository to the
+paper^[https://github.com/grantmcdermott/skeptic-priors] containing all the
+instructions to install the required software and then run the analysis. If we
+take a closer look at this repository, we will notice that many of the tools
+previously mentioned were used to capture the compatutational environment and
+make it available to other researchers:
+
 - Packages and their versions were listed and saved into an `renv.lock` file;

### 2023-12-14 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, interior hidden, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
 \makeatother
@@
 easily installable by someone that wishes to reproduce a study, such as
-{[}Docker{]}.\{proglang\}, a containerization solution. This paper will
-present the Nix package manager, and an {[}R{]}.\{proglang\} package
-called \pkg{rix} that lowers Nix's learning curve for users of the
-{[}R{]}.\{proglang\} programming language.}
+\proglang{Docker}, a containerization solution. This paper will present
+the Nix package manager, and an \proglang{R} package called \pkg{rix}
+that lowers Nix's learning curve for users of the \proglang{R}
+programming language.}
 
@@
 \begin{verbatim}
-options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"))
+options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/
+                               __linux__/jammy/2023-06-30"))
 \end{verbatim}
 
-The \texttt{.Rprofile} file is a file that gets read by R when starting
-a new session, which means that every call to the
-\texttt{install.packages()} function will now install the packages from
-this snapshotted mirror.
+The \texttt{.Rprofile} file gets read by R when starting a new session,
+which means that every call to the \texttt{install.packages()} function
+will now install the packages from this snapshotted mirror.
 
@@
 install the right version manually, but here too, there are tools that
-simplify the process such as \citet{rlib2023}.
+simplify the process such as \proglang{rig} by the R infrastructure team
+\citeyearpar{rlib2023}. One could thus install the right version of
+\proglang{R} using \proglang{rig} allows to easily install different
+versions of \proglang{R}, one could use to install the right version to
+reproduce an analysis, and then use one of the listed packages above to
+install the right library of of \proglang{R} packages. However, this
+involves many manual steps and is thus error prone. It is also
+time-consuming to do so, especially if one wants to reproduce many
+studies.
 
-There are exceptions of course, a great example of a paper that provides
-everything needed to reproduce its results is \citet{mcdermott2021}. The
-author of this paper set up an accompagnying Github repository to the
-paper\footnote{https://github.com/grantmcdermott/skeptic-priors}
+The final step towards the gold standard of reproducibility is to
+package the right version of \proglang{R} and \proglang{R} packages
+inside a \proglang{Docker} image. \proglang{Docker} is a
+containerisation tool: using \proglang{Docker} it is possible to package
+some \emph{data product} with its dependencies into an image. From this
+image, containers can be run to reproduce the data product with its
+exact dependencies. A statistical analysis, from the simplest to the
+most complex one, can be seen as such a data product that comes with
+many software dependencies. \emph{Dockerizing} an analysis consists in
+first building an image: in the build step, the dependencies of the
+analysis have to be installed and this can be achieved using the tools
+mentioned above. The scripts to run the analysis and the data are also
+added to the image at build time. Instances of that image, called
+containers, can then be executed, which give access to the exact
+development environment originally used for the study. It is also
+possible to script the building of the analysis itself, such that in one
+single call to the \texttt{docker\ run} command, a complete analysis is
+reproduced.
+
+The Rocker project initiated by \citet{boettiger2017} provides many
+pre-built \proglang{Docker} to the \proglang{R} community of users.
+These images can be used as bases to build other images containing
+statistical analyses more easily than starting from a bare-bones image.
+
+An optional step towards the gold standard is to use a build automation
+tool such as \proglang{Make} to run the whole analysis when the
+container is executed. Build automation tools make it easier to run
+arbitrary code in a series of well-defined steps. \proglang{R}
+programmers can use the \pkg{targets} by \citet{landau2021} as a build
+automation tool.
+
+\citet{mcdermott2021} is an example of a scientific study that reached
+the gold standard of reproducibility. The author of this paper set up an
+accompagnying Github repository to the paper\footnote{https://github.com/grantmcdermott/skeptic-priors}
 containing all the instructions to install the required software and
 then run the analysis. If we take a closer look at this repository, we
-will notice that several tools were used to capture the compatutational
-environment and make it available to other researchers:
+will notice that many of the tools previously mentioned were used to
+capture the compatutational environment and make it available to other
+researchers:
 
@@
 \tightlist
-\item
-  The version of \proglang{R} was stated;
 \item
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, leftrule=.75mm, colback=white, rightrule=.15mm, left=2mm, opacityback=0, breakable, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, breakable, left=2mm, colback=white, arc=.35mm, rightrule=.15mm, toprule=.15mm, bottomrule=.15mm]
 

### 2023-12-14 paper.qmd
started nix

@@
 
-As we go down this list, we get closer to the gold standard of a perfectly
-reproducible study.
+However, reaching this gold standard is quite costly: one needs to learn a tool
+to deal with package versions, then [Docker]{.proglang} for all the other
+software, including the programming language that was used. Ideally, a build
+automation tool such as [make]{.proglang} should be added to the list. It should
+also be noted that these requirements are not limited to the [R]{.proglang}
+programming language. A very similar approach should be taken if one uses 
+[Python]{.proglang} for statistical analysis instead.
 
-However, reaching this gold standard is quite costly: one needs to learn a tool
-to deal with package versions, then a build automation tool such as
-[make]{.proglang} and ideally [Docker]{.proglang} should be added to the list.
-[Docker]{.proglang} is a tool that makes it possible to run arbitrary code in a
-completely controlled and isolated environment as a way to capture the complete
-underlying system libraries. Simply put, a `Dockerfile` provides a complete
-description of the computational environment as well as the required instructions
-to build the complete project. One can then use [Docker]{.proglang} to build
-a so-called *image* which contains all the required software and instructions
-to build the project. From this image, it is then possible to run a *container*
-which actually runs the analysis.
+As an alternative to the above approach, I will present the [Nix]{.proglang}
+package manager, which available for all major operating system and focuses on
+installing and building software in a reproducible manner. To make
+[Nix]{.proglang} more accessible to [R]{.proglang} programmers, I developed the
+[rix]{.pkg} which I will also present in this article.
 
-### Prior art {#subsec-prior}
+## The [Nix]{.proglang} package manager
+
+[Nix]{.proglang} is a package manager that can be used to install and build
+software in a completely reproducible manner. With [Nix]{.proglang}
+it is possible to install [R]{.proglang} and [R]{.proglang} packages as well
+(whether these come from CRAN or Bioconductor)
+
+is that its repository of software is huge. As of writing, it contains more than
+80.000 packages, and the entirety of CRAN and Bioconductor is available through
+Nix’s repositories. This means that using Nix, it is possible to install not
+only R, but also all the packages required for your project. The obvious
+question is why use Nix instead of simply installing R and R packages as usual.
+The answer is that Nix makes sure to install every dependency of any package, up
+to required system libraries. For example, the `{xlsx}` package requires the
+Java programming language to be installed on your computer to successfully
+install. This can be difficult to achieve, and `{xlsx}` bullied many R
+developers throughout the years (especially those using a Linux distribution,
+`sudo R CMD javareconf` still plagues my nightmares). But with Nix, it suffices
+to declare that we want the `{xlsx}` package for our project, and Nix figures
+out automatically that Java is required and installs and configures it. It all
+just happens without any required intervention from the user. The second
+advantage of Nix is that it is possible to pin a certain revision of the Nix
+packages’ repository (called nixpkgs) for our project. Pinning a revision
+ensures that every package that Nix installs will always be at exactly the same
+versions, regardless of when in the future the packages get installed.
+
+With Nix, it is essentially possible to replace {renv} and Docker combined, or
+if you’re using mainly Python, you can replace `conda` or `requirements.txt`
+files. If you need other tools or languages like Python or Julia, this can also
+be done easily. Nix is available for Linux, macOS and Windows (via WSL2).
+Important remark: since using Nix on Windows must go through WSL, when we refer
+to "Linux" in the context of Nix, this includes Windows by default as well. It
+is also possible to build multi-language environments, containing R and Python,
+a LaTeX distribution and packages and so on.
+
+## The Nix programming language
+
+Nix is not just useful because it is possible to install many packages and even
+install older packages, but also because it comes with a complete functional
+programming language. This programming language is used to write *expressions*,
+and these expressions in turn are used to build software. Essentially, when you
+install a package using Nix, an expression gets downloaded from the Nix package
+repository (more on that in the next section), and it gets evaluated by the Nix
+package manager. This expression contains a so-called *derivation*. A derivation
+defines a build: some inputs, some commands, and then an output. Most of the
+time, a derivation downloads source code, builds the software from the source
+and then outputs a compiled binary. Derivations are extremely flexible, and you
+could write a derivation to build a complete environment and then build a
+complete reproducible pipeline. The output could be any of the discussed data
+products.
+
+Learning the Nix programming language is a good idea if you want to contribute
+to the Nix package repository, but you might not have to learn it in-depth if
+you simply wish to use it to build reproducible environments, as we will learn
+now. If you wish to learn about the programming language, I highly recommend [a
+tour of
+Nix](https://nixcloud.io/tour/?id=introduction/nix)^[https://nixcloud.io/tour/?id=introduction/nix].
+
+## The Nix package repository
+
+So, there’s the Nix package manager, the Nix programming language and the Nix
+package repository (henceforth nixpkgs). To look for packages click
+[here](https://search.nixos.org/packages)^[https://search.nixos.org/packages].
+The source code of all the packages (so the whole set of Nix expressions) can be
+found on [this Github
+repository](https://github.com/NixOS/nixpkgs)^[https://github.com/NixOS/nixpkgs].
+For example,
+[here](https://github.com/NixOS/nixpkgs/blob/master/pkgs/development/libraries/quarto/default.nix)^[https://github.com/NixOS/nixpkgs/blob/master/pkgs/development/libraries/quarto/default.nix]
+is the Nix expression that contains the derivation to build `quarto`. As you can
+see, the derivation uses the the pre-built Quarto binaries instead of building
+it from source. Adding packages to nixpkgs (or updating them) can be done by
+opening pull requests. For example,
+[here](https://github.com/NixOS/nixpkgs/pull/259443)^[https://github.com/NixOS/nixpkgs/pull/259443]
+is a pull request to make Quarto available to all platforms (before this PR
+Quarto was only available for Linux). PRs get reviewed and approuved by
+maintainers that also have the right to merge the PR into master. Once merged,
+the new or updated package is available for download. Because nixpkgs is a
+“just” Github repository, it is possible to use a specific commit hash to
+install the packages as they were at a specific point in time. For example, if
+you use this commit, 7c9cc5a6e, you’ll get the packages as of the 19th of
+October 2023, but if you used this one instead: 976fa3369, you’ll get packages
+from the 19th of August 2023. Using specific hashes is called "pinning" and you
+can read more about it [here](https://nixos.wiki/wiki/FAQ/Pinning_Nixpkgs). We
+will make extensive use of pinning.
 

### 2023-12-18 paper.qmd
some text b b

@@
 [Nix]{.proglang} is a package manager that can be used to install and build
-software in a completely reproducible manner. With [Nix]{.proglang}
-it is possible to install [R]{.proglang} and [R]{.proglang} packages as well
-(whether these come from CRAN or Bioconductor)
+software in a completely reproducible manner. As of writing, it contains more
+than 80.000 packages, and the entirety of CRAN and Bioconductor is available
+through [Nix]{.proglang}’s repositories. This means that using [Nix]{.proglang},
+it is possible to install not only [R]{.proglang}, but also all the packages
+required for a project. The reason why one should use [Nix]{.proglang} to
+install [R]{.proglang} packages, and not use the usual, built-in,
+`install.packages()` function instead, is that when installing a package with
+[Nix]{.proglang}, [Nix]{.proglang} makes sure to install every dependency of
+every package, whether this dependency is another R package or a system-level
+dependency.
 
-is that its repository of software is huge. As of writing, it contains more than
-80.000 packages, and the entirety of CRAN and Bioconductor is available through
-Nix’s repositories. This means that using Nix, it is possible to install not
-only R, but also all the packages required for your project. The obvious
-question is why use Nix instead of simply installing R and R packages as usual.
-The answer is that Nix makes sure to install every dependency of any package, up
-to required system libraries. For example, the `{xlsx}` package requires the
-Java programming language to be installed on your computer to successfully
-install. This can be difficult to achieve, and `{xlsx}` bullied many R
-developers throughout the years (especially those using a Linux distribution,
-`sudo R CMD javareconf` still plagues my nightmares). But with Nix, it suffices
-to declare that we want the `{xlsx}` package for our project, and Nix figures
-out automatically that Java is required and installs and configures it. It all
-just happens without any required intervention from the user. The second
-advantage of Nix is that it is possible to pin a certain revision of the Nix
-packages’ repository (called nixpkgs) for our project. Pinning a revision
-ensures that every package that Nix installs will always be at exactly the same
-versions, regardless of when in the future the packages get installed.
+For example, the `{xlsx}` package requires the Java programming language to be
+installed on your computer to successfully install. This can be difficult to
+achieve, and `{xlsx}` bullied many R developers throughout the years (especially
+those using a Linux distribution, `sudo R CMD javareconf` still plagues my
+nightmares). But with Nix, it suffices to declare that we want the `{xlsx}`
+package for our project, and Nix figures out automatically that Java is required
+and installs and configures it. It all just happens without any required
+intervention from the user. The second advantage of Nix is that it is possible
+to pin a certain revision of the Nix packages’ repository (called nixpkgs) for
+our project. Pinning a revision ensures that every package that Nix installs
+will always be at exactly the same versions, regardless of when in the future
+the packages get installed.
 

### 2023-12-21 paper.qmd
nixpkgs

@@
 
-For example, the `{xlsx}` package requires the Java programming language to be
-installed on your computer to successfully install. This can be difficult to
-achieve, and `{xlsx}` bullied many R developers throughout the years (especially
-those using a Linux distribution, `sudo R CMD javareconf` still plagues my
-nightmares). But with Nix, it suffices to declare that we want the `{xlsx}`
-package for our project, and Nix figures out automatically that Java is required
-and installs and configures it. It all just happens without any required
-intervention from the user. The second advantage of Nix is that it is possible
-to pin a certain revision of the Nix packages’ repository (called nixpkgs) for
-our project. Pinning a revision ensures that every package that Nix installs
-will always be at exactly the same versions, regardless of when in the future
-the packages get installed.
+For example, the [xlsx]{.pkg} package requires the [Java]{.proglang} programming
+language to be installed on your computer to successfully install. Depending on
+the system one tries to install [xlsx]{.pkg}, installing [Java]{.proglang} might
+not easy, nor even possible. But with [Nix]{.proglang}, it suffices to declare
+that the [xlsx]{.pkg} package is needed for the project, and [Nix]{.proglang}
+figures out automatically that [Java]{.proglang} is required and installs and
+configures it. It all just happens without any required intervention from the
+user.
+
+But where do these packages come from? When installing a package using Nix an
+expression written in the Nix programming language gets downloaded from the
+`nixpkgs` Github repository and evaluated. The evaluation of these expressions
+results in the installation of the package, alongside its dependencies. Because
+`nixpkgs` is a collection of expressions hosted on Github, it is possible to pin
+a specific revision of `nixpkgs` to ensure reproducibility of our project.
+Pinning a revision ensures that every package that Nix installs will always be
+at exactly the same versions, regardless of when in the future the packages get
+installed.
 

### 2023-12-22 paper.qmd
intro to nix done?

@@
 install [R]{.proglang} packages, and not use the usual, built-in,
-`install.packages()` function instead, is that when installing a package with
-[Nix]{.proglang}, [Nix]{.proglang} makes sure to install every dependency of
-every package, whether this dependency is another R package or a system-level
-dependency.
+`install.packages()` [R]{.proglang} function instead, is that when installing a
+package with [Nix]{.proglang}, [Nix]{.proglang} makes sure to install every
+dependency of every package, whether this dependency is another [R]{.proglang}
+package or a system-level dependency.
 
-For example, the [xlsx]{.pkg} package requires the [Java]{.proglang} programming
-language to be installed on your computer to successfully install. Depending on
-the system one tries to install [xlsx]{.pkg}, installing [Java]{.proglang} might
-not easy, nor even possible. But with [Nix]{.proglang}, it suffices to declare
-that the [xlsx]{.pkg} package is needed for the project, and [Nix]{.proglang}
-figures out automatically that [Java]{.proglang} is required and installs and
-configures it. It all just happens without any required intervention from the
-user.
+For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
+[Java]{.proglang} programming language to be installed to successfully install.
+Depending on the what system one tries to install [xlsx]{.pkg}, installing
+[Java]{.proglang} might not easy, nor even possible. But with [Nix]{.proglang},
+it suffices to declare that the [xlsx]{.pkg} package is needed for the project,
+and [Nix]{.proglang} figures out automatically that [Java]{.proglang} is
+required and installs and configures it. It all just happens without any
+required intervention from the user.
 
-But where do these packages come from? When installing a package using Nix an
-expression written in the Nix programming language gets downloaded from the
-`nixpkgs` Github repository and evaluated. The evaluation of these expressions
-results in the installation of the package, alongside its dependencies. Because
-`nixpkgs` is a collection of expressions hosted on Github, it is possible to pin
-a specific revision of `nixpkgs` to ensure reproducibility of our project.
-Pinning a revision ensures that every package that Nix installs will always be
-at exactly the same versions, regardless of when in the future the packages get
-installed.
+But where do these packages come from? When installing a package using
+[Nix]{.proglang}, an expression written in the [Nix]{.proglang} programming
+language gets downloaded from the `nixpkgs` Github repository and evaluated.
+This expression contains a so-called *derivation*. A derivation defines a build:
+its dependencies, commands to build and install the package in question, and
+then an output. Most of the time, a derivation downloads source code, builds the
+software from the source and then outputs a compiled binary. Derivations are
+extremely flexible, and by writing one's own, it is possible to define and build
+environment for a project in a reproducible manner.
 
-With Nix, it is essentially possible to replace {renv} and Docker combined, or
-if you’re using mainly Python, you can replace `conda` or `requirements.txt`
-files. If you need other tools or languages like Python or Julia, this can also
-be done easily. Nix is available for Linux, macOS and Windows (via WSL2).
-Important remark: since using Nix on Windows must go through WSL, when we refer
-to "Linux" in the context of Nix, this includes Windows by default as well. It
-is also possible to build multi-language environments, containing R and Python,
-a LaTeX distribution and packages and so on.
+Because the whole set of [Nix]{.proglang} expression is hosted on Github, it is
+possible to pin a specific revision of `nixpkgs` to ensure reproducibility of
+our project. Pinning a revision ensures that every package that Nix installs
+will always be at exactly the same versions, regardless of when in the future
+the packages get installed.
 
-## The Nix programming language
-
-Nix is not just useful because it is possible to install many packages and even
-install older packages, but also because it comes with a complete functional
-programming language. This programming language is used to write *expressions*,
-and these expressions in turn are used to build software. Essentially, when you
-install a package using Nix, an expression gets downloaded from the Nix package
-repository (more on that in the next section), and it gets evaluated by the Nix
-package manager. This expression contains a so-called *derivation*. A derivation
-defines a build: some inputs, some commands, and then an output. Most of the
-time, a derivation downloads source code, builds the software from the source
-and then outputs a compiled binary. Derivations are extremely flexible, and you
-could write a derivation to build a complete environment and then build a
-complete reproducible pipeline. The output could be any of the discussed data
-products.
-
-Learning the Nix programming language is a good idea if you want to contribute
-to the Nix package repository, but you might not have to learn it in-depth if
-you simply wish to use it to build reproducible environments, as we will learn
-now. If you wish to learn about the programming language, I highly recommend [a
-tour of
-Nix](https://nixcloud.io/tour/?id=introduction/nix)^[https://nixcloud.io/tour/?id=introduction/nix].
-
-## The Nix package repository
-
-So, there’s the Nix package manager, the Nix programming language and the Nix
-package repository (henceforth nixpkgs). To look for packages click
-[here](https://search.nixos.org/packages)^[https://search.nixos.org/packages].
-The source code of all the packages (so the whole set of Nix expressions) can be
-found on [this Github
-repository](https://github.com/NixOS/nixpkgs)^[https://github.com/NixOS/nixpkgs].
-For example,
-[here](https://github.com/NixOS/nixpkgs/blob/master/pkgs/development/libraries/quarto/default.nix)^[https://github.com/NixOS/nixpkgs/blob/master/pkgs/development/libraries/quarto/default.nix]
-is the Nix expression that contains the derivation to build `quarto`. As you can
-see, the derivation uses the the pre-built Quarto binaries instead of building
-it from source. Adding packages to nixpkgs (or updating them) can be done by
-opening pull requests. For example,
-[here](https://github.com/NixOS/nixpkgs/pull/259443)^[https://github.com/NixOS/nixpkgs/pull/259443]
-is a pull request to make Quarto available to all platforms (before this PR
-Quarto was only available for Linux). PRs get reviewed and approuved by
-maintainers that also have the right to merge the PR into master. Once merged,
-the new or updated package is available for download. Because nixpkgs is a
-“just” Github repository, it is possible to use a specific commit hash to
-install the packages as they were at a specific point in time. For example, if
-you use this commit, 7c9cc5a6e, you’ll get the packages as of the 19th of
-October 2023, but if you used this one instead: 976fa3369, you’ll get packages
-from the 19th of August 2023. Using specific hashes is called "pinning" and you
-can read more about it [here](https://nixos.wiki/wiki/FAQ/Pinning_Nixpkgs). We
-will make extensive use of pinning.
-
-[R]{.proglang} provides a very flexible implementation of the general GLM framework in the function [glm]{.fct} @ChambersHastie1992 in the [stats]{.pkg} package. Its most important arguments are
-
-```r
-glm(formula, data, subset, na.action, weights, offset,
-  family = gaussian, start = NULL, control = glm.control(…),
-  model = TRUE, y = TRUE, x = FALSE, …)
-```
-
-where `formula` plus `data` is the now standard way of specifying regression relationships in [R]{.proglang}/[S]{.proglang} introduced in @ChambersHastie1992.
-The remaining arguments in the first line (`subset`, `na.action`, `weights`, and `offset`) are also standard for setting up formula-based regression models in [R]{.proglang}/[S]{.proglang}.
-The arguments in the second line control aspects specific to GLMs while the arguments in the last line specify which components are returned in the fitted model object (of class [glm]{.class} which inherits from [lm]{.class}).
-For further arguments to [glm]{.fct} (including alternative specifications of starting values) see `?glm`.
-For estimating a Poisson model `family = poisson` has to be specified.
-
-::: callout
-As the synopsis above is a code listing that is not meant to be executed, one can use either the dedicated `{Code}` environment or a simple `{verbatim}` environment for this.
-Again, spaces before and after should be avoided.
-
-Finally, there might be a reference to a `{table}` such as @tbl-overview.
-Usually, these are placed at the top of the page (`[t!]`), centered (`\centering`), with a caption below the table, column headers and captions in sentence style, and if possible avoiding vertical lines.
-:::
-
-| Type           | Distribution | Method   | Description                                                                                                                                                                                  |
-|---------|---------|---------|-----------------------------------------------|
-| GLM            | Poisson      | ML       | Poisson regression: classical GLM, estimated by maximum likelihood (ML)                                                                                                                      |
-|                |              | Quasi    | "Quasi-Poisson regression'': same mean function, estimated by quasi-ML (QML) or equivalently generalized estimating equations (GEE), inference adjustment via estimated dispersion parameter |
-|                |              | Adjusted | "Adjusted Poisson regression'': same mean function, estimated by QML/GEE, inference adjustment via sandwich covariances                                                                      |
-|                | NB           | ML       | NB regression: extended GLM, estimated by ML including additional shape parameter                                                                                                            |
-| Zero-augmented | Poisson      | ML       | Zero-inflated Poisson (ZIP), hurdle Poisson                                                                                                                                                  |
-|                | NB           | ML       | Zero-inflated NB (ZINB), hurdle NB                                                                                                                                                           |
-: Overview of various count regression models.
-The table is usually placed at the top of the page (`[t!]`), centered (`centering`), has a caption below the table, column headers and captions are in sentence style, and if possible vertical lines should be avoided.
-{#tbl-overview}
-
-## Illustrations {#sec-illustrations}
-
-For a simple illustration of basic Poisson and NB count regression the
-`quine` data from the [MASS]{.pkg} package is used. This provides the number
-of `Days` that children were absent from school in Australia in a
-particular year, along with several covariates that can be employed as regressors. The data can be loaded by
-
-```{R}
-#| prompt: true
-data(mtcars)
-```
-and a basic frequency distribution of the response variable is displayed in
-@fig-quine.
-
-:::{.callout}
-For code input and output, the style files provide dedicated environments.
-Either the "agnostic" `{CodeInput}` and `{CodeOutput}` can be used
-or, equivalently, the environments `{Sinput}` and `{Soutput}` as
-produced by [Sweave]{.fct} or [knitr]{.pkg} when using the `render_sweave()`
-hook. Please make sure that all code is properly spaced, e.g., using
-`y = a + b * x` and _not_ `y=a+b*x`. Moreover, code input should use "the usual" command prompt in the respective software system. For
-[R]{.proglang} code, the prompt `R> ` should be used with `+  ` as
-the continuation prompt. Generally, comments within the code chunks should be
-avoided -- and made in the regular {{< latex >}} text instead. Finally, empty lines before and after code input/output should be avoided (see above).
-:::
-
-## Summary and discussion {#sec-summary}
-
-:::{.callout}
-
-As usual…
-
-:::
-
-## Computational details {.unnumbered}
-
-:::{.callout}
-
-If necessary or useful, information about certain computational details
-such as version numbers, operating systems, or compilers could be included
-in an unnumbered section. Also, auxiliary packages (say, for visualizations,
-maps, tables, …) that are not cited in the main text can be credited here.
-
-:::
-
-The results in this paper were obtained using [R]{.proglang}~3.4.1 with the
-[MASS]{.pkg}~7.3.47 package. [R]{.proglang} itself and all packages used are available from the Comprehensive [R]{.proglang} Archive Network (CRAN) at
-[https://CRAN.R-project.org/].
+With Nix, it is essentially possible to replace [renv]{.pkg} and Docker
+combined, or in the case of [Python]{.proglang}, replace `requirements.txt`
+files which list package dependencies for [Python]{.proglang} projects. It is
+also possible to build multi-language environments, containing R and Python, a
+LaTeX distribution, and even install one's favourite text editor to edit the
+project's source code.
 

### 2023-12-22 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, frame hidden, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 \makeatother
@@
 
-As we go down this list, we get closer to the gold standard of a
-perfectly reproducible study.
+However, reaching this gold standard is quite costly: one needs to learn
+a tool to deal with package versions, then \proglang{Docker} for all the
+other software, including the programming language that was used.
+Ideally, a build automation tool such as \proglang{make} should be added
+to the list. It should also be noted that these requirements are not
+limited to the \proglang{R} programming language. A very similar
+approach should be taken if one uses \proglang{Python} for statistical
+analysis instead.
 
-However, reaching this gold standard is quite costly: one needs to learn
-a tool to deal with package versions, then a build automation tool such
-as \proglang{make} and ideally \proglang{Docker} should be added to the
-list. \proglang{Docker} is a tool that makes it possible to run
-arbitrary code in a completely controlled and isolated environment as a
-way to capture the complete underlying system libraries. Simply put, a
-\texttt{Dockerfile} provides a complete description of the computational
-environment as well as the required instructions to build the complete
-project. One can then use \proglang{Docker} to build a so-called
-\emph{image} which contains all the required software and instructions
-to build the project. From this image, it is then possible to run a
-\emph{container} which actually runs the analysis.
+As an alternative to the above approach, I will present the
+\proglang{Nix} package manager, which available for all major operating
+system and focuses on installing and building software in a reproducible
+manner. To make \proglang{Nix} more accessible to \proglang{R}
+programmers, I developed the \pkg{rix} which I will also present in this
+article.
 
-\subsection{Prior art}\label{subsec-prior}
+\section{The Nix package manager}\label{the-nix-package-manager}
 
-\proglang{R} provides a very flexible implementation of the general GLM
-framework in the function \fct{glm} \citet{ChambersHastie1992} in the
-\pkg{stats} package. Its most important arguments are
+\proglang{Nix} is a package manager that can be used to install and
+build software in a completely reproducible manner. As of writing, it
+contains more than 80.000 packages, and the entirety of CRAN and
+Bioconductor is available through \proglang{Nix}'s repositories. This
+means that using \proglang{Nix}, it is possible to install not only
+\proglang{R}, but also all the packages required for a project. The
+reason why one should use \proglang{Nix} to install \proglang{R}
+packages, and not use the usual, built-in, \texttt{install.packages()}
+\proglang{R} function instead, is that when installing a package with
+\proglang{Nix}, \proglang{Nix} makes sure to install every dependency of
+every package, whether this dependency is another \proglang{R} package
+or a system-level dependency.
 
-\begin{verbatim}
-glm(formula, data, subset, na.action, weights, offset,
-  family = gaussian, start = NULL, control = glm.control(…),
-  model = TRUE, y = TRUE, x = FALSE, …)
-\end{verbatim}
+For example, the \pkg{xlsx} \proglang{R} package requires the
+\proglang{Java} programming language to be installed to successfully
+install. Depending on the what system one tries to install \pkg{xlsx},
+installing \proglang{Java} might not easy, nor even possible. But with
+\proglang{Nix}, it suffices to declare that the \pkg{xlsx} package is
+needed for the project, and \proglang{Nix} figures out automatically
+that \proglang{Java} is required and installs and configures it. It all
+just happens without any required intervention from the user.
 
-where \texttt{formula} plus \texttt{data} is the now standard way of
-specifying regression relationships in \proglang{R}/\proglang{S}
-introduced in \citet{ChambersHastie1992}. The remaining arguments in the
-first line (\texttt{subset}, \texttt{na.action}, \texttt{weights}, and
-\texttt{offset}) are also standard for setting up formula-based
-regression models in \proglang{R}/\proglang{S}. The arguments in the
-second line control aspects specific to GLMs while the arguments in the
-last line specify which components are returned in the fitted model
-object (of class \class{glm} which inherits from \class{lm}). For
-further arguments to \fct{glm} (including alternative specifications of
-starting values) see \texttt{?glm}. For estimating a Poisson model
-\texttt{family\ =\ poisson} has to be specified.
+But where do these packages come from? When installing a package using
+\proglang{Nix}, an expression written in the \proglang{Nix} programming
+language gets downloaded from the \texttt{nixpkgs} Github repository and
+evaluated. This expression contains a so-called \emph{derivation}. A
+derivation defines a build: its dependencies, commands to build and
+install the package in question, and then an output. Most of the time, a
+derivation downloads source code, builds the software from the source
+and then outputs a compiled binary. Derivations are extremely flexible,
+and by writing one's own, it is possible to define and build environment
+for a project in a reproducible manner.
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, breakable, left=2mm, colback=white, arc=.35mm, rightrule=.15mm, toprule=.15mm, bottomrule=.15mm]
+Because the whole set of \proglang{Nix} expression is hosted on Github,
+it is possible to pin a specific revision of \texttt{nixpkgs} to ensure
+reproducibility of our project. Pinning a revision ensures that every
+package that Nix installs will always be at exactly the same versions,
+regardless of when in the future the packages get installed.
 
-As the synopsis above is a code listing that is not meant to be
-executed, one can use either the dedicated \texttt{\{Code\}} environment
-or a simple \texttt{\{verbatim\}} environment for this. Again, spaces
-before and after should be avoided.
-
-Finally, there might be a reference to a \texttt{\{table\}} such as
-Table~\ref{tbl-overview}. Usually, these are placed at the top of the
-page (\texttt{{[}t!{]}}), centered (\texttt{\textbackslash{}centering}),
-with a caption below the table, column headers and captions in sentence
-style, and if possible avoiding vertical lines.
-
-\end{tcolorbox}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.6351}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Type
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Distribution
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Method
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Description
-\end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Type
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Distribution
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Method
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Description
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-GLM & Poisson & ML & Poisson regression: classical GLM, estimated by
-maximum likelihood (ML) \\
-& & Quasi & ``Quasi-Poisson regression'\,': same mean function,
-estimated by quasi-ML (QML) or equivalently generalized estimating
-equations (GEE), inference adjustment via estimated dispersion
-parameter \\
-& & Adjusted & ``Adjusted Poisson regression'\,': same mean function,
-estimated by QML/GEE, inference adjustment via sandwich covariances \\
-& NB & ML & NB regression: extended GLM, estimated by ML including
-additional shape parameter \\
-Zero-augmented & Poisson & ML & Zero-inflated Poisson (ZIP), hurdle
-Poisson \\
-& NB & ML & Zero-inflated NB (ZINB), hurdle NB \\
-\caption{Overview of various count regression models. The table is
-usually placed at the top of the page (\texttt{{[}t!{]}}), centered
-(\texttt{centering}), has a caption below the table, column headers and
-captions are in sentence style, and if possible vertical lines should be
-avoided.}\label{tbl-overview}\tabularnewline
-\end{longtable}
-
-\section{Illustrations}\label{sec-illustrations}
-
-For a simple illustration of basic Poisson and NB count regression the
-\texttt{quine} data from the \pkg{MASS} package is used. This provides
-the number of \texttt{Days} that children were absent from school in
-Australia in a particular year, along with several covariates that can
-be employed as regressors. The data can be loaded by
-
-\begin{verbatim}
-R> data(mtcars)
-\end{verbatim}
-
-and a basic frequency distribution of the response variable is displayed
-in \textbf{?@fig-quine}.
-
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, breakable, left=2mm, colback=white, arc=.35mm, rightrule=.15mm, toprule=.15mm, bottomrule=.15mm]
-
-For code input and output, the style files provide dedicated
-environments. Either the ``agnostic'' \texttt{\{CodeInput\}} and
-\texttt{\{CodeOutput\}} can be used or, equivalently, the environments
-\texttt{\{Sinput\}} and \texttt{\{Soutput\}} as produced by \fct{Sweave}
-or \pkg{knitr} when using the \texttt{render\_sweave()} hook. Please
-make sure that all code is properly spaced, e.g., using
-\texttt{y\ =\ a\ +\ b\ *\ x} and \emph{not} \texttt{y=a+b*x}. Moreover,
-code input should use ``the usual'' command prompt in the respective
-software system. For \proglang{R} code, the prompt
-\texttt{R\textgreater{}} should be used with \texttt{+} as the
-continuation prompt. Generally, comments within the code chunks should
-be avoided -- and made in the regular {\LaTeX} text instead. Finally,
-empty lines before and after code input/output should be avoided (see
-above).
-
-\end{tcolorbox}
-
-\section{Summary and discussion}\label{sec-summary}
-
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, breakable, left=2mm, colback=white, arc=.35mm, rightrule=.15mm, toprule=.15mm, bottomrule=.15mm]
-
-As usual\ldots{}
-
-\end{tcolorbox}
-
-\section*{Computational details}\label{computational-details}
-\addcontentsline{toc}{section}{Computational details}
-
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, breakable, left=2mm, colback=white, arc=.35mm, rightrule=.15mm, toprule=.15mm, bottomrule=.15mm]
-
-If necessary or useful, information about certain computational details
-such as version numbers, operating systems, or compilers could be
-included in an unnumbered section. Also, auxiliary packages (say, for
-visualizations, maps, tables, \ldots) that are not cited in the main
-text can be credited here.
-
-\end{tcolorbox}
-
-The results in this paper were obtained using
-\proglang{R}\textasciitilde3.4.1 with the
-\pkg{MASS}\textasciitilde7.3.47 package. \proglang{R} itself and all
-packages used are available from the Comprehensive \proglang{R} Archive
-Network (CRAN) at {[}https://CRAN.R-project.org/{]}.
+With Nix, it is essentially possible to replace \pkg{renv} and Docker
+combined, or in the case of \proglang{Python}, replace
+\texttt{requirements.txt} files which list package dependencies for
+\proglang{Python} projects. It is also possible to build multi-language
+environments, containing R and Python, a LaTeX distribution, and even
+install one's favourite text editor to edit the project's source code.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, breakable, left=2mm, colback=white, arc=.35mm, rightrule=.15mm, toprule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, rightrule=.15mm, arc=.35mm, leftrule=.75mm, toprule=.15mm, bottomrule=.15mm, colback=white, opacityback=0]
 

### 2023-12-22 paper.tex
rendered file

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, frame hidden, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 \makeatother
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, rightrule=.15mm, arc=.35mm, leftrule=.75mm, toprule=.15mm, bottomrule=.15mm, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, left=2mm, toprule=.15mm, colback=white, leftrule=.75mm, breakable, rightrule=.15mm, bottomrule=.15mm, opacityback=0]
 

### 2023-12-22 paper.qmd
GA
added an example of a dev env
added rix:
addel library(rix)
overwrite
I > we

@@
 installing and building software in a reproducible manner. To make
-[Nix]{.proglang} more accessible to [R]{.proglang} programmers, I developed the
-[rix]{.pkg} which I will also present in this article.
+[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
+[rix]{.pkg} which we will present in this article.
 
-## The [Nix]{.proglang} package manager
+## The Nix package manager
 
@@
 project's source code.
+
+## Reproducible development environments with Nix
+
+```{r}
+library(rix)
+
+rix(r_ver = "4.2.2",
+    r_pkgs = c("dplyr", "chronicler"),
+    project_path = tempdir(),
+    print = TRUE)
+```
 

### 2023-12-23 paper.qmd
expression that install latest R

@@
 
-As an alternative to the above approach, I will present the [Nix]{.proglang}
+As an alternative to the above approach, we will present the [Nix]{.proglang}
 package manager, which available for all major operating system and focuses on
@@
 ## Reproducible development environments with Nix
+
+Nix expressions are written in the [Nix]{.proglang} programming language,
+which is a purely functional programming language. Here is a simple expression
+that builds a shell environment containing version 4.3.1 of R:
+
+```{r, echo = F}
+library(rix)
+
+rix(r_ver = "4.3.1",
+    project_path = tempdir(),
+    print = TRUE)
+```
 

### 2023-12-23 paper.qmd
added overwrite equals T

@@
     project_path = tempdir(),
+    overwrite = TRUE,
     print = TRUE)
 ```
+
+To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so
+the first lines of this expression define a variable called `pkgs` and
+`system_pkgs`. `pkgs` is defined as the set of packages to be used in
+this derivation. In this example, the package expressions are going to be pulled
+from the `nixpkgs`
 
@@
     project_path = tempdir(),
+    overwrite = TRUE,
     print = TRUE)

### 2023-12-23 paper.qmd
more on rix

@@
 
-```{r, echo = F}
+```
+let
+  pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/976fa3369d722e76f37c77493d99829540d43845.tar.gz") {};
+  system_packages = builtins.attrValues {
+    inherit (pkgs) R ;
+  };
+in
+  pkgs.mkShell {
+    buildInputs = [ system_packages ];
+    shellHook = "R --vanilla";
+  }
+```
+
+To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so the
+first lines of this expression define a variable called `pkgs` and
+`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
+derivation. In this example, the package expressions are going to be pulled from
+the official `nixpkgs` Github respository, and in particular the commit starting
+with `976fa336` is going to be used. It sould be noted that is thus possible to
+use a personal fork of `nixpkgs` instead of the official repository.
+`system_pkgs` lists the packages that we wish to install and use with this
+environment, in this example only the [R]{.proglang} programming language will
+be installed, alongside all the required dependecnies to actually install and
+run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
+containing the set of packages defined by the variable `system_packages` and its
+`shellHook` is set to `R --vanilla`: this means that when entering the shell
+[R]{.proglang} will be started with the `--vanilla` flag (thus ignore any
+startup options).
+
+This expression can be saved in a file called `default.nix`, and the shell can
+be built using the `nix-build` command. Once `nix-build` is done building the
+development environment according to the specifications declared in the
+`default.nix` file, the user can enter the development shell using `nix-shell`.
+The user will thus be dropped into an interactive shell with all the packages
+declared in the `default.nix` file and can now use this shell for work.
+
+Obviously, writing such an expression will be quite complicated for someone that
+is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
+of being able to define a completely reproducible development environment in a
+single text file and then use that file to build and use that environment
+anywhere is quite interesting. Because we thought this, we wrote the [rix]{.pkg}
+which comes with the `rix()` function that helps generate such expressions quite
+easily. For example, here is how one could use [rix]{.pkg} to generate a
+`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
+[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
+expression here because of space):
+
+```{r, eval = F}
 library(rix)
@@
 rix(r_ver = "4.3.1",
+    r_pkgs = c("dplyr", "chronicler"),
     project_path = tempdir(),
@@
 
-To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so
-the first lines of this expression define a variable called `pkgs` and
-`system_pkgs`. `pkgs` is defined as the set of packages to be used in
-this derivation. In this example, the package expressions are going to be pulled
-from the `nixpkgs`
+Here, users do not need to look for the right commit of `nixpkgs`: it is enough
+to specify the version of [R]{.proglang} they wish to use and the packages that
+were current at the time will get used. But [rix]{.pkg} is quite flexible and
+makes it easy to install specific versions of packages as well (which will get
+installed from the CRAN archives) as well as packages hosted on Github and
+\LaTeX packages as well:
 
-```{r}
-library(rix)
+```{r, eval = F}
+rix(r_ver = "4.3.1",
+    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
+    system_pkgs = c("quarto", "git"),
+    tex_pkgs = c(
+          "amsmath",
+          "framed",
+          "fvextra",
+          "environ",
+          "fontawesome5",
+          "orcidlink",
+          "pdfcol",
+          "tcolorbox",
+          "tikzfill"
+    ),
+    git_pkgs = list(
+                 list(package_name = "rix",
+                   repo_url = "https://github.com/b-rodrigues/rix/",
+                   branch_name = "master",
+                   commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
+                 list(package_name = "fusen",
+                   repo_url = "https://github.com/ThinkR-open/fusen",
+                   branch_name = "main",
+                   commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
+    ),
+    ide = "rstudio",
+    project_path = tempfile(),
+    overwrite = TRUE)
+```
 
-rix(r_ver = "4.2.2",
-    r_pkgs = c("dplyr", "chronicler"),
-    project_path = tempdir(),
-    overwrite = TRUE,
-    print = TRUE)
-```
+The above call to `rix()` generates an expression that installs several
+[R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
+older version of the package. If the user didn't not specify the version, the
+version available as of the `nixpkgs` commit would be used), but also install
+several TeXLive packages for authoring documents with \LaTeX, the development
+versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
+the RStudio integrated development environment.
 

### 2023-12-23 paper.qmd
appendix

@@
 
-## The Nix package manager
+## The Nix package manager {#sec-nix}
 
@@
 
-## Reproducible development environments with Nix
+## Reproducible development environments with Nix {#sec-repro-nix}
 
@@
 [R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
-older version of the package. If the user didn't not specify the version, the
+older version of the package. Had the user not specified the version, the
 version available as of the `nixpkgs` commit would be used), but also install
@@
 versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
-the RStudio integrated development environment.
+the RStudio integrated development environment. It should be noted that as of
+writing, RStudio is not available on macOS through `nixpkgs`, so the above
+expression will only work on Windows and Linux distributions. To make the
+expression work on macOS, the "rstudio" string could be replaced by "code",
+which would install VS Code instead, or "other", to use any other editor
+such as Emacs or Vi.
 
@@
 
-## More technical details {#sec-techdetails .unnumbered}
+## Appendix {#sec-appendix .unnumbered}
+
+Here is the Nix expression generated from the call to the `rix()`
+function of section #sec-repro-nix:
+
+```{r}
+rix(r_ver = "4.3.1",
+    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
+    system_pkgs = c("quarto", "git"),
+    tex_pkgs = c(
+          "amsmath",
+          "framed",
+          "fvextra",
+          "environ",
+          "fontawesome5",
+          "orcidlink",
+          "pdfcol",
+          "tcolorbox",
+          "tikzfill"
+    ),
+    git_pkgs = list(
+                 list(package_name = "rix",
+                   repo_url = "https://github.com/b-rodrigues/rix/",
+                   branch_name = "master",
+                   commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
+                 list(package_name = "fusen",
+                   repo_url = "https://github.com/ThinkR-open/fusen",
+                   branch_name = "main",
+                   commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
+    ),
+    ide = "rstudio",
+    project_path = tempfile(),
+    overwrite = TRUE)
+```
 

### 2023-12-23 paper.qmd
github actions

@@
 single text file and then use that file to build and use that environment
-anywhere is quite interesting. Because we thought this, we wrote the [rix]{.pkg}
-which comes with the `rix()` function that helps generate such expressions quite
-easily. For example, here is how one could use [rix]{.pkg} to generate a
-`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
-[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
-expression here because of space):
+anywhere is quite interesting. Because we believed that lowering the cost of
+writing Nix expressions would lead to a wider adoption of the Nix package
+manager for reproducibility, we wrote the [rix]{.pkg} which comes with the
+`rix()` function that helps generate such expressions quite easily. For example,
+here is how one could use [rix]{.pkg} to generate a `default.nix` file that
+installs version 4.3.1 of [R]{.proglang} and the [dplyr]{.pkg} and
+[chronicler]{.pkg} packages (we don't show the generated expression here because
+of space):
 
@@
 
+[rix]{.pkg} also provides installation intructions for Nix, but [rix]{.pkg} can
+generate Nix expression even if Nix is not installed on one's system. This is
+quite useful if one wants to generate an expression and use that expression in a
+CI/CD environment such as Github Actions. For example, the repository that
+contains the source code this article^[https://github.com/b-rodrigues/rix_paper]
+uses Github Actions for compilation. Essentially, each time one of the authors
+pushed to the master branch, a Github Actions runner would be started, Nix would
+get installed on said runner, and an environment containing the right packages
+was generated from a `default.nix` file that is hosted on the same repository.
+The paper would then get compiled in this reproducible environment.
+
 ## Acknowledgments {.unnumbered}
@@
 ```{r}
+library(rix)
+
 rix(r_ver = "4.3.1",

### 2023-12-23 paper.qmd
targets github actions

@@
 
+If one wishes to run pipelines written using the [targets]{.pkg} [@landau2021]
+on Github Actions, [rix]{.pkg} featrues a function called `tar_nix_ga()`,
+inspired by the `tar_github_actions()` function from the [targets]{.pkg}
+package. This function writes a YAML file under `.github/workflows/` which then
+runs the [targets]{.pkg} pipeline using Github Actions at each push, in a
+completely reproducible environment. For such an example, consult the
+code in this repository^[https://github.com/b-rodrigues/rix_cicd].
+
 ## Acknowledgments {.unnumbered}

### 2023-12-23 paper.qmd
added print

@@
     project_path = tempfile(),
+    print = TRUE,
     overwrite = TRUE)

### 2023-12-23 paper.qmd
removed overwrite

--- 2023-12-23 paper.qmd
@@
     project_path = tempfile(),
-    overwrite = TRUE)
+    print = TRUE)
 ```

### 2023-12-23 paper.qmd
tempfile -> tempdir

@@
     r_pkgs = c("dplyr", "chronicler"),
-    project_path = tempdir(),
+    project_path = ".",
     overwrite = TRUE,
@@
     ide = "rstudio",
-    project_path = tempfile(),
+    project_path = ".",
     overwrite = TRUE)
@@
     ide = "rstudio",
-    project_path = tempfile(),
+    project_path = tempdir(),
+    overwrite = TRUE,
     print = TRUE)

### 2023-12-28 paper.qmd
typos and rewriting

@@
----
+--
 title: "Reproducible development environments with rix"
@@
   it is not enough to write clean code and document it well. One
-  must also make sure to list all the dependencies of the analysis
-  clearly and ideally provide an easy way to install said dependencies.
+  must also make sure to clear list all the dependencies of the analysis
+  and ideally provide an easy way to install them.
   There are several tools that can be used to list dependencies and to
   make them easily installable by someone that wishes to reproduce a study,
-  such as [Docker]{.proglang}, a containerization solution. This paper will present
-  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that lowers
-  Nix's learning curve for users of the [R]{.proglang} programming language.
+  such as [Docker]{.proglang}, a containerization solution. This paper will
+  present another tool that can be used to make projects reproducible,
+  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that
+  makes using Nix much simpler for users of [R]{.proglang}.
 
@@
 the ends of this continuum, we only have access to the paper describing the
-studies, which is not reproducible at all. Then, if in addition, to this paper
-we make the original source code of the analysis that was written to compute the
-results of the study available, reproducibility is improved, albeit only by a
-little. Adding the original data improves reproducibility yet again. Finally, if
-to all this we add what Roger Peng named the *linked and executable code and
-data*, we reach the gold standard of full replication.
+studies, which is not reproducible at all. If in addition, to this paper authors
+make the original source code of the analysis available, reproducibility is
+improved, albeit only by a little. Adding the original data improves
+reproducibility yet again. Finally, if to all this the authors add what Roger
+Peng named the *linked and executable code and data*, we reach the gold standard
+of reproducibility.
 
@@
 analysis was done, even though it has been shown that running the same analysis
-with the same software but on different operating systems could lead to different
-results, as described in @neupane2019. Authors also only very rarely provide
-instructions to install the required tools and software in order to reproduce
-their studies.
+with the same software but on different operating systems could lead to
+different results, as described in @neupane2019. Authors also only very rarely
+provide instructions to install the required tools and software in order to
+reproduce their studies.
 
-Tools can be used at each of these steps to reach the gold standard of full
-replication. Let's first consider the task of listing the software used.
-[R]{.proglang} provides the `sessionInfo()` function whose output can be saved
-into a file. Below is an example output of `sessionInfo()`:
+There are tools available that authors can use to reach the gold stardard of a
+fully reproducible study. Let us first consider the task of listing the software
+used. [R]{.proglang} provides the `sessionInfo()` function whose output can be
+saved into a file. Below is an example output of `sessionInfo()`:
 
@@
 If an author provides this information, other people trying to reproduce the
-study (or the author him- or herself in the future) can read this file
-and see which version of [R]{.proglang} was used, and which packages (and their
+study (or the author him- or herself in the future) can read this file and see
+which version of [R]{.proglang} was used, and which packages (and their
 versions) were used as well. However, others would still need to install the
-correct software themselves. An alternative to this is instead to use the [renv]{.pkg}
-package which generates a so-called `renv.lock` file which also lists
-[R]{.proglang} and package versions. Here is an example of such an `renv.lock` file:
+correct software themselves. An alternative to this is instead to use the
+[renv]{.pkg} package which generates a so-called `renv.lock` file which also
+lists [R]{.proglang} and package versions. Here is an example of such an
+`renv.lock` file:
 
@@
 
-    ***and many more packages***
+    ***lines below omitted***
 ```
@@
 by running `renv::restore()`. The exact same packages get installed in an
-isolated, project-specific, library which doesn't interfere with the other,
-main, library of the user. [renv]{.pkg} does not restore the [R]{.proglang}
-itself though, so installing the right version of R needs to be handled
+isolated, project-specific, library which doesn't interfere with the main
+library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
+used though, so installing the right version of R needs to be handled
 separately. Before continuing, it should be noted that other packages exist
 which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
-by @simonsohn2023 which makes it rather easy to install packages as they were
-on CRAN at a given date. For example, the code snippet below install the
+by @simonsohn2023 which makes it rather easy to install packages as they were on
+CRAN at a given date. For example, the code snippet below installs the
 [purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:
@@
 These packages also get installed in a project-specfic library so there is no
-interferences between these packages and other versions of the same packages
-that one might use for other projects. Because [groundhog]{.pkg} does not
-install [R]{.proglang} itself, users should either install the required version
+interference between these packages and other versions of the same packages that
+one might use for other projects. Because [groundhog]{.pkg} does not install
+[R]{.proglang} itself, users should either install the required version
 themselves, or they should use the `tolerate.R.version` argument as shown in the
 example above. Otherwise, [groundhog]{.pkg} would not continue with the
-installation of the packages. Another such package, developed by @chan2023 is
-[rang]{.pkg}, which also installs packages as they were on a given date.
-Yet another way to install packages as they were on a give date is to use
-the Posit Package Manager, which provides snapshots of CRAN. For example,
-to install the required packages for an analysis as they were on the 30th of June
+installation of the packages. Another alternative to [renv]{.pkg}, developed by
+@chan2023 is [rang]{.pkg}, which also installs packages as they were on a given
+date. Yet another way to install packages as they were on a give date is to use
+the Posit Package Manager, which provides snapshots of CRAN. For example, to
+install the required packages for an analysis as they were on the 30th of June
 2023, one could add the following line to the `.Rprofile` file:
@@
 that every call to the `install.packages()` function will now install the
-packages from this snapshotted mirror.
+packages from this snapshotted mirror. However, unless the user explicitely and
+manually manages the different projects' libraries, using the Posit Package
+Manager will install all the package for all the projects at the given date.
 
@@
 version of [R]{.proglang}. Of course, it would be possible to install the right
-version manually, but here too, there are tools that simplify the process such
-as [rig]{.proglang} by the R infrastructure team [-@rlib2023]. One could thus
-install the right version of [R]{.proglang} using [rig]{.proglang} allows to
-easily install different versions of [R]{.proglang}, one could use to install
-the right version to reproduce an analysis, and then use one of the listed
-packages above to install the right library of of [R]{.proglang} packages.
-However, this involves many manual steps and is thus error prone. It is also
-time-consuming to do so, especially if one wants to reproduce many studies.
+version manually, but here too, there are tools that simplify the process. One
+such tool is [rig]{.proglang} by the R infrastructure team [-@rlib2023].
+[rig]{.proglang} makes it easy to install different versions of [R]{.proglang},
+so one could use it to install the right version to reproduce an analysis, and
+then use one of the listed packages above to install the right library of of
+[R]{.proglang} packages. However, this involves many manual steps and is thus
+error prone. It is also time-consuming to do so, especially if one wants to
+reproduce many studies.
 
-The final step towards the gold standard of reproducibility is to package the
-right version of [R]{.proglang} and [R]{.proglang} packages inside a
+The final tool one could use to reach the gold standard of reproducibility is to
+package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
 [Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
 [Docker]{.proglang} it is possible to package some *data product* with its
-dependencies into an image. From this image, containers can be run to reproduce
-the data product with its exact dependencies. A statistical analysis, from the
-simplest to the most complex one, can be seen as such a data product that comes
-with many software dependencies. *Dockerizing* an analysis consists in first
-building an image: in the build step, the dependencies of the analysis have to
-be installed and this can be achieved using the tools mentioned above. The
-scripts to run the analysis and the data are also added to the image at build
-time. Instances of that image, called containers, can then be executed, which
-give access to the exact development environment originally used for the study.
-It is also possible to script the building of the analysis itself, such that in
-one single call to the `docker run` command, a complete analysis is reproduced.
+dependencies into an image. A statistical analysis, from the simplest to the
+most complex one, can be seen as such a data product that comes with many
+software dependencies. *Dockerizing* an analysis consists in first building an
+image: in the build step, the dependencies of the analysis have to be installed
+and this can be achieved using the tools mentioned above. The scripts to run the
+analysis and the data are also added to the image at build time. Instances of
+that image, called containers, can then be executed, which give access to the
+exact development environment originally used for the study. This execution step
+is called the "run step". During this run step, the analysis scripts can be
+executed to product the output.
+
+The advantage of using [Docker]{.proglang} is not only about having the right
+version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
+analysis scripts: [Docker]{.proglang} images are actually almost complete
+operating systems, which means that system-level dependecies also get bundled
+into the image. This is especially useful if packages that rely on external
+libraries get used for the analysis, as these external libraries are thus also
+made available to future reproducers of the study.
 
 The Rocker project initiated by @boettiger2017 provides many pre-built
-[Docker]{.proglang} to the [R]{.proglang} community of users. These images
-can be used as bases to build other images containing statistical analyses
-more easily than starting from a bare-bones image.
+[Docker]{.proglang} to the [R]{.proglang} community of users, already containing
+specific [R]{.proglang} versions and even packages. These images can be used as
+bases to build other images containing statistical analyses more easily than
+starting from a bare-bones image.
 
@@
 such as [Make]{.proglang} to run the whole analysis when the container is
-executed. Build automation tools make it easier to run arbitrary code in a
-series of well-defined steps. [R]{.proglang} programmers can use the
-[targets]{.pkg} by @landau2021 as a build automation tool.
+executed, instead of executing scripts. Build automation tools make it easier to
+run arbitrary code in a series of well-defined steps. [R]{.proglang} programmers
+can use the [targets]{.pkg} by @landau2021 as a build automation tool.
 
@@
 programming language. A very similar approach should be taken if one uses
-[Python]{.proglang} for statistical analysis instead.
+[Python]{.proglang} for statistical analysis instead. Furthermore, the previous
+discussion also implies that for an analysis to actually be reproducible at all,
+it must be made using free and open source software, as proprietary software
+cannot be freely shared with others that wish to reproduce a study, and there's
+no guarantee that others will have access to the proprietary software used.
 
 As an alternative to the above approach, we will present the [Nix]{.proglang}
-package manager, which available for all major operating system and focuses on
-installing and building software in a reproducible manner. To make
-[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
-[rix]{.pkg} which we will present in this article.
+package manager, which is available for all major operating systems and focuses
+on installing and building software in a reproducible manner. This means that
+this single tool can replace not only [Docker]{.proglang} but also [renv]{.pkg}
+and a build automation tool such as [Make]{.proglang}. To make [Nix]{.proglang}
+more accessible to [R]{.proglang} programmers, we developed the [rix]{.pkg}
+package which we will present in this article.
 
@@
 software in a completely reproducible manner. As of writing, it contains more
-than 80.000 packages, and the entirety of CRAN and Bioconductor is available
-through [Nix]{.proglang}’s repositories. This means that using [Nix]{.proglang},
-it is possible to install not only [R]{.proglang}, but also all the packages
-required for a project. The reason why one should use [Nix]{.proglang} to
-install [R]{.proglang} packages, and not use the usual, built-in,
-`install.packages()` [R]{.proglang} function instead, is that when installing a
-package with [Nix]{.proglang}, [Nix]{.proglang} makes sure to install every
-dependency of every package, whether this dependency is another [R]{.proglang}
-package or a system-level dependency.
+than 80.000 packages, including the entirety of CRAN and Bioconductor. This
+means that using [Nix]{.proglang}, it is possible to install not only
+[R]{.proglang}, but also all the packages required for a project. The reason why
+one should use [Nix]{.proglang} to install [R]{.proglang} packages, and not use
+the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
+that when installing a package with [Nix]{.proglang}, [Nix]{.proglang} makes
+sure to install every dependency of every package, whether this dependency is
+another [R]{.proglang} package or a system-level dependency.
 
@@
 [Java]{.proglang} programming language to be installed to successfully install.
-Depending on the what system one tries to install [xlsx]{.pkg}, installing
-[Java]{.proglang} might not easy, nor even possible. But with [Nix]{.proglang},
-it suffices to declare that the [xlsx]{.pkg} package is needed for the project,
-and [Nix]{.proglang} figures out automatically that [Java]{.proglang} is
-required and installs and configures it. It all just happens without any
-required intervention from the user.
+Depending on what system one tries to install [xlsx]{.pkg}, installing
+[Java]{.proglang} might not be easy, nor even possible. But with
+[Nix]{.proglang}, it suffices to declare that the [xlsx]{.pkg} package is needed
+for the project, and [Nix]{.proglang} figures out automatically that
+[Java]{.proglang} is required and installs and configures it. It all just
+happens without any required intervention from the user.
 
@@
 extremely flexible, and by writing one's own, it is possible to define and build
-environment for a project in a reproducible manner.
+an environment for a project in a reproducible manner.
 
-Because the whole set of [Nix]{.proglang} expression is hosted on Github, it is
-possible to pin a specific revision of `nixpkgs` to ensure reproducibility of
-our project. Pinning a revision ensures that every package that Nix installs
-will always be at exactly the same versions, regardless of when in the future
-the packages get installed.
+But why is installing software with [Nix]{.proglang} reproducible? Because the
+whole set of [Nix]{.proglang} expressions is hosted on Github, it is possible to
+use a specific commit of `nixpkgs` to ensure reproducibility of our project.
+Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
+Pinning a revision ensures that every package that Nix installs will always be
+at exactly the same versions, regardless of when in the future the packages get
+built. This is because the [Nix]{.proglang} expressions that get downloaded when
+pinning are the expressions as they were at that commit of `nixpkgs`.
 
-With Nix, it is essentially possible to replace [renv]{.pkg} and Docker
-combined, or in the case of [Python]{.proglang}, replace `requirements.txt`
-files which list package dependencies for [Python]{.proglang} projects. It is
-also possible to build multi-language environments, containing R and Python, a
-LaTeX distribution, and even install one's favourite text editor to edit the
-project's source code.
+With Nix, it is essentially possible to replace [renv]{.pkg} and
+[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
+`requirements.txt` files which list package dependencies for [Python]{.proglang}
+projects. It is also possible to build multi-language environments, containing R
+and Python, a LaTeX distribution, and even install one's favorite text editor to
+edit the project's source code. Essentially, one can define a complete
+reproducible environment and use it to work on a project interactively, or use
+that environment non-interactively to run an analysis sometime in the future.
 
@@
 
-Nix expressions are written in the [Nix]{.proglang} programming language,
+[Nix]{.proglang} expressions are written in the [Nix]{.proglang} programming language,
 which is a purely functional programming language. Here is a simple expression
-that builds a shell environment containing version 4.3.1 of R:
+that builds a shell environment containing version 4.3.1 of [R]{.proglang}:
 
@@
 derivation. In this example, the package expressions are going to be pulled from
-the official `nixpkgs` Github respository, and in particular the commit starting
-with `976fa336` is going to be used. It sould be noted that is thus possible to
+the official `nixpkgs` Github repository, and in particular the commit starting
+with `976fa336` is going to be used. It should be noted that is also possible to
 use a personal fork of `nixpkgs` instead of the official repository.
@@
 environment, in this example only the [R]{.proglang} programming language will
-be installed, alongside all the required dependecnies to actually install and
+be installed, alongside all the required dependencies to actually install and
 run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
 containing the set of packages defined by the variable `system_packages` and its
-`shellHook` is set to `R --vanilla`: this means that when entering the shell
-[R]{.proglang} will be started with the `--vanilla` flag (thus ignore any
+`shellHook` is set to `R --vanilla`: this means that when entering the shell,
+[R]{.proglang} will be started with the `--vanilla` flag (thus ignoring any
 startup options).
@@
 This expression can be saved in a file called `default.nix`, and the shell can
-be built using the `nix-build` command. Once `nix-build` is done building the
-development environment according to the specifications declared in the
-`default.nix` file, the user can enter the development shell using `nix-shell`.
-The user will thus be dropped into an interactive shell with all the packages
-declared in the `default.nix` file and can now use this shell for work.
+be built using the `nix-build` command on a system where [Nix]{.proglang} is
+available.^[To install [Nix]{.proglang} we recommend the installer from
+Determinate systems which can be found on the following link:
+\url{https://determinate.systems/posts/determinate-nix-installer}.] Once
+`nix-build` is done building the development environment according to the
+specifications declared in the `default.nix` file, the user can enter the
+development shell using `nix-shell`. The user will thus be dropped into an
+interactive shell with all the packages declared in the `default.nix` file and
+can now use this shell for work. This is similar to activating environments in
+the [Python]{.proglang} ecosystem.
 
@@
 single text file and then use that file to build and use that environment
-anywhere is quite interesting. Because we believed that lowering the cost of
-writing Nix expressions would lead to a wider adoption of the Nix package
-manager for reproducibility, we wrote the [rix]{.pkg} which comes with the
-`rix()` function that helps generate such expressions quite easily. For example,
-here is how one could use [rix]{.pkg} to generate a `default.nix` file that
-installs version 4.3.1 of [R]{.proglang} and the [dplyr]{.pkg} and
-[chronicler]{.pkg} packages (we don't show the generated expression here because
-of space):
+anywhere is quite appealing. Because we believe that lowering the cost of
+writing [Nix]{.proglang} expressions would lead to a wider adoption of the
+[Nix]{.proglang} package manager for reproducibility, we wrote [rix]{.pkg}
+which comes with the `rix()` function that helps generate such expressions quite
+easily. For example, here is how one could use [rix]{.pkg} to generate a
+`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
+[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
+expression here because of space):
 
@@
 expression work on macOS, the "rstudio" string could be replaced by "code",
-which would install VS Code instead, or "other", to use any other editor
-such as Emacs or Vi.
+which would install VS Code instead, or "other", to use any other editor such as
+Emacs or Vi. You can see the resulting `default.nix` file in the @sec-appendix.
 
-[rix]{.pkg} also provides installation intructions for Nix, but [rix]{.pkg} can
-generate Nix expression even if Nix is not installed on one's system. This is
-quite useful if one wants to generate an expression and use that expression in a
-CI/CD environment such as Github Actions. For example, the repository that
-contains the source code this article^[https://github.com/b-rodrigues/rix_paper]
-uses Github Actions for compilation. Essentially, each time one of the authors
-pushed to the master branch, a Github Actions runner would be started, Nix would
-get installed on said runner, and an environment containing the right packages
-was generated from a `default.nix` file that is hosted on the same repository.
-The paper would then get compiled in this reproducible environment.
+[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
+[rix]{.pkg} can generate Nix expression even if Nix is not installed on one's
+system. This is quite useful if one wants to generate an expression and use that
+expression in a CI/CD environment such as Github Actions. For example, the
+repository that contains the source code to this
+article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
+compilation. Essentially, each time one of the authors pushed to the master
+branch, a Github Actions runner would be started, [Nix]{.proglang} would get
+installed on said runner, and an environment containing the right packages was
+generated from a `default.nix` file that is hosted on the same repository. The
+paper would then get compiled in this reproducible environment.
 
-If one wishes to run pipelines written using the [targets]{.pkg} [@landau2021]
-on Github Actions, [rix]{.pkg} featrues a function called `tar_nix_ga()`,
-inspired by the `tar_github_actions()` function from the [targets]{.pkg}
-package. This function writes a YAML file under `.github/workflows/` which then
-runs the [targets]{.pkg} pipeline using Github Actions at each push, in a
-completely reproducible environment. For such an example, consult the
-code in this repository^[https://github.com/b-rodrigues/rix_cicd].
+We also provide a way to start a temporary shell with [R]{.proglang} and
+[rix]{.proglang} available, which makes it possible to start building
+reproducible environments on a system that does provide [Nix]{.proglang} but not
+[R]{.proglang}.
+
+If one wishes to run pipelines written using the [targets]{.pkg} package
+[@landau2021] on Github Actions, [rix]{.pkg} features a function called
+`tar_nix_ga()`, inspired by the `tar_github_actions()` function from the
+[targets]{.pkg} package. This function writes a YAML file under
+`.github/workflows/` which then runs the [targets]{.pkg} pipeline using Github
+Actions at each push, in a completely reproducible environment. For such an
+example, consult the code in this
+repository^[https://github.com/b-rodrigues/rix_cicd].
 

### 2023-12-28 paper.qmd
fixed abstract

@@
---
+---
 title: "Reproducible development environments with rix"

### 2024-01-29 paper.qmd
fixes #2
with_nix

@@
 
-We also provide a way to start a temporary shell with [R]{.proglang} and
-[rix]{.proglang} available, which makes it possible to start building
-reproducible environments on a system that does provide [Nix]{.proglang} but not
-[R]{.proglang}.
+## Additional features of the rix package
 
@@
 repository^[https://github.com/b-rodrigues/rix_cicd].
+
+The final feature that we provide is the ability to run one arbitrary function
+from what we call a *subshell*. Suppose that you are running the latest version
+of R, and that you wish or need to run one particular function from an older
+release of R, or from an older package. Using the `with_nix()` function, it is
+possible to run this single function inside the right envirnoment, and recover
+its output in the main session. To illustrate, consider the following two lines
+of code:
+
+```{r, eval = FALSE}
+set.seed(1234)
+sample(seq(1, 10), 5)
+```
+
+On versions of R released after version 3.6, these two lines return the
+following vector, `r set.seed(1234);sample(seq(1, 10), 5);set.seed(NULL)`, in
+versions prior to 3.6, the result will be different. This is a documented change
+as shown in the R version 3.6 release
+notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]
+
+When trying to reproduce a study made on versions prior to 3.6, a replicator
+will need to take this into account. Using the `with_nix()` function,
+the original results can be replicated even from the latest version of R.
+
+The first step is to create an environment with the old release of R:
+
+```{r}
+path_subshell <- tempdir()
+
+rix(
+  r_ver = "3.5.3",
+  overwrite = TRUE,
+  project_path = path_subshell
+)
+```
+
+We recommend `path_subshell` to be a subfolder of the project. Using `rix()`
+we generate a `default.nix` file that will provide a shell with R version 3.5.3
+when built. We then use `rix_init()`, which generates an `.Rprofile` file in
+the same subfolder as where the subshell's `default.nix` is located. This `.Rprofile`
+file ensures that the R session that will be spawned from this subshell does not
+get any interference from the main session it gets called from:
+
+```{r}
+rix_init(
+   project_path = path_subshell,
+   rprofile_action = "overwrite",
+   message_type = "simple"
+ )
+```
+
+We can now use this shell to generate our vector of random numbers using `with_nix()`:
+
+```{r}
+out_subshell <- with_nix(
+    expr = function(){
+      set.seed(1234)
+      a <- sample(seq(1, 10), 5)
+      set.seed(NULL)
+      return(a)
+    },
+    program = "R",
+    exec_mode = "non-blocking",
+    project_path = path_subshell,
+    message_type = "simple"
+  )
+
+```
+
+The `expr` argument of `with_nix()` can be arbitrarly complex, but must
+be wrapped inside a `function()` call. The value of `out_subshell` is
+`r out_subshell`, as is indeed different than the one generated from a
+recent version of R.
 
@@
 
-```{r}
+```{r, echo = FALSE}
 library(rix)

### 2024-01-30 paper.qmd
load rix
hiding messages

@@
 
+```{r, include = FALSE}
+set.seed(1234)
+out_r <- sample(seq(1, 10), 5)
+set.seedt(NULL)
+```
+
 On versions of R released after version 3.6, these two lines return the
-following vector, `r set.seed(1234);sample(seq(1, 10), 5);set.seed(NULL)`, in
-versions prior to 3.6, the result will be different. This is a documented change
-as shown in the R version 3.6 release
+following vector: `r out_r`, but in versions prior to 3.6, the result will be
+different. This is a documented change as shown in the R version 3.6 release
 notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]
@@
 
-```{r}
+```{r, include = FALSE}
+library(rix)
+
 path_subshell <- tempdir()
@@
 
-We recommend `path_subshell` to be a subfolder of the project. Using `rix()`
-we generate a `default.nix` file that will provide a shell with R version 3.5.3
-when built. We then use `rix_init()`, which generates an `.Rprofile` file in
-the same subfolder as where the subshell's `default.nix` is located. This `.Rprofile`
-file ensures that the R session that will be spawned from this subshell does not
-get any interference from the main session it gets called from:
+```{r, eval = F}
+library(rix)
 
-```{r}
+path_subshell <- tempdir()
+
+rix(
+  r_ver = "3.5.3",
+  overwrite = TRUE,
+  project_path = path_subshell
+)
+```
+
+We recommend `path_subshell` to be a subfolder of the project. Using `rix()` we
+generate a `default.nix` file that will provide a shell with R version 3.5.3
+when built. We then use `rix_init()`, which generates an `.Rprofile` file in the
+same subfolder as where the subshell's `default.nix` is located. This
+`.Rprofile` file ensures that the R session that will be spawned from this
+subshell does not get any interference from the main session it gets called
+from:
+
+```{r, eval = F}
 rix_init(
@@
 
-We can now use this shell to generate our vector of random numbers using `with_nix()`:
+```{r, include = F}
+rix_init(
+   project_path = path_subshell,
+   rprofile_action = "overwrite",
+   message_type = "simple"
+ )
+```
 
-```{r}
+We can now use this shell to generate our vector of random numbers using
+`with_nix()`:
+
+```{r, eval = F}
 out_subshell <- with_nix(
@@
 
-The `expr` argument of `with_nix()` can be arbitrarly complex, but must
-be wrapped inside a `function()` call. The value of `out_subshell` is
-`r out_subshell`, as is indeed different than the one generated from a
-recent version of R.
+```{r, include = FALSE}
+out_subshell <- with_nix(
+    expr = function(){
+      set.seed(1234)
+      a <- sample(seq(1, 10), 5)
+      set.seed(NULL)
+      return(a)
+    },
+    program = "R",
+    exec_mode = "non-blocking",
+    project_path = path_subshell,
+    message_type = "simple"
+  )
+
+```
+
+We should say that `rix_init()` and `with_nix()` print many status messages to
+the console while running, which we don't show here for space reasons.
+
+The `expr` argument of `with_nix()` can be arbitrarly complex, but must be
+wrapped inside a `function()` call. The value of `out_subshell` is `r
+out_subshell`, as is indeed different than the one generated from a recent
+version of R.
 

### 2024-01-30 paper.qmd
typo
review and cite dolstra

@@
 versions) were used as well. However, others would still need to install the
-correct software themselves. An alternative to this is instead to use the
-[renv]{.pkg} package which generates a so-called `renv.lock` file which also
-lists [R]{.proglang} and package versions. Here is an example of such an
-`renv.lock` file:
+correct versions of each package themselves. An alternative to this is instead
+to use the [renv]{.pkg} package which generates a so-called `renv.lock` file
+which also lists [R]{.proglang} and package versions. Here is an example of such
+an `renv.lock` file:
 
@@
 by running `renv::restore()`. The exact same packages get installed in an
-isolated, project-specific, library which doesn't interfere with the main
+isolated, project-specific library which doesn't interfere with the main
 library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
@@
 ```
-options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/
-                               __linux__/jammy/2023-06-30"))
+options(repos =
+  c(REPO_NAME =
+    "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
+  )
+)
 ```
@@
 so one could use it to install the right version to reproduce an analysis, and
-then use one of the listed packages above to install the right library of of
+then use one of the listed packages above to install the right library of
 [R]{.proglang} packages. However, this involves many manual steps and is thus
@@
 dependencies into an image. A statistical analysis, from the simplest to the
-most complex one, can be seen as such a data product that comes with many
-software dependencies. *Dockerizing* an analysis consists in first building an
-image: in the build step, the dependencies of the analysis have to be installed
-and this can be achieved using the tools mentioned above. The scripts to run the
-analysis and the data are also added to the image at build time. Instances of
-that image, called containers, can then be executed, which give access to the
-exact development environment originally used for the study. This execution step
-is called the "run step". During this run step, the analysis scripts can be
-executed to product the output.
+most complex one, can be seen as such a data product that requires many software
+dependencies. *Dockerizing* an analysis consists in first building an image: in
+the build step, the dependencies of the analysis have to be installed and this
+can be achieved using the tools mentioned above. The scripts to run the analysis
+and the data are also added to the image at build time. It is possible to then
+execute the code of a study inside a so-called container, which is an instance
+of an image. Executing, or running, these containers is done non-interactively,
+and so it is possible to instantiate the correct environment to run a study in a
+single command.
 
@@
 operating systems, which means that system-level dependecies also get bundled
-into the image. This is especially useful if packages that rely on external
-libraries get used for the analysis, as these external libraries are thus also
-made available to future reproducers of the study.
+into the image. This is especially useful for packages that rely on external
+libraries to run, as these external libraries are thus also made available to
+future reproducers of the study within the [Docker]{.proglang} image. The image
+can also be shared.
 
@@
 - A `Makefile` was used to run the whole analysis and compile the paper;
-- A `Dockerfile` was used to provide the complete computational environment, including the right version of [R]{.proglang} and run the whole analysis easily.
+- A `Dockerfile` was used to provide the complete computational environment,
+  including the right version of [R]{.proglang} and run the whole analysis
+  easily.
 
@@
 software, including the programming language that was used. Ideally, a build
-automation tool such as [make]{.proglang} should be added to the list. It should
+automation tool such as [Make]{.proglang} should be added to the list. It should
 also be noted that these requirements are not limited to the [R]{.proglang}
@@
 this single tool can replace not only [Docker]{.proglang} but also [renv]{.pkg}
-and a build automation tool such as [Make]{.proglang}. To make [Nix]{.proglang}
-more accessible to [R]{.proglang} programmers, we developed the [rix]{.pkg}
-package which we will present in this article.
+and even a build automation tool such as [Make]{.proglang}. To make
+[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
+[rix]{.pkg} package which we will present in this article.
 
@@
 built. This is because the [Nix]{.proglang} expressions that get downloaded when
-pinning are the expressions as they were at that commit of `nixpkgs`.
+pinning are the expressions as they were at that commit of `nixpkgs`. This is
+not the only reason [Nix]{.proglang} is very useful for reproducibility, but the
+[Nix]{.proglang} package manager itself was designed with reproducibility in
+mind. [Nix]{.proglang} is functional package manager, meaning that it uses
+concepts from functional programming to build software: no-side effects or
+global variables are allowed, and the output is always the same if given the
+same inputs, regardless of where or when a package gets built.
+We refer to @dolstra2004nix for a more technical discussion on [Nix]{.proglang}.
 
@@
 out_r <- sample(seq(1, 10), 5)
-set.seedt(NULL)
+set.seed(NULL)
 ```
@@
 The `expr` argument of `with_nix()` can be arbitrarly complex, but must be
-wrapped inside a `function()` call. The value of `out_subshell` is `r
-out_subshell`, as is indeed different than the one generated from a recent
-version of R.
+wrapped inside a `function()` call. The value of `out_subshell` is `r out_subshell`,
+as is indeed different than the one generated from a recent version of R.
+
+It is possible to use `with_nix()` from an R session that was installed itself
+via Nix, or from a regular R installation.
 

### 2024-02-01 paper.qmd
Add files via upload

@@
 
+But even assuming that authors provided clear instructions to install the exact
+requirements of a study, it might happen that these instructions do not even
+work on different platforms, or on different versions of the same platform. And
+reproducibility is not the only activity that might get impacted by a lack of
+clear instructions: if an author works on several machines, the exact same
+computational environment should be used on all the machines, and the same is
+true when a team of authors collaborate.
+
 There are tools available that authors can use to reach the gold stardard of a
-fully reproducible study. Let us first consider the task of listing the software
-used. [R]{.proglang} provides the `sessionInfo()` function whose output can be
-saved into a file. Below is an example output of `sessionInfo()`:
+fully reproducible study, or to successfully deploy the same development
+environment on several machines. Let us first consider the task of listing the
+software used. [R]{.proglang} provides the `sessionInfo()` function whose output
+can be saved into a file. Below is an example output of `sessionInfo()`:
 
@@
 If an author provides this information, other people trying to reproduce the
-study (or the author him- or herself in the future) can read this file and see
-which version of [R]{.proglang} was used, and which packages (and their
-versions) were used as well. However, others would still need to install the
-correct versions of each package themselves. An alternative to this is instead
-to use the [renv]{.pkg} package which generates a so-called `renv.lock` file
-which also lists [R]{.proglang} and package versions. Here is an example of such
-an `renv.lock` file:
+study (or the author him- or herself in the future), or co-authors, can read
+this file and see which version of [R]{.proglang} was used, and which packages
+(and their versions) were used as well. However, others would still need to
+install the correct versions of each package manually. This can be in certain
+cases challenging, especially if packages that depend on system-level
+dependencies are required. For example, the [nloptr]{.pkg} package requires
+either [cmake]{.proglang} on Linux or macOS, or a binary of [nlopt]{.proglang},
+but this is not the case of Windows.
+
+A better alternative to simply listing packages and their versions is to use the
+[renv]{.pkg} package which generates a so-called `renv.lock` file which also
+lists [R]{.proglang} and package versions. Here is an example of such an
+`renv.lock` file:
 
@@
 used though, so installing the right version of R needs to be handled
-separately. Before continuing, it should be noted that other packages exist
-which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
-by @simonsohn2023 which makes it rather easy to install packages as they were on
+separately. [renv]{.pkg} will also not install system-level dependencies such
+as [cmake]{.pkg} for [nloptr]{.pkg} on Linux or macOS, so these still need to
+be handled separately, as described in [renv]{.pkg}’s Readme: 
+\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.
+
+Before continuing, it should be noted that other packages exist which provide
+similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg} by
+@simonsohn2023 which makes it rather easy to install packages as they were on
 CRAN at a given date. For example, the code snippet below installs the
@@
 can be achieved using the tools mentioned above. The scripts to run the analysis
-and the data are also added to the image at build time. It is possible to then
-execute the code of a study inside a so-called container, which is an instance
-of an image. Executing, or running, these containers is done non-interactively,
-and so it is possible to instantiate the correct environment to run a study in a
-single command.
+and the data are also added to the image at build time. These steps are written
+down in a simple text file called a `Dockerfile`. This `Dockerfile` is the
+definition of the [Docker]{.prolgang} image.
+
+It is possible to then execute the code of a study inside a so-called container,
+which is an instance of an image. Executing, or running, these containers is
+done non-interactively, and so it is possible to instantiate the correct
+environment to run a study in a single command.
 
@@
 version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
-analysis scripts: [Docker]{.proglang} images are actually almost complete
+analysis scripts: [Docker]{.proglang} images are actually almost complete Linux
 operating systems, which means that system-level dependecies also get bundled
 into the image. This is especially useful for packages that rely on external
-libraries to run, as these external libraries are thus also made available to
-future reproducers of the study within the [Docker]{.proglang} image. The image
-can also be shared.
+libraries to run, such as for [nloptr]{.pkg}, as these external libraries are
+thus also made available to future reproducers of the study within the
+[Docker]{.proglang} image. The image can also be shared.
 
@@
 
-An optional step towards the gold standard is to use a build automation tool
-such as [Make]{.proglang} to run the whole analysis when the container is
-executed, instead of executing scripts. Build automation tools make it easier to
-run arbitrary code in a series of well-defined steps. [R]{.proglang} programmers
-can use the [targets]{.pkg} by @landau2021 as a build automation tool.
+While [Docker]{.proglang} offers a great solution to the reproducibility
+problem, it can be awkward to work interactively from a running container:
+containers can be changed at run-time, but any change will be lost once the
+container is stopped. Also, while it is possible to launch graphical programs
+from a container, this is generally not easy to set up and only works, as far as
+we know, only on Linux operating systems. In most cases, users of
+[Docker]{.proglang} will instead use the container to run a web server and the
+web based application of a popular integrated development environment (IDE). So
+for instance, it would be possible to launch the web interface of the RStudio
+IDE, but not the operating system client of RStudio, which is usually what an R
+programmer would use (or VS Code or Emacs...). A common approach seems to be to
+work on a study interactively using a standard installation of R and R packages,
+and after the study is done, take some time to write the required `Dockerfile`
+to provide an easy way for future replicators to reproduce the results of the
+study. But this approach does not solve the problem of deploying the same
+environment on several machines, and so ideally authors would develop using
+[Docker]{.proglang}. Another challenge of using [Docker]{.proglang} is that,
+as stated already, [Docker]{.proglang} images are actually almost complete
+Linux operating systems, so experience with Linux is highly recommended if
+one wishes to properly write a `Dockerfile`.
+
+A final, and optional, step towards the gold standard is to use a build
+automation tool such as [Make]{.proglang} to run the whole analysis when the
+container is executed, instead of executing scripts. Build automation tools make
+it easier to run arbitrary code in a series of well-defined steps.
+[R]{.proglang} programmers can use the [targets]{.pkg} by @landau2021 as a build
+automation tool.
 

### 2024-02-07 paper.qmd
Update paper.qmd

@@
   it is not enough to write clean code and document it well. One
-  must also make sure to clear list all the dependencies of the analysis
+  must also make sure to clearly list all the dependencies of the analysis
   and ideally provide an easy way to install them.
@@
 version manually, but here too, there are tools that simplify the process. One
-such tool is [rig]{.proglang} by the R infrastructure team [-@rlib2023].
-[rig]{.proglang} makes it easy to install different versions of [R]{.proglang},
+such tool is [rig]{.pkg} by the R infrastructure team [-@rlib2023].
+[rig]{.pkg} makes it easy to install different versions of [R]{.proglang},
 so one could use it to install the right version to reproduce an analysis, and
@@
 [R]{.proglang} packages. However, this involves many manual steps and is thus
-error prone. It is also time-consuming to do so, especially if one wants to
-reproduce many studies.
+error prone and time-consuming.
 
@@
 package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
-[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
-[Docker]{.proglang} it is possible to package some *data product* with its
+[Docker]{.pkg} image. [Docker]{.pkg} is a containerisation tool: using
+[Docker]{.pkg} it is possible to package some *data product* with its
 dependencies into an image. A statistical analysis, from the simplest to the
@@
 down in a simple text file called a `Dockerfile`. This `Dockerfile` is the
-definition of the [Docker]{.prolgang} image.
+definition of the [Docker]{.pkg} image.
 
 It is possible to then execute the code of a study inside a so-called container,
-which is an instance of an image. Executing, or running, these containers is
+which is a running instance of an image. Executing, or running, these containers is
 done non-interactively, and so it is possible to instantiate the correct
@@
 
-The advantage of using [Docker]{.proglang} is not only about having the right
+The advantage of using [Docker]{.pkg} is not only about having the right
 version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
-analysis scripts: [Docker]{.proglang} images are actually almost complete Linux
-operating systems, which means that system-level dependecies also get bundled
+analysis scripts: [Docker]{.pkg} images are actually almost complete Linux
+operating systems, which means that system-level dependencies also get bundled
 into the image. This is especially useful for packages that rely on external
@@
 thus also made available to future reproducers of the study within the
-[Docker]{.proglang} image. The image can also be shared.
+[Docker]{.pkg} image. The image can also be shared.
 
 The Rocker project initiated by @boettiger2017 provides many pre-built
-[Docker]{.proglang} to the [R]{.proglang} community of users, already containing
+[Docker]{.pkg} to the [R]{.proglang} community of users, already containing
 specific [R]{.proglang} versions and even packages. These images can be used as
@@
 
-While [Docker]{.proglang} offers a great solution to the reproducibility
+While [Docker]{.pkgs} offers a great solution to the reproducibility
 problem, it can be awkward to work interactively from a running container:
@@
 we know, only on Linux operating systems. In most cases, users of
-[Docker]{.proglang} will instead use the container to run a web server and the
-web based application of a popular integrated development environment (IDE). So
-for instance, it would be possible to launch the web interface of the RStudio
-IDE, but not the operating system client of RStudio, which is usually what an R
-programmer would use (or VS Code or Emacs...). A common approach seems to be to
+[Docker]{.pkg} will instead use the container to run a web server and the
+web based application of a popular integrated development environment (IDE).
+A common approach seems to be to
 work on a study interactively using a standard installation of R and R packages,
@@
 environment on several machines, and so ideally authors would develop using
-[Docker]{.proglang}. Another challenge of using [Docker]{.proglang} is that,
-as stated already, [Docker]{.proglang} images are actually almost complete
+a containerized development environment. Another challenge of using [Docker]{.pkg} is that,
+as stated already, [Docker]{.pkg} images are actually almost complete
 Linux operating systems, so experience with Linux is highly recommended if
-one wishes to properly write a `Dockerfile`.
+one wishes to properly write a `Dockerfile`. Finally, one also needs to be careful
+and write the `Dockerfile` in such a way that the build process will always result
+in the same image. This is only possible if a specific digest of a base
+image is used instead of a tagged version, but in practice, this is rarely done.
 
 A final, and optional, step towards the gold standard is to use a build
-automation tool such as [Make]{.proglang} to run the whole analysis when the
+automation tool such as [Make]{.pkg} to run the whole analysis when the
 container is executed, instead of executing scripts. Build automation tools make
@@
 However, reaching this gold standard is quite costly: one needs to learn a tool
-to deal with package versions, then [Docker]{.proglang} for all the other
+to deal with package versions, then [Docker]{.pkg} for all the other
 software, including the programming language that was used. Ideally, a build
@@
 on installing and building software in a reproducible manner. This means that
-this single tool can replace not only [Docker]{.proglang} but also [renv]{.pkg}
-and even a build automation tool such as [Make]{.proglang}. To make
+this single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg}
+and even a build automation tool such as [Make]{.pkg}. To make
 [Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
@@
 
-[Nix]{.proglang} is a package manager that can be used to install and build
+[Nix]{.pkg} is a package manager that can be used to install and build
 software in a completely reproducible manner. As of writing, it contains more
-than 80.000 packages, including the entirety of CRAN and Bioconductor. This
+than 80.000 packages, including almost the entirety of CRAN and Bioconductor. This
 means that using [Nix]{.proglang}, it is possible to install not only
@@
 the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
-that when installing a package with [Nix]{.proglang}, [Nix]{.proglang} makes
+that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
 sure to install every dependency of every package, whether this dependency is
@@
 [Java]{.proglang} might not be easy, nor even possible. But with
-[Nix]{.proglang}, it suffices to declare that the [xlsx]{.pkg} package is needed
+[Nix]{.pkg}, it suffices to declare that the [xlsx]{.pkg} package is needed
 for the project, and [Nix]{.proglang} figures out automatically that
 [Java]{.proglang} is required and installs and configures it. It all just
-happens without any required intervention from the user.
+happens without any required intervention from the user. In this context, 
+[xlsx]{.pkg} can be refered to as a component closure, and quoting @dolstra2004nix:
 
-But where do these packages come from? When installing a package using
-[Nix]{.proglang}, an expression written in the [Nix]{.proglang} programming
+> The idea is to always deploy component closures: if we deploy a
+> component, then we must also deploy its dependencies, their dependencies, and so on. That is, we must
+> always deploy a set of components that is closed under
+> the ‘‘depends on’’ relation. Since closures are selfcontained, they are the units of complete software
+> deployment. After all, if a set of components is not
+> closed, it is not safe to deploy, since using them might
+> cause other components to be referenced that are missing on the target system.
+
+But how does [Nix]{.pkg} achieve this? Where do these packages, or closures, come from? When installing a package using
+[Nix]{.pkg}, an expression written in the [Nix]{.proglang} programming
 language gets downloaded from the `nixpkgs` Github repository and evaluated.
@@
 extremely flexible, and by writing one's own, it is possible to define and build
-an environment for a project in a reproducible manner.
+an environment for a project in a reproducible manner. The dependencies of these
+derivations are also defined in other expressions, which will also get installed
+if needed.
 
 But why is installing software with [Nix]{.proglang} reproducible? Because the
-whole set of [Nix]{.proglang} expressions is hosted on Github, it is possible to
+whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to
 use a specific commit of `nixpkgs` to ensure reproducibility of our project.
@@
 at exactly the same versions, regardless of when in the future the packages get
-built. This is because the [Nix]{.proglang} expressions that get downloaded when
+built. This is because the [Nix]{.pkgs} expressions that get downloaded when
 pinning are the expressions as they were at that commit of `nixpkgs`. This is
-not the only reason [Nix]{.proglang} is very useful for reproducibility, but the
-[Nix]{.proglang} package manager itself was designed with reproducibility in
-mind. [Nix]{.proglang} is functional package manager, meaning that it uses
+not the only reason [Nix]{.pkg} is very useful for reproducibility, but the
+[Nix]{.pkg} package manager itself was designed with reproducibility in
+mind. [Nix]{.pkg} is functional package manager, meaning that it uses
 concepts from functional programming to build software: no-side effects or
 global variables are allowed, and the output is always the same if given the
-same inputs, regardless of where or when a package gets built.
-We refer to @dolstra2004nix for a more technical discussion on [Nix]{.proglang}.
+same inputs, regardless of where or when a package gets built. [Nix]{.pkg} 
+also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
+to be installed at the same time on the same system. Each variant of a package
+has a unique identifier, so that it is possible to have two versions of [R]{.proglang}
+installed and use the correct one where adequate.
+We refer to @dolstra2004nix for a more technical discussion on [Nix]{.pkg}.
 
@@
 
+[Nix]{.pkg} has a fairly high entry cost though. [Nix]{.pkg} is a complex piece of software that
+comes with its own programming language, which is also called [Nix]{.proglang}. Its purpose is 
+to solve a complex problem: defining instructions on how to build software packages 
+and manage configurations in a declarative way. This makes sure that software gets 
+installed in fully reproducible manner, on any operating system or hardware. To increase
+[Nix]{.pkg}’s adoption, we wrote the [rix]{.pkg} package.
+
 ## Reproducible development environments with Nix {#sec-repro-nix}
 
-[Nix]{.proglang} expressions are written in the [Nix]{.proglang} programming language,
+[Nix]{.pkg} expressions are written in the [Nix]{.proglang} programming language,
 which is a purely functional programming language. Here is a simple expression
@@
 
-To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so the
+To define a variable in the [Nix]{.proglang} language, the `let` keyword is used, so the
 first lines of this expression define a variable called `pkgs` and

### 2024-02-09 paper.qmd
Update paper.qmd

@@
 > always deploy a set of components that is closed under
-> the ‘‘depends on’’ relation. Since closures are selfcontained, they are the units of complete software
+> the ''depends on'' relation. Since closures are selfcontained, they are the units of complete software
 > deployment. After all, if a set of components is not
@@
 global variables are allowed, and the output is always the same if given the
-same inputs, regardless of where or when a package gets built. [Nix]{.pkg} 
-also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
+same inputs, regardless of where or when a package gets built (an important
+side-note is in order here: while the functional aspect of [Nix]{.pkg} is quite
+useful to ensure reproducibility, it can lead to situations where writing the
+required [Nix]{.proglang} expression to build it a can be much more difficult than usual: this is especially true for
+packages that must download assets off the internet during installation, such as 
+several BioConductor packages like [musData]{.pkg} for example. But this is only
+a concern for the packagers, not users of the package that simply wish to install
+and use the package). [Nix]{.pkg} also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
 to be installed at the same time on the same system. Each variant of a package

### 2024-02-09 paper.qmd
Update paper.qmd

--- 2024-02-07 paper.qmd
@@
 and even a build automation tool such as [Make]{.pkg}. To make
-[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
+[Nix]{.pkg} more accessible to [R]{.proglang} programmers, we developed the
 [rix]{.pkg} package which we will present in this article.
@@
 than 80.000 packages, including almost the entirety of CRAN and Bioconductor. This
-means that using [Nix]{.proglang}, it is possible to install not only
+means that using [Nix]{.pkg}, it is possible to install not only
 [R]{.proglang}, but also all the packages required for a project. The reason why
-one should use [Nix]{.proglang} to install [R]{.proglang} packages, and not use
+one should use [Nix]{.pkg} to install [R]{.proglang} packages, and not use
 the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
@@
 global variables are allowed, and the output is always the same if given the
-same inputs, regardless of where or when a package gets built. [Nix]{.pkg} 
-also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
+same inputs, regardless of where or when a package gets built (an important
+side-note is in order here: while the functional aspect of [Nix]{.pkg} is quite
+useful to ensure reproducibility, it can lead to situations where writing the
+required [Nix]{.proglang} expression to build it can be much more difficult than usual: this is especially true for
+packages that must download assets off the internet during installation, such as 
+several BioConductor packages like [musData]{.pkg} for example. But this is only
+a concern for the packagers, not users of the package that simply wish to install
+and use the package). [Nix]{.pkg} also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
 to be installed at the same time on the same system. Each variant of a package
@@
 and Python, a LaTeX distribution, and even install one's favorite text editor to
-edit the project's source code. Essentially, one can define a complete
-reproducible environment and use it to work on a project interactively, or use
-that environment non-interactively to run an analysis sometime in the future.
+edit the project's source code. Essentially, one can define a complete, project-specific,
+and reproducible environment and use it interactively, or non-interactively. It will
+always be possible to rebuild this environmen in the future (as long as the `nixpkgs`
+GitHub repository is online) to run an analysis sometime in the future.
 
@@
 
-[Nix]{.pkg} expressions are written in the [Nix]{.proglang} programming language,
+As already stated, [Nix]{.pkg} expressions are written in the [Nix]{.proglang} programming language,
 which is a purely functional programming language. Here is a simple expression
@@
 the official `nixpkgs` Github repository, and in particular the commit starting
-with `976fa336` is going to be used. It should be noted that is also possible to
-use a personal fork of `nixpkgs` instead of the official repository.
+with `976fa336` is going to be used (it should be noted that is also possible to
+use a personal fork of `nixpkgs` instead of the official repository).
 `system_pkgs` lists the packages that we wish to install and use with this
-environment, in this example only the [R]{.proglang} programming language will
-be installed, alongside all the required dependencies to actually install and
-run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
+environment; in this example only the [R]{.proglang} programming language will
+be installed, alongside all of its dependencies, and all of the dependencies’
+dependencies. Finally, the `mkShell` function is used to build a shell
 containing the set of packages defined by the variable `system_packages` and its
@@
 interactive shell with all the packages declared in the `default.nix` file and
-can now use this shell for work. This is similar to activating environments in
+can now use this shell for work. This is similar to activating virtual environments in
 the [Python]{.proglang} ecosystem.
@@
 writing [Nix]{.proglang} expressions would lead to a wider adoption of the
-[Nix]{.proglang} package manager for reproducibility, we wrote [rix]{.pkg}
-which comes with the `rix()` function that helps generate such expressions quite
+[Nix]{.pkg} package manager for reproducibility, we wrote [rix]{.pkg}. [rix]{.pkg}
+comes with the `rix()` function that helps generate such expressions quite
 easily. For example, here is how one could use [rix]{.pkg} to generate a
@@
     project_path = ".",
-    overwrite = TRUE,
-    print = TRUE)
+    overwrite = TRUE)
 ```
@@
 Here, users do not need to look for the right commit of `nixpkgs`: it is enough
-to specify the version of [R]{.proglang} they wish to use and the packages that
-were current at the time will get used. But [rix]{.pkg} is quite flexible and
+to specify the version of [R]{.proglang} they wish to use and the right
+commit will be selected (a lookup table linking versions of R and `nixpkgs` commits
+is included in the package). The versions of [R]{.proglang} packages that
+were current at the time will get installed in that environment. But [rix]{.pkg} is quite flexible and
 makes it easy to install specific versions of packages as well (which will get
-installed from the CRAN archives) as well as packages hosted on Github and
+installed from CRAN) as well as packages hosted on Github and
 \LaTeX packages as well:
@@
 generated from a `default.nix` file that is hosted on the same repository. The
-paper would then get compiled in this reproducible environment.
+paper would then get compiled using [Quarto]{.pkg} in this reproducible environment.
+
+Instead of first dropping into a [Nix]{.pkg} shell it is also possible to directly
+run a program from the environment:
+
+```
+cd /absolute/path/to/project/ && nix-shell default.nix --run "Rscript analysis.R"
+```
+
+This will run `Rscript` and execute the `analysis.R` script (which in this example
+should be in the same directory as `default.nix`).
 
@@
 
-We recommend `path_subshell` to be a subfolder of the project. Using `rix()` we
+We recommend `path_subshell` to be a subfolder of the main project. Using `rix()` we
 generate a `default.nix` file that will provide a shell with R version 3.5.3
@@
 wrapped inside a `function()` call. The value of `out_subshell` is `r out_subshell`,
-as is indeed different than the one generated from a recent version of R.
+and is indeed different than the one generated from a recent version of R.
 
@@
 via Nix, or from a regular R installation.
+
+## Conclusion
 

### 2024-03-28 paper.qmd
a bit of conclusion

@@
 
+There are many tools to make one’s analyses reproducible, but Nix distinguishes
+itself from these in that its main concern is in deploying
+the complete set of packages that are *closed under the ''depends on''
+relation*: Nix does not simply install one piece of software,
+but its dependencies, and their dependencies. This makes Nix extremely
+useful for reproducible research, but because it solves a complex problem,
+it is a complex tool. With [rix]{.pkg}, we wish to make using Nix more
+accessible by providing a familiar interface and 
+user experience to [R]{.proglang} users.
+
 ## Acknowledgments {.unnumbered}

### 2024-05-16 paper.qmd
Update paper.qmd

@@
 abstract: |
-  In order to create an analysis that is easily reproducible,
+  In order create an analysis that is easily reproducible,
   it is not enough to write clean code and document it well. One
@@
 
-There are many tools to make one’s analyses reproducible, but Nix distinguishes
-itself from these in that its main concern is in deploying
+There are many tools to make one’s analyses reproducible, but Nix distinguishes itself from 
+these in that its main concern is in deploying
 the complete set of packages that are *closed under the ''depends on''

### 2024-06-07 paper.qmd
Update paper.qmd

@@
   In order create an analysis that is easily reproducible,
+coucou
   it is not enough to write clean code and document it well. One

### 2024-06-20 paper.qmd
Update paper.qmd

--- 2024-05-16 paper.qmd
@@
 ---
-title: "Reproducible development environments with rix"
+title: "Reproducible development environments with rix hello"
 format:

### 2024-10-01 paper.qmd
Update paper.qmd

--- 2024-05-16 paper.qmd
@@
 the ends of this continuum, we only have access to the paper describing the
-studies, which is not reproducible at all. If in addition, to this paper authors
+studies, which is not reproducible at all. If in addition to this paper authors
 make the original source code of the analysis available, reproducibility is
-improved, albeit only by a little. Adding the original data improves
-reproducibility yet again. Finally, if to all this the authors add what Roger
+improved. Adding the original data improves reproducibility yet again. 
+Finally, if to all this the authors add what Roger
 Peng named the *linked and executable code and data*, we reach the gold standard
@@
     program = "R",
-    exec_mode = "non-blocking",
     project_path = path_subshell,

### 2024-10-01 paper.qmd
formatting
typos and update to latest version of packages

@@
 abstract: |
-  In order create an analysis that is easily reproducible,
+  In order create an analysis that is reproducible,
   it is not enough to write clean code and document it well. One
@@
 make the original source code of the analysis available, reproducibility is
-improved. Adding the original data improves reproducibility yet again. 
-Finally, if to all this the authors add what Roger
-Peng named the *linked and executable code and data*, we reach the gold standard
-of reproducibility.
+improved. Adding the original data improves reproducibility yet again. Finally,
+if to all this the authors add what Roger Peng referred to as the *linked and
+executable code and data*, we reach the gold standard of reproducibility.
 
-What is this *linked and executable code and data*? Another way to name this
-crucial piece of the reproducibility puzzle is *computational environment*. The
-computational environment is all the software required to actually run the
-analysis. Here too, we can speak of a continuum. One could simply name and list
-the software used: for example, the [R]{.proglang} programming language.
-Sometimes, authors have the courtesy to also state the version of [R]{.proglang}
-used. Some authors go further, and also list the packages used, and ideally with
-their versions as well. Authors rarely state the operating system on which the
-analysis was done, even though it has been shown that running the same analysis
-with the same software but on different operating systems could lead to
-different results, as described in @neupane2019. Authors also only very rarely
-provide instructions to install the required tools and software in order to
-reproduce their studies.
+*Linked and executable code and data* refer to the compiled source code scripts,
+but for an interpreted language such as [R]{}{.proglang}, this would rather
+refer to the *computational environment*. The computational environment is all
+the software required to run the analysis. Here too, we can speak of a
+continuum. One could simply name and list the software used: for example, the
+[R]{.proglang} programming language. Sometimes, authors have the courtesy to
+also state the version of [R]{.proglang} used. Some authors go further, and also
+list the packages used, and ideally with their versions as well. Authors rarely
+state the operating system on which the analysis was done, even though it has
+been shown that running the same analysis with the same software but on
+different operating systems could lead to different results, as described in
+@neupane2019. Authors also only very rarely provide instructions to install the
+required tools and software in order to reproduce their studies.
 
@@
 
-There are tools available that authors can use to reach the gold stardard of a
+There are tools available that authors can use to reach the gold standard of a
 fully reproducible study, or to successfully deploy the same development
@@
 cases challenging, especially if packages that depend on system-level
-dependencies are required. For example, the [nloptr]{.pkg} package requires
-either [cmake]{.proglang} on Linux or macOS, or a binary of [nlopt]{.proglang},
-but this is not the case of Windows.
+dependencies are required. For example, the [nloptr]{.pkg} package requires a
+binary of [nlopt]{.proglang} to be available, or if not, [cmake]{.proglang} on
+Linux or macOS.
 
@@
 that every call to the `install.packages()` function will now install the
-packages from this snapshotted mirror. However, unless the user explicitely and
+packages from this snapshotted mirror. However, unless the user explicitly and
 manually manages the different projects' libraries, using the Posit Package
@@
 
-The advantage of using [Docker]{.pkg} is not only about having the right
-version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
-analysis scripts: [Docker]{.pkg} images are actually almost complete Linux
-operating systems, which means that system-level dependencies also get bundled
-into the image. This is especially useful for packages that rely on external
-libraries to run, such as for [nloptr]{.pkg}, as these external libraries are
-thus also made available to future reproducers of the study within the
-[Docker]{.pkg} image. The image can also be shared.
+The advantage of using [Docker]{.pkg} is not only about having the right version
+of [R]{.proglang} and [R]{.proglang} packages bundled alongside the analysis
+scripts: [Docker]{.pkg} images are actually almost complete Linux operating
+systems, which means that system-level dependencies also get bundled into the
+image. This is especially useful for packages that rely on external system-level
+dependencies to run, such as for [nloptr]{.pkg}, as these external system-level
+dependencies are thus also made available to future replicators of the study
+within the [Docker]{.pkg} image. The image can also be shared.
 
@@
 
-While [Docker]{.pkgs} offers a great solution to the reproducibility
-problem, it can be awkward to work interactively from a running container:
-containers can be changed at run-time, but any change will be lost once the
-container is stopped. Also, while it is possible to launch graphical programs
-from a container, this is generally not easy to set up and only works, as far as
-we know, only on Linux operating systems. In most cases, users of
-[Docker]{.pkg} will instead use the container to run a web server and the
-web based application of a popular integrated development environment (IDE).
-A common approach seems to be to
-work on a study interactively using a standard installation of R and R packages,
-and after the study is done, take some time to write the required `Dockerfile`
-to provide an easy way for future replicators to reproduce the results of the
-study. But this approach does not solve the problem of deploying the same
-environment on several machines, and so ideally authors would develop using
-a containerized development environment. Another challenge of using [Docker]{.pkg} is that,
-as stated already, [Docker]{.pkg} images are actually almost complete
-Linux operating systems, so experience with Linux is highly recommended if
-one wishes to properly write a `Dockerfile`. Finally, one also needs to be careful
-and write the `Dockerfile` in such a way that the build process will always result
-in the same image. This is only possible if a specific digest of a base
-image is used instead of a tagged version, but in practice, this is rarely done.
+While [Docker]{.pkgs} offers a great solution to the reproducibility problem, it
+can be awkward to work interactively from a running container: containers can be
+changed at run-time, but any change will be lost once the container is stopped.
+Also, while it is possible to launch graphical programs from a container, this
+is generally not easy to set up and only works, as far as we know, only on Linux
+operating systems. In most cases, users of [Docker]{.pkg} will instead use the
+container to run a web server and the web based application of a popular
+integrated development environment (IDE).
+
+A common approach seems to be to work on a study interactively using a standard
+installation of R and R packages, and after the study is done, take some time to
+write the required `Dockerfile` to provide an easy way for future replicators to
+reproduce the results of the study. But this approach does not solve the problem
+of deploying the same environment on several machines, and so ideally authors
+would develop using a containerized development environment. Another challenge
+of using [Docker]{.pkg} is that, as stated already, [Docker]{.pkg} images are
+actually almost complete Linux operating systems, so experience with Linux is
+highly recommended if one wishes to properly write a `Dockerfile`. Finally, one
+also needs to be careful and write the `Dockerfile` in such a way that the build
+process will always result in the same image. This is only possible if a
+specific digest of a base image is used instead of a tagged version, but in
+practice, this is rarely done.
 
@@
 @mcdermott2021 is an example of a scientific study that reached the gold
-standard of reproducibility. The author of this paper set up an accompagnying
+standard of reproducibility. The author of this paper set up an accompanying
 Github repository to the
@@
 - A `Dockerfile` was used to provide the complete computational environment,
-  including the right version of [R]{.proglang} and run the whole analysis
+  including the right version of [R]{.proglang}, to run the whole analysis
   easily.
@@
 However, reaching this gold standard is quite costly: one needs to learn a tool
-to deal with package versions, then [Docker]{.pkg} for all the other
-software, including the programming language that was used. Ideally, a build
-automation tool such as [Make]{.proglang} should be added to the list. It should
-also be noted that these requirements are not limited to the [R]{.proglang}
-programming language. A very similar approach should be taken if one uses
-[Python]{.proglang} for statistical analysis instead. Furthermore, the previous
-discussion also implies that for an analysis to actually be reproducible at all,
-it must be made using free and open source software, as proprietary software
-cannot be freely shared with others that wish to reproduce a study, and there's
-no guarantee that others will have access to the proprietary software used.
+to deal with package versions, then [Docker]{.pkg} for all the other software,
+including the programming language that was used. Ideally, a build automation
+tool such as [Make]{.pkg} should be added to the list. It should also be
+noted that these requirements are not limited to the [R]{.proglang} programming
+language. A very similar approach should be taken if one uses
+[Python]{.proglang}, or any other programming language, for statistical analysis
+instead. Furthermore, the previous discussion also implies that for an analysis
+to actually be reproducible at all, it must be made using free and open source
+software, as proprietary software cannot be freely shared with others that wish
+to reproduce a study, and there's no guarantee that others will have access to
+the proprietary software used.
 
@@
 
-[Nix]{.pkg} is a package manager that can be used to install and build
-software in a completely reproducible manner. As of writing, it contains more
-than 80.000 packages, including almost the entirety of CRAN and Bioconductor. This
-means that using [Nix]{.pkg}, it is possible to install not only
-[R]{.proglang}, but also all the packages required for a project. The reason why
-one should use [Nix]{.pkg} to install [R]{.proglang} packages, and not use
-the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
-that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
-sure to install every dependency of every package, whether this dependency is
-another [R]{.proglang} package or a system-level dependency.
+[Nix]{.pkg} is a package manager that can be used to install and build software
+in a completely reproducible manner. As of writing, it contains more than
+100.000 packages, including almost the entirety of CRAN and Bioconductor. This
+means that using [Nix]{.pkg}, it is possible to install not only [R]{.proglang},
+but also all the packages required for a project. The reason why one should use
+[Nix]{.pkg} to install [R]{.proglang} packages, and not use the usual, built-in,
+`install.packages()` [R]{.proglang} function instead, is that when installing a
+package with [Nix]{.pkg}, [Nix]{.pkg} makes sure to install every dependency of
+every package, whether this dependency is another [R]{.proglang} package or a
+system-level dependency.
 
@@
 Depending on what system one tries to install [xlsx]{.pkg}, installing
-[Java]{.proglang} might not be easy, nor even possible. But with
-[Nix]{.pkg}, it suffices to declare that the [xlsx]{.pkg} package is needed
-for the project, and [Nix]{.proglang} figures out automatically that
-[Java]{.proglang} is required and installs and configures it. It all just
-happens without any required intervention from the user. In this context, 
-[xlsx]{.pkg} can be refered to as a component closure, and quoting @dolstra2004nix:
+[Java]{.proglang} might not be easy, nor even possible. But with [Nix]{.pkg}, it
+suffices to declare that the [xlsx]{.pkg} package is needed for the project, and
+[Nix]{.proglang} figures out automatically that [Java]{.proglang} is required
+and installs and configures it. It all just happens without any required
+intervention from the user. In this context, [xlsx]{.pkg} can be referred to as
+a component closure, and quoting @dolstra2004nix:
 
-> The idea is to always deploy component closures: if we deploy a
-> component, then we must also deploy its dependencies, their dependencies, and so on. That is, we must
-> always deploy a set of components that is closed under
-> the ''depends on'' relation. Since closures are selfcontained, they are the units of complete software
-> deployment. After all, if a set of components is not
-> closed, it is not safe to deploy, since using them might
-> cause other components to be referenced that are missing on the target system.
+> The idea is to always deploy component closures: if we deploy a component,
+> then we must also deploy its dependencies, their dependencies, and so on. That
+> is, we must always deploy a set of components that is closed under the
+> ''depends on'' relation. Since closures are selfcontained, they are the units
+> of complete software deployment. After all, if a set of components is not
+> closed, it is not safe to deploy, since using them might cause other
+> components to be referenced that are missing on the target system.
 
-But how does [Nix]{.pkg} achieve this? Where do these packages, or closures, come from? When installing a package using
-[Nix]{.pkg}, an expression written in the [Nix]{.proglang} programming
-language gets downloaded from the `nixpkgs` Github repository and evaluated.
-This expression contains a so-called *derivation*. A derivation defines a build:
-its dependencies, commands to build and install the package in question, and
-then an output. Most of the time, a derivation downloads source code, builds the
-software from the source and then outputs a compiled binary. Derivations are
-extremely flexible, and by writing one's own, it is possible to define and build
-an environment for a project in a reproducible manner. The dependencies of these
-derivations are also defined in other expressions, which will also get installed
-if needed.
+But how does [Nix]{.pkg} achieve this? Where do these packages, or closures,
+come from? When installing a package using [Nix]{.pkg}, an expression written in
+the [Nix]{.proglang} programming language gets downloaded from the `nixpkgs`
+Github repository and evaluated. This expression contains a so-called
+*derivation*. A derivation defines a build: its dependencies, commands to build
+and install the package in question, and then an output. Most of the time, a
+derivation downloads source code, builds the software from the source and then
+outputs a compiled binary. Derivations are extremely flexible, and by writing
+one's own, it is possible to define and build an environment for a project in a
+reproducible manner. The dependencies of these derivations are also defined in
+other expressions, which will also get installed if needed. This is why the 
+[Nix]{.pkg} manager *knows* that [Java]{.proglang} is needed for [xlsx]{.pkg}.
 
 But why is installing software with [Nix]{.proglang} reproducible? Because the
-whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to
-use a specific commit of `nixpkgs` to ensure reproducibility of our project.
-Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
-Pinning a revision ensures that every package that Nix installs will always be
-at exactly the same versions, regardless of when in the future the packages get
+whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to use
+a specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
+specific commit of `nixpkgs` is called *pinning a revision (or commit)*. Pinning
+a revision ensures that every package that Nix installs will always be at
+exactly the same versions, regardless of when in the future the packages get
 built. This is because the [Nix]{.pkgs} expressions that get downloaded when
-pinning are the expressions as they were at that commit of `nixpkgs`. This is
-not the only reason [Nix]{.pkg} is very useful for reproducibility, but the
-[Nix]{.pkg} package manager itself was designed with reproducibility in
-mind. [Nix]{.pkg} is functional package manager, meaning that it uses
-concepts from functional programming to build software: no-side effects or
-global variables are allowed, and the output is always the same if given the
-same inputs, regardless of where or when a package gets built (an important
-side-note is in order here: while the functional aspect of [Nix]{.pkg} is quite
-useful to ensure reproducibility, it can lead to situations where writing the
-required [Nix]{.proglang} expression to build it can be much more difficult than usual: this is especially true for
+pinning are the expressions as they were at that commit of `nixpkgs`.
+
+While pinning revisions is essential for reproducibility, this is not the only
+reason [Nix]{.pkg} is very useful for reproducibility, but the [Nix]{.pkg}
+package manager itself was designed with reproducibility in mind. [Nix]{.pkg} is
+a functional package manager, meaning that it uses concepts from functional
+programming to build software: no-side effects or global variables are allowed,
+and the output is always the same if given the same inputs, regardless of where
+or when a package gets built (an important side-note is in order here: while the
+functional aspect of [Nix]{.pkg} is quite useful to ensure reproducibility, it
+can lead to situations where writing the required [Nix]{.proglang} expression to
+build it can be much more difficult than usual: this is especially true for
 packages that must download assets off the internet during installation, such as
 several BioConductor packages like [musData]{.pkg} for example. But this is only
-a concern for the packagers, not users of the package that simply wish to install
-and use the package). [Nix]{.pkg} also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
-to be installed at the same time on the same system. Each variant of a package
-has a unique identifier, so that it is possible to have two versions of [R]{.proglang}
-installed and use the correct one where adequate.
-We refer to @dolstra2004nix for a more technical discussion on [Nix]{.pkg}.
+a concern for the packagers, not users of the package that simply wish to
+install and use the package). [Nix]{.pkg} also allows for multiple versions (or
+variants in the [Nix]{.pkg} jargon) to be installed at the same time on the same
+system. Each variant of a package has a unique identifier, so that it is
+possible to have two versions of [R]{.proglang} installed and use the correct
+one where adequate. We refer to @dolstra2004nix for a more technical discussion
+on [Nix]{.pkg}.
 
@@
 and Python, a LaTeX distribution, and even install one's favorite text editor to
-edit the project's source code. Essentially, one can define a complete, project-specific,
-and reproducible environment and use it interactively, or non-interactively. It will
-always be possible to rebuild this environmen in the future (as long as the `nixpkgs`
-GitHub repository is online) to run an analysis sometime in the future.
+edit the project's source code. Essentially, one can define a complete,
+project-specific, and reproducible environment and use it interactively, or
+non-interactively. It will always be possible to rebuild this environment in the
+future (as long as the `nixpkgs` GitHub repository is online) to run an analysis
+sometime in the future.
 
-[Nix]{.pkg} has a fairly high entry cost though. [Nix]{.pkg} is a complex piece of software that
-comes with its own programming language, which is also called [Nix]{.proglang}. Its purpose is 
-to solve a complex problem: defining instructions on how to build software packages 
-and manage configurations in a declarative way. This makes sure that software gets 
-installed in fully reproducible manner, on any operating system or hardware. To increase
+[Nix]{.pkg} has a fairly high entry cost though. [Nix]{.pkg} is a complex piece
+of software that comes with its own programming language, which is also called
+[Nix]{.proglang}. Its purpose is to solve a complex problem: defining
+instructions on how to build software packages and manage configurations in a
+declarative way. This makes sure that software gets installed in fully
+reproducible manner, on any operating system or hardware. To increase
 [Nix]{.pkg}’s adoption, we wrote the [rix]{.pkg} package.
@@
 
-As already stated, [Nix]{.pkg} expressions are written in the [Nix]{.proglang} programming language,
-which is a purely functional programming language. Here is a simple expression
-that builds a shell environment containing version 4.3.1 of [R]{.proglang}:
+As already stated, [Nix]{.pkg} expressions are written in the [Nix]{.proglang}
+programming language, which is a purely functional programming language. Here is
+a simple expression that builds a shell environment containing version 4.3.1 of
+[R]{.proglang}:
 
@@
 let
-  pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/976fa3369d722e76f37c77493d99829540d43845.tar.gz") {};
+  pkgs = import (fetchTarball 
+    "https://github.com/NixOS/nixpkgs/archive/976fa336....tar.gz"
+  ) {};
   system_packages = builtins.attrValues {
@@
 
-To define a variable in the [Nix]{.proglang} language, the `let` keyword is used, so the
-first lines of this expression define a variable called `pkgs` and
-`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
+To define a variable in the [Nix]{.proglang} language, the `let` keyword is
+used, so the first lines of this expression define the variables called `pkgs`
+and `system_pkgs`. `pkgs` is defined as the set of packages to be used in this
 derivation. In this example, the package expressions are going to be pulled from
@@
 be installed, alongside all of its dependencies, and all of the dependencies’
-dependencies. Finally, the `mkShell` function is used to build a shell
-containing the set of packages defined by the variable `system_packages` and its
-`shellHook` is set to `R --vanilla`: this means that when entering the shell,
-[R]{.proglang} will be started with the `--vanilla` flag (thus ignoring any
-startup options).
+dependencies. Finally, the `mkShell` function is used to build a development
+shell containing the set of packages defined by the variable `system_packages`
+and its `shellHook` is set to `R --vanilla`: this means that when entering the
+shell, [R]{.proglang} will be started with the `--vanilla` flag (thus ignoring
+any startup options).
 
@@
 interactive shell with all the packages declared in the `default.nix` file and
-can now use this shell for work. This is similar to activating virtual environments in
-the [Python]{.proglang} ecosystem.
+can now use this shell for work. This is similar to activating virtual
+environments in the [Python]{.proglang} ecosystem.
 
@@
 writing [Nix]{.proglang} expressions would lead to a wider adoption of the
-[Nix]{.pkg} package manager for reproducibility, we wrote [rix]{.pkg}. [rix]{.pkg}
-comes with the `rix()` function that helps generate such expressions quite
-easily. For example, here is how one could use [rix]{.pkg} to generate a
+[Nix]{.pkg} package manager for reproducibility, we developed [rix]{.pkg}.
+[rix]{.pkg} comes with the `rix()` function that helps generate such expressions
+quite easily. For example, here is how one could use [rix]{.pkg} to generate a
 `default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
@@
 Here, users do not need to look for the right commit of `nixpkgs`: it is enough
-to specify the version of [R]{.proglang} they wish to use and the right
-commit will be selected (a lookup table linking versions of R and `nixpkgs` commits
-is included in the package). The versions of [R]{.proglang} packages that
-were current at the time will get installed in that environment. But [rix]{.pkg} is quite flexible and
-makes it easy to install specific versions of packages as well (which will get
-installed from CRAN) as well as packages hosted on Github and
-\LaTeX packages as well:
+to specify the version of [R]{.proglang} they wish to use and the right commit
+will be selected (a lookup table linking versions of R and `nixpkgs` commits is
+included in the package). The versions of [R]{.proglang} packages that were
+current at the time will get installed in that environment. But [rix]{.pkg} is
+quite flexible and makes it easy to install specific versions of packages (which
+will get installed from CRAN) as well as packages hosted on Github and \LaTeX
+packages:
 
@@
 older version of the package. Had the user not specified the version, the
-version available as of the `nixpkgs` commit would be used), but also install
+version available as of the `nixpkgs` commit would be used), but also installs
 several TeXLive packages for authoring documents with \LaTeX, the development
@@
 which would install VS Code instead, or "other", to use any other editor such as
-Emacs or Vi. You can see the resulting `default.nix` file in the @sec-appendix.
+Emacs or Vim. You can see the resulting `default.nix` file in the @sec-appendix.
 
@@
 article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
-compilation. Essentially, each time one of the authors pushed to the master
-branch, a Github Actions runner would be started, [Nix]{.proglang} would get
-installed on said runner, and an environment containing the right packages was
-generated from a `default.nix` file that is hosted on the same repository. The
-paper would then get compiled using [Quarto]{.pkg} in this reproducible environment.
+compilation. Each time one of the authors pushed to the master branch, a Github
+Actions runner would be started, [Nix]{.proglang} would get installed on said
+runner, and an environment containing the right packages was generated from a
+`default.nix` file that is hosted on the same repository. The paper would then
+get compiled using [Quarto]{.pkg} in this reproducible environment.
 
@@
 
-This will run `Rscript` and execute the `analysis.R` script (which in this example
-should be in the same directory as `default.nix`).
+This will run `Rscript` and execute the `analysis.R` script (which in this
+example should be in the same directory as `default.nix`).
 
@@
 release of R, or from an older package. Using the `with_nix()` function, it is
-possible to run this single function inside the right envirnoment, and recover
+possible to run this single function inside the right environment, and recover
 its output in the main session. To illustrate, consider the following two lines
@@
 When trying to reproduce a study made on versions prior to 3.6, a replicator
-will need to take this into account. Using the `with_nix()` function,
-the original results can be replicated even from the latest version of R.
+will need to take this into account. Using the `with_nix()` function, the
+original results can be replicated even from the latest version of R.
 
@@
 
-We recommend `path_subshell` to be a subfolder of the main project. Using `rix()` we
-generate a `default.nix` file that will provide a shell with R version 3.5.3
-when built. We then use `rix_init()`, which generates an `.Rprofile` file in the
-same subfolder as where the subshell's `default.nix` is located. This
-`.Rprofile` file ensures that the R session that will be spawned from this
-subshell does not get any interference from the main session it gets called
-from:
+We recommend `path_subshell` to be a subfolder of the main project. Using
+`rix()` we generate a `default.nix` file that will provide a shell with R
+version 3.5.3 when built. When calling `rix()`, another function is also called,
+`rix_init()`, which generates an `.Rprofile` file in the same subfolder as where
+the subshell's `default.nix` is located. This `.Rprofile` file ensures that the
+R session that will be spawned from this subshell does not get any interference
+from the main session it gets called from.
 
@@
 
-```{r, include = F}
+```{r, eval = F}
 rix_init(

### 2024-10-01 paper.qmd
bootstrap env

@@
   present another tool that can be used to make projects reproducible,
-  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that
-  makes using Nix much simpler for users of [R]{.proglang}.
+  the [Nix]{.proglang} package manager, and an [R]{.proglang} package 
+  called [rix]{.pkg} that makes using [Nix]{.proglang} much simpler 
+  for users of [R]{.proglang}.
 
@@
 *Linked and executable code and data* refer to the compiled source code scripts,
-but for an interpreted language such as [R]{}{.proglang}, this would rather
+but for an interpreted language such as [R]{.proglang}, this would rather
 refer to the *computational environment*. The computational environment is all
@@
 [rix]{.pkg} comes with the `rix()` function that helps generate such expressions
-quite easily. For example, here is how one could use [rix]{.pkg} to generate a
-`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
-[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
-expression here because of space):
+quite easily. [rix]{.pkg} is available on CRAN and can be installed as any other
+package, but it is also possible to use it to bootstrap an [R]{.proglang}
+development environment on a system where [R]{.proglang} is not available (but
+[Nix]{.proglang} is). One simply needs to run this line inside a terminal:
+
+```
+nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
+```
+
+to be dropped in a temporary [R]{.proglang} session where [rix]{.pkg} is
+available. From there, it is possible to generate new [Nix]{.proglang}
+expressions to build new environments. For example, here is how one could use
+[rix]{.pkg} to generate a `default.nix` file that installs version 4.3.1 of
+[R]{.proglang} and the [dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't
+show the generated expression here because of space):
 
@@
 
-```{r, eval = F}
-rix_init(
-   project_path = path_subshell,
-   rprofile_action = "overwrite",
-   message_type = "simple"
- )
-```
-
-```{r, eval = F}
-rix_init(
-   project_path = path_subshell,
-   rprofile_action = "overwrite",
-   message_type = "simple"
- )
-```
-
 We can now use this shell to generate our vector of random numbers using
@@
 
-We should say that `rix_init()` and `with_nix()` print many status messages to
-the console while running, which we don't show here for space reasons.
+We should say that `with_nix()` prints many status messages to the console while
+running, which we don't show here for space reasons.
 
-The `expr` argument of `with_nix()` can be arbitrarly complex, but must be
-wrapped inside a `function()` call. The value of `out_subshell` is `r out_subshell`,
-and is indeed different than the one generated from a recent version of R.
+The `expr` argument of `with_nix()` can be arbitrarily complex, but must be
+wrapped inside a `function()` call. The value of `out_subshell` is `r
+out_subshell`, and is indeed different than the one generated from a recent
+version of R.
 
@@
 
-There are many tools to make one’s analyses reproducible, but Nix distinguishes itself from 
-these in that its main concern is in deploying
-the complete set of packages that are *closed under the ''depends on''
-relation*: Nix does not simply install one piece of software,
-but its dependencies, and their dependencies. This makes Nix extremely
-useful for reproducible research, but because it solves a complex problem,
-it is a complex tool. With [rix]{.pkg}, we wish to make using Nix more
-accessible by providing a familiar interface and 
-user experience to [R]{.proglang} users.
+There are many tools to make one’s analyses reproducible, but Nix distinguishes
+itself from these in that its main concern is in deploying the complete set of
+packages that are *closed under the ''depends on'' relation*: Nix does not
+simply install one piece of software, but its dependencies, and their
+dependencies. This makes Nix extremely useful for reproducible research, but
+because it solves a complex problem, it is a complex tool. With [rix]{.pkg}, we
+wish to make using Nix more accessible by providing a familiar interface and
+user experience to [R]{.proglang} users. Because it is possible to build
+reproducible development shells using [Nix]{.proglang}, it can accomodate any
+use case: running scripts, pipelines, interactive [shiny]{.pkg} applications,
+and serve [plumber]{.pkg} application programming interfaces.
 

### 2025-02-18 paper.qmd
Update paper.qmd

@@
   present another tool that can be used to make projects reproducible,
-  the [Nix]{.proglang} package manager, and an [R]{.proglang} package 
-  called [rix]{.pkg} that makes using [Nix]{.proglang} much simpler 
+  the  [Nix]{.pkg} package manager, and an [R]{.proglang} package 
+  called [rix]{.pkg} that makes using  [Nix]{.pkg} much simpler 
   for users of [R]{.proglang}.
@@
 keywords: [reproducibility, R, Nix]
-keywords-formatted: [reproducibility, "[R]{.proglang}", "[Nix]{.proglang}"]
+keywords-formatted: [reproducibility, "[R]{.proglang}", " [Nix]{.pkg}"]
 
@@
 *Linked and executable code and data* refer to the compiled source code scripts,
-but for an interpreted language such as [R]{.proglang}, this would rather
-refer to the *computational environment*. The computational environment is all
+but for the purposes of this article, we will assume this meas the
+*computational environment*. The computational environment is all
 the software required to run the analysis. Here too, we can speak of a
@@
 language. A very similar approach should be taken if one uses
-[Python]{.proglang}, or any other programming language, for statistical analysis
-instead. Furthermore, the previous discussion also implies that for an analysis
+[Python]{.proglang}, or any other programming language. 
+Furthermore, the previous discussion also implies that for an analysis
 to actually be reproducible at all, it must be made using free and open source
@@
 
-As an alternative to the above approach, we will present the [Nix]{.proglang}
+As an alternative to the above approach, we will present the  [Nix]{.pkg}
 package manager, which is available for all major operating systems and focuses
@@
 suffices to declare that the [xlsx]{.pkg} package is needed for the project, and
-[Nix]{.proglang} figures out automatically that [Java]{.proglang} is required
-and installs and configures it. It all just happens without any required
-intervention from the user. In this context, [xlsx]{.pkg} can be referred to as
-a component closure, and quoting @dolstra2004nix:
+ [Nix]{.pkg} will install [Java]{.proglang} and configure it. This is possible
+because the maintainers of the [R]{.proglang} language for  [Nix]{.pkg}
+declared [Java]{.proglang} as a dependency of [xlsx]{.pkg}. 
+From the perspective of the end-users, this all happens automatically, thanks
+to the efforts of the maintainers.
+
+[xlsx]{.pkg} can thus be referred to as a component closure, 
+and quoting @dolstra2004nix:
 
@@
 come from? When installing a package using  [Nix]{.pkg}, an expression written in
-the [Nix]{.proglang} programming language gets downloaded from the `nixpkgs`
-Github repository and evaluated. This expression contains a so-called
-*derivation*. A derivation defines a build: its dependencies, commands to build
+the  [Nix]{.pkg} programming language gets downloaded from the `nixpkgs`
+Github repository and evaluated. `nixpkgs` is the name of the repository that 
+contains the more than 100'000 pieces of software that can be installed with the 
+[Nix]{.pkg} manager. Expressions contain so-called *derivations*. 
+A derivation defines a build: its dependencies, commands to build
 and install the package in question, and then an output. Most of the time, a
@@
 reproducible manner. The dependencies of these derivations are also defined in
-other expressions, which will also get installed if needed. This is why the 
-[Nix]{.pkg} manager *knows* that [Java]{.proglang} is needed for [xlsx]{.pkg}.
+other expressions, which will also get installed if needed. As explained
+above, the maintainers of [R]{.proglang} for  [Nix]{.pkg} declare the right
+dependencies for the packages to turn them into component closures.
 
-But why is installing software with [Nix]{.proglang} reproducible? Because the
-whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to use
+But why is installing software with  [Nix]{.pkg} reproducible? Because the
+whole set of  [Nix]{.pkg} expressions is hosted on GitHhub, it is possible to use
 a specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
@@
 functional aspect of  [Nix]{.pkg} is quite useful to ensure reproducibility, it
-can lead to situations where writing the required [Nix]{.proglang} expression to
+can lead to situations where writing the required  [Nix]{.pkg} expression to
 build it can be much more difficult than usual: this is especially true for
-packages that must download assets off the internet during installation, such as
+packages that must download assets from the internet during installation, such as
 several BioConductor packages like [musData]{.pkg} for example. But this is only
@@
 `requirements.txt` files which list package dependencies for [Python]{.proglang}
-projects. It is also possible to build multi-language environments, containing R
+projects and virtual environments. 
+It is also possible to build multi-language environments, containing R
 and Python, a LaTeX distribution, and even install one's favorite text editor to
@@
 derivation. In this example, the package expressions are going to be pulled from
-the official `nixpkgs` Github repository, and in particular the commit starting
-with `976fa336` is going to be used (it should be noted that is also possible to
-use a personal fork of `nixpkgs` instead of the official repository).
+the official `nixpkgs` GitHub repository, and in particular the commit starting
+with `976fa336` is going to be used.
 `system_pkgs` lists the packages that we wish to install and use with this
@@
 This expression can be saved in a file called `default.nix`, and the shell can
-be built using the `nix-build` command on a system where [Nix]{.proglang} is
-available.^[To install [Nix]{.proglang} we recommend the installer from
+be built using the `nix-build` command on a system where  [Nix]{.pkg} is
+available.^[To install  [Nix]{.pkg} we recommend the installer from
 Determinate systems which can be found on the following link:
@@
 development environment on a system where [R]{.proglang} is not available (but
-[Nix]{.proglang} is). One simply needs to run this line inside a terminal:
+[Nix]{.pkg} is). One simply needs to run this line inside a terminal:
 
 ```
-nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
+nix-shell -p R rPackages.rix"
 ```
@@
 to be dropped in a temporary [R]{.proglang} session where [rix]{.pkg} is
-available. From there, it is possible to generate new [Nix]{.proglang}
+available. From there, it is possible to generate new  [Nix]{.pkg}
 expressions to build new environments. For example, here is how one could use
@@
 
-Here, users do not need to look for the right commit of `nixpkgs`: it is enough
-to specify the version of [R]{.proglang} they wish to use and the right commit
-will be selected (a lookup table linking versions of R and `nixpkgs` commits is
-included in the package). The versions of [R]{.proglang} packages that were
-current at the time will get installed in that environment. But [rix]{.pkg} is
-quite flexible and makes it easy to install specific versions of packages (which
-will get installed from CRAN) as well as packages hosted on Github and \LaTeX
-packages:
+[rix]{.pkg} can also generate much more complex expressions:
 
@@
     ),
-    ide = "rstudio",
+    ide = "positron",
     project_path = ".",
@@
 [R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
-older version of the package. Had the user not specified the version, the
-version available as of the `nixpkgs` commit would be used), but also installs
+older version of the package and also installs
 several TeXLive packages for authoring documents with \LaTeX, the development
 versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
-the RStudio integrated development environment. It should be noted that as of
-writing, RStudio is not available on macOS through `nixpkgs`, so the above
-expression will only work on Windows and Linux distributions. To make the
-expression work on macOS, the "rstudio" string could be replaced by "code",
-which would install VS Code instead, or "other", to use any other editor such as
-Emacs or Vim. You can see the resulting `default.nix` file in the @sec-appendix.
+the Positron integrated development environment. You can see the resulting `default.nix` file in the @sec-appendix.
 
@@
 compilation. Each time one of the authors pushed to the master branch, a Github
-Actions runner would be started, [Nix]{.proglang} would get installed on said
+Actions runner would be started,  [Nix]{.pkg} would get installed on said
 runner, and an environment containing the right packages was generated from a
@@
 example should be in the same directory as `default.nix`).
+
+## rix uses a fork of Nixpkgs
+
+Expressions generated with [rix]{.pkg} will point to a fork of `nixpkgs` instead
+of the official, upstream repository.
 
@@
 user experience to [R]{.proglang} users. Because it is possible to build
-reproducible development shells using [Nix]{.proglang}, it can accomodate any
+reproducible development shells using  [Nix]{.pkg}, it can accomodate any
 use case: running scripts, pipelines, interactive [shiny]{.pkg} applications,

### 2025-02-19 paper.qmd
continue on 630

@@
 but for the purposes of this article, we will assume this meas the
-*computational environment*. The computational environment is all
-the software required to run the analysis. Here too, we can speak of a
-continuum. One could simply name and list the software used: for example, the
-[R]{.proglang} programming language. Sometimes, authors have the courtesy to
-also state the version of [R]{.proglang} used. Some authors go further, and also
-list the packages used, and ideally with their versions as well. Authors rarely
-state the operating system on which the analysis was done, even though it has
-been shown that running the same analysis with the same software but on
-different operating systems could lead to different results, as described in
-@neupane2019. Authors also only very rarely provide instructions to install the
-required tools and software in order to reproduce their studies.
+*computational environment*. The computational environment is all the software
+required to run the analysis. Here too, we can speak of a continuum. One could
+simply name and list the software used: for example, the [R]{.proglang}
+programming language. Sometimes, authors have the courtesy to also state the
+version of [R]{.proglang} used. Some authors go further, and also list the
+packages used, and ideally with their versions as well. Authors rarely state the
+operating system on which the analysis was done, even though it has been shown
+that running the same analysis with the same software but on different operating
+systems could lead to different results, as described in @neupane2019. Authors
+also only very rarely provide instructions to install the required tools and
+software in order to reproduce their studies.
 
@@
 library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
-used though, so installing the right version of R needs to be handled
+used though, so installing the right version of [R]{.proglang}needs to be handled
 separately. [renv]{.pkg} will also not install system-level dependencies such
@@
 
-The `.Rprofile` file gets read by R when starting a new session, which means
+The `.Rprofile` file gets read by [R]{.proglang}when starting a new session, which means
 that every call to the `install.packages()` function will now install the
@@
 version manually, but here too, there are tools that simplify the process. One
-such tool is [rig]{.pkg} by the R infrastructure team [-@rlib2023].
-[rig]{.pkg} makes it easy to install different versions of [R]{.proglang},
-so one could use it to install the right version to reproduce an analysis, and
-then use one of the listed packages above to install the right library of
-[R]{.proglang} packages. However, this involves many manual steps and is thus
-error prone and time-consuming.
+such tool is [rig]{.pkg} by the [R]{.proglang}infrastructure team [-@rlib2023]. [rig]{.pkg}
+makes it easy to install different versions of [R]{.proglang}, so one could use
+it to install the right version to reproduce an analysis, and then use one of
+the listed packages above to install the right library of [R]{.proglang}
+packages. However, this involves many manual steps and is thus error prone and
+time-consuming.
 
@@
 It is possible to then execute the code of a study inside a so-called container,
-which is a running instance of an image. Executing, or running, these containers is
-done non-interactively, and so it is possible to instantiate the correct
+which is a running instance of an image. Executing, or running, these containers
+is done non-interactively, and so it is possible to instantiate the correct
 environment to run a study in a single command.
@@
 A common approach seems to be to work on a study interactively using a standard
-installation of R and R packages, and after the study is done, take some time to
+installation of [R]{.proglang}and [R]{.proglang}packages, and after the study is done, take some time to
 write the required `Dockerfile` to provide an easy way for future replicators to
@@
 including the programming language that was used. Ideally, a build automation
-tool such as [Make]{.pkg} should be added to the list. It should also be
-noted that these requirements are not limited to the [R]{.proglang} programming
+tool such as [Make]{.pkg} should be added to the list. It should also be noted
+that these requirements are not limited to the [R]{.proglang} programming
 language. A very similar approach should be taken if one uses
-[Python]{.proglang}, or any other programming language. 
-Furthermore, the previous discussion also implies that for an analysis
-to actually be reproducible at all, it must be made using free and open source
-software, as proprietary software cannot be freely shared with others that wish
-to reproduce a study, and there's no guarantee that others will have access to
-the proprietary software used.
+[Python]{.proglang}, or any other programming language. Furthermore, the
+previous discussion also implies that for an analysis to actually be
+reproducible at all, it must be made using free and open source software, as
+proprietary software cannot be freely shared with others that wish to reproduce
+a study, and there's no guarantee that others will have access to the
+proprietary software used.
 
-As an alternative to the above approach, we will present the  [Nix]{.pkg}
-package manager, which is available for all major operating systems and focuses
-on installing and building software in a reproducible manner. This means that
-this single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg}
-and even a build automation tool such as [Make]{.pkg}. To make
-[Nix]{.pkg} more accessible to [R]{.proglang} programmers, we developed the
-[rix]{.pkg} package which we will present in this article.
+As an alternative to the above approach, we will present the [Nix]{.pkg} package
+manager, which is available for all major operating systems and focuses on
+installing and building software in a reproducible manner. This means that this
+single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg} and even a
+build automation tool such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
+to [R]{.proglang} programmers, we developed the [rix]{.pkg} package which we
+will present in this article.
 
-## The Nix package manager {#sec-nix}
+## The [Nix]{.pkg} package manager {#sec-nix}
 
@@
 means that using [Nix]{.pkg}, it is possible to install not only [R]{.proglang},
-but also all the packages required for a project. The reason why one should use
-[Nix]{.pkg} to install [R]{.proglang} packages, and not use the usual, built-in,
-`install.packages()` [R]{.proglang} function instead, is that when installing a
-package with  [Nix]{.pkg},  [Nix]{.pkg} makes sure to install every dependency of
-every package, whether this dependency is another [R]{.proglang} package or a
-system-level dependency.
+but also all the packages required for a project. [Nix]{.pkg} is actually the
+package manager of a Linux distribution called NixOS, but can be installed as a
+stand-alone tool for other Linux distributions, macOS, and on the Windows
+Subsystem for Linux on Windows^[Which means that in practical terms, non-Nixos
+Linux distributions and Windows can be considered as the same platform.].
+
+The reason why one should use [Nix]{.pkg} to install [R]{.proglang} packages,
+and not use the usual, built-in, `install.packages()` [R]{.proglang} function
+instead, is that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
+sure to install every dependency of every package, whether this dependency is
+another [R]{.proglang} package or a system-level dependency.
 
@@
 [Nix]{.pkg} will install [Java]{.proglang} and configure it. This is possible
-because the maintainers of the [R]{.proglang} language for  [Nix]{.pkg}
-declared [Java]{.proglang} as a dependency of [xlsx]{.pkg}. 
-From the perspective of the end-users, this all happens automatically, thanks
-to the efforts of the maintainers.
+because the maintainers of the [R]{.proglang} language for [Nix]{.pkg} declared
+[Java]{.proglang} as a dependency of [xlsx]{.pkg}. From the perspective of the
+end-users, this all happens automatically, thanks to the efforts of the
+maintainers.
 
-[xlsx]{.pkg} can thus be referred to as a component closure, 
-and quoting @dolstra2004nix:
+[xlsx]{.pkg} can thus be referred to as a component closure, and quoting
+@dolstra2004nix:
 
@@
 come from? When installing a package using [Nix]{.pkg}, an expression written in
-the  [Nix]{.pkg} programming language gets downloaded from the `nixpkgs`
-Github repository and evaluated. `nixpkgs` is the name of the repository that 
-contains the more than 100'000 pieces of software that can be installed with the 
-[Nix]{.pkg} manager. Expressions contain so-called *derivations*. 
-A derivation defines a build: its dependencies, commands to build
-and install the package in question, and then an output. Most of the time, a
-derivation downloads source code, builds the software from the source and then
-outputs a compiled binary. Derivations are extremely flexible, and by writing
-one's own, it is possible to define and build an environment for a project in a
-reproducible manner. The dependencies of these derivations are also defined in
-other expressions, which will also get installed if needed. As explained
-above, the maintainers of [R]{.proglang} for  [Nix]{.pkg} declare the right
-dependencies for the packages to turn them into component closures.
+the [Nix]{.pkg} programming language gets downloaded from the `nixpkgs` Github
+repository and evaluated. `nixpkgs` is the name of the repository that contains
+the more than 100'000 pieces of software that can be installed with the
+[Nix]{.pkg} manager. Expressions contain so-called *derivations*. A derivation
+defines a build: its dependencies, commands to build and install the package in
+question, and then an output. Most of the time, a derivation downloads source
+code, builds the software from the source and then outputs a compiled binary.
+Derivations are extremely flexible, and by writing one's own, it is possible to
+define and build an environment for a project in a reproducible manner. The
+dependencies of these derivations are also defined in other expressions, which
+will also get installed if needed. As explained above, the maintainers of
+[R]{.proglang} for [Nix]{.pkg} declare the right dependencies for the packages
+to turn them into component closures.
 
-But why is installing software with  [Nix]{.pkg} reproducible? Because the
-whole set of  [Nix]{.pkg} expressions is hosted on GitHhub, it is possible to use
-a specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
+But why is installing software with [Nix]{.pkg} reproducible? Because the whole
+set of [Nix]{.pkg} expressions is hosted on GitHhub, it is possible to use a
+specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
 specific commit of `nixpkgs` is called *pinning a revision (or commit)*. Pinning
-a revision ensures that every package that Nix installs will always be at
+a revision ensures that every package that [Nix]{.pkg} installs will always be at
 exactly the same versions, regardless of when in the future the packages get
@@
 build it can be much more difficult than usual: this is especially true for
-packages that must download assets from the internet during installation, such as
-several BioConductor packages like [musData]{.pkg} for example. But this is only
-a concern for the packagers, not users of the package that simply wish to
+packages that must download assets from the internet during installation, such
+as several BioConductor packages like [musData]{.pkg} for example. But this is
+only a concern for the packagers, not users of the package that simply wish to
 install and use the package). [Nix]{.pkg} also allows for multiple versions (or
@@
 `requirements.txt` files which list package dependencies for [Python]{.proglang}
-projects and virtual environments. 
-It is also possible to build multi-language environments, containing R
-and Python, a LaTeX distribution, and even install one's favorite text editor to
-edit the project's source code. Essentially, one can define a complete,
-project-specific, and reproducible environment and use it interactively, or
-non-interactively. It will always be possible to rebuild this environment in the
-future (as long as the `nixpkgs` GitHub repository is online) to run an analysis
-sometime in the future.
+projects and virtual environments. It is also possible to build multi-language
+environments, containing [R]{.proglang}and Python, a LaTeX distribution, and even install
+one's favorite text editor to edit the project's source code. Essentially, one
+can define a complete, project-specific, and reproducible environment and use it
+interactively, or non-interactively. It will always be possible to rebuild this
+environment in the future (as long as the `nixpkgs` GitHub repository is online)
+to run an analysis sometime in the future.
 
@@
 
-## Reproducible development environments with Nix {#sec-repro-nix}
+## Reproducible development environments with [Nix]{.pkg} {#sec-repro-nix}
 
@@
   system_packages = builtins.attrValues {
-    inherit (pkgs) R ;
+    inherit (pkgs) [R]{.proglang};
   };
@@
 the official `nixpkgs` GitHub repository, and in particular the commit starting
-with `976fa336` is going to be used.
-`system_pkgs` lists the packages that we wish to install and use with this
-environment; in this example only the [R]{.proglang} programming language will
-be installed, alongside all of its dependencies, and all of the dependencies’
-dependencies. Finally, the `mkShell` function is used to build a development
-shell containing the set of packages defined by the variable `system_packages`
-and its `shellHook` is set to `R --vanilla`: this means that when entering the
-shell, [R]{.proglang} will be started with the `--vanilla` flag (thus ignoring
-any startup options).
+with `976fa336` is going to be used. `system_pkgs` lists the packages that we
+wish to install and use with this environment; in this example only the
+[R]{.proglang} programming language will be installed, alongside all of its
+dependencies, and all of the dependencies’ dependencies. Finally, the `mkShell`
+function is used to build a development shell containing the set of packages
+defined by the variable `system_packages` and its `shellHook` is set to `R
+--vanilla`: this means that when entering the shell, [R]{.proglang} will be
+started with the `--vanilla` flag (thus ignoring any startup options).
 
@@
 ```
-nix-shell -p R rPackages.rix"
+nix-shell -p [R]{.proglang}rPackages.rix"
 ```
@@
 to be dropped in a temporary [R]{.proglang} session where [rix]{.pkg} is
-available. From there, it is possible to generate new  [Nix]{.pkg}
-expressions to build new environments. For example, here is how one could use
-[rix]{.pkg} to generate a `default.nix` file that installs version 4.3.1 of
-[R]{.proglang} and the [dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't
-show the generated expression here because of space):
+available.^[`nix-shell -p` starts an interactive shell where the package
+specified with the `-p` flag are present.] From there, it is possible to
+generate new [Nix]{.pkg} expressions to build new environments. For example,
+here is how one could use [rix]{.pkg} to generate a `default.nix` file that
+installs version 4.3.1 of [R]{.proglang} and the [dplyr]{.pkg} and
+[chronicler]{.pkg} packages (we don't show the generated expression here because
+of space):
 
@@
 [rix]{.pkg}'s documentation also provides installation instructions for Nix, but
-[rix]{.pkg} can generate Nix expression even if Nix is not installed on one's
+[rix]{.pkg} can generate [Nix]{.pkg} expression even if [Nix]{.pkg} is not installed on one's
 system. This is quite useful if one wants to generate an expression and use that
@@
 
-Instead of first dropping into a  [Nix]{.pkg} shell it is also possible to directly
-run a program from the environment:
+Instead of first dropping into a [Nix]{.pkg} shell it is also possible to
+directly run a program from the environment:
 
@@
 
-Expressions generated with [rix]{.pkg} will point to a fork of `nixpkgs` instead
-of the official, upstream repository.
+As previously explained, when invoked, the [Nix]{.pkg} package manager will use
+expressions hosted on the `nixpkgs` GitHub repository to build software.
+However, when generating expressions with [rix]{.pkg} our fork
+`rstats-on-nix/nixpkgs` is used instead.
+
+Using a fork instead of the official upstream `nixpkgs` repository has several
+advantages.
+
+The first thing we need to stress, is that it is important to keep in mind that
+the [Nix]{.pkg} package manager is first and foremost the package manager of a
+Linux distribution called NixOS, and as such, several technical and governance
+choice have to be made that limit what we, the authors of [rix]{.pkg} want to offer
+to the users of `{rix}`. For our needs, having a fork gives us more flexibility.
+
+For example, while it is technically possible for [Nix]{.pkg} to provide many versions
+(or variants, in [Nix]{.pkg} jargon) of the same package (for example, you can
+install the latest [Emacs]{.pkg} by installing the `emacs` package, or
+[Emacs]{.pkg} 28 by installing `emacs28`) this ultimately depends on whether the
+maintainer wishes to package several variants, or whether it is practical. In
+the case of the [R]{.proglang} ecosystem, with more than 20'000 CRAN and
+Bioconductor packages, that is not possible for the maintainers to provide
+several variants of the same packages. So for a given `nixpkgs` commit, it is
+not possible to easily install a specific version of an [R]{.proglang} package
+that is not included in that particular `nixpkgs` commit. Instead, [rix]{.pkg}
+can install it from source with this syntax:
+
+```{r, eval = FALSE}
+rix(..., r_pkgs = "dplyr@1.0.7", ...)
+```
+
+However, some packages might need specific fixes to be successfully installed
+by [Nix]{.pkg} (as explained in the case of [xlsx]{.pkg} previously) and thus
+trying to install packages like this might fail.
+
+Furthermore, it isn't practical to update the whole of the [R]{.proglang}
+packages set on [Nix]{.pkg} every day: so while CRAN and Bioconductor get
+updates daily, the [R]{.proglang} packages set on [Nix]{.pkg} gets updated only
+around new releases of R. Again, this is a consequence of [Nix]{.pkg} being
+first and foremost the package manager of a Linux distribution with its own
+governance; daily updates of more than 20'000 package is simply not feasible.
+
+Because of these reasons, we decided to fork `nixpkgs`: because it is a fork, we
+can afford to do things in a way that could not be possible or practical for
+upstream.
+
+The first thing this fork allows us to do is offer a daily snapshot of CRAN.
+Every day, thanks to Github Actions, the [R]{.proglang} packages set gets
+updated, and the result commited to a dated branch. This has been going on since
+the 14th of December 2024 (see
+[here](https://github.com/rstats-on-nix/nixpkgs/tree/2024-12-14)). So when users
+set a date as in `rix(date = "2024-12-14", ...)` the specified dated branch from
+the fork is going to get used.
+
+But this doesn’t mean that we recommend you
+use any date from the `rstats-on-nix/nixpkgs` fork: instead, each Monday,
+another action uses this fork and tries to build a set of popular packages on
+Linux and macOS, and only if this succeeds is the date added through a PR to the
+list of available dates on `{rix}`!
+
+The reason this is done like this is to manage another *risk* of the upstream
+`nixpkgs`. As you know, `nixpkgs` is huge, and though the utmost care is taken
+by contributors and the PR review process is very strict, it can happen that
+updating packages breaks other packages. For example recently RStudio was in a
+broken state due to an issue in one its dependencies, `boost`. This is not the
+fault of anyone in particular: it's just that packages get updated and packages
+that depend on them should get updated as well: but if that doesn't happen
+quickly enough, the `nixpkgs` maintainer faces a conundrum. Either he or she
+doesn't update the package because it breaks others, but not updating a package
+could be a security vulnerability, or he or she updates the package, but now
+others, perhaps less critical packages are broken and need to be fixed, either
+by their upstream developers, or by the `nixpkgs` maintainer of said packages.
+In the case of RStudio a fix was proposed and promptly merged, but if you wanted
+to install RStudio during the time it took to fix it, you would have faced
+an error message, which isn't great if all you want is use [Nix]{.pkg} shells as
+development environments.
+
+So for us, having a fork allows us to backport these fixes and so if you try to
+install RStudio using the latest available date, which is `"2025-02-10"`, it’s
+going to work, whereas if you tried to build it on that date using upstream
+`nixpkgs` you'd be facing an error!
+
+We spent quite some time backporting fixes: we went back all the way to
+2019. The way this works, is that we start by checking out a `nixpkgs` commit on
+selected dates, then we "update" the [R]{.proglang}packages set by using the Posit CRAN and
+Bioconductor daily snapshots. Then, we backport as many fixes as possible, and
+ensure that a selection of popular packages work on both x86-linux (which
+includes Windows, through WSL) and aarch64-darwin (the M-series of Macs). Then
+we commit everything to a dated branch of the `rstats-on-nix/nixpkgs` fork. You
+can check out all the available dates by running: `rix::available_dates()`.
+We're pretty confindent that you should not face any issues when using [Nix]{.pkg} to
+build reproducible environments for R. However, should you face a problem, don't
+hesitate to open an issue!
+
+We have now packages and [R]{.proglang}versions working on Linux and macOS from March 2019
+to now. See [this
+repository](https://github.com/rstats-on-nix/daily_cran/blob/master/readme.md)
+that contains the scripts that allowed us to do it. Backporting fixes was
+especially important for Apple Silicon computers, as it took some time for this
+platform to work correctly on Nix. By backporting fixes, we can now provide
+olders versions of these packages for Apple Silicon as well!
+
+Using this approach, our fork now contains many more versions of working R
+packages than upstream. [rix]{.pkg} will thus likely keep pointing towards our fork
+in the future, and not upstream anymore. This should provide a much better user
+experience. An issue with our fork though, is that by backporting fixes, we
+essentially create new [Nix]{.pkg} packages that are not included in upstream, and thus,
+these are not built by Hydra, Nix's CI platform which builds binary packages. In
+practice this means that anyone using our fork will have to compile many
+packages from source. Now this is pretty bad, as building packages from source
+takes quite some time. But fear not, because thanks to
+[Cachix](https://www.cachix.org/) we now also have a dedicated binary cache of
+packages that complements the default, public [Nix]{.pkg} cache! We provide instructions
+on how to use Cachix, it's very easy, it's just running 2 additional commands
+after installing Nix. Using Cachix speeds up the installation process of
+packages tremendously. I want to give my heartfelt thanks to
+[Domen Kožar](https://www.cachix.org/about) for sponsoring the cache!
+
+Another thing we do with our fork is run an action every day at midnight, that
+monitors the *health* of the [R]{.proglang}packages set. Of course, we don't build every
+CRAN package, merely a handful, but these are among the most popular or the most
+*at-risk* of being in a broken state. See
+[here](https://github.com/rstats-on-nix/monitor_health/actions).
 
@@
 
-On versions of R released after version 3.6, these two lines return the
+On versions of [R]{.proglang}released after version 3.6, these two lines return the
 following vector: `r out_r`, but in versions prior to 3.6, the result will be
-different. This is a documented change as shown in the R version 3.6 release
+different. This is a documented change as shown in the [R]{.proglang}version 3.6 release
 notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]
@@
 
-It is possible to use `with_nix()` from an R session that was installed itself
-via Nix, or from a regular R installation.
+It is possible to use `with_nix()` from an [R]{.proglang}session that was installed itself
+via Nix, or from a regular [R]{.proglang}installation.
 
@@
 
-There are many tools to make one’s analyses reproducible, but Nix distinguishes
+There are many tools to make one’s analyses reproducible, but [Nix]{.pkg} distinguishes
 itself from these in that its main concern is in deploying the complete set of
-packages that are *closed under the ''depends on'' relation*: Nix does not
+packages that are *closed under the ''depends on'' relation*: [Nix]{.pkg} does not
 simply install one piece of software, but its dependencies, and their
-dependencies. This makes Nix extremely useful for reproducible research, but
+dependencies. This makes [Nix]{.pkg} extremely useful for reproducible research, but
 because it solves a complex problem, it is a complex tool. With [rix]{.pkg}, we
-wish to make using Nix more accessible by providing a familiar interface and
+wish to make using [Nix]{.pkg} more accessible by providing a familiar interface and
 user experience to [R]{.proglang} users. Because it is possible to build
@@
 
-Here is the Nix expression generated from the call to the `rix()`
+Here is the [Nix]{.pkg} expression generated from the call to the `rix()`
 function of section #sec-repro-nix:

### 2025-02-19 paper.qmd
Update paper.qmd

@@
 
-## The [Nix]{.pkg} package manager {#sec-nix}
+## The Nix package manager {#sec-nix}
 
@@
 projects and virtual environments. It is also possible to build multi-language
-environments, containing [R]{.proglang}and Python, a LaTeX distribution, and even install
+environments, containing [R]{.proglang} and Python, a LaTeX distribution, and even install
 one's favorite text editor to edit the project's source code. Essentially, one
@@
 
-## Reproducible development environments with [Nix]{.pkg} {#sec-repro-nix}
+## Reproducible development environments with Nix{#sec-repro-nix}
 
@@
 ```
-nix-shell -p [R]{.proglang}rPackages.rix"
+nix-shell -p R rPackages.rix
 ```

### 2025-02-20 paper.qmd
done?

@@
 packages that must download assets from the internet during installation, such
-as several BioConductor packages like [musData]{.pkg} for example. But this is
+as several Bioconductor packages like [musData]{.pkg} for example. But this is
 only a concern for the packagers, not users of the package that simply wish to
@@
 The first thing this fork allows us to do is offer a daily snapshot of CRAN.
-Every day, thanks to Github Actions, the [R]{.proglang} packages set gets
-updated, and the result commited to a dated branch. This has been going on since
-the 14th of December 2024 (see
+Every day, the [R]{.proglang} packages set gets updated, and the result commited
+to a dated branch using scripts running on GitHub Actions (a continuous
+integration and deployment platform). This has been going on since the 14th of
+December 2024 (see
 [here](https://github.com/rstats-on-nix/nixpkgs/tree/2024-12-14)). So when users
@@
 
-But this doesn’t mean that we recommend you
-use any date from the `rstats-on-nix/nixpkgs` fork: instead, each Monday,
-another action uses this fork and tries to build a set of popular packages on
-Linux and macOS, and only if this succeeds is the date added through a PR to the
-list of available dates on `{rix}`!
+But this doesn’t mean that we recommend users use any date from the
+`rstats-on-nix/nixpkgs` fork: instead, each Monday, another action uses this
+fork and tries to build a set of popular packages on Linux and macOS, and only
+if this succeeds is the date added through a PR to the list of available dates
+on `{rix}`. This means that we also monitor the health of popular [R]{.proglang}
+packages and can propose a fix upstream if a package stops building
+successfully, which is something that can sometimes happen. For example,
+recently RStudio was in a broken state due to an issue in one of its
+dependencies, `boost`. This is not the fault of anyone in particular: it's just
+that packages get updated and packages that depend on them should get updated as
+well: but if that doesn't happen quickly enough, the `nixpkgs` maintainer faces
+a conundrum. Either he or she doesn't update the package because it breaks
+others, but not updating a package could be a security vulnerability, or he or
+she updates the package, but now other, perhaps less critical packages are
+broken and need to be fixed, either by their upstream developers, or by the
+`nixpkgs` maintainer of said packages. In the case of RStudio a fix was proposed
+and promptly merged, but if users wanted to install RStudio during the time it
+took to fix it, they would have faced an error message.
 
-The reason this is done like this is to manage another *risk* of the upstream
-`nixpkgs`. As you know, `nixpkgs` is huge, and though the utmost care is taken
-by contributors and the PR review process is very strict, it can happen that
-updating packages breaks other packages. For example recently RStudio was in a
-broken state due to an issue in one its dependencies, `boost`. This is not the
-fault of anyone in particular: it's just that packages get updated and packages
-that depend on them should get updated as well: but if that doesn't happen
-quickly enough, the `nixpkgs` maintainer faces a conundrum. Either he or she
-doesn't update the package because it breaks others, but not updating a package
-could be a security vulnerability, or he or she updates the package, but now
-others, perhaps less critical packages are broken and need to be fixed, either
-by their upstream developers, or by the `nixpkgs` maintainer of said packages.
-In the case of RStudio a fix was proposed and promptly merged, but if you wanted
-to install RStudio during the time it took to fix it, you would have faced
-an error message, which isn't great if all you want is use [Nix]{.pkg} shells as
-development environments.
+However, thanks to our fork, we can back-port fixes and so if users try to
+install RStudio using the date `"2025-02-10"` (which is when it was broken
+on upstream `nixpkgs`), it'll work.
 
-So for us, having a fork allows us to backport these fixes and so if you try to
-install RStudio using the latest available date, which is `"2025-02-10"`, it’s
-going to work, whereas if you tried to build it on that date using upstream
-`nixpkgs` you'd be facing an error!
+We spent quite some time backporting fixes: we went back all the way back to
+March 2019. To do so, we start by checking out a `nixpkgs` commit on selected
+dates, then we "update" the [R]{.proglang} packages set by using the Posit CRAN
+and Bioconductor daily snapshots at that date. Then, we backport as many fixes
+as possible, and ensure that a selection of popular packages work on both
+x86-linux (which includes Windows, through WSL2) and aarch64-darwin (which is
+how Apple Silicon Apple computers are designated as). Then we commit everything
+to a dated branch of the `rstats-on-nix/nixpkgs` fork. Users can consult all the
+available dates by running: `rix::available_dates()`.
 
-We spent quite some time backporting fixes: we went back all the way to
-2019. The way this works, is that we start by checking out a `nixpkgs` commit on
-selected dates, then we "update" the [R]{.proglang}packages set by using the Posit CRAN and
-Bioconductor daily snapshots. Then, we backport as many fixes as possible, and
-ensure that a selection of popular packages work on both x86-linux (which
-includes Windows, through WSL) and aarch64-darwin (the M-series of Macs). Then
-we commit everything to a dated branch of the `rstats-on-nix/nixpkgs` fork. You
-can check out all the available dates by running: `rix::available_dates()`.
-We're pretty confindent that you should not face any issues when using [Nix]{.pkg} to
-build reproducible environments for R. However, should you face a problem, don't
-hesitate to open an issue!
-
-We have now packages and [R]{.proglang}versions working on Linux and macOS from March 2019
-to now. See [this
-repository](https://github.com/rstats-on-nix/daily_cran/blob/master/readme.md)
-that contains the scripts that allowed us to do it. Backporting fixes was
-especially important for Apple Silicon computers, as it took some time for this
-platform to work correctly on Nix. By backporting fixes, we can now provide
-olders versions of these packages for Apple Silicon as well!
-
-Using this approach, our fork now contains many more versions of working R
-packages than upstream. [rix]{.pkg} will thus likely keep pointing towards our fork
-in the future, and not upstream anymore. This should provide a much better user
-experience. An issue with our fork though, is that by backporting fixes, we
-essentially create new [Nix]{.pkg} packages that are not included in upstream, and thus,
-these are not built by Hydra, Nix's CI platform which builds binary packages. In
-practice this means that anyone using our fork will have to compile many
-packages from source. Now this is pretty bad, as building packages from source
-takes quite some time. But fear not, because thanks to
-[Cachix](https://www.cachix.org/) we now also have a dedicated binary cache of
-packages that complements the default, public [Nix]{.pkg} cache! We provide instructions
-on how to use Cachix, it's very easy, it's just running 2 additional commands
-after installing Nix. Using Cachix speeds up the installation process of
-packages tremendously. I want to give my heartfelt thanks to
-[Domen Kožar](https://www.cachix.org/about) for sponsoring the cache!
-
-Another thing we do with our fork is run an action every day at midnight, that
-monitors the *health* of the [R]{.proglang}packages set. Of course, we don't build every
-CRAN package, merely a handful, but these are among the most popular or the most
-*at-risk* of being in a broken state. See
-[here](https://github.com/rstats-on-nix/monitor_health/actions).
+An issue with forking `nixpkgs` though, is that by backporting fixes, we
+essentially create new [Nix]{.pkg} packages that are not included in upstream,
+and thus, these are not built by Hydra, Nix's CI platform which builds binary
+packages. In practice this means that anyone using our fork will have to compile
+many packages from source, which can be very time consuming. However, we also
+provide a binary cache of packages sponsored by
+[Cachix](https://www.cachix.org/) which complements the default, public
+[Nix]{.pkg} cache. We provide instructions on how to use Cachix in the package's
+Readme file. Using Cachix speeds up the installation process of packages
+tremendously, as they simply get downloaded instead of being built (we build
+a selection of packages only however, not the whole of CRAN and Bioconductor).
 
@@
 `.github/workflows/` which then runs the [targets]{.pkg} pipeline using Github
-Actions at each push, in a completely reproducible environment. For such an
+Actions on each push, in a completely reproducible environment. For such an
 example, consult the code in this
@@
 The final feature that we provide is the ability to run one arbitrary function
-from what we call a *subshell*. Suppose that you are running the latest version
+in what we call a *subshell*. Suppose that you are running the latest version
 of R, and that you wish or need to run one particular function from an older
@@
 
-On versions of [R]{.proglang}released after version 3.6, these two lines return the
-following vector: `r out_r`, but in versions prior to 3.6, the result will be
-different. This is a documented change as shown in the [R]{.proglang}version 3.6 release
-notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]
+On versions of [R]{.proglang} released after version 3.6, these two lines return
+the following vector: `r out_r`, but in versions prior to 3.6, the result will
+be different. This is a documented change as shown in the [R]{.proglang} version
+3.6 release notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]
 
@@
 
-It is possible to use `with_nix()` from an [R]{.proglang}session that was installed itself
-via Nix, or from a regular [R]{.proglang}installation.
+It is possible to use `with_nix()` from an [R]{.proglang} session that was
+installed itself via Nix, or from a regular [R]{.proglang} installation.
 
@@
 
-There are many tools to make one’s analyses reproducible, but [Nix]{.pkg} distinguishes
-itself from these in that its main concern is in deploying the complete set of
-packages that are *closed under the ''depends on'' relation*: [Nix]{.pkg} does not
-simply install one piece of software, but its dependencies, and their
-dependencies. This makes [Nix]{.pkg} extremely useful for reproducible research, but
-because it solves a complex problem, it is a complex tool. With [rix]{.pkg}, we
-wish to make using [Nix]{.pkg} more accessible by providing a familiar interface and
-user experience to [R]{.proglang} users. Because it is possible to build
-reproducible development shells using  [Nix]{.pkg}, it can accomodate any
-use case: running scripts, pipelines, interactive [shiny]{.pkg} applications,
-and serve [plumber]{.pkg} application programming interfaces.
+There are many tools to make one’s analyses reproducible, but [Nix]{.pkg}
+distinguishes itself from these in that its main concern is in deploying the
+complete set of packages that are *closed under the ''depends on'' relation*:
+[Nix]{.pkg} does not simply install one piece of software, but its dependencies,
+and their dependencies. This makes [Nix]{.pkg} extremely useful for reproducible
+research, but because it solves a complex problem, it is a complex tool. With
+[rix]{.pkg}, we wish to make using [Nix]{.pkg} more accessible to users of
+[R]{.proglang} by providing a familiar interface and user experience. Because it
+is possible to build reproducible development shells using [Nix]{.pkg}, it can
+accomodate any use case: running scripts, pipelines, interactive [shiny]{.pkg}
+applications, and serve [plumber]{.pkg} application programming interfaces.
 

### 2025-02-20 paper.qmd
typos

@@
 abstract: |
-  In order create an analysis that is reproducible,
-  it is not enough to write clean code and document it well. One
-  must also make sure to clearly list all the dependencies of the analysis
-  and ideally provide an easy way to install them.
-  There are several tools that can be used to list dependencies and to
-  make them easily installable by someone that wishes to reproduce a study,
-  such as [Docker]{.proglang}, a containerization solution. This paper will
-  present another tool that can be used to make projects reproducible,
-  the  [Nix]{.pkg} package manager, and an [R]{.proglang} package 
-  called [rix]{.pkg} that makes using  [Nix]{.pkg} much simpler 
-  for users of [R]{.proglang}.
+  In order create an analysis that is reproducible, it is not enough to write
+  clean code and document it well. One must also make sure to clearly list all
+  the dependencies of the analysis and ideally provide an easy way to install
+  them. There are several tools that can be used to list dependencies and to
+  make them easily installable by someone that wishes to reproduce a study, such
+  as [Docker]{.proglang}, a containerization solution. This paper will present
+  another tool that can be used to make projects reproducible, the [Nix]{.pkg}
+  package manager, and an [R]{.proglang} package called [rix]{.pkg} that makes
+  using [Nix]{.pkg} much simpler for users of [R]{.proglang}.
 
 keywords: [reproducibility, R, Nix]
-keywords-formatted: [reproducibility, "[R]{.proglang}", " [Nix]{.pkg}"]
+keywords-formatted: [reproducibility, "[R]{.proglang}", "[Nix]{.pkg}"]
+number-sections: true
 
@@
   system_packages = builtins.attrValues {
-    inherit (pkgs) [R]{.proglang};
+    inherit (pkgs) R;
   };
@@
 [R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
-older version of the package and also installs
-several TeXLive packages for authoring documents with \LaTeX, the development
-versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
-the Positron integrated development environment. You can see the resulting `default.nix` file in the @sec-appendix.
+older version of the package and also installs several TeXLive packages for
+authoring documents with \LaTeX, the development versions of [rix]{.pkg} and
+[fusen]{.pkg} (downloaded from Github), and finally the Positron integrated
+development environment. You can see the resulting `default.nix` file in the
+@sec-appendix.
 

### 2025-10-16 paper.qmd
Update paper.qmd

@@
 ---
-title: "Reproducible development environments with rix"
+title: "A Framework for Reproducible Data Science: Managing Environments and Pipelines with Nix, rix, and rixpress"
 format:
@@
 abstract: |
-  In order create an analysis that is reproducible, it is not enough to write
-  clean code and document it well. One must also make sure to clearly list all
-  the dependencies of the analysis and ideally provide an easy way to install
-  them. There are several tools that can be used to list dependencies and to
-  make them easily installable by someone that wishes to reproduce a study, such
-  as [Docker]{.proglang}, a containerization solution. This paper will present
-  another tool that can be used to make projects reproducible, the [Nix]{.pkg}
-  package manager, and an [R]{.proglang} package called [rix]{.pkg} that makes
-  using [Nix]{.pkg} much simpler for users of [R]{.proglang}.
+  To create a reproducible analysis, it is not enough to write clean, well-documented code. 
+  One must also ensure that all software dependencies are clearly declared and, ideally, 
+  provide a simple way to install them. Several tools exist for this purpose, such as the 
+  containerization solution [Docker]{.proglang}. A complete solution, however, requires not only 
+  managing the computational environment but also orchestrating the execution of the analysis itself
+  in a robust and automated way. This paper presents a comprehensive framework based on the [Nix]{.pkg} 
+  package manager to solve both problems. We first introduce the [R]{.proglang} package [rix]{.pkg}, 
+  which simplifies the creation of reproducible environments with [Nix]{.pkg}. We then introduce 
+  [rixpress]{.pkg}, a second [R]{.proglang} package that leverages these environments to define and 
+  execute complex, multi-language analytical pipelines. Together, these tools provide a powerful, 
+  unified approach to achieving high standards of computational reproducibility.
 
@@
 
-A final, and optional, step towards the gold standard is to use a build
+A final step towards the gold standard is to use a build
 automation tool such as [Make]{.pkg} to run the whole analysis when the
@@
 to deal with package versions, then [Docker]{.pkg} for all the other software,
-including the programming language that was used. Ideally, a build automation
-tool such as [Make]{.pkg} should be added to the list. It should also be noted
+including the programming language that was used, as well as a build automation
+tool. It should also be noted
 that these requirements are not limited to the [R]{.proglang} programming
@@
 build automation tool such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
-to [R]{.proglang} programmers, we developed the [rix]{.pkg} package which we
+to [R]{.proglang} programmers, we developed the [rix]{.pkg} and [rixpress]{.pkgs} packages which we
 will present in this article.
@@
 
+## Orchestrating the Workflow with `rixpress` {#sec-rixpress}
+
+Defining a reproducible environment with [rix]{.pkg} solves the first major challenge of reproducibility.
+The second is to reliably and efficiently execute the analysis workflow within that environment.
+This is the role of the [rixpress]{.pkg} package.
+
+As mentioned in the introduction, a build automation tool like [targets]{.pkg} is invaluable for managing complex analyses.
+It automatically tracks dependencies between code and data objects, caching results and only re-computing steps that have changed.
+One can run a [targets]{.pkg} pipeline inside a Nix environment to make it reproducible.
+However, this approach has limitations: the entire pipeline must run in a single environment, and orchestrating steps across different languages (e.g., [R]{.proglang} and [Python]{.proglang}) requires manual handling via packages like [reticulate]{.pkg}.
+
+[rixpress]{.pkg} overcomes these limitations by using Nix not just as a package manager, but as the build automation engine itself.
+In a [rixpress]{.pkg} pipeline, each step is defined as a Nix derivation.
+This powerful abstraction provides two key benefits:
+
+1.
+True Polyglot Pipelines: Each step can be assigned its own, distinct Nix environment.
+A Python step can run in a pure Python environment, an [R]{.proglang} step in an [R]{.proglang} environment, and a Quarto rendering step in yet another, all within the same pipeline.
+
+2.
+Deep Reproducibility: Because each step is a Nix derivation, its execution is hermetically sealed and its output is cached in the Nix store based on the hash of all its inputs.
+This provides caching and reproducibility at a level deeper than most other tools.
+On top of this, all artifacts built by the pipeline are direct children of the computational environment: should the environment change in any way, all the artifacts would get rebuilt, ensuring that the environment and the artifacts are never out of sync.
+
+The user defines the pipeline in a simple [R]{.proglang} script, using functions heavily inspired by [targets]{.pkg}.
+Let's consider a simple pipeline that reads data with Python (`polars`), filters it, passes it to [R]{.proglang} (`dplyr`) for further analysis, and then compiles a Quarto document.
+
+First, the pipeline is defined in an [R]{.proglang} script as a list of `rxp_*` function calls:
+
+```{r, eval=FALSE}
+# In a file like 'pipeline.R'
+library(rixpress)
+
+# Define the sequence of steps
+pipeline_steps <- list(
+  # Step 1: Read data using Python in its own environment
+  rxp_py_file(
+    name = mtcars_pl,
+    path = "data/mtcars.csv",
+    read_function = "lambda x: polars.read_csv(x, separator='|')",
+    nix_env = "python-env.nix" # A Nix file defining the Python env
+  ),
+
+  # Step 2: Filter the data in Python
+  rxp_py(
+    name = mtcars_filtered,
+    expr = "mtcars_pl.filter(polars.col('am') == 1).to_pandas()",
+    nix_env = "python-env.nix"
+  ),
+
+  # Step 3: Convert the Python pandas DataFrame to an R data.frame
+  # This uses the main project environment ('default.nix')
+  rxp_py2r(
+    name = mtcars_r,
+    expr = mtcars_filtered
+  ),
+
+  # Step 4: Process the data in R
+  rxp_r(
+    name = mtcars_summary,
+    expr = dplyr::summarise(mtcars_r, avg_mpg = mean(mpg))
+  )
+)
+
+# Generate the 'pipeline.nix' file from the R list
+rixpress(pipeline_steps, project_path = ".")
+```
+
+The user first creates the necessary environment files (e.g., `default.nix`, `python-env.nix`) using `rix`. 
+Then, running the script above does not execute the pipeline but instead generates a `pipeline.nix` file, which is the complete, declarative definition of the workflow.
+
+To build the pipeline, the user runs:
+
+```{r, eval=FALSE}
+rxp_make()
+```
+
+[Nix]{.pkg} then executes each step in the correct order, building any dependencies as needed.
+Because the output of each step is cached, subsequent runs of `rxp_make()` will only re-execute steps whose code or inputs have changed.
+This provides the same efficiency as [targets]{.pkg} but with the added benefits of multi-language support and the bit-for-bit reproducibility of Nix.
+Artifacts from any step can be loaded into an interactive [R]{.proglang} session for inspection using `rxp_read("artifact_name")` or `rxp_load("artifact_name")`.
+
+For Python users, there is also a port of [rixpress]{.pkg}, called [ryxpress]{.pkg} that uses the same definition of the pipeline as an R script, but allows building the pipeline and viewing the outputs from a Python session.
+
 ## Additional features of the rix package

### 2025-10-17 paper.qmd
Update paper.qmd
getting ready

--- 2025-02-20 paper.qmd
@@
 ---
-title: "Reproducible development environments with rix"
+title: "A Framework for Reproducible Data Science using Nix"
 format:
@@
     url: https://www.brodrigues.co 
-  - name: Philipp Baumann
-    affiliations:
-      - Plus Affiliation
 abstract: |
-  In order create an analysis that is reproducible, it is not enough to write
-  clean code and document it well. One must also make sure to clearly list all
-  the dependencies of the analysis and ideally provide an easy way to install
-  them. There are several tools that can be used to list dependencies and to
-  make them easily installable by someone that wishes to reproduce a study, such
-  as [Docker]{.proglang}, a containerization solution. This paper will present
-  another tool that can be used to make projects reproducible, the [Nix]{.pkg}
-  package manager, and an [R]{.proglang} package called [rix]{.pkg} that makes
-  using [Nix]{.pkg} much simpler for users of [R]{.proglang}.
+  To create a reproducible analysis, it is not enough to write clean,
+  well-documented and tested code. One must also ensure that all software
+  dependencies are clearly declared and, ideally, provide a simple way to
+  install them. Several tools exist for this purpose, such as the
+  containerization solution [Docker]{.pkg}. A complete solution, however,
+  requires not only managing the computational environment but also
+  orchestrating the execution of the analysis itself in a robust and automated
+  way. This paper presents a comprehensive framework based on the [Nix]{.pkg}
+  package manager to solve both problems. We first introduce the [R]{.proglang}
+  package [rix]{.pkg}, which simplifies the creation of reproducible
+  environments with [Nix]{.pkg}. We then introduce [rixpress]{.pkg}, a second
+  [R]{.proglang} package that leverages these environments to define and execute
+  complex, multi-language analytical pipelines. Together, these tools provide a
+  powerful, unified approach to achieving high standards of computational
+  reproducibility.
 
@@
 
-@peng2011 introduced the idea of reproducibility being on a continuum: on one of
-the ends of this continuum, we only have access to the paper describing the
-studies, which is not reproducible at all. If in addition to this paper authors
-make the original source code of the analysis available, reproducibility is
-improved. Adding the original data improves reproducibility yet again. Finally,
-if to all this the authors add what Roger Peng referred to as the *linked and
-executable code and data*, we reach the gold standard of reproducibility.
+@peng2011 introduced the concept of reproducibility as a *continuum*. At one end
+lies the least reproducible state, where only a paper describing the study is
+available. Reproducibility improves when authors share the original source code,
+improves further when they include the underlying data, and reaches its highest
+level when what Roger Peng called *linked and executable code and data* are
+provided.
 
-*Linked and executable code and data* refer to the compiled source code scripts,
-but for the purposes of this article, we will assume this meas the
-*computational environment*. The computational environment is all the software
-required to run the analysis. Here too, we can speak of a continuum. One could
-simply name and list the software used: for example, the [R]{.proglang}
-programming language. Sometimes, authors have the courtesy to also state the
-version of [R]{.proglang} used. Some authors go further, and also list the
-packages used, and ideally with their versions as well. Authors rarely state the
-operating system on which the analysis was done, even though it has been shown
-that running the same analysis with the same software but on different operating
-systems could lead to different results, as described in @neupane2019. Authors
-also only very rarely provide instructions to install the required tools and
-software in order to reproduce their studies.
+By *linked and executable code and data*, Peng referred to compiled source code
+and runnable scripts. In this paper, we interpret this notion more broadly as
+the *computational environment*: the complete set of software required to
+execute an analysis. Here too, a continuum exists. At the minimal end, authors
+might only name the main software used—say, the [R]{.proglang} programming
+language. More careful authors might also specify the version of [R]{.proglang},
+or list the additional packages and their versions. Rarely, however, do authors
+specify the operating system on which the analysis was performed, even though
+differences in operating systems can lead to divergent results when using the
+same code and software versions, as shown by @neupane2019. It is even less
+common for authors to provide step-by-step installation instructions for the
+required software stack.
 
-But even assuming that authors provided clear instructions to install the exact
-requirements of a study, it might happen that these instructions do not even
-work on different platforms, or on different versions of the same platform. And
-reproducibility is not the only activity that might get impacted by a lack of
-clear instructions: if an author works on several machines, the exact same
-computational environment should be used on all the machines, and the same is
-true when a team of authors collaborate.
+Even when such instructions are given, they often fail across different
+platforms or versions of the same platform. This lack of portability not only
+hinders reproducibility but also complicates everyday research workflows.
+Researchers working across multiple machines must be able to recreate the same
+environment consistently, and collaborators must share identical computational
+setups to avoid inconsistencies.
 
-There are tools available that authors can use to reach the gold standard of a
-fully reproducible study, or to successfully deploy the same development
-environment on several machines. Let us first consider the task of listing the
-software used. [R]{.proglang} provides the `sessionInfo()` function whose output
-can be saved into a file. Below is an example output of `sessionInfo()`:
+Finally, once the execution environment is correctly configured, additional
+clarity is needed on how to *run* the project itself. Which packages should be
+loaded first? Which scripts should be executed, and in what order? Without clear
+documentation or a logically organized project structure, these operational
+details become yet another barrier to reproducibility.
 
-```{r, eval = FALSE}
-sessionInfo()
-```
+A range of tools now exist to help researchers approach the gold standard of
+full reproducibility, or to consistently deploy the same development environment
+across multiple machines. Let us first consider the most basic step in this
+process: listing the software used. In [R]{.proglang}, the `sessionInfo()`
+function provides a concise summary of the software environment, including the R
+version, platform details, and all loaded packages. Its output can be saved to a
+file and included as part of a study’s reproducibility record. Below is an
+example output from `sessionInfo()`:
 
-```
+```{=tex}
+\begin{CodeInput}
+R> sessionInfo()
+\end{CodeInput}
+\begin{CodeOutput}
 R version 4.3.2 (2023-10-31)
@@
 locale:
- [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
- [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
- [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
- [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
- [9] LC_ADDRESS=C               LC_TELEPHONE=C            
-[11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C
+ LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+ LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
+ LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
+ LC_PAPER=en_US.UTF-8       LC_NAME=C                 
+ LC_ADDRESS=C               LC_TELEPHONE=C            
+ LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C
 
@@
 attached base packages:
-[1] stats     graphics  grDevices utils     datasets  methods
-[7] base
+ stats     graphics  grDevices utils     datasets  methods   base
 
 other attached packages:
-[1] nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163
+ nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163
 
 loaded via a namespace (and not attached):
-[1] compiler_4.3.2 Matrix_1.6-1.1 tools_4.3.2    splines_4.3.2
-[5] grid_4.3.2     lattice_0.21-9
+ compiler_4.3.2  Matrix_1.6-1.1  tools_4.3.2     splines_4.3.2
+ grid_4.3.2      lattice_0.21-9 
+\end{CodeOutput}
 ```
 
-If an author provides this information, other people trying to reproduce the
-study (or the author him- or herself in the future), or co-authors, can read
-this file and see which version of [R]{.proglang} was used, and which packages
-(and their versions) were used as well. However, others would still need to
-install the correct versions of each package manually. This can be in certain
-cases challenging, especially if packages that depend on system-level
-dependencies are required. For example, the [nloptr]{.pkg} package requires a
-binary of [nlopt]{.proglang} to be available, or if not, [cmake]{.proglang} on
-Linux or macOS.
+When an author includes this information, others attempting to reproduce the
+study (future readers, collaborators, or even the author at a later
+time) can easily see which version of [R]{.proglang} and which packages (with
+their versions) were used. However, reproducing the environment still requires
+manually installing the correct package versions. This can be difficult,
+especially when packages depend on system-level libraries. For example, the
+[nloptr]{.pkg} package requires a precompiled [nlopt]{.proglang} binary, or
+alternatively, [cmake]{.proglang} when building from source on Linux or macOS.
 
-A better alternative to simply listing packages and their versions is to use the
-[renv]{.pkg} package which generates a so-called `renv.lock` file which also
-lists [R]{.proglang} and package versions. Here is an example of such an
-`renv.lock` file:
+A more robust approach than simply listing package versions is to use the
+[renv]{.pkg} package. [renv]{.pkg} captures the project’s software state and
+writes it to a `renv.lock` file, which includes the exact versions of
+[R]{.proglang} and all required packages. This lockfile serves as a blueprint
+for restoring the environment automatically, ensuring that others can recreate
+the same setup with minimal effort. Below is an example of an `renv.lock` file:
 
-```
+```{=tex}
+\begin{Code}
 {
@@
       ]
+    }
 
-    ***lines below omitted***
+    ... lines below omitted ...
+  }
+}
+\end{Code}
 ```
 
-This file lists every package alongside their versions and the repository from
-which they were downloaded. Generating this file only requires one to run the
-`renv::init()` function. Someone else can then restore the same package library
-by running `renv::restore()`. The exact same packages get installed in an
-isolated, project-specific library which doesn't interfere with the main
-library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
-used though, so installing the right version of [R]{.proglang}needs to be handled
-separately. [renv]{.pkg} will also not install system-level dependencies such
-as [cmake]{.pkg} for [nloptr]{.pkg} on Linux or macOS, so these still need to
-be handled separately, as described in [renv]{.pkg}’s Readme: 
-\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.
+This lockfile lists each package alongside its version and the repository from
+which it was downloaded. Creating it requires only a single command:
+`renv::init()`. Others can then reproduce the same package library by running
+`renv::restore()`, which installs the exact package versions in an isolated,
+project-specific library that does not interfere with the user’s global R
+library.
+
+However, [renv]{.pkg} does not restore the version of [R]{.proglang} itself.
+Installing the correct R version must therefore be done separately. Similarly,
+`renv` does not handle system-level dependencies such as [cmake]{.proglang},
+which is required by [nloptr]{.pkg} on Linux and macOS. These dependencies must
+be managed manually or through other tools. Further details are discussed in the
+[renv]{.pkg} documentation: \url{https://rstudio.github.io/renv/articles/renv.html#caveats}.
 
@@
 
-```
-groundhog.library("
-    library(purrr)
-    library(ggplot2)",
-    "2017-10-04",
-    tolerate.R.version = "4.2.2")
+```{=tex}
+\begin{CodeInput}
+R> groundhog.library("
++   library('purrr')
++   library('ggplot2')",
++   "2017-10-04",
++   tolerate.R.version = "4.2.2")
+\end{CodeInput}
 ```
 
-These packages also get installed in a project-specfic library so there is no
-interference between these packages and other versions of the same packages that
-one might use for other projects. Because [groundhog]{.pkg} does not install
-[R]{.proglang} itself, users should either install the required version
-themselves, or they should use the `tolerate.R.version` argument as shown in the
-example above. Otherwise, [groundhog]{.pkg} would not continue with the
-installation of the packages. Another alternative to [renv]{.pkg}, developed by
-@chan2023 is [rang]{.pkg}, which also installs packages as they were on a given
-date. Yet another way to install packages as they were on a give date is to use
-the Posit Package Manager, which provides snapshots of CRAN. For example, to
-install the required packages for an analysis as they were on the 30th of June
-2023, one could add the following line to the `.Rprofile` file:
+Packages installed with [groundhog]{.pkg} are placed in a project-specific
+library, ensuring they do not interfere with other versions of the same packages
+used in different projects. Since [groundhog]{.pkg} does not install
+[R]{.proglang} itself, users must either install the required R version manually
+or use the `tolerate.R.version` argument, as illustrated in the example above.
+Without this argument, [groundhog]{.pkg} will not proceed with package
+installation if the R version does not match the expected one.
 
-```
-options(repos =
-  c(REPO_NAME =
-    "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
-  )
-)
+An alternative to [renv]{.pkg} is [rang]{.pkg}, developed by @chan2023, which
+similarly allows users to install packages as they existed on a specific date.
+Another approach is to use the Posit Package Manager, which provides dated
+snapshots of CRAN. For instance, to ensure that packages are installed as they
+were on June 30, 2023, one could include the following line in the `.Rprofile`
+file:
+
+```{=tex}
+\begin{CodeInput}
+R> options(repos =
++    c(REPO_NAME =
++      "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
++    )
++  )
+\end{CodeInput}
 ```
 
-The `.Rprofile` file gets read by [R]{.proglang}when starting a new session, which means
-that every call to the `install.packages()` function will now install the
-packages from this snapshotted mirror. However, unless the user explicitly and
-manually manages the different projects' libraries, using the Posit Package
-Manager will install all the package for all the projects at the given date.
+The `.Rprofile` file is read by [R]{.proglang} at the start of each new session,
+meaning that every call to `install.packages()` will install packages from the
+specified snapshot mirror. However, unless users explicitly manage separate
+libraries for different projects, the Posit Package Manager will install all
+packages for all projects as they existed on that snapshot date.
 
-The next step in reaching the gold standard of reproducibility would be to not
-only install the right packages used for the analysis, but also the right
-version of [R]{.proglang}. Of course, it would be possible to install the right
-version manually, but here too, there are tools that simplify the process. One
-such tool is [rig]{.pkg} by the [R]{.proglang}infrastructure team [-@rlib2023]. [rig]{.pkg}
-makes it easy to install different versions of [R]{.proglang}, so one could use
-it to install the right version to reproduce an analysis, and then use one of
-the listed packages above to install the right library of [R]{.proglang}
-packages. However, this involves many manual steps and is thus error prone and
-time-consuming.
+The next step toward the gold standard of reproducibility is ensuring not only
+that the correct package versions are installed, but also that the correct
+version of [R]{.proglang} itself is used. While this can be done manually,
+dedicated tools make the process more efficient. One such tool is [rig]{.pkg},
+developed by the [R]{.proglang} Infrastructure Team [-@rlib2023]. [rig]{.pkg}
+simplifies installing and managing multiple R versions, allowing users to match
+the exact version needed for a given analysis. Once the correct R version is
+installed, one can then use tools such as [renv]{.pkg}, [groundhog]{.pkg}, or
+[rang]{.pkg} to recreate the associated package library.
 
-The final tool one could use to reach the gold standard of reproducibility is to
-package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
-[Docker]{.pkg} image. [Docker]{.pkg} is a containerisation tool: using
-[Docker]{.pkg} it is possible to package some *data product* with its
-dependencies into an image. A statistical analysis, from the simplest to the
-most complex one, can be seen as such a data product that requires many software
-dependencies. *Dockerizing* an analysis consists in first building an image: in
-the build step, the dependencies of the analysis have to be installed and this
-can be achieved using the tools mentioned above. The scripts to run the analysis
-and the data are also added to the image at build time. These steps are written
-down in a simple text file called a `Dockerfile`. This `Dockerfile` is the
-definition of the [Docker]{.pkg} image.
+However, this multi-step process remains manual, error-prone, and
+time-consuming, highlighting the need for more comprehensive and automated
+solutions to achieve full computational reproducibility.
 
-It is possible to then execute the code of a study inside a so-called container,
-which is a running instance of an image. Executing, or running, these containers
-is done non-interactively, and so it is possible to instantiate the correct
-environment to run a study in a single command.
+The final tool for achieving the gold standard of reproducibility is to bundle
+the exact version of [R]{.proglang} and its packages within a [Docker]{.pkg}
+image. [Docker]{.pkg} is a containerization platform that allows packaging a
+*data product* together with all its dependencies into a self-contained image. A
+statistical analysis can be viewed as such a data product, requiring a specific
+set of software dependencies to run. *Dockerizing* an analysis involves building
+an image that installs these dependencies, often using the tools mentioned
+earlier, and includes the analysis scripts and data (but data can also be
+dynamically made available at run-time, to avoid having to copy it inside the
+image). The steps to create this image are specified in a simple text file
+called a `Dockerfile`, which defines the structure and content of the
+[Docker]{.pkg} image.
 
-The advantage of using [Docker]{.pkg} is not only about having the right version
-of [R]{.proglang} and [R]{.proglang} packages bundled alongside the analysis
-scripts: [Docker]{.pkg} images are actually almost complete Linux operating
-systems, which means that system-level dependencies also get bundled into the
-image. This is especially useful for packages that rely on external system-level
-dependencies to run, such as for [nloptr]{.pkg}, as these external system-level
-dependencies are thus also made available to future replicators of the study
-within the [Docker]{.pkg} image. The image can also be shared.
+Once the image is built, the analysis can be executed inside a *container*,
+which is a running instance of the image. Containers are typically run
+non-interactively, allowing the complete computational environment to be
+instantiated and executed with a single command.
 
-The Rocker project initiated by @boettiger2017 provides many pre-built
-[Docker]{.pkg} to the [R]{.proglang} community of users, already containing
-specific [R]{.proglang} versions and even packages. These images can be used as
-bases to build other images containing statistical analyses more easily than
-starting from a bare-bones image.
+The strength of [Docker]{.pkg} lies not only in its ability to encapsulate the
+correct versions of [R]{.proglang} and its packages but also in its inclusion of
+system-level dependencies. Since [Docker]{.pkg} images are nearly complete Linux
+systems, they naturally bundle libraries and tools required by certain R
+packages (for example, [nloptr]{.pkg}’s dependency on the [nlopt]{.proglang}
+library). This ensures that future replicators of a study have access to the
+same software environment, including all necessary system components. Moreover,
+these images can be easily shared, enabling seamless reproducibility across
+machines and collaborators.
 
-While [Docker]{.pkgs} offers a great solution to the reproducibility problem, it
-can be awkward to work interactively from a running container: containers can be
-changed at run-time, but any change will be lost once the container is stopped.
-Also, while it is possible to launch graphical programs from a container, this
-is generally not easy to set up and only works, as far as we know, only on Linux
-operating systems. In most cases, users of [Docker]{.pkg} will instead use the
-container to run a web server and the web based application of a popular
-integrated development environment (IDE).
+The Rocker project, introduced by @boettiger2017, provides a collection of
+pre-built [Docker]{.pkg} images for the [R]{.proglang} community. These images
+come with specific R versions and, in some cases, preinstalled packages, making
+them convenient base images for building reproducible analysis environments
+without starting from scratch.
 
-A common approach seems to be to work on a study interactively using a standard
-installation of [R]{.proglang}and [R]{.proglang}packages, and after the study is done, take some time to
-write the required `Dockerfile` to provide an easy way for future replicators to
-reproduce the results of the study. But this approach does not solve the problem
-of deploying the same environment on several machines, and so ideally authors
-would develop using a containerized development environment. Another challenge
-of using [Docker]{.pkg} is that, as stated already, [Docker]{.pkg} images are
-actually almost complete Linux operating systems, so experience with Linux is
-highly recommended if one wishes to properly write a `Dockerfile`. Finally, one
-also needs to be careful and write the `Dockerfile` in such a way that the build
-process will always result in the same image. This is only possible if a
-specific digest of a base image is used instead of a tagged version, but in
-practice, this is rarely done.
+Despite its strengths, [Docker]{.pkg} can be cumbersome for interactive use.
+Although containers can be modified at runtime, any changes are lost once the
+container stops unless they are explicitly saved. Running graphical applications
+from containers is also possible, but it is generally difficult to configure and
+tends to work reliably only on Linux systems. For this reason, [Docker]{.pkg} is
+most often used to run web-based applications (such as integrated development
+environments, IDEs, like the web version of RStudio) inside containers, offering
+a reproducible yet flexible environment for interactive work.
 
-A final, and optional, step towards the gold standard is to use a build
-automation tool such as [Make]{.pkg} to run the whole analysis when the
-container is executed, instead of executing scripts. Build automation tools make
-it easier to run arbitrary code in a series of well-defined steps.
-[R]{.proglang} programmers can use the [targets]{.pkg} by @landau2021 as a build
-automation tool.
+Some IDEs provide built-in support for executing code directly within a running
+container, allowing a more seamless workflow. However, these setups typically
+require additional configuration, and many general-purpose IDEs are not well
+optimized for statistical or data science workflows (though Positron may soon
+change that).
 
-@mcdermott2021 is an example of a scientific study that reached the gold
-standard of reproducibility. The author of this paper set up an accompanying
-Github repository to the
-paper^[https://github.com/grantmcdermott/skeptic-priors] containing all the
-instructions to install the required software and then run the analysis. If we
-take a closer look at this repository, we will notice that many of the tools
-previously mentioned were used to capture the compatutational environment and
-make it available to other researchers:
+A common practice is to conduct a study interactively using a standard
+installation of [R]{.proglang} and its packages, and only after completing the
+analysis, create a `Dockerfile` to allow others to reproduce the results easily.
+For instance, the [dockerfiler]{.pkg} package [@dockerfiler] can automatically
+generate a `Dockerfile` from an existing `renv.lock` file, simplifying this
+process. While this approach enables post-hoc reproducibility, it does not solve
+the challenge of consistently deploying the same environment across multiple
+machines as co-authors are working on the project. Ideally, analyses would be
+developed directly within a containerized environment from the start, ensuring
+that the computational setup is reproducible throughout the project lifecycle.
 
-- Packages and their versions were listed and saved into an `renv.lock` file;
-- A `Makefile` was used to run the whole analysis and compile the paper;
-- A `Dockerfile` was used to provide the complete computational environment,
-  including the right version of [R]{.proglang}, to run the whole analysis
-  easily.
+Another challenge of using [Docker]{.pkg} is that its images are effectively
+minimal Linux systems, so familiarity with Linux is highly recommended for
+writing reliable and efficient `Dockerfile`s. To guarantee that a build
+consistently produces the same image, the `Dockerfile` should reference specific
+image *digests* rather than general *tags*. In practice, however, tagged
+versions are more commonly used, which can introduce potential reproducibility
+issues. [dockerfiler]{.pkg} is an R package that lowers the
+barrier to entry by automating parts of the `Dockerfile` creation process, but
+they cannot fully replace the need for basic Linux knowledge.
 
-However, reaching this gold standard is quite costly: one needs to learn a tool
-to deal with package versions, then [Docker]{.pkg} for all the other software,
-including the programming language that was used. Ideally, a build automation
-tool such as [Make]{.pkg} should be added to the list. It should also be noted
-that these requirements are not limited to the [R]{.proglang} programming
-language. A very similar approach should be taken if one uses
-[Python]{.proglang}, or any other programming language. Furthermore, the
-previous discussion also implies that for an analysis to actually be
-reproducible at all, it must be made using free and open source software, as
-proprietary software cannot be freely shared with others that wish to reproduce
-a study, and there's no guarantee that others will have access to the
-proprietary software used.
+A final step toward the gold standard of reproducibility is to automate the
+analysis execution using a build system such as [Make]{.pkg}, rather than
+relying on ad hoc script execution. Build automation tools enable analyses to be
+defined as a series of reproducible, well-ordered steps that can be executed
+consistently. Within [R]{.proglang}, a modern and powerful alternative is the
+[targets]{.pkg} package by @landau2021, which provides declarative workflows for
+reproducible data analysis.
 
-As an alternative to the above approach, we will present the [Nix]{.pkg} package
-manager, which is available for all major operating systems and focuses on
-installing and building software in a reproducible manner. This means that this
-single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg} and even a
-build automation tool such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
-to [R]{.proglang} programmers, we developed the [rix]{.pkg} package which we
-will present in this article.
+@mcdermott2021 provides a clear example of a scientific study that achieved the
+gold standard of reproducibility. The author created an accompanying GitHub
+repository^[https://github.com/grantmcdermott/skeptic-priors] containing
+detailed instructions to install the required software and execute the analysis.
+Examining this repository reveals the use of multiple tools to capture and share
+the computational environment:
+
+- Packages and their versions were recorded in an `renv.lock` file;
+- A `Makefile` was used to automate the full analysis and compile the paper;
+- A `Dockerfile` provided a complete computational environment, including the
+  correct version of [R]{.proglang}, enabling easy execution of the entire
+  workflow.
+
+Reaching this gold standard, however, is resource-intensive. It requires
+learning a tool for managing package versions, mastering [Docker]{.pkg} for the
+programming language and other software dependencies, and using a build
+automation tool. Importantly, these considerations are not unique to
+[R]{.proglang}: similar approaches are needed for [Python]{.proglang} or any
+other programming language, and complexity further increases if several
+programming languages are needed for a single project.
+
+As an alternative, we present the [Nix]{.pkg} package manager, which supports
+all major operating systems and emphasizes reproducible software installation
+and building. [Nix]{.pkg} can replace [Docker]{.pkg}, [renv]{.pkg}, and even
+build automation tools such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
+for [R]{.proglang} users, we developed the [rix]{.pkg} and [rixpress]{.pkgs}
+packages, which are introduced and discussed in this article.
 
@@
 
-[Nix]{.pkg} is a package manager that can be used to install and build software
-in a completely reproducible manner. As of writing, it contains more than
-100.000 packages, including almost the entirety of CRAN and Bioconductor. This
-means that using [Nix]{.pkg}, it is possible to install not only [R]{.proglang},
-but also all the packages required for a project. [Nix]{.pkg} is actually the
-package manager of a Linux distribution called NixOS, but can be installed as a
-stand-alone tool for other Linux distributions, macOS, and on the Windows
-Subsystem for Linux on Windows^[Which means that in practical terms, non-Nixos
-Linux distributions and Windows can be considered as the same platform.].
+[Nix]{.pkg} is a package manager designed to install and build software in a
+fully reproducible manner. As of this writing, it provides access to over
+120000 packages, including nearly all of CRAN and Bioconductor. This allows
+users to install not only [R]{.proglang} itself but also all packages required
+for a given project. Although [Nix]{.pkg} is the package manager of the NixOS
+Linux distribution, it can also be installed as a standalone tool on other Linux
+distributions, macOS, and the Windows Subsystem for Linux^[Effectively treating
+non-NixOS Linux distributions and Windows as equivalent platforms in practice.].
 
-The reason why one should use [Nix]{.pkg} to install [R]{.proglang} packages,
-and not use the usual, built-in, `install.packages()` [R]{.proglang} function
-instead, is that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
-sure to install every dependency of every package, whether this dependency is
-another [R]{.proglang} package or a system-level dependency.
+The advantage of using [Nix]{.pkg} to install [R]{.proglang} packages, instead
+of the standard `install.packages()` function, is that [Nix]{.pkg} ensures all
+dependencies of each package are installed, whether they are other
+[R]{.proglang} packages or system-level libraries.
 
-For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
-[Java]{.proglang} programming language to be installed to successfully install.
-Depending on what system one tries to install [xlsx]{.pkg}, installing
-[Java]{.proglang} might not be easy, nor even possible. But with [Nix]{.pkg}, it
-suffices to declare that the [xlsx]{.pkg} package is needed for the project, and
-[Nix]{.pkg} will install [Java]{.proglang} and configure it. This is possible
-because the maintainers of the [R]{.proglang} language for [Nix]{.pkg} declared
-[Java]{.proglang} as a dependency of [xlsx]{.pkg}. From the perspective of the
-end-users, this all happens automatically, thanks to the efforts of the
-maintainers.
+For example, the [xlsx]{.pkg} package requires [Java]{.proglang} to be
+installed. On some systems, installing [Java]{.proglang} may be difficult or
+even impossible manually. With [Nix]{.pkg}, however, the user only needs to
+declare that [xlsx]{.pkg} is required for the project; [Nix]{.pkg} automatically
+installs and configures [Java]{.proglang} as a dependency. This works because
+the maintainers of the [R]{.proglang} language packages for [Nix]{.pkg} have
+declared [Java]{.proglang} as a dependency of [xlsx]{.pkg}, allowing the process
+to happen seamlessly from the end-user perspective.
 
@@
 
-But how does [Nix]{.pkg} achieve this? Where do these packages, or closures,
-come from? When installing a package using [Nix]{.pkg}, an expression written in
-the [Nix]{.pkg} programming language gets downloaded from the `nixpkgs` Github
-repository and evaluated. `nixpkgs` is the name of the repository that contains
-the more than 100'000 pieces of software that can be installed with the
-[Nix]{.pkg} manager. Expressions contain so-called *derivations*. A derivation
-defines a build: its dependencies, commands to build and install the package in
-question, and then an output. Most of the time, a derivation downloads source
-code, builds the software from the source and then outputs a compiled binary.
-Derivations are extremely flexible, and by writing one's own, it is possible to
-define and build an environment for a project in a reproducible manner. The
-dependencies of these derivations are also defined in other expressions, which
-will also get installed if needed. As explained above, the maintainers of
-[R]{.proglang} for [Nix]{.pkg} declare the right dependencies for the packages
-to turn them into component closures.
+But how does [Nix]{.pkg} achieve this reproducibility? Where do these packages,
+or *closures*, come from? When installing a package with [Nix]{.pkg}, an
+expression written in the [Nix]{.pkg} language is downloaded from the `nixpkgs`
+GitHub repository and evaluated. These expressions define *derivations*, which
+describe how to build a package: specifying its dependencies, the commands to
+build and install it, and the resulting output. Typically, a derivation
+downloads the source code, compiles it, and outputs a binary. Derivations are
+highly flexible; by writing custom derivations, users can define and build
+reproducible environments instead of a single program. The dependencies of each
+derivation are themselves defined in other expressions and are automatically
+installed if needed. In the case of [R]{.proglang}, maintainers declare the
+appropriate dependencies to create fully self-contained component closures.
 
-But why is installing software with [Nix]{.pkg} reproducible? Because the whole
-set of [Nix]{.pkg} expressions is hosted on GitHhub, it is possible to use a
-specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
-specific commit of `nixpkgs` is called *pinning a revision (or commit)*. Pinning
-a revision ensures that every package that [Nix]{.pkg} installs will always be at
-exactly the same versions, regardless of when in the future the packages get
-built. This is because the [Nix]{.pkgs} expressions that get downloaded when
-pinning are the expressions as they were at that commit of `nixpkgs`.
+Why is installing software with [Nix]{.pkg} reproducible? Because the entire set
+of [Nix]{.pkg} expressions is hosted on GitHub, it is possible to reference a
+specific commit of `nixpkgs`, a process called *pinning a revision*. Pinning a
+revision ensures that all packages installed by [Nix]{.pkg} are always the exact
+same versions, regardless of when in the future they are built, since the
+expressions downloaded correspond to that specific commit of `nixpkgs`.
 
-While pinning revisions is essential for reproducibility, this is not the only
-reason [Nix]{.pkg} is very useful for reproducibility, but the [Nix]{.pkg}
-package manager itself was designed with reproducibility in mind. [Nix]{.pkg} is
-a functional package manager, meaning that it uses concepts from functional
-programming to build software: no-side effects or global variables are allowed,
-and the output is always the same if given the same inputs, regardless of where
-or when a package gets built (an important side-note is in order here: while the
-functional aspect of [Nix]{.pkg} is quite useful to ensure reproducibility, it
-can lead to situations where writing the required [Nix]{.pkg} expression to
-build it can be much more difficult than usual: this is especially true for
-packages that must download assets from the internet during installation, such
-as several Bioconductor packages like [musData]{.pkg} for example. But this is
-only a concern for the packagers, not users of the package that simply wish to
-install and use the package). [Nix]{.pkg} also allows for multiple versions (or
-variants in the [Nix]{.pkg} jargon) to be installed at the same time on the same
-system. Each variant of a package has a unique identifier, so that it is
-possible to have two versions of [R]{.proglang} installed and use the correct
-one where adequate. We refer to @dolstra2004nix for a more technical discussion
-on [Nix]{.pkg}.
+Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports
+reproducibility. [Nix]{.pkg} is a functional package manager: it applies
+principles from functional programming, disallowing side effects and global
+variables, and ensuring that the output of a build is always the same given the
+same inputs, regardless of time or location. A side note: while this functional
+approach greatly enhances reproducibility, it can make writing derivations more
+complex, particularly for packages that need to download assets during
+installation (e.g., some Bioconductor packages like [musData]{.pkg}). However,
+this concern is primarily for package maintainers, not end-users.
 
-With Nix, it is essentially possible to replace [renv]{.pkg} and
-[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
-`requirements.txt` files which list package dependencies for [Python]{.proglang}
-projects and virtual environments. It is also possible to build multi-language
-environments, containing [R]{.proglang} and Python, a LaTeX distribution, and even install
-one's favorite text editor to edit the project's source code. Essentially, one
-can define a complete, project-specific, and reproducible environment and use it
-interactively, or non-interactively. It will always be possible to rebuild this
-environment in the future (as long as the `nixpkgs` GitHub repository is online)
-to run an analysis sometime in the future.
+Additionally, [Nix]{.pkg} supports multiple versions, or *variants*, of a
+package on the same system. Each variant has a unique identifier, allowing
+multiple versions of [R]{.proglang} to coexist and ensuring the correct version
+is used for each project. For a more technical discussion of [Nix]{.pkg}, we
+refer to @dolstra2004nix.
 
-[Nix]{.pkg} has a fairly high entry cost though. [Nix]{.pkg} is a complex piece
-of software that comes with its own programming language, which is also called
-[Nix]{.proglang}. Its purpose is to solve a complex problem: defining
-instructions on how to build software packages and manage configurations in a
-declarative way. This makes sure that software gets installed in fully
-reproducible manner, on any operating system or hardware. To increase
-[Nix]{.pkg}’s adoption, we wrote the [rix]{.pkg} package.
+With [Nix]{.pkg}, it is possible to effectively replace both [renv]{.pkg} and
+[Docker]{.pkg} for [R]{.proglang} projects—or, in the case of
+[Python]{.proglang}, to replace `requirements.txt` files and virtual
+environments. [Nix]{.pkg} also supports building multi-language environments,
+which can include [R]{.proglang} and [Python]{.proglang}, a \LaTeX
+distribution or any other of the 120,000 available tools. This enables the
+creation of a complete, project-specific, and reproducible environment that can
+be used interactively or non-interactively. As long as the `nixpkgs` GitHub
+repository remains online, the environment can be rebuilt in the future to rerun
+analyses reliably.
 
-## Reproducible development environments with Nix{#sec-repro-nix}
+However, [Nix]{.pkg} has a steep learning curve. It is a complex system that
+comes with its own programming language, also called [Nix]{.proglang}, designed
+to solve the challenging problem of declaratively defining how software is built
+and configured. This ensures that installations are fully reproducible across
+operating systems and hardware platforms. To make [Nix]{.pkg} more accessible to
+[R]{.proglang} users that wish to use project-specific and reproducible
+development environments, we developed the [rix]{.pkg} package.
 
-As already stated,  [Nix]{.pkg} expressions are written in the  [Nix]{.proglang}
-programming language, which is a purely functional programming language. Here is
-a simple expression that builds a shell environment containing version 4.3.1 of
-[R]{.proglang}:
+## Reproducible development environments with Nix {#sec-repro-nix}
 
-```
+As mentioned, [Nix]{.pkg} expressions are written in the [Nix]{.proglang}
+programming language, which is purely functional. Here is a simple example that
+creates a shell environment containing version 4.3.1 of [R]{.proglang}:
+
+```{=tex}
+\begin{CodeInput}
 let
   pkgs = import (fetchTarball
-    "https://github.com/NixOS/nixpkgs/archive/976fa336....tar.gz"
+    "https://github.com/NixOS/nixpkgs/archive/976fa336.tar.gz"
   ) {};
@@
   }
+\end{CodeInput}
 ```
 
-To define a variable in the [Nix]{.proglang} language, the `let` keyword is
-used, so the first lines of this expression define the variables called `pkgs`
-and `system_pkgs`. `pkgs` is defined as the set of packages to be used in this
-derivation. In this example, the package expressions are going to be pulled from
-the official `nixpkgs` GitHub repository, and in particular the commit starting
-with `976fa336` is going to be used. `system_pkgs` lists the packages that we
-wish to install and use with this environment; in this example only the
-[R]{.proglang} programming language will be installed, alongside all of its
-dependencies, and all of the dependencies’ dependencies. Finally, the `mkShell`
-function is used to build a development shell containing the set of packages
-defined by the variable `system_packages` and its `shellHook` is set to `R
---vanilla`: this means that when entering the shell, [R]{.proglang} will be
-started with the `--vanilla` flag (thus ignoring any startup options).
+In this expression, the `let` keyword is used to define variables. The variable
+`pkgs` imports the set of packages from the `nixpkgs` repository at the
+specified commit `976fa336`. The variable `system_packages` lists the packages
+to include in the environment; in this case, it is just the [R]{.proglang}
+programming language, along with all its dependencies and their transitive
+dependencies. The `mkShell` function then creates a development shell with the
+specified packages. The `shellHook` is set to `"R --vanilla"`, meaning that
+entering the shell automatically starts [R]{.proglang} in vanilla mode, ignoring
+any startup options.
 
-This expression can be saved in a file called `default.nix`, and the shell can
-be built using the `nix-build` command on a system where  [Nix]{.pkg} is
-available.^[To install  [Nix]{.pkg} we recommend the installer from
-Determinate systems which can be found on the following link:
-\url{https://determinate.systems/posts/determinate-nix-installer}.] Once
-`nix-build` is done building the development environment according to the
-specifications declared in the `default.nix` file, the user can enter the
-development shell using `nix-shell`. The user will thus be dropped into an
-interactive shell with all the packages declared in the `default.nix` file and
-can now use this shell for work. This is similar to activating virtual
-environments in the [Python]{.proglang} ecosystem.
+This expression can be saved in a file called `default.nix`. The environment can
+then be built on a system with [Nix]{.pkg} installed using the `nix-build`
+command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
+installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
+Once the build completes, the user can enter the interactive shell with
+`nix-shell`. This shell contains all the packages specified in `default.nix` and
+can be used for development, similar to activating a virtual environment in the
+[Python]{.proglang} ecosystem.
 
-Obviously, writing such an expression will be quite complicated for someone that
-is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
-of being able to define a completely reproducible development environment in a
-single text file and then use that file to build and use that environment
-anywhere is quite appealing. Because we believe that lowering the cost of
-writing [Nix]{.proglang} expressions would lead to a wider adoption of the
-[Nix]{.pkg} package manager for reproducibility, we developed [rix]{.pkg}.
-[rix]{.pkg} comes with the `rix()` function that helps generate such expressions
-quite easily. [rix]{.pkg} is available on CRAN and can be installed as any other
-package, but it is also possible to use it to bootstrap an [R]{.proglang}
-development environment on a system where [R]{.proglang} is not available (but
-[Nix]{.pkg} is). One simply needs to run this line inside a terminal:
+Writing [Nix]{.pkg} expressions can be challenging for users unfamiliar with the
+[Nix]{.proglang} language. However, the ability to define a fully reproducible
+development environment in a single text file and then rebuild it anywhere is
+highly appealing. To lower the barrier to adoption of [Nix]{.pkg} for
+reproducibility, we developed the [rix]{.pkg} package.
 
-```
-nix-shell -p R rPackages.rix
+[rix]{.pkg} provides the `rix()` function, which simplifies generating
+[Nix]{.pkg} expressions. It is available on CRAN and can be installed like any
+other R package. Additionally, it can bootstrap an [R]{.proglang} development
+environment on a system where [R]{.proglang} is not yet installed but
+[Nix]{.pkg} is available. This can be done by running (inside of a terminal):
+
+```{=tex}
+\begin{CodeInput}
+$> nix-shell -I \
+  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/refs/heads/2025-10-17.tar.gz -p \
+  R rPackages.rix
+\end{CodeInput}
 ```
 
-to be dropped in a temporary [R]{.proglang} session where [rix]{.pkg} is
-available.^[`nix-shell -p` starts an interactive shell where the package
-specified with the `-p` flag are present.] From there, it is possible to
-generate new [Nix]{.pkg} expressions to build new environments. For example,
-here is how one could use [rix]{.pkg} to generate a `default.nix` file that
-installs version 4.3.1 of [R]{.proglang} and the [dplyr]{.pkg} and
-[chronicler]{.pkg} packages (we don't show the generated expression here because
-of space):
+(the `-I` flag allows one to pass a specific revision of `nixpkgs`, ensuring
+temporary shells are also reproducible).
 
-```{r, eval = F}
-library(rix)
+This command opens a temporary [R]{.proglang} session with [rix]{.pkg}
+available.^[`nix-shell -p` starts an interactive shell with the specified
+packages.] From there, users can generate new [Nix]{.pkg} expressions for
+building environments. For example, the following generates a `default.nix` file
+that installs [R]{.proglang} 4.3.1 along with the [dplyr]{.pkg} and
+[chronicler]{.pkg} packages:
 
-rix(r_ver = "4.3.1",
-    r_pkgs = c("dplyr", "chronicler"),
-    project_path = ".",
-    overwrite = TRUE)
+```{=tex}
+\begin{CodeInput}
+R> library('rix')
+
+R> rix(r_ver = "4.3.1",
++    r_pkgs = c("dplyr", "chronicler"),
++    project_path = ".",
++    overwrite = TRUE)
+\end{CodeInput}
 ```
 
-[rix]{.pkg} can also generate much more complex expressions:
+[rix]{.pkg} can also handle more complex setups:
 
-```{r, eval = F}
-rix(r_ver = "4.3.1",
-    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
-    system_pkgs = c("quarto", "git"),
-    tex_pkgs = c(
-          "amsmath",
-          "framed",
-          "fvextra",
-          "environ",
-          "fontawesome5",
-          "orcidlink",
-          "pdfcol",
-          "tcolorbox",
-          "tikzfill"
-    ),
-    git_pkgs = list(
-                 list(package_name = "rix",
-                   repo_url = "https://github.com/b-rodrigues/rix/",
-                   branch_name = "master",
-                   commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
-                 list(package_name = "fusen",
-                   repo_url = "https://github.com/ThinkR-open/fusen",
-                   branch_name = "main",
-                   commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
-    ),
-    ide = "positron",
-    project_path = ".",
-    overwrite = TRUE)
+```{=tex}
+\begin{CodeInput}
+R> rix(r_ver = "4.3.1",
++    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
++    system_pkgs = c("quarto", "git"),
++    tex_pkgs = c(
++          "amsmath",
++          "framed",
++          "fvextra",
++          "environ",
++          "fontawesome5",
++          "orcidlink",
++          "pdfcol",
++          "tcolorbox",
++          "tikzfill"
++    ),
++    git_pkgs = list(
++      list(
++        package_name = "rix",
++        repo_url = "https://github.com/b-rodrigues/rix/",
++        branch_name = "master",
++        commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
++      list(
++        package_name = "fusen",
++        repo_url = "https://github.com/ThinkR-open/fusen",
++        branch_name = "main",
++        commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
++    ),
++    ide = "positron",
++    project_path = ".",
++    overwrite = TRUE)
+\end{CodeInput}
 ```
 
-The above call to `rix()` generates an expression that installs several
-[R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
-older version of the package and also installs several TeXLive packages for
-authoring documents with \LaTeX, the development versions of [rix]{.pkg} and
-[fusen]{.pkg} (downloaded from Github), and finally the Positron integrated
-development environment. You can see the resulting `default.nix` file in the
-@sec-appendix.
+This call to `rix()` generates an environment that installs several
+[R]{.proglang} packages (including [AER]{.pkg} version 1.2-8), several TeXLive
+packages for \LaTeX document authoring, development versions of [rix]{.pkg} and
+[fusen]{.pkg} from GitHub, and the Positron editor.
 
-[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
-[rix]{.pkg} can generate [Nix]{.pkg} expression even if [Nix]{.pkg} is not installed on one's
-system. This is quite useful if one wants to generate an expression and use that
-expression in a CI/CD environment such as Github Actions. For example, the
-repository that contains the source code to this
-article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
-compilation. Each time one of the authors pushed to the master branch, a Github
-Actions runner would be started,  [Nix]{.pkg} would get installed on said
-runner, and an environment containing the right packages was generated from a
-`default.nix` file that is hosted on the same repository. The paper would then
-get compiled using [Quarto]{.pkg} in this reproducible environment.
+[rix]{.pkg} can generate [Nix]{.pkg} expressions even if [Nix]{.pkg} is not
+installed on the system. This is useful for continuous integration and
+continuous deployment (CI/CD) workflows on platforms such as GitHub Actions. For
+instance, the repository containing the source code for this
+article^[https://github.com/b-rodrigues/rix_paper] uses GitHub Actions to
+compile the paper. Each time a push is made to the master branch, a runner
+installs [Nix]{.pkg}, generates the environment from the hosted `default.nix`
+file, and compiles the paper using [Quarto]{.pkg} within the reproducible
+environment. This ensure that *exactly* the same environment is used on the
+author's computer and on the CI/CD without any additional, platform-specific,
+configuration.
 
-Instead of first dropping into a [Nix]{.pkg} shell it is also possible to
-directly run a program from the environment:
+Instead of first entering a [Nix]{.pkg} shell, it is also possible to run a
+program directly from the environment:
 
-```
-cd /absolute/path/to/project/ && nix-shell default.nix --run "Rscript analysis.R"
+```{=tex}
+\begin{CodeInput}
+cd /path/to/project/ && nix-shell default.nix --run "Rscript analysis.R"
+\end{CodeInput}
 ```
 
-This will run `Rscript` and execute the `analysis.R` script (which in this
-example should be in the same directory as `default.nix`).
+This command runs `Rscript` and executes the `analysis.R` script, which in this
+example should be located in the same directory as `default.nix`.
 
-## rix uses a fork of Nixpkgs
+## The rstats-on-nix fork of nixpkgs {#sec-fork}
 
-As previously explained, when invoked, the [Nix]{.pkg} package manager will use
-expressions hosted on the `nixpkgs` GitHub repository to build software.
-However, when generating expressions with [rix]{.pkg} our fork
-`rstats-on-nix/nixpkgs` is used instead.
+As explained earlier, [Nix]{.pkg} uses expressions from the `nixpkgs` GitHub
+repository to build software. However, when generating expressions with
+[rix]{.pkg}, our fork `rstats-on-nix/nixpkgs` is used instead.
 
-Using a fork instead of the official upstream `nixpkgs` repository has several
-advantages.
+Using a fork offers several advantages. First, it provides flexibility that the
+official `nixpkgs` repository cannot always accommodate. [Nix]{.pkg} is
+primarily the package manager for the NixOS Linux distribution, and governance
+and technical choices made upstream can limit what [rix]{.pkg} aims to provide.
 
-The first thing we need to stress, is that it is important to keep in mind that
-the [Nix]{.pkg} package manager is first and foremost the package manager of a
-Linux distribution called NixOS, and as such, several technical and governance
-choice have to be made that limit what we, the authors of [rix]{.pkg} want to offer
-to the users of `{rix}`. For our needs, having a fork gives us more flexibility.
+For instance, while [Nix]{.pkg} can theoretically support multiple versions (or
+*variants*) of the same package, in practice maintainers cannot provide several
+variants for all [R]{.proglang} packages, given the size of the ecosystem (over
+20,000 CRAN and Bioconductor packages). This makes it difficult to install a
+specific version of an [R]{.proglang} package not included in a particular
+`nixpkgs` commit. With [rix]{.pkg}, users can install a specific package version
+from source, e.g.:
 
-For example, while it is technically possible for [Nix]{.pkg} to provide many versions
-(or variants, in [Nix]{.pkg} jargon) of the same package (for example, you can
-install the latest [Emacs]{.pkg} by installing the `emacs` package, or
-[Emacs]{.pkg} 28 by installing `emacs28`) this ultimately depends on whether the
-maintainer wishes to package several variants, or whether it is practical. In
-the case of the [R]{.proglang} ecosystem, with more than 20'000 CRAN and
-Bioconductor packages, that is not possible for the maintainers to provide
-several variants of the same packages. So for a given `nixpkgs` commit, it is
-not possible to easily install a specific version of an [R]{.proglang} package
-that is not included in that particular `nixpkgs` commit. Instead, [rix]{.pkg}
-can install it from source with this syntax:
-
-```{r, eval = FALSE}
-rix(..., r_pkgs = "dplyr@1.0.7", ...)
+```{=tex}
+\begin{CodeInput}
+R> rix(..., r_pkgs = "dplyr@1.0.7", ...)
+\end{CodeInput}
 ```
 
-However, some packages might need specific fixes to be successfully installed
-by [Nix]{.pkg} (as explained in the case of [xlsx]{.pkg} previously) and thus
-trying to install packages like this might fail.
+However, installing from source might fail, especially if the package needs
+to be compiled.
 
-Furthermore, it isn't practical to update the whole of the [R]{.proglang}
-packages set on [Nix]{.pkg} every day: so while CRAN and Bioconductor get
-updates daily, the [R]{.proglang} packages set on [Nix]{.pkg} gets updated only
-around new releases of R. Again, this is a consequence of [Nix]{.pkg} being
-first and foremost the package manager of a Linux distribution with its own
-governance; daily updates of more than 20'000 package is simply not feasible.
+Additionally, updating the full [R]{.proglang} package set on [Nix]{.pkg} daily
+is impractical. While CRAN and Bioconductor update daily, the [R]{.proglang}
+packages in `nixpkgs` are only updated with new R releases. This limitation is
+due to [Nix]{.pkg}’s governance as a Linux distribution package manager.
 
-Because of these reasons, we decided to fork `nixpkgs`: because it is a fork, we
-can afford to do things in a way that could not be possible or practical for
-upstream.
+Our fork allows us to circumvent these limitations. For example, we provide a
+daily snapshot of CRAN. Each day, the [R]{.proglang} package set is updated and
+committed to a dated branch using GitHub Actions. Users can select a specific
+date with:
 
-The first thing this fork allows us to do is offer a daily snapshot of CRAN.
-Every day, the [R]{.proglang} packages set gets updated, and the result commited
-to a dated branch using scripts running on GitHub Actions (a continuous
-integration and deployment platform). This has been going on since the 14th of
-December 2024 (see
-[here](https://github.com/rstats-on-nix/nixpkgs/tree/2024-12-14)). So when users
-set a date as in `rix(date = "2024-12-14", ...)` the specified dated branch from
-the fork is going to get used.
-
-But this doesn’t mean that we recommend users use any date from the
-`rstats-on-nix/nixpkgs` fork: instead, each Monday, another action uses this
-fork and tries to build a set of popular packages on Linux and macOS, and only
-if this succeeds is the date added through a PR to the list of available dates
-on `{rix}`. This means that we also monitor the health of popular [R]{.proglang}
-packages and can propose a fix upstream if a package stops building
-successfully, which is something that can sometimes happen. For example,
-recently RStudio was in a broken state due to an issue in one of its
-dependencies, `boost`. This is not the fault of anyone in particular: it's just
-that packages get updated and packages that depend on them should get updated as
-well: but if that doesn't happen quickly enough, the `nixpkgs` maintainer faces
-a conundrum. Either he or she doesn't update the package because it breaks
-others, but not updating a package could be a security vulnerability, or he or
-she updates the package, but now other, perhaps less critical packages are
-broken and need to be fixed, either by their upstream developers, or by the
-`nixpkgs` maintainer of said packages. In the case of RStudio a fix was proposed
-and promptly merged, but if users wanted to install RStudio during the time it
-took to fix it, they would have faced an error message.
-
-However, thanks to our fork, we can back-port fixes and so if users try to
-install RStudio using the date `"2025-02-10"` (which is when it was broken
-on upstream `nixpkgs`), it'll work.
-
-We spent quite some time backporting fixes: we went back all the way back to
-March 2019. To do so, we start by checking out a `nixpkgs` commit on selected
-dates, then we "update" the [R]{.proglang} packages set by using the Posit CRAN
-and Bioconductor daily snapshots at that date. Then, we backport as many fixes
-as possible, and ensure that a selection of popular packages work on both
-x86-linux (which includes Windows, through WSL2) and aarch64-darwin (which is
-how Apple Silicon Apple computers are designated as). Then we commit everything
-to a dated branch of the `rstats-on-nix/nixpkgs` fork. Users can consult all the
-available dates by running: `rix::available_dates()`.
-
-An issue with forking `nixpkgs` though, is that by backporting fixes, we
-essentially create new [Nix]{.pkg} packages that are not included in upstream,
-and thus, these are not built by Hydra, Nix's CI platform which builds binary
-packages. In practice this means that anyone using our fork will have to compile
-many packages from source, which can be very time consuming. However, we also
-provide a binary cache of packages sponsored by
-[Cachix](https://www.cachix.org/) which complements the default, public
-[Nix]{.pkg} cache. We provide instructions on how to use Cachix in the package's
-Readme file. Using Cachix speeds up the installation process of packages
-tremendously, as they simply get downloaded instead of being built (we build
-a selection of packages only however, not the whole of CRAN and Bioconductor).
-
-## Additional features of the rix package
-
-If one wishes to run pipelines written using the [targets]{.pkg} package
-[@landau2021] on Github Actions, [rix]{.pkg} features a function called
-`tar_nix_ga()`, inspired by the `tar_github_actions()` function from the
-[targets]{.pkg} package. This function writes a YAML file under
-`.github/workflows/` which then runs the [targets]{.pkg} pipeline using Github
-Actions on each push, in a completely reproducible environment. For such an
-example, consult the code in this
-repository^[https://github.com/b-rodrigues/rix_cicd].
-
-The final feature that we provide is the ability to run one arbitrary function
-in what we call a *subshell*. Suppose that you are running the latest version
-of R, and that you wish or need to run one particular function from an older
-release of R, or from an older package. Using the `with_nix()` function, it is
-possible to run this single function inside the right environment, and recover
-its output in the main session. To illustrate, consider the following two lines
-of code:
-
-```{r, eval = FALSE}
-set.seed(1234)
-sample(seq(1, 10), 5)
+```{=tex}
+\begin{CodeInput}
+R> rix(date = "2024-12-14", ...)
+\end{CodeInput}
 ```
 
-```{r, include = FALSE}
-set.seed(1234)
-out_r <- sample(seq(1, 10), 5)
-set.seed(NULL)
+We strive to provide an available date per week: each Monday, a GitHub Action
+tests popular packages on Linux and macOS, and only if all tests succeed is the
+date added to the list of available dates in [rix]{.pkg}. This ensures users can
+reliably install packages, and allows us to backport fixes if needed. For
+example, when RStudio was temporarily broken due to a dependency issue
+(`boost`), a pull request was submitted to the official `nixpkgs` repository. We
+backported the fix to our fork, making RStudio available to users of our fork
+earlier than upstream, as merging PRs in the official repository can take some
+time.
+
+We have backported fixes to our `nixpkgs` fork as far back as March 2019. The
+process involves checking out a `nixpkgs` commit on the selected date, updating
+the [R]{.proglang} package set using Posit CRAN and Bioconductor snapshots,
+backporting fixes, and ensuring popular packages work on both x86-linux
+(including WSL2) and aarch64-darwin (Apple Silicon). These changes are committed
+to a dated branch in `rstats-on-nix/nixpkgs`. Users can see all available dates
+with `rix::available_dates()`.
+
+A drawback of forking `nixpkgs` is that backported packages are not included
+upstream and thus are not prebuilt by Nix’s CI platform, Hydra. Users may need
+to build many packages from source, which can be time-consuming. To mitigate
+this, we provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
+complementing the public Nix cache. Instructions for using Cachix are in
+[rix]{.pkg}’s documentation. Using the cache significantly speeds up
+installations, as prebuilt packages are downloaded rather than compiled.
+
+## Orchestrating the workflow with rixpress {#sec-rixpress}
+
+Defining a reproducible environment with [rix]{.pkg} addresses the first major
+challenge of reproducibility. The second challenge is reliably and efficiently
+executing the analysis workflow within that environment, which is the role of
+the [rixpress]{.pkg} package.
+
+As mentioned in the introduction, a build automation tool like [targets]{.pkg}
+is invaluable for managing complex analyses. It tracks dependencies between code
+and data, caches results, and only recomputes steps that have changed. One can
+run a [targets]{.pkg} pipeline inside a Nix environment to make it reproducible.
+However, this approach has limitations: the entire pipeline must run in a single
+environment, and orchestrating steps across different languages (e.g.,
+[R]{.proglang} and [Python]{.proglang}) requires manual handling via packages
+like [reticulate]{.pkg}.
+
+[rixpress]{.pkg} overcomes these limitations by using Nix not just as a package
+manager, but as the build automation engine itself. In a [rixpress]{.pkg}
+pipeline, each step is defined as a Nix derivation, providing two key benefits:
+
+  1. True Polyglot Pipelines: Each step can have its own Nix environment. A
+  Python step can run in a pure Python environment, an [R]{.proglang} step in an
+  [R]{.proglang} environment, and a Quarto rendering step in yet another, all
+  within the same pipeline.
+  2. Deep Reproducibility: Each step is a hermetically sealed Nix derivation
+  whose output is cached in the Nix store based on the hash of all its inputs.
+  All artifacts are direct children of the computational environment (because
+  the computational environment is actually a dependency of the artifacts),
+  ensuring they are rebuilt if the environment changes, keeping environment and
+  outputs always in sync.
+
+The user defines the pipeline in an [R]{.proglang} script using functions
+inspired by [targets]{.pkg}. For example, a simple pipeline might read data with
+[Python]{.proglang} (`polars`), filter it, process it with [R]{.proglang}
+(`dplyr`), and compile a Quarto document. The pipeline is defined as a list,
+typically in a file called `gen-pipeline.R` and its execution environment in a
+file called `gen-env.R`. Templates of these files can be created using
+`rxp_init()`. On a machine that has [Nix]{.pkg} installed it is possible to
+bootstrap an environment using a temporary shell:
+
+```{=tex}
+\begin{CodeInput}
+$> nix-shell -I \
+  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/refs/heads/2025-10-17.tar.gz -p \
+  R rPackages.rix rPackages.rixpress
+\end{CodeInput}
+```
+One can start the R interpreter, and then call `rxp_init()`:
+
+```{=tex}
+\begin{CodeInput}
+R> rixpress::rxp_init()
+\end{CodeInput}
 ```
 
-On versions of [R]{.proglang} released after version 3.6, these two lines return
-the following vector: `r out_r`, but in versions prior to 3.6, the result will
-be different. This is a documented change as shown in the [R]{.proglang} version
-3.6 release notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]
+Let us modify `gen-pipeline.R`:
 
-When trying to reproduce a study made on versions prior to 3.6, a replicator
-will need to take this into account. Using the `with_nix()` function, the
-original results can be replicated even from the latest version of R.
+```{=tex}
+\begin{CodeInput}
+R> library('rixpress')
 
-The first step is to create an environment with the old release of R:
+R> pipeline_steps <- list(
++  # Step 1: Read data using Python
++  rxp_py_file(
++    name = mtcars_pl,
++    path = "data/mtcars.csv",
++    read_function = "lambda x: polars.read_csv(x, separator = '|')"
++  ),
++
++  # Step 2: Filter the data in Python
++  rxp_py(
++    name = mtcars_filtered,
++    expr = "mtcars_pl.filter(polars.col('am') == 1).to_pandas()"
++  ),
++
++  # Step 3: Convert Python pandas DataFrame to R data.frame
++  # This uses reticulate for conversion. It would also be possible
++  # to pass objects to and from languages by serializing them
++  # to json for instance, or another universal format
++  rxp_py2r(
++    name = mtcars_r,
++    expr = mtcars_filtered
++  ),
++
++  # Step 4: Process data in R
++  rxp_r(
++    name = mtcars_summary,
++    expr = dplyr::summarise(mtcars_r, avg_mpg = mean(mpg))
++  )
++)
 
-```{r, include = FALSE}
-library(rix)
-
-path_subshell <- tempdir()
-
-rix(
-  r_ver = "3.5.3",
-  overwrite = TRUE,
-  project_path = path_subshell
-)
+# Generate the 'pipeline.nix' file from the R list
+R> rxp_populate(pipeline_steps)
+\end{CodeInput}
 ```
 
-```{r, eval = F}
-library(rix)
+The required execution environment, defined in `gen-env.R`, is as follows:
 
-path_subshell <- tempdir()
+```{=tex}
+\begin{CodeInput}
+R> library(rix)
 
-rix(
-  r_ver = "3.5.3",
-  overwrite = TRUE,
-  project_path = path_subshell
-)
+R> rix(
++   date = "2025-10-14",
++   r_pkgs = c("dplyr", "ggdag", "quarto", "reticulate", "rixpress"),
++   py_conf = list(
++     py_version = "3.13",
++     py_pkgs = c("polars", "pyarrow", "pandas")
++   ),
++   ide = "none",
++   project_path = ".",
++   overwrite = TRUE
++ )
+\end{CodeInput}
 ```
 
-We recommend `path_subshell` to be a subfolder of the main project. Using
-`rix()` we generate a `default.nix` file that will provide a shell with R
-version 3.5.3 when built. When calling `rix()`, another function is also called,
-`rix_init()`, which generates an `.Rprofile` file in the same subfolder as where
-the subshell's `default.nix` is located. This `.Rprofile` file ensures that the
-R session that will be spawned from this subshell does not get any interference
-from the main session it gets called from.
+Using the same temporary shell, users can now generate the `default.nix` file
+by calling `source("gen-env.R")`. Users can now quit the temporary shell, and
+build the environment by calling `nix-shell` in a terminal. Once the
+development (or execution) environment has been built, users can enter it,
+and call `source("gen-pipeline.R")` to generate a file called `pipeline.nix`
+which is the [Nix]{.pkg} expression that defines the entire reproducible
+pipeline.
 
-We can now use this shell to generate our vector of random numbers using
-`with_nix()`:
+The environment also includes the [ggdag]{.pkg} package [@ggdag], so it's
+possible to view a visual representation of the pipeline (yellow nodes are
+[Python]{.pkg} and blue nodes are [R]{.pkg} derivations respectively):
 
-```{r, eval = F}
-out_subshell <- with_nix(
-    expr = function(){
-      set.seed(1234)
-      a <- sample(seq(1, 10), 5)
-      set.seed(NULL)
-      return(a)
-    },
-    program = "R",
-    project_path = path_subshell,
-    message_type = "simple"
-  )
+![Graphical representation of the pipeline](dag.png)
 
+Running the script above does not execute the pipeline; it generates a
+`pipeline.nix` file, which declaratively defines the workflow.
+
+To build and execute the pipeline:
+
+```{=tex}
+\begin{CodeInput}
+R> rxp_make()
+\end{CodeInput}
 ```
 
-```{r, include = FALSE}
-out_subshell <- with_nix(
-    expr = function(){
-      set.seed(1234)
-      a <- sample(seq(1, 10), 5)
-      set.seed(NULL)
-      return(a)
-    },
-    program = "R",
-    project_path = path_subshell,
-    message_type = "simple"
-  )
+```{=tex}
+\begin{CodeInput}
+Build process started...
 
++ mtcars_pl building
++ mtcars_filtered building
++ mtcars_r building
++ mtcars_summary building
+✓ mtcars_filtered built
+✓ mtcars_pl built
+✓ mtcars_r built
+✓ mtcars_summary built
+✓ pipeline completed [4 completed, 0 errored]
+Build successful! Run `rxp_inspect()` for a summary.
+Use `rxp_read("derivation_name")` to read objects or
+`rxp_load("derivation_name")` to load them into the global environment.
+\end{CodeInput}
 ```
 
-We should say that `with_nix()` prints many status messages to the console while
-running, which we don't show here for space reasons.
+[Nix]{.pkg} executes each step in order, building dependencies as needed.
+Outputs are cached, so subsequent runs only recompute steps with changed inputs
+or code. The build process can also be executed in parallel, if the the pipeline
+has independent branches. This provides the efficiency of [targets]{.pkg} with
+multi-language support and bit-for-bit reproducibility. Artifacts can be
+inspected interactively in [R]{.proglang} using `rxp_read("artifact_name")` or
+`rxp_load("artifact_name")`.
 
-The `expr` argument of `with_nix()` can be arbitrarily complex, but must be
-wrapped inside a `function()` call. The value of `out_subshell` is `r
-out_subshell`, and is indeed different than the one generated from a recent
-version of R.
+For Python users, a port called [ryxpress]{.pkg} allows building the same
+pipelines and inspecting outputs from Python sessions.
 
-It is possible to use `with_nix()` from an [R]{.proglang} session that was
-installed itself via Nix, or from a regular [R]{.proglang} installation.
+[rixpress]{.pkg} includes several additional features not covered here for the
+sake of brevity.
 
-## Conclusion
+It is also possible to configure popular IDEs to work interactively and
+seamlessly with both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth,
+reproducible workflow from within the development environment. Detailed setup
+instructions are provided in the vignettes of both packages.
 
-There are many tools to make one’s analyses reproducible, but [Nix]{.pkg}
-distinguishes itself from these in that its main concern is in deploying the
-complete set of packages that are *closed under the ''depends on'' relation*:
-[Nix]{.pkg} does not simply install one piece of software, but its dependencies,
-and their dependencies. This makes [Nix]{.pkg} extremely useful for reproducible
-research, but because it solves a complex problem, it is a complex tool. With
-[rix]{.pkg}, we wish to make using [Nix]{.pkg} more accessible to users of
-[R]{.proglang} by providing a familiar interface and user experience. Because it
-is possible to build reproducible development shells using [Nix]{.pkg}, it can
-accomodate any use case: running scripts, pipelines, interactive [shiny]{.pkg}
-applications, and serve [plumber]{.pkg} application programming interfaces.
+## Conclusion {#sec-conclusion}
+
+Many tools exist to improve reproducibility, but [Nix]{.pkg} stands out because
+it deploys complete software environments *closed under the "depends on"
+relation*: it installs not only a package, but all its dependencies and their
+dependencies. This makes [Nix]{.pkg} particularly powerful for reproducible
+research.
+
+However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
+[rix]{.pkg}, we aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
+by providing a familiar interface and workflow. By building reproducible
+development shells with [Nix]{.pkg}, researchers can accommodate a wide range of
+use cases: running scripts and pipelines, developing interactive [shiny]{.pkg}
+applications, or serving [plumber]{.pkg} APIs.
+
+Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
+pipelines. By leveraging Nix as a build automation engine, [rixpress]{.pkg}
+allows polyglot workflows where each step runs in its own hermetically sealed
+environment. This ensures deep reproducibility, efficient caching, and seamless
+orchestration of multi-language analyses, making it easier to manage complex
+projects with confidence that results can be reproduced exactly.
 
@@
 
-:::{.callout}
-
-All acknowledgments (note the AE spelling) should be collected in this
-unnumbered section before the references. It may contain the usual information
-about funding and feedback from colleagues/reviewers/etc. Furthermore,
-information such as relative contributions of the authors may be added here
-(if any).
-
-:::
+We thank the rOpenSci reviewers and contributors who provided valuable feedback
+on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, we are
+grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of `{rix}`,
+and to William Landau and Anthony Martinez for their reviews of `{rixpress}`. We
+also acknowledge the contributions of Richard J. Acton, Jordi Rosell, Elio
+Campitelli, László Kupcsik, and Michael Heming for `{rix}`. Their expertise and
+feedback greatly improved the quality and usability of these packages.
 
@@
 :::
-
-{{< pagebreak >}}
-
-## Appendix {#sec-appendix .unnumbered}
-
-Here is the [Nix]{.pkg} expression generated from the call to the `rix()`
-function of section #sec-repro-nix:
-
-```{r, echo = FALSE}
-library(rix)
-
-rix(r_ver = "4.3.1",
-    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
-    system_pkgs = c("quarto", "git"),
-    tex_pkgs = c(
-          "amsmath",
-          "framed",
-          "fvextra",
-          "environ",
-          "fontawesome5",
-          "orcidlink",
-          "pdfcol",
-          "tcolorbox",
-          "tikzfill"
-    ),
-    git_pkgs = list(
-                 list(package_name = "rix",
-                   repo_url = "https://github.com/b-rodrigues/rix/",
-                   branch_name = "master",
-                   commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
-                 list(package_name = "fusen",
-                   repo_url = "https://github.com/ThinkR-open/fusen",
-                   branch_name = "main",
-                   commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
-    ),
-    ide = "rstudio",
-    project_path = tempdir(),
-    overwrite = TRUE,
-    print = TRUE)
-```
-
-:::{.callout}
-
-Appendices can be included after the bibliography (with a page break). Each
-section within the appendix should have a proper section title (rather than
-just _Appendix_).
-
-For more technical style details, please check out JSS's style FAQ at
-[https://www.jstatsoft.org/pages/view/style#frequently-asked-questions]
-which includes the following topics:
-
-- Title vs. sentence case.
-- Graphics formatting.
-- Naming conventions.
-- Turning JSS manuscripts into [R]{.proglang} package vignettes.
-- Trouble shooting.
-- Many other potentially helpful details…
-
-:::
-
-## Using BibTeX {#sec-bibtex .unnumbered}
-
-:::{.callout}
-
-References need to be provided in a {{< bibtex >}} file (`.bib`). All
-references should be made with `@cite` syntax. This commands yield different
-formats of author-year citations and allow to include additional details (e.g.,pages, chapters, \dots) in brackets. In case you are not familiar with these
-commands see the JSS style FAQ for details.
-
-Cleaning up {{< bibtex >}} files is a somewhat tedious task -- especially
-when acquiring the entries automatically from mixed online sources. However,
-it is important that informations are complete and presented in a consistent
-style to avoid confusions. JSS requires the following format.
-
-- item JSS-specific markup (`\proglang`, `\pkg`, `\code`) should be used in the references.
-- item Titles should be in title case.
-- item Journal titles should not be abbreviated and in title case.
-- item DOIs should be included where available.
-- item Software should be properly cited as well. For [R]{.proglang} packages `citation("pkgname")` typically provides a good starting point.
-
-:::

### 2025-10-17 paper.tex
getting ready

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
   article]{jss}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
 \makeatother
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \newcommand{\fct}[1]{\code{#1()}}
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
-\definecolor{quarto-callout-color}{HTML}{909090}
-\definecolor{quarto-callout-note-color}{HTML}{0758E5}
-\definecolor{quarto-callout-important-color}{HTML}{CC1914}
-\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
-\definecolor{quarto-callout-tip-color}{HTML}{00A047}
-\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
-\definecolor{quarto-callout-color-frame}{HTML}{acacac}
-\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
-\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
-\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
-\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
-\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
-\makeatother
 \makeatletter
@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
-  pdftitle={Reproducible development environments with rix},
-  pdfauthor={Bruno Rodrigues; Philipp Baumann},
+  pdftitle={A Framework for Reproducible Data Science using Nix},
+  pdfauthor={Bruno Rodrigues},
   pdfkeywords={reproducibility, R, Nix},
@@
 \author{Bruno Rodrigues~\orcidlink{0000-0002-3211-3689}\\Ministry of
-Research and Higher education, Luxembourg \And Philipp Baumann\\Plus
-Affiliation}
-\Plainauthor{Bruno Rodrigues, Philipp Baumann} %% comma-separated
+Research and Higher education, Luxembourg}
+\Plainauthor{Bruno Rodrigues} %% comma-separated
 
-\title{Reproducible development environments with rix}
-\Plaintitle{Reproducible development environments with
-rix} %% without formatting
+\title{A Framework for Reproducible Data Science using Nix}
+\Plaintitle{A Framework for Reproducible Data Science using
+Nix} %% without formatting
 
 %% an abstract and keywords
-\Abstract{In order to create an analysis that is easily reproducible, it
-is not enough to write clean code and document it well. One must also
-make sure to list all the dependencies of the analysis clearly and
-ideally provide an easy way to install said dependencies. There are
-several tools that can be used to list dependencies and to make them
-easily installable by someone that wishes to reproduce a study, such as
-\proglang{Docker}, a containerization solution. This paper will present
-the Nix package manager, and an \proglang{R} package called \pkg{rix}
-that lowers Nix's learning curve for users of the \proglang{R}
-programming language.}
+\Abstract{To create a reproducible analysis, it is not enough to write
+clean, well-documented and tested code. One must also ensure that all
+software dependencies are clearly declared and, ideally, provide a
+simple way to install them. Several tools exist for this purpose, such
+as the containerization solution \pkg{Docker}. A complete solution,
+however, requires not only managing the computational environment but
+also orchestrating the execution of the analysis itself in a robust and
+automated way. This paper presents a comprehensive framework based on
+the \pkg{Nix} package manager to solve both problems. We first introduce
+the \proglang{R} package \pkg{rix}, which simplifies the creation of
+reproducible environments with \pkg{Nix}. We then introduce
+\pkg{rixpress}, a second \proglang{R} package that leverages these
+environments to define and execute complex, multi-language analytical
+pipelines. Together, these tools provide a powerful, unified approach to
+achieving high standards of computational reproducibility.}
 
 %% at least one keyword must be supplied
-\Keywords{reproducibility, \proglang{R}, \proglang{Nix}}
+\Keywords{reproducibility, \proglang{R}, \pkg{Nix}}
 \Plainkeywords{reproducibility, R, Nix}
@@
 \\~
-Philipp Baumann\\
-\\~
 
@@
 \maketitle
+
 \section{Introduction: Reproducibility is also about
@@
 
-\citet{peng2011} introduced the idea of reproducibility being on a
-continuum: on one of the ends of this continuum, we only have access to
-the paper describing the studies, which is not reproducible at all.
-Then, if in addition, to this paper we make the original source code of
-the analysis that was written to compute the results of the study
-available, reproducibility is improved, albeit only by a little. Adding
-the original data improves reproducibility yet again. Finally, if to all
-this we add what Roger Peng named the \emph{linked and executable code
-and data}, we reach the gold standard of full replication.
+\citet{peng2011} introduced the concept of reproducibility as a
+\emph{continuum}. At one end lies the least reproducible state, where
+only a paper describing the study is available. Reproducibility improves
+when authors share the original source code, improves further when they
+include the underlying data, and reaches its highest level when what
+Roger Peng called \emph{linked and executable code and data} are
+provided.
 
-What is this \emph{linked and executable code and data}? Another way to
-name this crucial piece of the reproducibility puzzle is
-\emph{computational environment}. The computational environment is all
-the software required to actually run the analysis. Here too, we can
-speak of a continuum. One could simply name and list the software used:
-for example, the \proglang{R} programming language. Sometimes, authors
-have the courtesy to also state the version of \proglang{R} used. Some
-authors go further, and also list the packages used, and ideally with
-their versions as well. Authors rarely state the operating system on
-which the analysis was done, even though it has been shown that running
-the same analysis with the same software but on different operating
-systems could lead to different results, as described in
-\citet{neupane2019}. Authors also only very rarely provide instructions
-to install the required tools and software in order to reproduce their
-studies.
+By \emph{linked and executable code and data}, Peng referred to compiled
+source code and runnable scripts. In this paper, we interpret this
+notion more broadly as the \emph{computational environment}: the
+complete set of software required to execute an analysis. Here too, a
+continuum exists. At the minimal end, authors might only name the main
+software used---say, the \proglang{R} programming language. More careful
+authors might also specify the version of \proglang{R}, or list the
+additional packages and their versions. Rarely, however, do authors
+specify the operating system on which the analysis was performed, even
+though differences in operating systems can lead to divergent results
+when using the same code and software versions, as shown by
+\citet{neupane2019}. It is even less common for authors to provide
+step-by-step installation instructions for the required software stack.
 
-Tools can be used at each of these steps to reach the gold standard of
-full replication. Let's first consider the task of listing the software
-used. \proglang{R} provides the \texttt{sessionInfo()} function whose
-output can be saved into a file. Below is an example output of
-\texttt{sessionInfo()}:
+Even when such instructions are given, they often fail across different
+platforms or versions of the same platform. This lack of portability not
+only hinders reproducibility but also complicates everyday research
+workflows. Researchers working across multiple machines must be able to
+recreate the same environment consistently, and collaborators must share
+identical computational setups to avoid inconsistencies.
 
-\begin{verbatim}
-sessionInfo()
-\end{verbatim}
+Finally, once the execution environment is correctly configured,
+additional clarity is needed on how to \emph{run} the project itself.
+Which packages should be loaded first? Which scripts should be executed,
+and in what order? Without clear documentation or a logically organized
+project structure, these operational details become yet another barrier
+to reproducibility.
 
-\begin{verbatim}
+A range of tools now exist to help researchers approach the gold
+standard of full reproducibility, or to consistently deploy the same
+development environment across multiple machines. Let us first consider
+the most basic step in this process: listing the software used. In
+\proglang{R}, the \texttt{sessionInfo()} function provides a concise
+summary of the software environment, including the R version, platform
+details, and all loaded packages. Its output can be saved to a file and
+included as part of a study's reproducibility record. Below is an
+example output from \texttt{sessionInfo()}:
+
+\begin{CodeInput}
+R> sessionInfo()
+\end{CodeInput}
+\begin{CodeOutput}
 R version 4.3.2 (2023-10-31)
@@
 attached base packages:
-[1] stats     graphics  grDevices utils     datasets  methods  
-[7] base
+ stats     graphics  grDevices utils     datasets  methods   base
 
 other attached packages:
-[1] nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163
+ nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163
 
 loaded via a namespace (and not attached):
-[1] compiler_4.3.2 Matrix_1.6-1.1 tools_4.3.2    splines_4.3.2 
-[5] grid_4.3.2     lattice_0.21-9
-\end{verbatim}
+ compiler_4.3.2  Matrix_1.6-1.1  tools_4.3.2     splines_4.3.2
+ grid_4.3.2      lattice_0.21-9 
+\end{CodeOutput}
 
-If an author provides this information, other people trying to reproduce
-the study (or the author him- or herself in the future) can read this
-file and see which version of \proglang{R} was used, and which packages
-(and their versions) were used as well. However, others would still need
-to install the correct software themselves. An alternative to this is
-instead to use the \pkg{renv} package which generates a so-called
-\texttt{renv.lock} file which also lists \proglang{R} and package
-versions. Here is an example of such an \texttt{renv.lock} file:
+When an author includes this information, others attempting to reproduce
+the study (future readers, collaborators, or even the author at a later
+time) can easily see which version of \proglang{R} and which packages
+(with their versions) were used. However, reproducing the environment
+still requires manually installing the correct package versions. This
+can be difficult, especially when packages depend on system-level
+libraries. For example, the \pkg{nloptr} package requires a precompiled
+\proglang{nlopt} binary, or alternatively, \proglang{cmake} when
+building from source on Linux or macOS.
 
-\begin{verbatim}
+A more robust approach than simply listing package versions is to use
+the \pkg{renv} package. \pkg{renv} captures the project's software state
+and writes it to a \texttt{renv.lock} file, which includes the exact
+versions of \proglang{R} and all required packages. This lockfile serves
+as a blueprint for restoring the environment automatically, ensuring
+that others can recreate the same setup with minimal effort. Below is an
+example of an \texttt{renv.lock} file:
+
+\begin{Code}
 {
@@
       ]
+    }
 
-    ***and many more packages***
-\end{verbatim}
+    ... lines below omitted ...
+  }
+}
+\end{Code}
 
-This file lists every package alongside their versions and the
-repository from which they were downloaded. Generating this file only
-requires one to run the \texttt{renv::init()} function. Someone else can
-then restore the same package library by running
-\texttt{renv::restore()}. The exact same packages get installed in an
-isolated, project-specific, library which doesn't interfere with the
-other, main, library of the user. \pkg{renv} does not restore the
-\proglang{R} itself though, so installing the right version of R needs
-to be handled separately. Before continuing, it should be noted that
-other packages exist which provide similar functionality to \pkg{renv}:
-there is \pkg{groundhog} by \citet{simonsohn2023} which makes it rather
-easy to install packages as they were on CRAN at a given date. For
-example, the code snippet below install the \pkg{purrr} and
-\pkg{ggplot2} packages as they were on April 4th, 2017:
+This lockfile lists each package alongside its version and the
+repository from which it was downloaded. Creating it requires only a
+single command: \texttt{renv::init()}. Others can then reproduce the
+same package library by running \texttt{renv::restore()}, which installs
+the exact package versions in an isolated, project-specific library that
+does not interfere with the user's global R library.
 
-\begin{verbatim}
-groundhog.library("
-    library(purrr)
-    library(ggplot2)",
-    "2017-10-04",
-    tolerate.R.version = "4.2.2")
-\end{verbatim}
+However, \pkg{renv} does not restore the version of \proglang{R} itself.
+Installing the correct R version must therefore be done separately.
+Similarly, \texttt{renv} does not handle system-level dependencies such
+as \proglang{cmake}, which is required by \pkg{nloptr} on Linux and
+macOS. These dependencies must be managed manually or through other
+tools. Further details are discussed in the \pkg{renv} documentation:
+\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.
 
-These packages also get installed in a project-specfic library so there
-is no interferences between these packages and other versions of the
-same packages that one might use for other projects. Because
-\pkg{groundhog} does not install \proglang{R} itself, users should
-either install the required version themselves, or they should use the
-\texttt{tolerate.R.version} argument as shown in the example above.
-Otherwise, \pkg{groundhog} would not continue with the installation of
-the packages. Another such package, developed by \citet{chan2023} is
-\pkg{rang}, which also installs packages as they were on a given date.
-Yet another way to install packages as they were on a give date is to
-use the Posit Package Manager, which provides snapshots of CRAN. For
-example, to install the required packages for an analysis as they were
-on the 30th of June 2023, one could add the following line to the
-\texttt{.Rprofile} file:
+Before continuing, it should be noted that other packages exist which
+provide similar functionality to \pkg{renv}: there is \pkg{groundhog} by
+\citet{simonsohn2023} which makes it rather easy to install packages as
+they were on CRAN at a given date. For example, the code snippet below
+installs the \pkg{purrr} and \pkg{ggplot2} packages as they were on
+April 4th, 2017:
 
-\begin{verbatim}
-options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/
-                               __linux__/jammy/2023-06-30"))
-\end{verbatim}
+\begin{CodeInput}
+R> groundhog.library("
++   library('purrr')
++   library('ggplot2')",
++   "2017-10-04",
++   tolerate.R.version = "4.2.2")
+\end{CodeInput}
 
-The \texttt{.Rprofile} file gets read by R when starting a new session,
-which means that every call to the \texttt{install.packages()} function
-will now install the packages from this snapshotted mirror.
+Packages installed with \pkg{groundhog} are placed in a project-specific
+library, ensuring they do not interfere with other versions of the same
+packages used in different projects. Since \pkg{groundhog} does not
+install \proglang{R} itself, users must either install the required R
+version manually or use the \texttt{tolerate.R.version} argument, as
+illustrated in the example above. Without this argument, \pkg{groundhog}
+will not proceed with package installation if the R version does not
+match the expected one.
 
-The next step in reaching the gold standard of reproducibility would be
-to not only install the right packages used for the analysis, but also
-the right version of \proglang{R}. Of course, it would be possible to
-install the right version manually, but here too, there are tools that
-simplify the process such as \proglang{rig} by the R infrastructure team
-\citeyearpar{rlib2023}. One could thus install the right version of
-\proglang{R} using \proglang{rig} allows to easily install different
-versions of \proglang{R}, one could use to install the right version to
-reproduce an analysis, and then use one of the listed packages above to
-install the right library of of \proglang{R} packages. However, this
-involves many manual steps and is thus error prone. It is also
-time-consuming to do so, especially if one wants to reproduce many
-studies.
+An alternative to \pkg{renv} is \pkg{rang}, developed by
+\citet{chan2023}, which similarly allows users to install packages as
+they existed on a specific date. Another approach is to use the Posit
+Package Manager, which provides dated snapshots of CRAN. For instance,
+to ensure that packages are installed as they were on June 30, 2023, one
+could include the following line in the \texttt{.Rprofile} file:
 
-The final step towards the gold standard of reproducibility is to
-package the right version of \proglang{R} and \proglang{R} packages
-inside a \proglang{Docker} image. \proglang{Docker} is a
-containerisation tool: using \proglang{Docker} it is possible to package
-some \emph{data product} with its dependencies into an image. From this
-image, containers can be run to reproduce the data product with its
-exact dependencies. A statistical analysis, from the simplest to the
-most complex one, can be seen as such a data product that comes with
-many software dependencies. \emph{Dockerizing} an analysis consists in
-first building an image: in the build step, the dependencies of the
-analysis have to be installed and this can be achieved using the tools
-mentioned above. The scripts to run the analysis and the data are also
-added to the image at build time. Instances of that image, called
-containers, can then be executed, which give access to the exact
-development environment originally used for the study. It is also
-possible to script the building of the analysis itself, such that in one
-single call to the \texttt{docker\ run} command, a complete analysis is
-reproduced.
+\begin{CodeInput}
+R> options(repos =
++    c(REPO_NAME =
++      "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
++    )
++  )
+\end{CodeInput}
 
-The Rocker project initiated by \citet{boettiger2017} provides many
-pre-built \proglang{Docker} to the \proglang{R} community of users.
-These images can be used as bases to build other images containing
-statistical analyses more easily than starting from a bare-bones image.
+The \texttt{.Rprofile} file is read by \proglang{R} at the start of each
+new session, meaning that every call to \texttt{install.packages()} will
+install packages from the specified snapshot mirror. However, unless
+users explicitly manage separate libraries for different projects, the
+Posit Package Manager will install all packages for all projects as they
+existed on that snapshot date.
 
-An optional step towards the gold standard is to use a build automation
-tool such as \proglang{Make} to run the whole analysis when the
-container is executed. Build automation tools make it easier to run
-arbitrary code in a series of well-defined steps. \proglang{R}
-programmers can use the \pkg{targets} by \citet{landau2021} as a build
-automation tool.
+The next step toward the gold standard of reproducibility is ensuring
+not only that the correct package versions are installed, but also that
+the correct version of \proglang{R} itself is used. While this can be
+done manually, dedicated tools make the process more efficient. One such
+tool is \pkg{rig}, developed by the \proglang{R} Infrastructure Team
+\citeyearpar{rlib2023}. \pkg{rig} simplifies installing and managing
+multiple R versions, allowing users to match the exact version needed
+for a given analysis. Once the correct R version is installed, one can
+then use tools such as \pkg{renv}, \pkg{groundhog}, or \pkg{rang} to
+recreate the associated package library.
 
-\citet{mcdermott2021} is an example of a scientific study that reached
-the gold standard of reproducibility. The author of this paper set up an
-accompagnying Github repository to the paper\footnote{https://github.com/grantmcdermott/skeptic-priors}
-containing all the instructions to install the required software and
-then run the analysis. If we take a closer look at this repository, we
-will notice that many of the tools previously mentioned were used to
-capture the compatutational environment and make it available to other
-researchers:
+However, this multi-step process remains manual, error-prone, and
+time-consuming, highlighting the need for more comprehensive and
+automated solutions to achieve full computational reproducibility.
+
+The final tool for achieving the gold standard of reproducibility is to
+bundle the exact version of \proglang{R} and its packages within a
+\pkg{Docker} image. \pkg{Docker} is a containerization platform that
+allows packaging a \emph{data product} together with all its
+dependencies into a self-contained image. A statistical analysis can be
+viewed as such a data product, requiring a specific set of software
+dependencies to run. \emph{Dockerizing} an analysis involves building an
+image that installs these dependencies, often using the tools mentioned
+earlier, and includes the analysis scripts and data (but data can also
+be dynamically made available at run-time, to avoid having to copy it
+inside the image). The steps to create this image are specified in a
+simple text file called a \texttt{Dockerfile}, which defines the
+structure and content of the \pkg{Docker} image.
+
+Once the image is built, the analysis can be executed inside a
+\emph{container}, which is a running instance of the image. Containers
+are typically run non-interactively, allowing the complete computational
+environment to be instantiated and executed with a single command.
+
+The strength of \pkg{Docker} lies not only in its ability to encapsulate
+the correct versions of \proglang{R} and its packages but also in its
+inclusion of system-level dependencies. Since \pkg{Docker} images are
+nearly complete Linux systems, they naturally bundle libraries and tools
+required by certain R packages (for example, \pkg{nloptr}'s dependency
+on the \proglang{nlopt} library). This ensures that future replicators
+of a study have access to the same software environment, including all
+necessary system components. Moreover, these images can be easily
+shared, enabling seamless reproducibility across machines and
+collaborators.
+
+The Rocker project, introduced by \citet{boettiger2017}, provides a
+collection of pre-built \pkg{Docker} images for the \proglang{R}
+community. These images come with specific R versions and, in some
+cases, preinstalled packages, making them convenient base images for
+building reproducible analysis environments without starting from
+scratch.
+
+Despite its strengths, \pkg{Docker} can be cumbersome for interactive
+use. Although containers can be modified at runtime, any changes are
+lost once the container stops unless they are explicitly saved. Running
+graphical applications from containers is also possible, but it is
+generally difficult to configure and tends to work reliably only on
+Linux systems. For this reason, \pkg{Docker} is most often used to run
+web-based applications (such as integrated development environments,
+IDEs, like the web version of RStudio) inside containers, offering a
+reproducible yet flexible environment for interactive work.
+
+Some IDEs provide built-in support for executing code directly within a
+running container, allowing a more seamless workflow. However, these
+setups typically require additional configuration, and many
+general-purpose IDEs are not well optimized for statistical or data
+science workflows (though Positron may soon change that).
+
+A common practice is to conduct a study interactively using a standard
+installation of \proglang{R} and its packages, and only after completing
+the analysis, create a \texttt{Dockerfile} to allow others to reproduce
+the results easily. For instance, the \pkg{dockerfiler} package
+\citep{dockerfiler} can automatically generate a \texttt{Dockerfile}
+from an existing \texttt{renv.lock} file, simplifying this process.
+While this approach enables post-hoc reproducibility, it does not solve
+the challenge of consistently deploying the same environment across
+multiple machines as co-authors are working on the project. Ideally,
+analyses would be developed directly within a containerized environment
+from the start, ensuring that the computational setup is reproducible
+throughout the project lifecycle.
+
+Another challenge of using \pkg{Docker} is that its images are
+effectively minimal Linux systems, so familiarity with Linux is highly
+recommended for writing reliable and efficient \texttt{Dockerfile}s. To
+guarantee that a build consistently produces the same image, the
+\texttt{Dockerfile} should reference specific image \emph{digests}
+rather than general \emph{tags}. In practice, however, tagged versions
+are more commonly used, which can introduce potential reproducibility
+issues. \pkg{dockerfiler} is an R package that lowers the barrier to
+entry by automating parts of the \texttt{Dockerfile} creation process,
+but they cannot fully replace the need for basic Linux knowledge.
+
+A final step toward the gold standard of reproducibility is to automate
+the analysis execution using a build system such as \pkg{Make}, rather
+than relying on ad hoc script execution. Build automation tools enable
+analyses to be defined as a series of reproducible, well-ordered steps
+that can be executed consistently. Within \proglang{R}, a modern and
+powerful alternative is the \pkg{targets} package by \citet{landau2021},
+which provides declarative workflows for reproducible data analysis.
+
+\citet{mcdermott2021} provides a clear example of a scientific study
+that achieved the gold standard of reproducibility. The author created
+an accompanying GitHub repository\footnote{https://github.com/grantmcdermott/skeptic-priors}
+containing detailed instructions to install the required software and
+execute the analysis. Examining this repository reveals the use of
+multiple tools to capture and share the computational environment:
 
@@
 \item
-  Packages and their versions were listed and saved into an
-  \texttt{renv.lock} file;
+  Packages and their versions were recorded in an \texttt{renv.lock}
+  file;
 \item
-  A \texttt{Makefile} was used to run the whole analysis and compile the
-  paper;
+  A \texttt{Makefile} was used to automate the full analysis and compile
+  the paper;
 \item
-  A \texttt{Dockerfile} was used to provide the complete computational
-  environment, including the right version of \proglang{R} and run the
-  whole analysis easily.
+  A \texttt{Dockerfile} provided a complete computational environment,
+  including the correct version of \proglang{R}, enabling easy execution
+  of the entire workflow.
 \end{itemize}
 
-However, reaching this gold standard is quite costly: one needs to learn
-a tool to deal with package versions, then \proglang{Docker} for all the
-other software, including the programming language that was used.
-Ideally, a build automation tool such as \proglang{make} should be added
-to the list. It should also be noted that these requirements are not
-limited to the \proglang{R} programming language. A very similar
-approach should be taken if one uses \proglang{Python} for statistical
-analysis instead.
+Reaching this gold standard, however, is resource-intensive. It requires
+learning a tool for managing package versions, mastering \pkg{Docker}
+for the programming language and other software dependencies, and using
+a build automation tool. Importantly, these considerations are not
+unique to \proglang{R}: similar approaches are needed for
+\proglang{Python} or any other programming language, and complexity
+further increases if several programming languages are needed for a
+single project.
 
-As an alternative to the above approach, I will present the
-\proglang{Nix} package manager, which available for all major operating
-system and focuses on installing and building software in a reproducible
-manner. To make \proglang{Nix} more accessible to \proglang{R}
-programmers, I developed the \pkg{rix} which I will also present in this
-article.
+As an alternative, we present the \pkg{Nix} package manager, which
+supports all major operating systems and emphasizes reproducible
+software installation and building. \pkg{Nix} can replace \pkg{Docker},
+\pkg{renv}, and even build automation tools such as \pkg{Make}. To make
+\pkg{Nix} more accessible for \proglang{R} users, we developed the
+\pkg{rix} and {rixpress} packages, which are introduced and discussed in
+this article.
 
-\section{The Nix package manager}\label{the-nix-package-manager}
+\section{The Nix package manager}\label{sec-nix}
 
-\proglang{Nix} is a package manager that can be used to install and
-build software in a completely reproducible manner. As of writing, it
-contains more than 80.000 packages, and the entirety of CRAN and
-Bioconductor is available through \proglang{Nix}'s repositories. This
-means that using \proglang{Nix}, it is possible to install not only
-\proglang{R}, but also all the packages required for a project. The
-reason why one should use \proglang{Nix} to install \proglang{R}
-packages, and not use the usual, built-in, \texttt{install.packages()}
-\proglang{R} function instead, is that when installing a package with
-\proglang{Nix}, \proglang{Nix} makes sure to install every dependency of
-every package, whether this dependency is another \proglang{R} package
-or a system-level dependency.
+\pkg{Nix} is a package manager designed to install and build software in
+a fully reproducible manner. As of this writing, it provides access to
+over 120000 packages, including nearly all of CRAN and Bioconductor.
+This allows users to install not only \proglang{R} itself but also all
+packages required for a given project. Although \pkg{Nix} is the package
+manager of the NixOS Linux distribution, it can also be installed as a
+standalone tool on other Linux distributions, macOS, and the Windows
+Subsystem for Linux\footnote{Effectively treating non-NixOS Linux
+  distributions and Windows as equivalent platforms in practice.}.
 
-For example, the \pkg{xlsx} \proglang{R} package requires the
-\proglang{Java} programming language to be installed to successfully
-install. Depending on the what system one tries to install \pkg{xlsx},
-installing \proglang{Java} might not easy, nor even possible. But with
-\proglang{Nix}, it suffices to declare that the \pkg{xlsx} package is
-needed for the project, and \proglang{Nix} figures out automatically
-that \proglang{Java} is required and installs and configures it. It all
-just happens without any required intervention from the user.
+The advantage of using \pkg{Nix} to install \proglang{R} packages,
+instead of the standard \texttt{install.packages()} function, is that
+\pkg{Nix} ensures all dependencies of each package are installed,
+whether they are other \proglang{R} packages or system-level libraries.
 
-But where do these packages come from? When installing a package using
-\proglang{Nix}, an expression written in the \proglang{Nix} programming
-language gets downloaded from the \texttt{nixpkgs} Github repository and
-evaluated. This expression contains a so-called \emph{derivation}. A
-derivation defines a build: its dependencies, commands to build and
-install the package in question, and then an output. Most of the time, a
-derivation downloads source code, builds the software from the source
-and then outputs a compiled binary. Derivations are extremely flexible,
-and by writing one's own, it is possible to define and build environment
-for a project in a reproducible manner.
+For example, the \pkg{xlsx} package requires \proglang{Java} to be
+installed. On some systems, installing \proglang{Java} may be difficult
+or even impossible manually. With \pkg{Nix}, however, the user only
+needs to declare that \pkg{xlsx} is required for the project; \pkg{Nix}
+automatically installs and configures \proglang{Java} as a dependency.
+This works because the maintainers of the \proglang{R} language packages
+for \pkg{Nix} have declared \proglang{Java} as a dependency of
+\pkg{xlsx}, allowing the process to happen seamlessly from the end-user
+perspective.
 
-Because the whole set of \proglang{Nix} expression is hosted on Github,
-it is possible to pin a specific revision of \texttt{nixpkgs} to ensure
-reproducibility of our project. Pinning a revision ensures that every
-package that Nix installs will always be at exactly the same versions,
-regardless of when in the future the packages get installed.
+\pkg{xlsx} can thus be referred to as a component closure, and quoting
+\citet{dolstra2004nix}:
 
-With Nix, it is essentially possible to replace \pkg{renv} and Docker
-combined, or in the case of \proglang{Python}, replace
-\texttt{requirements.txt} files which list package dependencies for
-\proglang{Python} projects. It is also possible to build multi-language
-environments, containing R and Python, a LaTeX distribution, and even
-install one's favourite text editor to edit the project's source code.
+\begin{quote}
+The idea is to always deploy component closures: if we deploy a
+component, then we must also deploy its dependencies, their
+dependencies, and so on. That is, we must always deploy a set of
+components that is closed under the `'depends on'\,' relation. Since
+closures are selfcontained, they are the units of complete software
+deployment. After all, if a set of components is not closed, it is not
+safe to deploy, since using them might cause other components to be
+referenced that are missing on the target system.
+\end{quote}
+
+But how does \pkg{Nix} achieve this reproducibility? Where do these
+packages, or \emph{closures}, come from? When installing a package with
+\pkg{Nix}, an expression written in the \pkg{Nix} language is downloaded
+from the \texttt{nixpkgs} GitHub repository and evaluated. These
+expressions define \emph{derivations}, which describe how to build a
+package: specifying its dependencies, the commands to build and install
+it, and the resulting output. Typically, a derivation downloads the
+source code, compiles it, and outputs a binary. Derivations are highly
+flexible; by writing custom derivations, users can define and build
+reproducible environments instead of a single program. The dependencies
+of each derivation are themselves defined in other expressions and are
+automatically installed if needed. In the case of \proglang{R},
+maintainers declare the appropriate dependencies to create fully
+self-contained component closures.
+
+Why is installing software with \pkg{Nix} reproducible? Because the
+entire set of \pkg{Nix} expressions is hosted on GitHub, it is possible
+to reference a specific commit of \texttt{nixpkgs}, a process called
+\emph{pinning a revision}. Pinning a revision ensures that all packages
+installed by \pkg{Nix} are always the exact same versions, regardless of
+when in the future they are built, since the expressions downloaded
+correspond to that specific commit of \texttt{nixpkgs}.
+
+Pinning is crucial, but it is not the only reason \pkg{Nix} supports
+reproducibility. \pkg{Nix} is a functional package manager: it applies
+principles from functional programming, disallowing side effects and
+global variables, and ensuring that the output of a build is always the
+same given the same inputs, regardless of time or location. A side note:
+while this functional approach greatly enhances reproducibility, it can
+make writing derivations more complex, particularly for packages that
+need to download assets during installation (e.g., some Bioconductor
+packages like \pkg{musData}). However, this concern is primarily for
+package maintainers, not end-users.
+
+Additionally, \pkg{Nix} supports multiple versions, or \emph{variants},
+of a package on the same system. Each variant has a unique identifier,
+allowing multiple versions of \proglang{R} to coexist and ensuring the
+correct version is used for each project. For a more technical
+discussion of \pkg{Nix}, we refer to \citet{dolstra2004nix}.
+
+With \pkg{Nix}, it is possible to effectively replace both \pkg{renv}
+and \pkg{Docker} for \proglang{R} projects---or, in the case of
+\proglang{Python}, to replace \texttt{requirements.txt} files and
+virtual environments. \pkg{Nix} also supports building multi-language
+environments, which can include \proglang{R} and \proglang{Python}, a
+\LaTeX distribution or any other of the 120,000 available tools. This
+enables the creation of a complete, project-specific, and reproducible
+environment that can be used interactively or non-interactively. As long
+as the \texttt{nixpkgs} GitHub repository remains online, the
+environment can be rebuilt in the future to rerun analyses reliably.
+
+However, \pkg{Nix} has a steep learning curve. It is a complex system
+that comes with its own programming language, also called
+\proglang{Nix}, designed to solve the challenging problem of
+declaratively defining how software is built and configured. This
+ensures that installations are fully reproducible across operating
+systems and hardware platforms. To make \pkg{Nix} more accessible to
+\proglang{R} users that wish to use project-specific and reproducible
+development environments, we developed the \pkg{rix} package.
+
+\section{Reproducible development environments with
+Nix}\label{sec-repro-nix}
+
+As mentioned, \pkg{Nix} expressions are written in the \proglang{Nix}
+programming language, which is purely functional. Here is a simple
+example that creates a shell environment containing version 4.3.1 of
+\proglang{R}:
+
+\begin{CodeInput}
+let
+  pkgs = import (fetchTarball
+    "https://github.com/NixOS/nixpkgs/archive/976fa336.tar.gz"
+  ) {};
+  system_packages = builtins.attrValues {
+    inherit (pkgs) R;
+  };
+in
+  pkgs.mkShell {
+    buildInputs = [ system_packages ];
+    shellHook = "R --vanilla";
+  }
+\end{CodeInput}
+
+In this expression, the \texttt{let} keyword is used to define
+variables. The variable \texttt{pkgs} imports the set of packages from
+the \texttt{nixpkgs} repository at the specified commit
+\texttt{976fa336}. The variable \texttt{system\_packages} lists the
+packages to include in the environment; in this case, it is just the
+\proglang{R} programming language, along with all its dependencies and
+their transitive dependencies. The \texttt{mkShell} function then
+creates a development shell with the specified packages. The
+\texttt{shellHook} is set to \texttt{"R\ -\/-vanilla"}, meaning that
+entering the shell automatically starts \proglang{R} in vanilla mode,
+ignoring any startup options.
+
+This expression can be saved in a file called \texttt{default.nix}. The
+environment can then be built on a system with \pkg{Nix} installed using
+the \texttt{nix-build} command.\footnote{For installing \pkg{Nix}, we
+  recommend the Determinate Systems installer:
+  \url{https://determinate.systems/posts/determinate-nix-installer}}.
+Once the build completes, the user can enter the interactive shell with
+\texttt{nix-shell}. This shell contains all the packages specified in
+\texttt{default.nix} and can be used for development, similar to
+activating a virtual environment in the \proglang{Python} ecosystem.
+
+Writing \pkg{Nix} expressions can be challenging for users unfamiliar
+with the \proglang{Nix} language. However, the ability to define a fully
+reproducible development environment in a single text file and then
+rebuild it anywhere is highly appealing. To lower the barrier to
+adoption of \pkg{Nix} for reproducibility, we developed the \pkg{rix}
+package.
+
+\pkg{rix} provides the \texttt{rix()} function, which simplifies
+generating \pkg{Nix} expressions. It is available on CRAN and can be
+installed like any other R package. Additionally, it can bootstrap an
+\proglang{R} development environment on a system where \proglang{R} is
+not yet installed but \pkg{Nix} is available. This can be done by
+running (inside of a terminal):
+
+\begin{CodeInput}
+$> nix-shell -I \
+  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/refs/heads/2025-10-17.tar.gz -p \
+  R rPackages.rix
+\end{CodeInput}
+
+(the \texttt{-I} flag allows one to pass a specific revision of
+\texttt{nixpkgs}, ensuring temporary shells are also reproducible).
+
+This command opens a temporary \proglang{R} session with \pkg{rix}
+available.\footnote{\texttt{nix-shell\ -p} starts an interactive shell
+  with the specified packages.} From there, users can generate new
+\pkg{Nix} expressions for building environments. For example, the
+following generates a \texttt{default.nix} file that installs
+\proglang{R} 4.3.1 along with the \pkg{dplyr} and \pkg{chronicler}
+packages:
+
+\begin{CodeInput}
+R> library('rix')
+
+R> rix(r_ver = "4.3.1",
++    r_pkgs = c("dplyr", "chronicler"),
++    project_path = ".",
++    overwrite = TRUE)
+\end{CodeInput}
+
+\pkg{rix} can also handle more complex setups:
+
+\begin{CodeInput}
+R> rix(r_ver = "4.3.1",
++    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
++    system_pkgs = c("quarto", "git"),
++    tex_pkgs = c(
++          "amsmath",
++          "framed",
++          "fvextra",
++          "environ",
++          "fontawesome5",
++          "orcidlink",
++          "pdfcol",
++          "tcolorbox",
++          "tikzfill"
++    ),
++    git_pkgs = list(
++      list(
++        package_name = "rix",
++        repo_url = "https://github.com/b-rodrigues/rix/",
++        branch_name = "master",
++        commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
++      list(
++        package_name = "fusen",
++        repo_url = "https://github.com/ThinkR-open/fusen",
++        branch_name = "main",
++        commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
++    ),
++    ide = "positron",
++    project_path = ".",
++    overwrite = TRUE)
+\end{CodeInput}
+
+This call to \texttt{rix()} generates an environment that installs
+several \proglang{R} packages (including \pkg{AER} version 1.2-8),
+several TeXLive packages for \LaTeX document authoring, development
+versions of \pkg{rix} and \pkg{fusen} from GitHub, and the Positron
+editor.
+
+\pkg{rix} can generate \pkg{Nix} expressions even if \pkg{Nix} is not
+installed on the system. This is useful for continuous integration and
+continuous deployment (CI/CD) workflows on platforms such as GitHub
+Actions. For instance, the repository containing the source code for
+this article\footnote{https://github.com/b-rodrigues/rix\_paper} uses
+GitHub Actions to compile the paper. Each time a push is made to the
+master branch, a runner installs \pkg{Nix}, generates the environment
+from the hosted \texttt{default.nix} file, and compiles the paper using
+\pkg{Quarto} within the reproducible environment. This ensure that
+\emph{exactly} the same environment is used on the author's computer and
+on the CI/CD without any additional, platform-specific, configuration.
+
+Instead of first entering a \pkg{Nix} shell, it is also possible to run
+a program directly from the environment:
+
+\begin{CodeInput}
+cd /path/to/project/ && nix-shell default.nix --run "Rscript analysis.R"
+\end{CodeInput}
+
+This command runs \texttt{Rscript} and executes the \texttt{analysis.R}
+script, which in this example should be located in the same directory as
+\texttt{default.nix}.
+
+\section{The rstats-on-nix fork of nixpkgs}\label{sec-fork}
+
+As explained earlier, \pkg{Nix} uses expressions from the
+\texttt{nixpkgs} GitHub repository to build software. However, when
+generating expressions with \pkg{rix}, our fork
+\texttt{rstats-on-nix/nixpkgs} is used instead.
+
+Using a fork offers several advantages. First, it provides flexibility
+that the official \texttt{nixpkgs} repository cannot always accommodate.
+\pkg{Nix} is primarily the package manager for the NixOS Linux
+distribution, and governance and technical choices made upstream can
+limit what \pkg{rix} aims to provide.
+
+For instance, while \pkg{Nix} can theoretically support multiple
+versions (or \emph{variants}) of the same package, in practice
+maintainers cannot provide several variants for all \proglang{R}
+packages, given the size of the ecosystem (over 20,000 CRAN and
+Bioconductor packages). This makes it difficult to install a specific
+version of an \proglang{R} package not included in a particular
+\texttt{nixpkgs} commit. With \pkg{rix}, users can install a specific
+package version from source, e.g.:
+
+\begin{CodeInput}
+R> rix(..., r_pkgs = "dplyr@1.0.7", ...)
+\end{CodeInput}
+
+However, installing from source might fail, especially if the package
+needs to be compiled.
+
+Additionally, updating the full \proglang{R} package set on \pkg{Nix}
+daily is impractical. While CRAN and Bioconductor update daily, the
+\proglang{R} packages in \texttt{nixpkgs} are only updated with new R
+releases. This limitation is due to \pkg{Nix}'s governance as a Linux
+distribution package manager.
+
+Our fork allows us to circumvent these limitations. For example, we
+provide a daily snapshot of CRAN. Each day, the \proglang{R} package set
+is updated and committed to a dated branch using GitHub Actions. Users
+can select a specific date with:
+
+\begin{CodeInput}
+R> rix(date = "2024-12-14", ...)
+\end{CodeInput}
+
+We strive to provide an available date per week: each Monday, a GitHub
+Action tests popular packages on Linux and macOS, and only if all tests
+succeed is the date added to the list of available dates in \pkg{rix}.
+This ensures users can reliably install packages, and allows us to
+backport fixes if needed. For example, when RStudio was temporarily
+broken due to a dependency issue (\texttt{boost}), a pull request was
+submitted to the official \texttt{nixpkgs} repository. We backported the
+fix to our fork, making RStudio available to users of our fork earlier
+than upstream, as merging PRs in the official repository can take some
+time.
+
+We have backported fixes to our \texttt{nixpkgs} fork as far back as
+March 2019. The process involves checking out a \texttt{nixpkgs} commit
+on the selected date, updating the \proglang{R} package set using Posit
+CRAN and Bioconductor snapshots, backporting fixes, and ensuring popular
+packages work on both x86-linux (including WSL2) and aarch64-darwin
+(Apple Silicon). These changes are committed to a dated branch in
+\texttt{rstats-on-nix/nixpkgs}. Users can see all available dates with
+\texttt{rix::available\_dates()}.
+
+A drawback of forking \texttt{nixpkgs} is that backported packages are
+not included upstream and thus are not prebuilt by Nix's CI platform,
+Hydra. Users may need to build many packages from source, which can be
+time-consuming. To mitigate this, we provide a binary cache sponsored by
+\href{https://www.cachix.org/}{Cachix}, complementing the public Nix
+cache. Instructions for using Cachix are in \pkg{rix}'s documentation.
+Using the cache significantly speeds up installations, as prebuilt
+packages are downloaded rather than compiled.
+
+\section{Orchestrating the workflow with rixpress}\label{sec-rixpress}
+
+Defining a reproducible environment with \pkg{rix} addresses the first
+major challenge of reproducibility. The second challenge is reliably and
+efficiently executing the analysis workflow within that environment,
+which is the role of the \pkg{rixpress} package.
+
+As mentioned in the introduction, a build automation tool like
+\pkg{targets} is invaluable for managing complex analyses. It tracks
+dependencies between code and data, caches results, and only recomputes
+steps that have changed. One can run a \pkg{targets} pipeline inside a
+Nix environment to make it reproducible. However, this approach has
+limitations: the entire pipeline must run in a single environment, and
+orchestrating steps across different languages (e.g., \proglang{R} and
+\proglang{Python}) requires manual handling via packages like
+\pkg{reticulate}.
+
+\pkg{rixpress} overcomes these limitations by using Nix not just as a
+package manager, but as the build automation engine itself. In a
+\pkg{rixpress} pipeline, each step is defined as a Nix derivation,
+providing two key benefits:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  True Polyglot Pipelines: Each step can have its own Nix environment. A
+  Python step can run in a pure Python environment, an \proglang{R} step
+  in an \proglang{R} environment, and a Quarto rendering step in yet
+  another, all within the same pipeline.
+\item
+  Deep Reproducibility: Each step is a hermetically sealed Nix
+  derivation whose output is cached in the Nix store based on the hash
+  of all its inputs. All artifacts are direct children of the
+  computational environment (because the computational environment is
+  actually a dependency of the artifacts), ensuring they are rebuilt if
+  the environment changes, keeping environment and outputs always in
+  sync.
+\end{enumerate}
+
+The user defines the pipeline in an \proglang{R} script using functions
+inspired by \pkg{targets}. For example, a simple pipeline might read
+data with \proglang{Python} (\texttt{polars}), filter it, process it
+with \proglang{R} (\texttt{dplyr}), and compile a Quarto document. The
+pipeline is defined as a list, typically in a file called
+\texttt{gen-pipeline.R} and its execution environment in a file called
+\texttt{gen-env.R}. Templates of these files can be created using
+\texttt{rxp\_init()}. On a machine that has \pkg{Nix} installed it is
+possible to bootstrap an environment using a temporary shell:
+
+\begin{CodeInput}
+$> nix-shell -I \
+  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/refs/heads/2025-10-17.tar.gz -p \
+  R rPackages.rix rPackages.rixpress
+\end{CodeInput}
+
+One can start the R interpreter, and then call \texttt{rxp\_init()}:
+
+\begin{CodeInput}
+R> rixpress::rxp_init()
+\end{CodeInput}
+
+Let us modify \texttt{gen-pipeline.R}:
+
+\begin{CodeInput}
+R> library('rixpress')
+
+R> pipeline_steps <- list(
++  # Step 1: Read data using Python
++  rxp_py_file(
++    name = mtcars_pl,
++    path = "data/mtcars.csv",
++    read_function = "lambda x: polars.read_csv(x, separator = '|')"
++  ),
++
++  # Step 2: Filter the data in Python
++  rxp_py(
++    name = mtcars_filtered,
++    expr = "mtcars_pl.filter(polars.col('am') == 1).to_pandas()"
++  ),
++
++  # Step 3: Convert Python pandas DataFrame to R data.frame
++  # This uses reticulate for conversion. It would also be possible
++  # to pass objects to and from languages by serializing them
++  # to json for instance, or another universal format
++  rxp_py2r(
++    name = mtcars_r,
++    expr = mtcars_filtered
++  ),
++
++  # Step 4: Process data in R
++  rxp_r(
++    name = mtcars_summary,
++    expr = dplyr::summarise(mtcars_r, avg_mpg = mean(mpg))
++  )
++)
+
+# Generate the 'pipeline.nix' file from the R list
+R> rxp_populate(pipeline_steps)
+\end{CodeInput}
+
+The required execution environment, defined in \texttt{gen-env.R}, is as
+follows:
+
+\begin{CodeInput}
+R> library(rix)
+
+R> rix(
++   date = "2025-10-14",
++   r_pkgs = c("dplyr", "ggdag", "quarto", "reticulate", "rixpress"),
++   py_conf = list(
++     py_version = "3.13",
++     py_pkgs = c("polars", "pyarrow", "pandas")
++   ),
++   ide = "none",
++   project_path = ".",
++   overwrite = TRUE
++ )
+\end{CodeInput}
+
+Using the same temporary shell, users can now generate the
+\texttt{default.nix} file by calling \texttt{source("gen-env.R")}. Users
+can now quit the temporary shell, and build the environment by calling
+\texttt{nix-shell} in a terminal. Once the development (or execution)
+environment has been built, users can enter it, and call
+\texttt{source("gen-pipeline.R")} to generate a file called
+\texttt{pipeline.nix} which is the \pkg{Nix} expression that defines the
+entire reproducible pipeline.
+
+The environment also includes the \pkg{ggdag} package \citep{ggdag}, so
+it's possible to view a visual representation of the pipeline (yellow
+nodes are \pkg{Python} and blue nodes are \pkg{R} derivations
+respectively):
+
+\begin{figure}[H]
+
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{dag.png}}
+
+}
+
+\caption{Graphical representation of the pipeline}
+
+\end{figure}%
+
+Running the script above does not execute the pipeline; it generates a
+\texttt{pipeline.nix} file, which declaratively defines the workflow.
+
+To build and execute the pipeline:
+
+\begin{CodeInput}
+R> rxp_make()
+\end{CodeInput}
+
+\begin{CodeInput}
+Build process started...
+
++ mtcars_pl building
++ mtcars_filtered building
++ mtcars_r building
++ mtcars_summary building
+✓ mtcars_filtered built
+✓ mtcars_pl built
+✓ mtcars_r built
+✓ mtcars_summary built
+✓ pipeline completed [4 completed, 0 errored]
+Build successful! Run `rxp_inspect()` for a summary.
+Use `rxp_read("derivation_name")` to read objects or
+`rxp_load("derivation_name")` to load them into the global environment.
+\end{CodeInput}
+
+\pkg{Nix} executes each step in order, building dependencies as needed.
+Outputs are cached, so subsequent runs only recompute steps with changed
+inputs or code. The build process can also be executed in parallel, if
+the the pipeline has independent branches. This provides the efficiency
+of \pkg{targets} with multi-language support and bit-for-bit
+reproducibility. Artifacts can be inspected interactively in
+\proglang{R} using \texttt{rxp\_read("artifact\_name")} or
+\texttt{rxp\_load("artifact\_name")}.
+
+For Python users, a port called \pkg{ryxpress} allows building the same
+pipelines and inspecting outputs from Python sessions.
+
+\pkg{rixpress} includes several additional features not covered here for
+the sake of brevity.
+
+It is also possible to configure popular IDEs to work interactively and
+seamlessly with both \pkg{rix} and \pkg{rixpress}, enabling a smooth,
+reproducible workflow from within the development environment. Detailed
+setup instructions are provided in the vignettes of both packages.
+
+\section{Conclusion}\label{sec-conclusion}
+
+Many tools exist to improve reproducibility, but \pkg{Nix} stands out
+because it deploys complete software environments \emph{closed under the
+``depends on'' relation}: it installs not only a package, but all its
+dependencies and their dependencies. This makes \pkg{Nix} particularly
+powerful for reproducible research.
+
+However, solving such a complex problem makes \pkg{Nix} a complex tool.
+With \pkg{rix}, we aim to make \pkg{Nix} more accessible to \proglang{R}
+users by providing a familiar interface and workflow. By building
+reproducible development shells with \pkg{Nix}, researchers can
+accommodate a wide range of use cases: running scripts and pipelines,
+developing interactive \pkg{shiny} applications, or serving
+\pkg{plumber} APIs.
+
+Furthermore, \pkg{rixpress} extends this reproducibility to entire
+analysis pipelines. By leveraging Nix as a build automation engine,
+\pkg{rixpress} allows polyglot workflows where each step runs in its own
+hermetically sealed environment. This ensures deep reproducibility,
+efficient caching, and seamless orchestration of multi-language
+analyses, making it easier to manage complex projects with confidence
+that results can be reproduced exactly.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, left=2mm, toprule=.15mm, colback=white, leftrule=.75mm, breakable, rightrule=.15mm, bottomrule=.15mm, opacityback=0]
-
-All acknowledgments (note the AE spelling) should be collected in this
-unnumbered section before the references. It may contain the usual
-information about funding and feedback from colleagues/reviewers/etc.
-Furthermore, information such as relative contributions of the authors
-may be added here (if any).
-
-\end{tcolorbox}
+We thank the rOpenSci reviewers and contributors who provided valuable
+feedback on the development of \pkg{rix} and \pkg{rixpress}. In
+particular, we are grateful to David Watkins and Jacob Wujiciak-Jens for
+their reviews of \texttt{\{rix\}}, and to William Landau and Anthony
+Martinez for their reviews of \texttt{\{rixpress\}}. We also acknowledge
+the contributions of Richard J. Acton, Jordi Rosell, Elio Campitelli,
+László Kupcsik, and Michael Heming for \texttt{\{rix\}}. Their expertise
+and feedback greatly improved the quality and usability of these
+packages.
 
@@
 
-\newpage{}
-
-\section*{More technical details}\label{sec-techdetails}
-\addcontentsline{toc}{section}{More technical details}
-
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, left=2mm, toprule=.15mm, colback=white, leftrule=.75mm, breakable, rightrule=.15mm, bottomrule=.15mm, opacityback=0]
-
-Appendices can be included after the bibliography (with a page break).
-Each section within the appendix should have a proper section title
-(rather than just \emph{Appendix}).
-
-For more technical style details, please check out JSS's style FAQ at
-{[}https://www.jstatsoft.org/pages/view/style\#frequently-asked-questions{]}
-which includes the following topics:
-
-\begin{itemize}
-\tightlist
-\item
-  Title vs.~sentence case.
-\item
-  Graphics formatting.
-\item
-  Naming conventions.
-\item
-  Turning JSS manuscripts into \proglang{R} package vignettes.
-\item
-  Trouble shooting.
-\item
-  Many other potentially helpful details\ldots{}
-\end{itemize}
-
-\end{tcolorbox}
-
-\section*{Using BibTeX}\label{sec-bibtex}
-\addcontentsline{toc}{section}{Using BibTeX}
-
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, left=2mm, toprule=.15mm, colback=white, leftrule=.75mm, breakable, rightrule=.15mm, bottomrule=.15mm, opacityback=0]
-
-References need to be provided in a \textsc{Bib}{\TeX} file
-(\texttt{.bib}). All references should be made with \texttt{@cite}
-syntax. This commands yield different formats of author-year citations
-and allow to include additional details (e.g.,pages, chapters, \dots) in
-brackets. In case you are not familiar with these commands see the JSS
-style FAQ for details.
-
-Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task --
-especially when acquiring the entries automatically from mixed online
-sources. However, it is important that informations are complete and
-presented in a consistent style to avoid confusions. JSS requires the
-following format.
-
-\begin{itemize}
-\tightlist
-\item
-  item JSS-specific markup (\texttt{\textbackslash{}proglang},
-  \texttt{\textbackslash{}pkg}, \texttt{\textbackslash{}code}) should be
-  used in the references.
-\item
-  item Titles should be in title case.
-\item
-  item Journal titles should not be abbreviated and in title case.
-\item
-  item DOIs should be included where available.
-\item
-  item Software should be properly cited as well. For \proglang{R}
-  packages \texttt{citation("pkgname")} typically provides a good
-  starting point.
-\end{itemize}
-
-\end{tcolorbox}
-
 \end{document}

### 2025-10-20 paper.qmd
updated example

@@
   1. True Polyglot Pipelines: Each step can have its own Nix environment. A
-  Python step can run in a pure Python environment, an [R]{.proglang} step in an
-  [R]{.proglang} environment, and a Quarto rendering step in yet another, all
-  within the same pipeline.
+     Python step can run in a pure Python environment, an [R]{.proglang} step in
+     an [R]{.proglang} environment, and a Quarto rendering step in yet another,
+     all within the same pipeline.
   2. Deep Reproducibility: Each step is a hermetically sealed Nix derivation
-  whose output is cached in the Nix store based on the hash of all its inputs.
-  All artifacts are direct children of the computational environment (because
-  the computational environment is actually a dependency of the artifacts),
-  ensuring they are rebuilt if the environment changes, keeping environment and
-  outputs always in sync.
-
-The user defines the pipeline in an [R]{.proglang} script using functions
-inspired by [targets]{.pkg}. For example, a simple pipeline might read data with
-[Python]{.proglang} (`polars`), filter it, process it with [R]{.proglang}
-(`dplyr`), and compile a Quarto document. The pipeline is defined as a list,
-typically in a file called `gen-pipeline.R` and its execution environment in a
-file called `gen-env.R`. Templates of these files can be created using
-`rxp_init()`. On a machine that has [Nix]{.pkg} installed it is possible to
-bootstrap an environment using a temporary shell:
-
-```{=tex}
-\begin{CodeInput}
-$> nix-shell -I \
-  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/refs/heads/2025-10-17.tar.gz -p \
-  R rPackages.rix rPackages.rixpress
-\end{CodeInput}
-```
-One can start the R interpreter, and then call `rxp_init()`:
-
-```{=tex}
-\begin{CodeInput}
-R> rixpress::rxp_init()
-\end{CodeInput}
-```
-
-Let us modify `gen-pipeline.R`:
+     whose output is cached in the Nix store based on the hash of all its
+     inputs. All artifacts are direct children of the computational environment
+     (because the computational environment is actually a dependency of the
+     artifacts), ensuring they are rebuilt if the environment changes, keeping
+     environment and outputs always in sync.
+
+To illustrate these capabilities with a realistic research task, we present a
+polyglot pipeline that simulates a Real Business Cycle (RBC) model in
+[Julia]{.proglang}, uses the resulting data to train an [XGBoost]{.pkg}
+forecasting model in [Python]{.proglang}, visualizes the results in
+[R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
+report.
+
+The user defines the pipeline in an [R]{.proglang} script (`gen-pipeline.R`)
+using functions inspired by [targets]{.pkg}. The underlying logic for each step
+is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
+`functions.R`), the full contents of which are detailed in the Appendix. The
+high-level orchestration script demonstrates how `{rixpress}` defines a
+granular, multi-step machine learning workflow that crosses language boundaries:
 
@@
 R> pipeline_steps <- list(
-+  # Step 1: Read data using Python
-+  rxp_py_file(
-+    name = mtcars_pl,
-+    path = "data/mtcars.csv",
-+    read_function = "lambda x: polars.read_csv(x, separator = '|')"
++  # STEP 0: Define RBC Model Parameters in Julia
++  rxp_jl(alpha, 0.3),
++  rxp_jl(beta, 1 / 1.01),
++  # ... (other parameters omitted for brevity) ...
++
++  # STEP 1: Julia - Simulate the RBC model
++  rxp_jl(
++    name = simulated_rbc_data,
++    expr = "simulate_rbc_model(alpha, beta, delta, rho, sigma, sigma_z)",
++    user_functions = "functions/functions.jl",
++    encoder = "arrow_write"
 +  ),
 +
-+  # Step 2: Filter the data in Python
++  # STEP 2.1: Python - Prepare features
 +  rxp_py(
-+    name = mtcars_filtered,
-+    expr = "mtcars_pl.filter(polars.col('am') == 1).to_pandas()"
++    name = processed_data,
++    expr = "prepare_features(simulated_rbc_data)",
++    user_functions = "functions/functions.py",
++    decoder = "pyarrow.feather.read_feather"
 +  ),
 +
-+  # Step 3: Convert Python pandas DataFrame to R data.frame
-+  # This uses reticulate for conversion. It would also be possible
-+  # to pass objects to and from languages by serializing them
-+  # to json for instance, or another universal format
-+  rxp_py2r(
-+    name = mtcars_r,
-+    expr = mtcars_filtered
++  # STEP 2.2: Python - Split data (X_train, y_train, etc.)
++  rxp_py(name = X_train, expr = "get_X_train(processed_data)", ...),
++  # ... (other data splits omitted for brevity) ...
++
++  # STEP 2.3: Python - Train the XGBoost model
++  rxp_py(
++    name = trained_model,
++    expr = "train_model(X_train, y_train)",
++    user_functions = "functions/functions.py"
 +  ),
 +
-+  # Step 4: Process data in R
++  # STEP 2.4: Python - Make predictions and format results
++  # ... (prediction and formatting steps omitted for brevity) ...
++  rxp_py(
++    name = final_predictions_df,
++    expr = "format_results(y_test, model_predictions)",
++    user_functions = "functions/functions.py",
++    encoder = "save_arrow"
++  ),
++
++  # STEP 3: R - Visualize the predictions
 +  rxp_r(
-+    name = mtcars_summary,
-+    expr = dplyr::summarise(mtcars_r, avg_mpg = mean(mpg))
++    name = output_plot,
++    expr = plot_predictions(final_predictions_df),
++    user_functions = "functions/functions.R",
++    decoder = arrow::read_feather
++  ),
++
++  # STEP 4: Quarto - Compile the final report
++  rxp_qmd(
++    name = final_report,
++    qmd_file = "readme.qmd"
 +  )
@@
 # Generate the 'pipeline.nix' file from the R list
-R> rxp_populate(pipeline_steps)
-\end{CodeInput}
-```
-
-The required execution environment, defined in `gen-env.R`, is as follows:
-
-```{=tex}
-\begin{CodeInput}
-R> library(rix)
-
-R> rix(
-+   date = "2025-10-14",
-+   r_pkgs = c("dplyr", "ggdag", "quarto", "reticulate", "rixpress"),
-+   py_conf = list(
-+     py_version = "3.13",
-+     py_pkgs = c("polars", "pyarrow", "pandas")
-+   ),
-+   ide = "none",
-+   project_path = ".",
-+   overwrite = TRUE
-+ )
-\end{CodeInput}
-```
-
-Using the same temporary shell, users can now generate the `default.nix` file
-by calling `source("gen-env.R")`. Users can now quit the temporary shell, and
-build the environment by calling `nix-shell` in a terminal. Once the
-development (or execution) environment has been built, users can enter it,
-and call `source("gen-pipeline.R")` to generate a file called `pipeline.nix`
-which is the [Nix]{.pkg} expression that defines the entire reproducible
-pipeline.
-
-The environment also includes the [ggdag]{.pkg} package [@ggdag], so it's
-possible to view a visual representation of the pipeline (yellow nodes are
-[Python]{.pkg} and blue nodes are [R]{.pkg} derivations respectively):
-
-![Graphical representation of the pipeline](dag.png)
-
-Running the script above does not execute the pipeline; it generates a
-`pipeline.nix` file, which declaratively defines the workflow.
-
-To build and execute the pipeline:
-
-```{=tex}
-\begin{CodeInput}
-R> rxp_make()
-\end{CodeInput}
-```
-
-```{=tex}
-\begin{CodeInput}
-Build process started...
-
-+ mtcars_pl building
-+ mtcars_filtered building
-+ mtcars_r building
-+ mtcars_summary building
-✓ mtcars_filtered built
-✓ mtcars_pl built
-✓ mtcars_r built
-✓ mtcars_summary built
-✓ pipeline completed [4 completed, 0 errored]
-Build successful! Run `rxp_inspect()` for a summary.
-Use `rxp_read("derivation_name")` to read objects or
-`rxp_load("derivation_name")` to load them into the global environment.
-\end{CodeInput}
-```
-
-[Nix]{.pkg} executes each step in order, building dependencies as needed.
-Outputs are cached, so subsequent runs only recompute steps with changed inputs
-or code. The build process can also be executed in parallel, if the the pipeline
-has independent branches. This provides the efficiency of [targets]{.pkg} with
-multi-language support and bit-for-bit reproducibility. Artifacts can be
-inspected interactively in [R]{.proglang} using `rxp_read("artifact_name")` or
-`rxp_load("artifact_name")`.
+R> rxp_populate(pipeline_steps, build = TRUE)
+\end{CodeInput}
+```
+
+The `rxp_populate()` function translates this [R]{.proglang} list into a
+`pipeline.nix` file, which declaratively defines the entire workflow. The
+package can then generate a visual representation of the pipeline's directed
+acyclic graph (DAG), as shown in Figure 1.
+
+![Graphical representation of the polyglot pipeline. Green nodes are
+[Julia]{.proglang}, yellow nodes are [Python]{.pkg}, and blue nodes are
+[R]{.proglang} derivations.](dag-polyglot.png){#fig-dag}
+
+To execute the pipeline, one can either set `build = TRUE` in `rxp_populate()`
+or call `rxp_make()` separately. [Nix]{.pkg} executes each step in order,
+building dependencies as needed. Outputs are cached, so subsequent runs only
+recompute steps with changed inputs or code. This provides the efficiency of
+[targets]{.pkg} with multi-language support and bit-for-bit reproducibility.
+Artifacts can be inspected interactively in [R]{.proglang} using
+`rxp_read("artifact_name")` or `rxp_load("artifact_name")`.
 
 For Python users, a port called [ryxpress]{.pkg} allows building the same
-pipelines and inspecting outputs from Python sessions.
-
-[rixpress]{.pkg} includes several additional features not covered here for the
-sake of brevity.
-
-It is also possible to configure popular IDEs to work interactively and
-seamlessly with both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth,
-reproducible workflow from within the development environment. Detailed setup
-instructions are provided in the vignettes of both packages.
+pipelines and inspecting outputs from Python sessions. [rixpress]{.pkg} also
+includes several additional features not covered here for brevity. It is also
+possible to configure popular IDEs to work interactively and seamlessly with
+both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth, reproducible workflow
+from within the development environment. Detailed setup instructions are
+provided in the vignettes of both packages.
 
@@
 :::
+
+## Appendix: A Complete Polyglot Example with rixpress {#sec-appendix}
+
+This appendix provides the complete source code for the polyglot pipeline
+example discussed in Section 5. The pipeline simulates a Real Business Cycle
+(RBC) model in [Julia]{.proglang}, trains an [XGBoost]{.pkg} model in
+[Python]{.proglang}, and visualizes the results in [R]{.proglang}.
+
+### The Environment Definition (`gen-env.R`)
+
+This script uses [rix]{.pkg} to define a `default.nix` file that contains all
+necessary dependencies for [R]{.proglang}, [Julia]{.proglang}, and
+[Python]{.proglang}.
+
+```{=tex}
+\begin{CodeInput}
+R> # This script defines the polyglot environment our pipeline will run in.
+R> library(rix)
+R>
+R> # Define the complete execution environment
+R> rix(
++   # Pin the environment to a specific date for reproducibility.
++   date = "2025-10-14",
++
++   # R Packages
++   r_pkgs = c(
++     "ggplot2", "dplyr", "arrow", "rix", "rixpress", "quarto"
++   ),
++ 
++   # Julia Configuration
++   jl_conf = list(
++     jl_version = "lts",
++     jl_pkgs = c("Distributions", "DataFrames", "Arrow", "Random")
++   ),
++ 
++   # Python Configuration
++   py_conf = list(
++     py_version = "3.13",
++     py_pkgs = c("pandas", "scikit-learn", "xgboost", "pyarrow")
++   ),
++ 
++   project_path = ".",
++   overwrite = TRUE
++ )
+\end{CodeInput}
+```
+
+### The RBC Model Simulation (`functions/functions.jl`)
+
+This Julia script implements the state-space solution to the RBC model as a pure
+function. The economic theory is based on @depaoli2009.
+
+```{=tex}
+\begin{Code}
+# This script contains the helper functions for the Julia portion of the pipeline.
+# It implements the correct, stable, and canonical log-linearized solution to
+# a standard Real Business Cycle (RBC) model, based on the provided lecture slides.
+
+# In an interactive session, you should uncomment the line below
+# to load the required packages. rixpress handles this automatically.
+#using LinearAlgebra, Distributions, DataFrames, Arrow, Random
+
+# 1. Main Simulation Function
+
+"""
+    simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
+
+    Takes RBC model parameters as input, solves for the state-space representation
+    using the method of undetermined coefficients (as per the slides), simulates
+    the model for 250 quarters, and returns a DataFrame.
+"""
+function simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
+
+    # --- STEADY-STATE VALUES AND CONSTRUCTED PARAMETERS (Slides p. 16, 31) ---
+    y_k = ((1/β) - 1 + δ) / α
+
+    # --- SOLVE FOR POLICY FUNCTION COEFFICIENTS (Slides p. 19-22) ---
+    # We solve for the coefficients of the policy functions:
+    # c_hat_t = c_ck * k_hat_{t-1} + c_cz * z_hat_t
+    # k_hat_t = c_kk * k_hat_{t-1} + c_kz * z_hat_t
+
+    # Solve the quadratic equation for c_ck (coefficient of capital on consumption)
+    # Based on Campbell (1994) and slide 22, we take the stable root.
+    # a*x^2 + b*x + c = 0
+    a_quad = y_k - δ
+    b_quad = -( (1-β*(1-δ))*(y_k - δ) + y_k*(1+α) + 1 )
+    c_quad = y_k * α
+
+    # The stable solution is the smaller positive root
+    c_ck = (-b_quad - sqrt(b_quad^2 - 4*a_quad*c_quad)) / (2*a_quad)
+
+    # Now solve for the other coefficients based on c_ck
+    c_kk = (y_k * α) / (y_k - δ - c_ck)
+    c_cz = (y_k * (1 - c_kk * (1 - α) * β * ρ)) /
+           ( (y_k - δ - c_ck) * (1 - β * ρ) + σ * (1 - c_ck) * (1 - β * ρ) )
+    c_kz = (c_cz * (1 - c_ck)) / (y_k * (1 - α))
+
+    # --- BUILD STATE-SPACE MATRICES FROM SOLVED COEFFICIENTS ---
+    # State vector: s_t = [k_{t-1}, z_t]'
+    # Transition:   s_t = T*s_{t-1} + R*ε_t
+
+    T = [c_kk  c_kz * ρ
+         0     ρ      ]
+
+    R = [c_kz ; 1]
+
+    # Observation Matrix for [output, consumption, investment]
+    # y_hat_t = α*k_{t-1} + z_t
+    # c_hat_t = c_ck*k_{t-1} + c_cz*z_t
+    # i_hat_t = (i_y_ss)^-1 * (k_t - (1-δ)k_{t-1})
+    #         = (i_y_ss)^-1 * ( (c_kk - (1-δ))k_{t-1} + c_kz*z_t )
+
+    i_y_ss = δ * (α / ((1/β) - 1 + δ))
+
+    C = [α      1
+         c_ck   c_cz
+         (c_kk - (1-δ))/i_y_ss   c_kz/i_y_ss]
+
+    # --- SIMULATE THE MODEL ---
+    Random.seed!(1234)
+    n_periods = 250
+    shocks = randn(n_periods) * σ_z
+
+    states = zeros(2, n_periods)
+    for t in 2:n_periods
+        # The state is [k_{t-1}, z_t]'
+        # To get the next state [k_t, z_{t+1}]', we first find k_t
+        k_t = T[1,1]*states[1, t] + T[1,2]*states[2, t] + R[1]*shocks[t]
+        z_tp1 = T[2,1]*states[1, t] + T[2,2]*states[2, t] + R[2]*shocks[t]
+
+        states[1, t] = k_t # This is now k_t, which is "k_lag" for t+1
+        states[2, t] = z_tp1 # This is now z_{t+1}
+    end
+    # Re-aligning states to be [k_{t-1}, z_t]
+    k_lag = [0; states[1, 1:end-1]]
+    z = [0; states[2, 1:end-1]]
+
+    observables = C * [k_lag'; z']
+
+    df = DataFrame(
+        period = 1:n_periods,
+        output = observables[1, :],
+        consumption = observables[2, :],
+        investment = observables[3, :],
+        capital = k_lag,
+        technology = z
+    )
+
+    return df
+end
+
+# 2. Encoder Function (Unchanged)
+"""
+    arrow_write(df::DataFrame, path::String)
+    Encoder function to save a Julia DataFrame to an Arrow file.
+"""
+function arrow_write(df::DataFrame, path::String)
+    Arrow.write(path, df)
+end
+\end{Code}
+```
+
+### The XGBoost Forecasting Model (`functions/functions.py`)
+
+This Python script contains a series of modular functions to handle feature
+engineering, data splitting, model training, and prediction.
+
+```{=tex}
+\begin{Code}
+# This script contains modular helper functions for the Python portion of the pipeline.
+# Each function performs a single, distinct task in the ML workflow.
+
+# 1. Load required Python packages
+# This is only needed if you run this script by hand,
+# in which case, uncomment lines 8 to 11.
+# With rixpress, the packages get loaded automatically.
+#import pandas as pd
+#import pyarrow.feather as feather
+#from sklearn.model_selection import train_test_split
+#import xgboost as xgb
+# This script contains the helper functions for the Python portion of the pipeline.
+# It defines the ML model training and prediction logic as a pure function.
+# This script contains the helper functions for the Python portion of the pipeline.
+# It defines the ML model training and prediction logic as a pure function.
+
+# Step 1: Feature Engineering
+def prepare_features(simulated_df: pd.DataFrame) -> pd.DataFrame:
+    """Takes the raw simulated data and creates lagged features and the target variable."""
+    df = simulated_df.copy()
+
+    # Create lagged features
+    for col in ['output', 'consumption', 'investment', 'capital', 'technology']:
+        df[f'{col}_lag1'] = df[col].shift(1)
+
+    # Drop the first row which now has NaNs
+    df.dropna(inplace=True)
+
+    return df
+
+# Step 2: Data Splitting Functions
+def get_X_train(processed_df: pd.DataFrame):
+    """Gets the training features (X_train) from the processed data."""
+    features = [col for col in processed_df.columns if '_lag1' in col]
+    X = processed_df[features]
+    train_size = int(0.75 * len(X))
+    return X[:train_size]
+
+def get_y_train(processed_df: pd.DataFrame):
+    """Gets the training target (y_train) from the processed data."""
+    y = processed_df['output']
+    train_size = int(0.75 * len(y))
+    return y[:train_size]
+
+def get_X_test(processed_df: pd.DataFrame):
+    """Gets the testing features (X_test) from the processed data."""
+    features = [col for col in processed_df.columns if '_lag1' in col]
+    X = processed_df[features]
+    train_size = int(0.75 * len(X))
+    return X[train_size:]
+
+def get_y_test(processed_df: pd.DataFrame):
+    """Gets the testing target (y_test) from the processed data."""
+    y = processed_df['output']
+    train_size = int(0.75 * len(y))
+    return y[train_size:]
+
+# Step 3: Model Training
+def train_model(X_train: pd.DataFrame, y_train: pd.Series):
+    """Initializes and trains the XGBoost model."""
+    model = xgb.XGBRegressor(
+        objective='reg:squarederror',
+        n_estimators=100,
+        learning_rate=0.1,
+        max_depth=3,
+        random_state=42
+    )
+    model.fit(X_train, y_train)
+    return model
+
+# Step 4: Prediction
+def make_predictions(model, X_test: pd.DataFrame):
+    """Makes predictions on the test set using the trained model."""
+    return model.predict(X_test)
+
+# Step 5: Format Final Results
+def format_results(y_test: pd.Series, predictions) -> pd.DataFrame:
+    """Combines test data and predictions into a final DataFrame for R."""
+    results_df = pd.DataFrame({
+        'period': y_test.index,
+        'actual_output': y_test.values,
+        'predicted_output': predictions
+    })
+    return results_df
+
+# Encoder Function (for saving the final output)
+def save_arrow(df: pd.DataFrame, path: str):
+    """Encoder function to save a pandas DataFrame to an Arrow file."""
+    feather.write_feather(df, path)
+
+\end{Code}
+```
+
+### The Visualization (`functions/functions.R`)
+
+This R script uses [ggplot2]{.pkg} to create a plot comparing the model's output
+against the XGBoost predictions.
+
+```{=tex}
+\begin{CodeInput}
+# This script contains the helper functions for the R portion of the pipeline.
+# It defines the visualization logic using ggplot2.
+
+# 1. Load required R packages
+# This is only needed if you run this script by hand,
+# in which case, uncomment lines 8 and 9.
+# With rixpress, the packages get loaded automatically.
+#library(ggplot2)
+#library(dplyr)
+
+#-------------------------------------------------------------------------------
+# 2. Main Function: Create the Visualization
+#-------------------------------------------------------------------------------
+
+#' Create a plot comparing actual vs. predicted output
+#'
+#' This function takes a data frame with actual and predicted time series data
+#' and generates a ggplot visualization.
+#'
+#' @param predictions_df A data frame containing columns: 'period',
+#'   'actual_output', and 'predicted_output'. This data frame will be the
+#'   output of the Python XGBoost step.
+#' @return A ggplot object.
+#'
+plot_predictions <- function(predictions_df) {
+  # Create the plot object
+  p <- ggplot(predictions_df, aes(x = period)) +
+
+    # Add a line for the actual output from the RBC model simulation
+    geom_line(
+      aes(y = actual_output, color = "Actual (RBC Model)"),
+      linewidth = 1
+    ) +
+
+    # Add a dashed line for the XGBoost model's predictions
+    geom_line(
+      aes(y = predicted_output, color = "Predicted (XGBoost)"),
+      linetype = "dashed",
+      linewidth = 1
+    ) +
+
+    # Define custom colors and legend labels
+    scale_color_manual(
+      name = "Series",
+      values = c("Actual (RBC Model)" = "blue", "Predicted (XGBoost)" = "red")
+    ) +
+
+    # Add informative labels and a title
+    labs(
+      title = "XGBoost Prediction of RBC Model Output",
+      subtitle = "Forecasting next-quarter output based on current-quarter economic variables",
+      x = "Time (Quarters)",
+      y = "Output (Log-deviations from steady state)"
+    ) +
+
+    # Use a clean theme
+    theme_minimal() +
+
+    # Improve legend position
+    theme(legend.position = "bottom")
+
+  # Return the final ggplot object
+  return(p)
+}
+\end{CodeInput}
+```

### 2025-10-20 paper.qmd
citations

@@
 CRAN at a given date. For example, the code snippet below installs the
-[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:
+[purrr]{.pkg} [@purrr] and [ggplot2]{.pkg} [@ggplot2] packages as they were on
+April 4th, 2017:
 
@@
 polyglot pipeline that simulates a Real Business Cycle (RBC) model in
-[Julia]{.proglang}, uses the resulting data to train an [XGBoost]{.pkg}
-forecasting model in [Python]{.proglang}, visualizes the results in
-[R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
+[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
+[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
+in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
 report.
@@
 
-![Graphical representation of the polyglot pipeline. Green nodes are
-[Julia]{.proglang}, yellow nodes are [Python]{.pkg}, and blue nodes are
-[R]{.proglang} derivations.](dag-polyglot.png){#fig-dag}
+![Graphical representation of the polyglot pipeline. Purple nodes are
+[Julia]{.proglang}, yellow nodes are [Python]{.pkg}, light blue nodes are
+[R]{.proglang}, and dark blue nodes are Quarto derivations.](dag.png){#fig-dag}
 
@@
 This Julia script implements the state-space solution to the RBC model as a pure
-function. The economic theory is based on @depaoli2009.
+function. The economic theory is based on @depaoli2009rbc.
 

### 2025-10-20 paper.qmd
figures$
appendix

@@
 package can then generate a visual representation of the pipeline's directed
-acyclic graph (DAG), as shown in Figure 1.
+acyclic graph:
 
@@
 [R]{.proglang}, and dark blue nodes are Quarto derivations.](dag.png){#fig-dag}
+
+Figure @fig-dag can be generated by calling `rxp_ggdag()`.
 
@@
 
-### The Environment Definition (`gen-env.R`)
-
-This script uses [rix]{.pkg} to define a `default.nix` file that contains all
+### The Environment Definition
+
+This code snippet uses [rix]{.pkg} to generate a `default.nix` file that contains all
 necessary dependencies for [R]{.proglang}, [Julia]{.proglang}, and
-[Python]{.proglang}.
+[Python]{.proglang}. It is typically saved in a file called `gen-env.R`:
 
@@
 \begin{CodeInput}
-R> # This script defines the polyglot environment our pipeline will run in.
-R> library(rix)
-R>
-R> # Define the complete execution environment
-R> rix(
-+   # Pin the environment to a specific date for reproducibility.
-+   date = "2025-10-14",
-+
-+   # R Packages
-+   r_pkgs = c(
-+     "ggplot2", "dplyr", "arrow", "rix", "rixpress", "quarto"
-+   ),
-+ 
-+   # Julia Configuration
-+   jl_conf = list(
-+     jl_version = "lts",
-+     jl_pkgs = c("Distributions", "DataFrames", "Arrow", "Random")
-+   ),
-+ 
-+   # Python Configuration
-+   py_conf = list(
-+     py_version = "3.13",
-+     py_pkgs = c("pandas", "scikit-learn", "xgboost", "pyarrow")
-+   ),
-+ 
-+   project_path = ".",
-+   overwrite = TRUE
-+ )
+# This script defines the polyglot environment our pipeline will run in.
+library(rix)
+
+# Define the complete execution environment
+rix(
+ # Pin the environment to a specific date for reproducibility.
+ date = "2025-10-14",
+
+ # R Packages
+ r_pkgs = c(
+   "ggplot2", "dplyr", "arrow", "rix", "rixpress", "quarto"
+ ),
+
+ # Julia Configuration
+ jl_conf = list(
+   jl_version = "lts",
+   jl_pkgs = c("Distributions", "DataFrames", "Arrow", "Random")
+ ),
+
+ # Python Configuration
+ py_conf = list(
+   py_version = "3.13",
+   py_pkgs = c("pandas", "scikit-learn", "xgboost", "pyarrow")
+ ),
+
+ project_path = ".",
+ overwrite = TRUE
+)
 \end{CodeInput}
@@
 
-### The RBC Model Simulation (`functions/functions.jl`)
+### The RBC Model Simulation
 
 This Julia script implements the state-space solution to the RBC model as a pure
-function. The economic theory is based on @depaoli2009rbc.
+function, and is saved in a subfolder of the project named `functions` and in a script
+`functions.jl`. The economic theory is based on @depaoli2009rbc.
 
@@
 
-### The XGBoost Forecasting Model (`functions/functions.py`)
+### The XGBoost Forecasting Model
 
 This Python script contains a series of modular functions to handle feature
-engineering, data splitting, model training, and prediction.
+engineering, data splitting, model training, and prediction, and is saved in
+`functions/functions.py`.
 
@@
 
-### The Visualization (`functions/functions.R`)
-
-This R script uses [ggplot2]{.pkg} to create a plot comparing the model's output
-against the XGBoost predictions.
+### The Visualization
+
+This R snippet uses [ggplot2]{.pkg} to create a plot comparing the model's output
+against the XGBoost predictions and is saved in `functions/functions.R`.
 
@@
 ```
+
+### Running the project
+
+The structure of the project should look like this:
+
+```bash
+.
+├── functions
+│   ├── functions.jl
+│   ├── functions.py
+│   └── functions.R
+├── gen-env.R
+└── gen-pipeline.R
+
+2 directories, 5 files
+```
+
+Assuming we are working on a system that has Nix installed, we can "drop" into
+a temporary Nix shell with R and `{rix}` available:
+
+```r
+nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
+```
+
+Once the shell is ready, start R (by simply typing `R`) and run `source("gen-env.R")`
+to generate the adequate `default.nix` file. This is the environment that we can use
+to work interactively with the code while developing, and in which the pipeline will
+be executed. We can then leave the R session and the shell entirely (by typing `exit`)
+and type `nix-shell` to drop into the environment as defined by the `default.nix`.
+The pipeline can now be built using `source("gen-pipeline.R")`.

### 2025-10-21 paper.qmd
Update paper.qmd

@@
 ---
-title: "A Framework for Reproducible Data Science using Nix"
+title: "Nix for Polyglot, Reproducible Data Science Workflows"
 format:
@@
 abstract: |
-  To create a reproducible analysis, it is not enough to write clean,
-  well-documented and tested code. One must also ensure that all software
-  dependencies are clearly declared and, ideally, provide a simple way to
-  install them. Several tools exist for this purpose, such as the
-  containerization solution [Docker]{.pkg}. A complete solution, however,
-  requires not only managing the computational environment but also
-  orchestrating the execution of the analysis itself in a robust and automated
-  way. This paper presents a comprehensive framework based on the [Nix]{.pkg}
-  package manager to solve both problems. We first introduce the [R]{.proglang}
-  package [rix]{.pkg}, which simplifies the creation of reproducible
-  environments with [Nix]{.pkg}. We then introduce [rixpress]{.pkg}, a second
-  [R]{.proglang} package that leverages these environments to define and execute
-  complex, multi-language analytical pipelines. Together, these tools provide a
-  powerful, unified approach to achieving high standards of computational
-  reproducibility.
+  Reproducible analysis requires more than clean, well-documented code: 
+  researchers must also manage software dependencies, computational environments, and workflow execution. 
+  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, but no single solution handles multi-language environments, 
+  system dependencies, and pipeline orchestration together. I present a two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
+  environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments 
+  spanning [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system dependencies. Building on these environments, [rixpress]{.pkg} orchestrates polyglot pipelines where each 
+  computational step runs in its own hermetically sealed environment, with automatic caching and dependency tracking. 
+  This approach enables researchers to develop analyses interactively while maintaining bit-for-bit reproducibility, supports 
+  collaboration across heterogeneous computational environments, and ensures that analyses remain executable years into the future.
 
@@
 ---
+
+## Introduction: Reproducibility is also about software {#sec-intro}
 
@@
 loaded first? Which scripts should be executed, and in what order? Without clear
-documentation or a logically organized project structure, these operational
-details become yet another barrier to reproducibility.
+documentation or automated orchestration, these operational details become yet
+another barrier to reproducibility.
+
+This paper focuses specifically on two critical but often overlooked aspects of
+reproducibility: *computational environment management* and *workflow
+orchestration*. We present a comprehensive framework that addresses both
+challenges using the [Nix]{.pkg} package manager, making it accessible to
+researchers through two [R]{.proglang} packages: [rix]{.pkg} and
+[rixpress]{.pkg}. Before introducing our approach, we survey existing tools and
+their limitations to contextualize our contribution.
+
+### Existing approaches to environment management {#sec-existing}
 
@@
 version, platform details, and all loaded packages. Its output can be saved to a
-file and included as part of a study’s reproducibility record. Below is an
+file and included as part of a study's reproducibility record. Below is an
 example output from `sessionInfo()`:
@@
  LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
- LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
- LC_PAPER=en_US.UTF-8       LC_NAME=C                 
- LC_ADDRESS=C               LC_TELEPHONE=C            
- LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C
-
-time zone: Etc/UTC
-tzcode source: system (glibc)
+ ...
 
@@
 their versions) were used. However, reproducing the environment still requires
-manually installing the correct package versions. This can be difficult,
-especially when packages depend on system-level libraries. For example, the
-[nloptr]{.pkg} package requires a precompiled [nlopt]{.proglang} binary, or
-alternatively, [cmake]{.proglang} when building from source on Linux or macOS.
+manually installing the correct package versions—a process that becomes
+particularly difficult when packages depend on system-level libraries. For
+example, the [sf]{.pkg} package requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} system libraries.
+Installing and configuring these dependencies varies substantially across
+operating systems and can be prohibitively difficult on some platforms.
+
+#### Package-level solutions: renv and alternatives
 
 A more robust approach than simply listing package versions is to use the
-[renv]{.pkg} package. [renv]{.pkg} captures the project’s software state and
+[renv]{.pkg} package. [renv]{.pkg} captures the project's software state and
 writes it to a `renv.lock` file, which includes the exact versions of
@@
 for restoring the environment automatically, ensuring that others can recreate
-the same setup with minimal effort. Below is an example of an `renv.lock` file:
-
+the same setup with minimal effort. Below is an abbreviated example of an
+`renv.lock` file:
 ```{=tex}
@@
       "Hash": "539dc0c0c05636812f1080f473d2c177",
-      "Requirements": [
-        "lattice"
-      ]
+      "Requirements": ["lattice"]
     }
-
-    ... lines below omitted ...
+    ... additional packages omitted ...
   }
@@
 `renv::restore()`, which installs the exact package versions in an isolated,
-project-specific library that does not interfere with the user’s global R
+project-specific library that does not interfere with the user's global R
 library.
 
-However, [renv]{.pkg} does not restore the version of [R]{.proglang} itself.
-Installing the correct R version must therefore be done separately. Similarly,
-`renv` does not handle system-level dependencies such as [cmake]{.proglang},
-which is required by [nloptr]{.pkg} on Linux and macOS. These dependencies must
-be managed manually or through other tools. Further details are discussed in the
-[renv]{.pkg} documentation: \url{https://rstudio.github.io/renv/articles/renv.html#caveats}.
-
-Before continuing, it should be noted that other packages exist which provide
-similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg} by
-@simonsohn2023 which makes it rather easy to install packages as they were on
-CRAN at a given date. For example, the code snippet below installs the
-[purrr]{.pkg} [@purrr] and [ggplot2]{.pkg} [@ggplot2] packages as they were on
-April 4th, 2017:
-
+However, [renv]{.pkg} has important limitations. It does not restore the version
+of [R]{.proglang} itself—installing the correct R version must be done
+separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
+[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
+[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
+resolve this conflict. Users must manually install system libraries, and the
+process differs across operating systems. Further details are discussed in the
+[renv]{.pkg} documentation:
+\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.
+
+Other packages provide similar functionality with different trade-offs.
+[groundhog]{.pkg} by @simonsohn2023 makes it easy to install packages as they
+existed on CRAN at a given date:
 ```{=tex}
@@
 
-Packages installed with [groundhog]{.pkg} are placed in a project-specific
-library, ensuring they do not interfere with other versions of the same packages
-used in different projects. Since [groundhog]{.pkg} does not install
-[R]{.proglang} itself, users must either install the required R version manually
-or use the `tolerate.R.version` argument, as illustrated in the example above.
-Without this argument, [groundhog]{.pkg} will not proceed with package
-installation if the R version does not match the expected one.
-
-An alternative to [renv]{.pkg} is [rang]{.pkg}, developed by @chan2023, which
-similarly allows users to install packages as they existed on a specific date.
-Another approach is to use the Posit Package Manager, which provides dated
-snapshots of CRAN. For instance, to ensure that packages are installed as they
-were on June 30, 2023, one could include the following line in the `.Rprofile`
-file:
+[groundhog]{.pkg} places packages in project-specific libraries but does not
+install [R]{.proglang} itself. Without the `tolerate.R.version` argument, it
+will refuse to proceed if the R version does not match expectations.
+[rang]{.pkg}, developed by @chan2023, offers similar date-based package
+installation. The Posit Package Manager provides another approach through dated
+CRAN snapshots, configured via the `.Rprofile` file:
 
@@
 
-The `.Rprofile` file is read by [R]{.proglang} at the start of each new session,
-meaning that every call to `install.packages()` will install packages from the
-specified snapshot mirror. However, unless users explicitly manage separate
-libraries for different projects, the Posit Package Manager will install all
-packages for all projects as they existed on that snapshot date.
-
-The next step toward the gold standard of reproducibility is ensuring not only
-that the correct package versions are installed, but also that the correct
-version of [R]{.proglang} itself is used. While this can be done manually,
-dedicated tools make the process more efficient. One such tool is [rig]{.pkg},
-developed by the [R]{.proglang} Infrastructure Team [-@rlib2023]. [rig]{.pkg}
-simplifies installing and managing multiple R versions, allowing users to match
-the exact version needed for a given analysis. Once the correct R version is
-installed, one can then use tools such as [renv]{.pkg}, [groundhog]{.pkg}, or
-[rang]{.pkg} to recreate the associated package library.
-
-However, this multi-step process remains manual, error-prone, and
-time-consuming, highlighting the need for more comprehensive and automated
-solutions to achieve full computational reproducibility.
-
-The final tool for achieving the gold standard of reproducibility is to bundle
-the exact version of [R]{.proglang} and its packages within a [Docker]{.pkg}
-image. [Docker]{.pkg} is a containerization platform that allows packaging a
-*data product* together with all its dependencies into a self-contained image. A
-statistical analysis can be viewed as such a data product, requiring a specific
-set of software dependencies to run. *Dockerizing* an analysis involves building
-an image that installs these dependencies, often using the tools mentioned
-earlier, and includes the analysis scripts and data (but data can also be
-dynamically made available at run-time, to avoid having to copy it inside the
-image). The steps to create this image are specified in a simple text file
-called a `Dockerfile`, which defines the structure and content of the
-[Docker]{.pkg} image.
+This approach installs all packages from the specified snapshot date, but unless
+users explicitly manage separate libraries for different projects, all projects
+will share the same package versions.
+
+These tools represent significant progress in reproducibility, but they share a
+common limitation: **none addresses system-level dependencies or supports
+multi-language environments**. A project using [R]{.proglang},
+[Python]{.proglang}, and system tools like [Quarto]{.pkg} or \LaTeX requires
+coordinating multiple package managers ([renv]{.pkg} for R, virtual environments
+for Python, manual installation for system tools), each with its own
+configuration and potential for conflicts.
+
+#### System-level solutions: Docker
+
+The most comprehensive approach to reproducibility is containerization with
+[Docker]{.pkg}. [Docker]{.pkg} packages a *data product* together with all its
+dependencies—including the operating system, system libraries, programming
+languages, and packages—into a self-contained image. A statistical analysis can
+be viewed as such a data product. The steps to create an image are specified in
+a `Dockerfile`, which defines the structure and content of the [Docker]{.pkg}
+image.
 
@@
 
-The strength of [Docker]{.pkg} lies not only in its ability to encapsulate the
-correct versions of [R]{.proglang} and its packages but also in its inclusion of
-system-level dependencies. Since [Docker]{.pkg} images are nearly complete Linux
-systems, they naturally bundle libraries and tools required by certain R
-packages (for example, [nloptr]{.pkg}’s dependency on the [nlopt]{.proglang}
-library). This ensures that future replicators of a study have access to the
-same software environment, including all necessary system components. Moreover,
-these images can be easily shared, enabling seamless reproducibility across
+The strength of [Docker]{.pkg} lies in its ability to bundle not only the
+correct versions of [R]{.proglang} and its packages but also system-level
+dependencies. Since [Docker]{.pkg} images are effectively minimal Linux systems,
+they naturally include libraries required by packages like [sf]{.pkg}. This
+ensures that future replicators have access to the same software environment,
+including all necessary system components. Moreover, these images can be easily
+shared via registries like Docker Hub, enabling seamless reproducibility across
 machines and collaborators.
@@
 
-Despite its strengths, [Docker]{.pkg} can be cumbersome for interactive use.
-Although containers can be modified at runtime, any changes are lost once the
-container stops unless they are explicitly saved. Running graphical applications
-from containers is also possible, but it is generally difficult to configure and
-tends to work reliably only on Linux systems. For this reason, [Docker]{.pkg} is
-most often used to run web-based applications (such as integrated development
-environments, IDEs, like the web version of RStudio) inside containers, offering
-a reproducible yet flexible environment for interactive work.
-
-Some IDEs provide built-in support for executing code directly within a running
-container, allowing a more seamless workflow. However, these setups typically
-require additional configuration, and many general-purpose IDEs are not well
-optimized for statistical or data science workflows (though Positron may soon
-change that).
-
-A common practice is to conduct a study interactively using a standard
-installation of [R]{.proglang} and its packages, and only after completing the
-analysis, create a `Dockerfile` to allow others to reproduce the results easily.
-For instance, the [dockerfiler]{.pkg} package [@dockerfiler] can automatically
-generate a `Dockerfile` from an existing `renv.lock` file, simplifying this
-process. While this approach enables post-hoc reproducibility, it does not solve
-the challenge of consistently deploying the same environment across multiple
-machines as co-authors are working on the project. Ideally, analyses would be
-developed directly within a containerized environment from the start, ensuring
-that the computational setup is reproducible throughout the project lifecycle.
-
-Another challenge of using [Docker]{.pkg} is that its images are effectively
-minimal Linux systems, so familiarity with Linux is highly recommended for
-writing reliable and efficient `Dockerfile`s. To guarantee that a build
-consistently produces the same image, the `Dockerfile` should reference specific
-image *digests* rather than general *tags*. In practice, however, tagged
-versions are more commonly used, which can introduce potential reproducibility
-issues. [dockerfiler]{.pkg} is an R package that lowers the
-barrier to entry by automating parts of the `Dockerfile` creation process, but
-they cannot fully replace the need for basic Linux knowledge.
-
-A final step toward the gold standard of reproducibility is to automate the
-analysis execution using a build system such as [Make]{.pkg}, rather than
-relying on ad hoc script execution. Build automation tools enable analyses to be
-defined as a series of reproducible, well-ordered steps that can be executed
-consistently. Within [R]{.proglang}, a modern and powerful alternative is the
-[targets]{.pkg} package by @landau2021, which provides declarative workflows for
-reproducible data analysis.
-
-@mcdermott2021 provides a clear example of a scientific study that achieved the
-gold standard of reproducibility. The author created an accompanying GitHub
-repository^[https://github.com/grantmcdermott/skeptic-priors] containing
-detailed instructions to install the required software and execute the analysis.
-Examining this repository reveals the use of multiple tools to capture and share
-the computational environment:
-
-- Packages and their versions were recorded in an `renv.lock` file;
-- A `Makefile` was used to automate the full analysis and compile the paper;
-- A `Dockerfile` provided a complete computational environment, including the
-  correct version of [R]{.proglang}, enabling easy execution of the entire
-  workflow.
-
-Reaching this gold standard, however, is resource-intensive. It requires
-learning a tool for managing package versions, mastering [Docker]{.pkg} for the
-programming language and other software dependencies, and using a build
-automation tool. Importantly, these considerations are not unique to
-[R]{.proglang}: similar approaches are needed for [Python]{.proglang} or any
-other programming language, and complexity further increases if several
-programming languages are needed for a single project.
-
-As an alternative, we present the [Nix]{.pkg} package manager, which supports
-all major operating systems and emphasizes reproducible software installation
-and building. [Nix]{.pkg} can replace [Docker]{.pkg}, [renv]{.pkg}, and even
-build automation tools such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
-for [R]{.proglang} users, we developed the [rix]{.pkg} and [rixpress]{.pkgs}
-packages, which are introduced and discussed in this article.
+Despite its strengths, [Docker]{.pkg} presents several challenges for
+statistical computing workflows:
+
+1. Poor interactive support: While containers can be modified at runtime,
+changes are lost when the container stops unless explicitly saved. Running
+graphical applications like RStudio Desktop from containers requires complex X11
+configuration that works reliably only on Linux. Instead, web-based IDEs (like RStudio
+Server) are used instead, but come with the inherent limitations of web-based solutions.
+
+2. Steep learning curve: [Docker]{.pkg} images are minimal Linux systems, so
+writing reliable `Dockerfile`s requires familiarity with Linux system
+administration. To ensure reproducibility, `Dockerfile`s should reference
+specific image *digests* rather than *tags*, but this practice is rarely
+followed. Packages like [dockerfiler]{.pkg} [@dockerfiler] help automate
+`Dockerfile` creation but cannot eliminate the need for basic Linux knowledge.
+
+3. Post-hoc reproducibility: A common workflow is to develop analyses
+interactively using standard R installations, then create a `Dockerfile`
+afterward to enable reproduction. While this achieves post-hoc reproducibility,
+it fails to address the challenge of deploying consistent environments across
+multiple machines *during* development, a critical need for collaborative
+projects.
+
+4. Single-language focus: While [Docker]{.pkg} can support multi-language
+environments, orchestrating them requires manual coordination. There is no
+standard way to specify that one step should run in a Python environment while
+another runs in R.
+
+### The workflow orchestration gap
+
+Even with a perfectly reproducible environment, researchers face another
+challenge: reliably executing the analysis workflow. Which scripts run first?
+What are the dependencies between steps? Manual execution is error-prone and
+poorly documented. Build automation tools like [Make]{.pkg} address this by
+defining analyses as ordered, reproducible steps.
+
+Within [R]{.proglang}, the [targets]{.pkg} package by @landau2021 provides a
+modern, declarative approach to workflow management. [targets]{.pkg} tracks
+dependencies between code and data, caches intermediate results, and only
+recomputes steps affected by changes. This dramatically improves efficiency for
+complex analyses.
+
+However, [targets]{.pkg} operates within a single R session. While it can call
+Python via [reticulate]{.pkg} or execute system commands, all steps share the
+same computational environment. For truly polyglot pipelines (where different
+steps may require incompatible dependencies or different language versions) this
+limitation becomes problematic. One might run [targets]{.pkg} inside a
+[Docker]{.pkg} container to ensure reproducibility, but this requires the entire
+pipeline to use a single environment.
+
+@mcdermott2021 provides an excellent example of achieving the gold standard of
+reproducibility. The accompanying GitHub
+repository^[https://github.com/grantmcdermott/skeptic-priors] demonstrates the
+state of the art:
+
+- Package versions recorded in an `renv.lock` file
+- A `Makefile` automating the full analysis
+- A `Dockerfile` providing a complete computational environment
+
+However, achieving this required mastering multiple complex tools: [renv]{.pkg}
+for package management, [Docker]{.pkg} for environment containerization, and
+[Make]{.pkg} for workflow orchestration. Each tool has its own syntax, concepts,
+and failure modes. The complexity compounds for multi-language projects—a Python
+project would require learning virtual environments or conda, and coordinating R
+and Python in a single analysis requires additional tooling.
+
+### A unified approach with Nix
+
+What researchers need is a tool that:
+
+1. Manages complete environments including programming languages, packages,
+and system dependencies;
+2. Supports multiple languages natively, not through workarounds;
+3. Works interactively without the complexity of containers;
+4. Provides step-level isolation so different pipeline steps can use
+different environments;
+5. Ensures bit-for-bit reproducibility through deterministic builds;
+6. Remains simple enough for researchers without systems administration
+expertise.
+
+The [Nix]{.pkg} package manager provides these capabilities. [Nix]{.pkg} ensures
+reproducible software installation by deploying *component closures*—packages
+bundled with all their dependencies and transitive dependencies [@dolstra2004nix].
+This creates self-contained software environments that work identically across
+machines and over time. [Nix]{.pkg} can replace [Docker]{.pkg} for environment
+isolation, [renv]{.pkg} for package management, and even [Make]{.pkg} for
+workflow orchestration—all within a single, unified framework.
+
+However, [Nix]{.pkg} has a steep learning curve. It is a complex system with its
+own purely functional programming language designed to declaratively define how
+software is built and configured. While this functional approach ensures
+reproducibility, it creates a significant barrier to adoption.
+
+To make [Nix]{.pkg} accessible to researchers, I developed two [R]{.proglang}
+packages:
+
+- [rix]{.pkg} generates [Nix]{.pkg} expressions from intuitive R function
+calls, eliminating the need to learn the [Nix]{.proglang} language. It handles
+environment definition including [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, system tools, and their
+dependencies.
+
+- **[rixpress]{.pkg}** orchestrates polyglot analytical pipelines using
+[Nix]{.pkg} as the build engine. Each pipeline step runs in its own
+hermetically sealed environment with automatic caching and dependency tracking.
+This enables true polyglot workflows where [R]{.proglang}, [Python]{.proglang} and [Julia]{.proglang} steps coexist
+with different dependencies.
+
+Together, these packages provide what existing tools cannot: a single framework
+for managing reproducible, multi-language computational environments and
+executing complex analytical workflows with step-level isolation. While
+[Nix]{.pkg} remains complex for advanced use cases, [rix]{.pkg} and
+[rixpress]{.pkg} abstract this complexity for common research workflows, making
+deep reproducibility accessible without requiring systems administration
+expertise.
+
+The remainder of this paper proceeds as follows. Section 2 introduces the
+[Nix]{.pkg} package manager and explains why its functional approach enables
+reproducibility. Section 3 presents [rix]{.pkg} and demonstrates environment
+definition for various use cases. Section 4 discusses our `rstats-on-nix` fork
+of the official package repository and how it addresses practical limitations.
+Section 5 introduces [rixpress]{.pkg} and demonstrates polyglot pipeline
+orchestration through a complete example. Section 6 concludes with discussion of
+appropriate use cases and future directions.
 

### 2025-10-21 paper.qmd
Update paper.qmd
Revise Nix package manager section for clarity and detail
Enhanced the section on the Nix package manager to provide a more detailed explanation of its features, advantages, and cross-platform support. Improved clarity on dependency management, reproducibility, and the challenges of using Nix.

@@
 ---
-
-## Introduction: Reproducibility is also about software {#sec-intro}
 
@@
 
-[Nix]{.pkg} is a package manager designed to install and build software in a
-fully reproducible manner. As of this writing, it provides access to over
-120000 packages, including nearly all of CRAN and Bioconductor. This allows
-users to install not only [R]{.proglang} itself but also all packages required
-for a given project. Although [Nix]{.pkg} is the package manager of the NixOS
-Linux distribution, it can also be installed as a standalone tool on other Linux
-distributions, macOS, and the Windows Subsystem for Linux^[Effectively treating
-non-NixOS Linux distributions and Windows as equivalent platforms in practice.].
-
-The advantage of using [Nix]{.pkg} to install [R]{.proglang} packages, instead
-of the standard `install.packages()` function, is that [Nix]{.pkg} ensures all
-dependencies of each package are installed, whether they are other
-[R]{.proglang} packages or system-level libraries.
-
-For example, the [xlsx]{.pkg} package requires [Java]{.proglang} to be
-installed. On some systems, installing [Java]{.proglang} may be difficult or
-even impossible manually. With [Nix]{.pkg}, however, the user only needs to
-declare that [xlsx]{.pkg} is required for the project; [Nix]{.pkg} automatically
-installs and configures [Java]{.proglang} as a dependency. This works because
-the maintainers of the [R]{.proglang} language packages for [Nix]{.pkg} have
-declared [Java]{.proglang} as a dependency of [xlsx]{.pkg}, allowing the process
-to happen seamlessly from the end-user perspective.
-
-[xlsx]{.pkg} can thus be referred to as a component closure, and quoting
-@dolstra2004nix:
+[Nix]{.pkg} is a powerful, cross-platform package manager designed for installing 
+and building software in a fully reproducible and reliable manner. Unlike traditional 
+package managers, [Nix]{.pkg} emphasizes immutable infrastructure and a functional
+approach, which significantly enhances the consistency of computational environments.
+As of writing, the primary Nix package collection, `nixpkgs`, provides access to over 
+120,000 packages, including nearly all of CRAN and Bioconductor. 
+This extensive coverage allows users to install not only [R]{.proglang} itself but 
+also all its required packages and system-level dependencies for any given project.
+
+While [Nix]{.pkg} is the default package manager for the NixOS 
+Linux distribution, it can be installed as a standalone tool on other Linux 
+distributions, macOS, and effectively on Windows via the Windows Subsystem for 
+Linux (WSL2), treating these as equivalent platforms in practice.
+
+A key advantage of using [Nix]{.pkg} to install [R]{.proglang} packages, as 
+opposed to the standard `install.packages()` function, lies in its comprehensive 
+dependency management. [Nix]{.pkg} ensures that all dependencies of each package are
+installed, irrespective of whether they are other [R]{.proglang} packages or 
+underlying system libraries.
+
+Consider, for example, the [sf]{.pkg} package, which is fundamental for spatial data analysis in [R]{.proglang}.
+[sf]{.pkg} relies on several complex external system libraries such as [GDAL]{.proglang}, 
+[GEOS]{.proglang}, and [PROJ]{.proglang}. Manually installing and configuring these libraries 
+to be compatible with [sf]{.pkg} can be challenging and platform-specific. With [Nix]{.pkg}, 
+however, the user merely needs to declare [sf]{.pkg} as a project requirement. 
+[Nix]{.pkg} then automatically installs and configures all necessary system libraries
+as transitive dependencies. This seamless process is possible because the maintainers of the 
+[R]{.proglang} programming language for [Nix]{.pkg}
+have declaratively specified these dependencies for [sf]{.pkg} in `nixpkgs`, allowing the entire software 
+stack to be provisioned effortlessly from an end-user perspective.
+
+This concept is central to [Nix]{.pkg}'s design, where packages are referred to as *component closures*. 
+As @dolstra2004nix explains:
 
@@
 
-But how does [Nix]{.pkg} achieve this reproducibility? Where do these packages,
-or *closures*, come from? When installing a package with [Nix]{.pkg}, an
-expression written in the [Nix]{.pkg} language is downloaded from the `nixpkgs`
-GitHub repository and evaluated. These expressions define *derivations*, which
-describe how to build a package: specifying its dependencies, the commands to
-build and install it, and the resulting output. Typically, a derivation
-downloads the source code, compiles it, and outputs a binary. Derivations are
-highly flexible; by writing custom derivations, users can define and build
-reproducible environments instead of a single program. The dependencies of each
-derivation are themselves defined in other expressions and are automatically
-installed if needed. In the case of [R]{.proglang}, maintainers declare the
-appropriate dependencies to create fully self-contained component closures.
-
-Why is installing software with [Nix]{.pkg} reproducible? Because the entire set
-of [Nix]{.pkg} expressions is hosted on GitHub, it is possible to reference a
-specific commit of `nixpkgs`, a process called *pinning a revision*. Pinning a
-revision ensures that all packages installed by [Nix]{.pkg} are always the exact
-same versions, regardless of when in the future they are built, since the
-expressions downloaded correspond to that specific commit of `nixpkgs`.
-
-Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports
-reproducibility. [Nix]{.pkg} is a functional package manager: it applies
-principles from functional programming, disallowing side effects and global
-variables, and ensuring that the output of a build is always the same given the
-same inputs, regardless of time or location. A side note: while this functional
-approach greatly enhances reproducibility, it can make writing derivations more
-complex, particularly for packages that need to download assets during
-installation (e.g., some Bioconductor packages like [musData]{.pkg}). However,
-this concern is primarily for package maintainers, not end-users.
-
-Additionally, [Nix]{.pkg} supports multiple versions, or *variants*, of a
-package on the same system. Each variant has a unique identifier, allowing
-multiple versions of [R]{.proglang} to coexist and ensuring the correct version
-is used for each project. For a more technical discussion of [Nix]{.pkg}, we
-refer to @dolstra2004nix.
-
-With [Nix]{.pkg}, it is possible to effectively replace both [renv]{.pkg} and
-[Docker]{.pkg} for [R]{.proglang} projects—or, in the case of
-[Python]{.proglang}, to replace `requirements.txt` files and virtual
-environments. [Nix]{.pkg} also supports building multi-language environments,
-which can include [R]{.proglang} and [Python]{.proglang}, a \LaTeX
-distribution or any other of the 120,000 available tools. This enables the
-creation of a complete, project-specific, and reproducible environment that can
-be used interactively or non-interactively. As long as the `nixpkgs` GitHub
-repository remains online, the environment can be rebuilt in the future to rerun
-analyses reliably.
-
-However, [Nix]{.pkg} has a steep learning curve. It is a complex system that
-comes with its own programming language, also called [Nix]{.proglang}, designed
-to solve the challenging problem of declaratively defining how software is built
-and configured. This ensures that installations are fully reproducible across
-operating systems and hardware platforms. To make [Nix]{.pkg} more accessible to
-[R]{.proglang} users that wish to use project-specific and reproducible
-development environments, we developed the [rix]{.pkg} package.
+The core of [Nix]{.pkg}'s reproducibility lies in its functional package management paradigm
+and its unique approach to builds. When installing software with [Nix]{.pkg}, it evaluates
+an *expression* written in the [Nix]{.proglang} language. These expressions define *derivations*,
+which are declarative blueprints describing how to build a package. A derivation specifies all inputs: 
+source code, build commands, and crucially, all direct and transitive dependencies. This declarative 
+nature ensures that every build process is fully specified and independent of the environment where 
+it's executed. Furthermore, all [Nix]{.pkg} expressions for official packages are hosted in the `nixpkgs`
+GitHub repository. By referencing a specific commit of `nixpkgs` (a process known as *pinning a revision*) users 
+guarantee that all packages installed are derived from the exact same set of build instructions.
+This means that if an environment is built today using a pinned revision, it will produce the 
+identical set of software versions and configurations if rebuilt tomorrow, next year, or on a 
+different machine, as long as the `nixpkgs` repository remains accessible.
+
+Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports reproducibility.
+[Nix]{.pkg} is a *functional* package manager in the programming sense: it treats software builds as pure functions.
+Given the same inputs (source code, dependencies, build scripts), a [Nix]{.pkg} build will always produce the 
+identical output, regardless of when or where it's executed. This is achieved by disallowing side effects and 
+ensuring that builds operate within hermetic (isolated) environments, with no hidden dependencies on global system 
+state. While this functional approach greatly enhances reproducibility, it can sometimes introduce complexity for
+package maintainers, especially for software that needs to download external assets during installation (e.g., 
+certain Bioconductor packages). For end-users, however, this complexity is largely abstracted away. Additionally, 
+[Nix]{.pkg} supports multiple versions or *variants* of a package on the same system. Each package installed into
+the "Nix store" (typically in `/nix/store`) has a unique identifier derived from a hash of all its inputs. This 
+isolation ensures that projects always use their intended software versions, allowing multiple versions of
+[R]{.proglang} to coexist and preventing "dependency hell" and global environment pollution.
+
+For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, we refer readers to @dolstra2004nix.
+
+By leveraging these principles, [Nix]{.pkg} can effectively replace and integrate the functionalities of 
+tools like [renv]{.pkg} and [Docker]{.pkg} for [R]{.proglang} projects, or `requirements.txt` files 
+and virtual environments for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
+multi-language environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
+a [LaTeX]{.proglang} distribution, or any of the numerous other tools available in `nixpkgs`. 
+This enables the creation of a truly complete, project-specific, and deeply reproducible environment
+that can be used interactively for development or non-interactively for automated analysis. 
+As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
+the environment can be rebuilt reliably in the future.
+
+[Nix]{.pkg} provides strong cross-platform support across major operating systems. It runs natively on Linux 
+(x86_64) with full functionality and optimal integration. On macOS (Intel and Apple Silicon), however, 
+reproducibility is less reliable due to dependencies on impure system frameworks and Xcode toolchains outside
+the Nix store. As a result, macOS users may need to update project pins after system or Xcode upgrades to 
+restore builds. On Windows, [Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though graphical
+applications may need extra setup. Nix environments on WSL can be used interactively via VS Code or Positron 
+for a smooth development experience.
+
+While [Nix]{.pkg} is a powerful system, it comes with notable challenges. The steepest is its learning curve:
+[Nix]{.pkg} uses its own declarative language (also called [Nix]{.proglang}), which can be difficult for 
+those new to functional or declarative paradigms. Complexity becomes most apparent when writing custom 
+derivations or troubleshooting intricate build issues. Build times can also be long, especially for large
+environments, when pre-built binaries aren’t available. Additionally, because [Nix]{.pkg} stores all package
+outputs immutably in the Nix store, disk usage tends to be higher than with traditional package managers. 
+Finally, while `nixpkgs` is extensive, some new or niche packages may be missing or fail to build across platforms.
+
+To make these capabilities more approachable for [R]{.proglang} users seeking reproducible, 
+project-specific environments without learning the full [Nix]{.proglang} language, I created the 
+[rix]{.pkg} and [rixpress]{.pkg} packages.
 

### 2025-10-21 paper.qmd
Update paper.qmd

@@
 As of writing, the primary Nix package collection, `nixpkgs`, provides access to over 
-120,000 packages, including nearly all of CRAN and Bioconductor. 
+120,000 packages, including nearly all of CRAN and Bioconductor, making it the largest,
+and among the most up-to-date package repository according to Repology:
+
+![nixpkgs is the largest and among the most up-to-date packages repository](repology.png){#fig-repology}
+
 This extensive coverage allows users to install not only [R]{.proglang} itself but 

### 2025-10-21 paper.qmd
Edit paper.qmd for clarity and consistency
Revised text for clarity and consistency throughout the paper, including updates to the narrative voice and the presentation of tools and frameworks.

@@
 reproducibility: *computational environment management* and *workflow
-orchestration*. We present a comprehensive framework that addresses both
+orchestration*. I present a comprehensive framework that addresses both
 challenges using the [Nix]{.pkg} package manager, making it accessible to
 researchers through two [R]{.proglang} packages: [rix]{.pkg} and
-[rixpress]{.pkg}. Before introducing our approach, we survey existing tools and
+[rixpress]{.pkg}. Before introducing our approach, I survey existing tools and
 their limitations to contextualize our contribution.
-
-### Existing approaches to environment management {#sec-existing}
 
@@
 operating systems and can be prohibitively difficult on some platforms.
-
-#### Package-level solutions: renv and alternatives
 
@@
 
-#### System-level solutions: Docker
-
 The most comprehensive approach to reproducibility is containerization with
@@
 
-### The workflow orchestration gap
-
 Even with a perfectly reproducible environment, researchers face another
@@
 and Python in a single analysis requires additional tooling.
-
-### A unified approach with Nix
 
@@
 restore builds. On Windows, [Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though graphical
-applications may need extra setup. Nix environments on WSL can be used interactively via VS Code or Positron 
-for a smooth development experience.
+applications may need extra setup. [Nix]{.pkg} environments on WSL can be used interactively via VS Code or Positron 
+for a smooth development experience. It is also important to note that [Nix]{.pkg} can be installed
+within a [Docker]{.pkg} image to deploy and reproduce environments reliably in containerized workflows.
 
@@
 
-This call to `rix()` generates an environment that installs several
-[R]{.proglang} packages (including [AER]{.pkg} version 1.2-8), several TeXLive
+This call to `rix()` generates a `default.nix` that builds a development shell
+which ships [R]{.proglang} and [R]{.proglang} packages (including [AER]{.pkg} version 1.2-8), several TeXLive
 packages for \LaTeX document authoring, development versions of [rix]{.pkg} and
@@
 is impractical. While CRAN and Bioconductor update daily, the [R]{.proglang}
-packages in `nixpkgs` are only updated with new R releases. This limitation is
+packages in `nixpkgs` are only updated with new [R]{.proglang} releases. This limitation is
 due to [Nix]{.pkg}’s governance as a Linux distribution package manager.
 
-Our fork allows us to circumvent these limitations. For example, we provide a
-daily snapshot of CRAN. Each day, the [R]{.proglang} package set is updated and
+The `rstats-on-nix` fork allows us to circumvent these limitations. For example, it provides
+daily snapshots of CRAN. Each day, the [R]{.proglang} package set is updated and
 committed to a dated branch using GitHub Actions. Users can select a specific
@@
 
-We strive to provide an available date per week: each Monday, a GitHub Action
+I strive to provide an available date per week: each Monday, a GitHub Action
 tests popular packages on Linux and macOS, and only if all tests succeed is the
@@
 
-We have backported fixes to our `nixpkgs` fork as far back as March 2019. The
+I have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as March 2019. The
 process involves checking out a `nixpkgs` commit on the selected date, updating
@@
 installations, as prebuilt packages are downloaded rather than compiled.
+
+The fork also allows me to catch issues (such as packages’ builds breaking) early
+on, and prepare fixes that can then be contributed upstream.
 
@@
 The `rxp_populate()` function translates this [R]{.proglang} list into a
-`pipeline.nix` file, which declaratively defines the entire workflow. The
-package can then generate a visual representation of the pipeline's directed
+`pipeline.nix` file, which declaratively defines the entire workflow.
+Data flows from derivation to derivation by being serialized into the 
+efficient and language-agnostic  Arrow format using the pair of `encoder/decoder` 
+functions. Other universal formats, such as `csv` or `json` could have been used.
+
+The package can then generate a visual representation of the pipeline's directed
 acyclic graph:
@@
 projects with confidence that results can be reproduced exactly.
+
+While [rix]{.pkg} and [rixpress]{.pkg} provide significant steps forward, it is 
+important to acknowledge current limitations. Both packages are relatively recent. 
+[rixpress]{.pkg}, in particular, requires further development, especially to enhance
+the ease of debugging complex pipelines. Its visualization capabilities are also still
+fairly limited; for instance, they do not currently indicate outdated derivations when
+a user modifies code that would necessitate a rebuild. These areas represent important 
+avenues for future work to further mature the framework and improve the user experience.
 

### 2025-10-21 run_polyglot_example.sh
Create run_polyglot_example.sh

@@ -1,0 +1,404 @@
+#!/usr/bin/env bash
+#
+# This script sets up and runs a reproducible, polyglot data science pipeline.
+# It requires Nix to be installed on your system (macOS or Linux).
+# For installation instructions, visit: https://nixos.org/download.html
+
+set -e # Exit immediately if a command exits with a non-zero status.
+
+if ! command -v nix-shell &> /dev/null; then
+    echo -e "${RED}Error: Nix is not installed or not in your PATH.${NC}"
+    echo "This script requires Nix to create a reproducible environment."
+    echo "Please install it using the recommended Determinate Systems installer:"
+    echo -e "${BLUE}https://determinate.systems/posts/determinate-nix-installer${NC}"
+    exit 1
+fi
+
+# --- Style and Color Definitions ---
+GREEN='\033[0;32m'
+YELLOW='\033[1;33m'
+RED='\033[0;31m'
+BLUE='\033[0;34m'
+NC='\033[0m' # No Color
+
+# --- Helper function to run commands with a spinner ---
+run_with_spinner() {
+    local cmd="$1"
+    local message="$2"
+    
+    # Run the command in the background, redirecting all output
+    eval "$cmd" > /tmp/spinner.log 2>&1 &
+    local pid=$!
+
+    # Animate spinner
+    local spin='-\|/'
+    local i=0
+    while kill -0 $pid 2>/dev/null; do
+        i=$(( (i+1) %4 ))
+        printf "\r${YELLOW}%s${NC} %s" "${spin:$i:1}" "$message"
+        sleep .1
+    done
+
+    # Wait for the command to finish and get its exit code
+    wait $pid
+    local exit_code=$?
+
+    # Check the exit code and print final status
+    if [ $exit_code -eq 0 ]; then
+        printf "\r${GREEN}✓${NC} %s\n" "$message"
+    else
+        printf "\r${RED}✗${NC} %s\n" "$message"
+        echo -e "${RED}An error occurred. Please check the log file for details: /tmp/spinner.log${NC}"
+        cat /tmp/spinner.log
+        exit 1
+    fi
+}
+
+# --- Main Script Logic ---
+
+# 1. Check if the directory is empty (or contains only this script)
+script_name=$(basename "$0")
+# List all items in the directory, but exclude the script itself.
[…]

### 2025-10-21 run_polyglot_example.sh
Update run_polyglot_example.sh

@@ -79,7 +79,7 @@ echo -e "${GREEN}✓${NC}"
 printf "Creating Julia helper script (functions.jl)... "
 cat > functions/functions.jl << 'EOF'
 # This script contains the helper functions for the Julia portion of the pipeline.
-using LinearAlgebra, Distributions, DataFrames, Arrow, Random
+#using LinearAlgebra, Distributions, DataFrames, Arrow, Random
 
 function simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
 
@@ -145,10 +145,10 @@ echo -e "${GREEN}✓${NC}"
 printf "Creating Python helper script (functions.py)... "
 cat > functions/functions.py << 'EOF'
 # This script contains modular helper functions for the Python portion of the pipeline.
-import pandas as pd
-import pyarrow.feather as feather
-from sklearn.model_selection import train_test_split
-import xgboost as xgb
+#import pandas as pd
+#import pyarrow.feather as feather
+#from sklearn.model_selection import train_test_split
+#import xgboost as xgb
 
 def prepare_features(simulated_df: pd.DataFrame) -> pd.DataFrame:
     df = simulated_df.copy()
@@ -210,8 +210,8 @@ printf "Creating R helper script (functions.R)... "
 cat > functions/functions.R << 'EOF'
 # This script contains the helper functions for the R portion of the pipeline.
 # It defines the visualization logic using ggplot2.
-library(ggplot2)
-library(dplyr)
+#library(ggplot2)
+#library(dplyr)
 
 plot_predictions <- function(predictions_df) {
   p <- ggplot(predictions_df, aes(x = period)) +

### 2025-10-22 paper.qmd
Update paper.qmd

@@
 process differs across operating systems. Further details are discussed in the
-[renv]{.pkg} documentation:
-\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.
+[renv]{.pkg} 
+documentation^[\url{https://rstudio.github.io/renv/articles/renv.html#caveats}].
 
@@
 calls, eliminating the need to learn the [Nix]{.proglang} language. It handles
-environment definition including [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, system tools, and their
-dependencies.
+environment definition including [R]{.proglang}, [Python]{.proglang}, 
+[Julia]{.proglang}, system tools, and their dependencies.
 
@@
 hermetically sealed environment with automatic caching and dependency tracking.
-This enables true polyglot workflows where [R]{.proglang}, [Python]{.proglang} and [Julia]{.proglang} steps coexist
-with different dependencies.
+This enables true polyglot workflows where [R]{.proglang}, [Python]{.proglang} and 
+[Julia]{.proglang} steps coexist with different dependencies.
 
@@
 $> nix-shell -I \
-  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/refs/heads/2025-10-17.tar.gz -p \
-  R rPackages.rix
++  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/2025-10-20.tar.gz -p \
++  R rPackages.rix
 \end{CodeInput}
@@
 
-[rix]{.pkg} can also handle more complex setups:
-
-```{=tex}
-\begin{CodeInput}
-R> rix(r_ver = "4.3.1",
-+    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
+[rix]{.pkg} can also handle more complex setups, and users can provide a date
+instead of a specific [R]{.proglang} version:
+
+```{=tex}
+\begin{CodeInput}
+R> rix(date = "2025-10-20",
++    r_pkgs = c("rix", "dplyr", "chronicler", "AER@1.2-8"),
 +    system_pkgs = c("quarto", "git"),
@@
 +    git_pkgs = list(
-+      list(
-+        package_name = "rix",
-+        repo_url = "https://github.com/b-rodrigues/rix/",
-+        branch_name = "master",
-+        commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
-+      list(
 +        package_name = "fusen",
 +        repo_url = "https://github.com/ThinkR-open/fusen",
-+        branch_name = "main",
-+        commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
++        commit = "60346860111be79fc2beb33c53e195f97504a667"
 +    ),
@@
 
-This call to `rix()` generates a `default.nix` that builds a development shell
-which ships [R]{.proglang} and [R]{.proglang} packages (including [AER]{.pkg} version 1.2-8), several TeXLive
-packages for \LaTeX document authoring, development versions of [rix]{.pkg} and
-[fusen]{.pkg} from GitHub, and the Positron editor.
+This call to rix() generates a `default.nix` file for a development shell that encapsulates 
+a complete and reproducible research environment. It provides [R]{.proglang} and its packages 
+(including [AER]{.pkg} at version 1.2-8), several TeXLive packages for \LaTeX document authoring, 
+development versions of [rix]{.pkg} and [fusen]{.pkg} pulled directly from GitHub at a specific commit,
+and the Positron editor. The reproducibility of this environment is guaranteed by pinning all 
+components to a single point in time: the [R]{.proglang} packages are resolved from the CRAN 
+snapshot of October 20, 2025, while all other system tools are fixed to the version of 
+`nixpkgs` from that same date. Typing `nix-shell` in a terminal in the folder that containts
+this `default.nix`, will drop the user into the shell. It should be noted that it is 
+also possible to configure IDEs to dynamically load this shell to provide the right
+tooling for developing.
 
@@
 in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
-report.
+report. The code to this example can be found on 
+[following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
+with additional details provided in the Appendix of this manuscript.
 
@@
 pipelines and inspecting outputs from Python sessions. [rixpress]{.pkg} also
-includes several additional features not covered here for brevity. It is also
-possible to configure popular IDEs to work interactively and seamlessly with
+includes several additional features not covered here for brevity. As mentioned
+previously, it is also possible to configure popular IDEs to work interactively and seamlessly with
 both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth, reproducible workflow
@@
 
-We thank the rOpenSci reviewers and contributors who provided valuable feedback
-on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, we are
+I thank the rOpenSci reviewers and contributors who provided valuable feedback
+on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, I am
 grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of `{rix}`,
-and to William Landau and Anthony Martinez for their reviews of `{rixpress}`. We
-also acknowledge the contributions of Richard J. Acton, Jordi Rosell, Elio
-Campitelli, László Kupcsik, and Michael Heming for `{rix}`. Their expertise and
-feedback greatly improved the quality and usability of these packages.
+and to William Landau and Anthony Martinez for their reviews of `{rixpress}`. I
+also acknowledge the contributions of Philipp Baumann, [rix]{.pkg}’s co-author,
+Richard J. Acton, Jordi Rosell, Elio Campitelli, László Kupcsik, and 
+Michael Heming for `{rix}`. Their expertise and feedback greatly 
+improved the quality and usability of these packages.
 
@@
 
-## Appendix: A Complete Polyglot Example with rixpress {#sec-appendix}
+## Appendix {#sec-appendix}
+
+### Reproducing this paper
+
+The source code of this paper is hosted on GitHub and can be found
+on following 
+[link](https://github.com/b-rodrigues/rix_paper)^[https://github.com/b-rodrigues/rix_paper].
+The paper can be easily compiled by running the following command:
+
+```{=tex}
+\begin{CodeInput}
+$> nix-shell --run "quarto render paper.qmd --to jss-pdf"
+\end{CodeInput}
+```
+
+This command utilizes the `default.nix` file to build the exact computational 
+environment required to compile the manuscript. To ensure continuous reproducibility,
+the paper is also automatically re-compiled via GitHub Actions on every push to the 
+repository, using the same Nix environment.
+
+### A Complete Polyglot Example with rixpress
 
@@
 
-### The Environment Definition
+It is possible to run this complete example by executing the shell
+script called `run_polyglot_example.sh` found in the repository of
+this 
+paper^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/run_polyglot_example.sh}].
+
+#### The Environment Definition
 
@@
 
-### The RBC Model Simulation
+#### The RBC Model Simulation
 
@@
 
-### The XGBoost Forecasting Model
+#### The XGBoost Forecasting Model
 

### 2025-10-22 paper.qmd
Update paper.qmd
Update nixpkgs URL in CodeInput section

@@
 process differs across operating systems. Further details are discussed in the
-[renv]{.pkg} 
-documentation^[\url{https://rstudio.github.io/renv/articles/renv.html#caveats}].
+[[renv]{.pkg} documentation](https://rstudio.github.io/renv/articles/renv.html#caveats).
 
@@
 $> nix-shell -I \
-+  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/2025-10-20.tar.gz -p \
++  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/tarball/2025-10-20 -p \
 +  R rPackages.rix

### 2025-10-22 paper.qmd
Update paper.qmd

@@
 
-## Introduction: Reproducibility is also about software {#sec-intro}
+## Introduction: Reproducibility is also about software {#secntro}
 
@@
 which is a running instance of the image. Containers are typically run
-non-interactively, allowing the complete computational environment to be
+nonnteractively, allowing the complete computational environment to be
 instantiated and executed with a single command.
@@
 This enables the creation of a truly complete, project-specific, and deeply reproducible environment
-that can be used interactively for development or non-interactively for automated analysis. 
+that can be used interactively for development or nonnteractively for automated analysis. 
 As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
@@
 command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
-installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
+installer: \url{https://determinate.systems/posts/determinate-nixnstaller}].
 Once the build completes, the user can enter the interactive shell with
@@
 
-The package can then generate a visual representation of the pipeline's directed
-acyclic graph:
-
 ![Graphical representation of the polyglot pipeline. Purple nodes are
@@
 
-Figure @fig-dag can be generated by calling `rxp_ggdag()`.
+The package can then generate a visual representation of the pipeline's directed
+acyclic graph by running `rxp_dag()`, as can be seen in Figure @fig-dag.
 
@@
 
-This appendix provides the complete source code for the polyglot pipeline
-example discussed in Section 5. The pipeline simulates a Real Business Cycle
-(RBC) model in [Julia]{.proglang}, trains an [XGBoost]{.pkg} model in
-[Python]{.proglang}, and visualizes the results in [R]{.proglang}.
-
-It is possible to run this complete example by executing the shell
-script called `run_polyglot_example.sh` found in the repository of
-this 
-paper^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/run_polyglot_example.sh}].
-
-#### The Environment Definition
-
-This code snippet uses [rix]{.pkg} to generate a `default.nix` file that contains all
-necessary dependencies for [R]{.proglang}, [Julia]{.proglang}, and
-[Python]{.proglang}. It is typically saved in a file called `gen-env.R`:
-
-```{=tex}
-\begin{CodeInput}
-# This script defines the polyglot environment our pipeline will run in.
-library(rix)
-
-# Define the complete execution environment
-rix(
- # Pin the environment to a specific date for reproducibility.
- date = "2025-10-14",
-
- # R Packages
- r_pkgs = c(
-   "ggplot2", "dplyr", "arrow", "rix", "rixpress", "quarto"
- ),
-
- # Julia Configuration
- jl_conf = list(
-   jl_version = "lts",
-   jl_pkgs = c("Distributions", "DataFrames", "Arrow", "Random")
- ),
-
- # Python Configuration
- py_conf = list(
-   py_version = "3.13",
-   py_pkgs = c("pandas", "scikit-learn", "xgboost", "pyarrow")
- ),
-
- project_path = ".",
- overwrite = TRUE
-)
-\end{CodeInput}
-```
-
-#### The RBC Model Simulation
-
-This Julia script implements the state-space solution to the RBC model as a pure
-function, and is saved in a subfolder of the project named `functions` and in a script
-`functions.jl`. The economic theory is based on @depaoli2009rbc.
-
-```{=tex}
-\begin{Code}
-# This script contains the helper functions for the Julia portion of the pipeline.
-# It implements the correct, stable, and canonical log-linearized solution to
-# a standard Real Business Cycle (RBC) model, based on the provided lecture slides.
-
-# In an interactive session, you should uncomment the line below
-# to load the required packages. rixpress handles this automatically.
-#using LinearAlgebra, Distributions, DataFrames, Arrow, Random
-
-# 1. Main Simulation Function
-
-"""
-    simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
-
-    Takes RBC model parameters as input, solves for the state-space representation
-    using the method of undetermined coefficients (as per the slides), simulates
-    the model for 250 quarters, and returns a DataFrame.
-"""
-function simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
-
-    # --- STEADY-STATE VALUES AND CONSTRUCTED PARAMETERS (Slides p. 16, 31) ---
-    y_k = ((1/β) - 1 + δ) / α
-
-    # --- SOLVE FOR POLICY FUNCTION COEFFICIENTS (Slides p. 19-22) ---
-    # We solve for the coefficients of the policy functions:
-    # c_hat_t = c_ck * k_hat_{t-1} + c_cz * z_hat_t
-    # k_hat_t = c_kk * k_hat_{t-1} + c_kz * z_hat_t
-
-    # Solve the quadratic equation for c_ck (coefficient of capital on consumption)
-    # Based on Campbell (1994) and slide 22, we take the stable root.
-    # a*x^2 + b*x + c = 0
-    a_quad = y_k - δ
-    b_quad = -( (1-β*(1-δ))*(y_k - δ) + y_k*(1+α) + 1 )
-    c_quad = y_k * α
-
-    # The stable solution is the smaller positive root
-    c_ck = (-b_quad - sqrt(b_quad^2 - 4*a_quad*c_quad)) / (2*a_quad)
-
-    # Now solve for the other coefficients based on c_ck
-    c_kk = (y_k * α) / (y_k - δ - c_ck)
-    c_cz = (y_k * (1 - c_kk * (1 - α) * β * ρ)) /
-           ( (y_k - δ - c_ck) * (1 - β * ρ) + σ * (1 - c_ck) * (1 - β * ρ) )
-    c_kz = (c_cz * (1 - c_ck)) / (y_k * (1 - α))
-
-    # --- BUILD STATE-SPACE MATRICES FROM SOLVED COEFFICIENTS ---
-    # State vector: s_t = [k_{t-1}, z_t]'
-    # Transition:   s_t = T*s_{t-1} + R*ε_t
-
-    T = [c_kk  c_kz * ρ
-         0     ρ      ]
-
-    R = [c_kz ; 1]
-
-    # Observation Matrix for [output, consumption, investment]
-    # y_hat_t = α*k_{t-1} + z_t
-    # c_hat_t = c_ck*k_{t-1} + c_cz*z_t
-    # i_hat_t = (i_y_ss)^-1 * (k_t - (1-δ)k_{t-1})
-    #         = (i_y_ss)^-1 * ( (c_kk - (1-δ))k_{t-1} + c_kz*z_t )
-
-    i_y_ss = δ * (α / ((1/β) - 1 + δ))
-
-    C = [α      1
-         c_ck   c_cz
-         (c_kk - (1-δ))/i_y_ss   c_kz/i_y_ss]
-
-    # --- SIMULATE THE MODEL ---
-    Random.seed!(1234)
-    n_periods = 250
-    shocks = randn(n_periods) * σ_z
-
-    states = zeros(2, n_periods)
-    for t in 2:n_periods
-        # The state is [k_{t-1}, z_t]'
-        # To get the next state [k_t, z_{t+1}]', we first find k_t
-        k_t = T[1,1]*states[1, t] + T[1,2]*states[2, t] + R[1]*shocks[t]
-        z_tp1 = T[2,1]*states[1, t] + T[2,2]*states[2, t] + R[2]*shocks[t]
-
-        states[1, t] = k_t # This is now k_t, which is "k_lag" for t+1
-        states[2, t] = z_tp1 # This is now z_{t+1}
-    end
-    # Re-aligning states to be [k_{t-1}, z_t]
-    k_lag = [0; states[1, 1:end-1]]
-    z = [0; states[2, 1:end-1]]
-
-    observables = C * [k_lag'; z']
-
-    df = DataFrame(
-        period = 1:n_periods,
-        output = observables[1, :],
-        consumption = observables[2, :],
-        investment = observables[3, :],
-        capital = k_lag,
-        technology = z
-    )
-
-    return df
-end
-
-# 2. Encoder Function (Unchanged)
-"""
-    arrow_write(df::DataFrame, path::String)
-    Encoder function to save a Julia DataFrame to an Arrow file.
-"""
-function arrow_write(df::DataFrame, path::String)
-    Arrow.write(path, df)
-end
-\end{Code}
-```
-
-#### The XGBoost Forecasting Model
-
-This Python script contains a series of modular functions to handle feature
-engineering, data splitting, model training, and prediction, and is saved in
-`functions/functions.py`.
-
-```{=tex}
-\begin{Code}
-# This script contains modular helper functions for the Python portion of the pipeline.
-# Each function performs a single, distinct task in the ML workflow.
-
-# 1. Load required Python packages
-# This is only needed if you run this script by hand,
-# in which case, uncomment lines 8 to 11.
-# With rixpress, the packages get loaded automatically.
-#import pandas as pd
-#import pyarrow.feather as feather
-#from sklearn.model_selection import train_test_split
-#import xgboost as xgb
-# This script contains the helper functions for the Python portion of the pipeline.
-# It defines the ML model training and prediction logic as a pure function.
-# This script contains the helper functions for the Python portion of the pipeline.
-# It defines the ML model training and prediction logic as a pure function.
-
-# Step 1: Feature Engineering
-def prepare_features(simulated_df: pd.DataFrame) -> pd.DataFrame:
-    """Takes the raw simulated data and creates lagged features and the target variable."""
-    df = simulated_df.copy()
-
-    # Create lagged features
-    for col in ['output', 'consumption', 'investment', 'capital', 'technology']:
-        df[f'{col}_lag1'] = df[col].shift(1)
-
-    # Drop the first row which now has NaNs
-    df.dropna(inplace=True)
-
-    return df
-
-# Step 2: Data Splitting Functions
-def get_X_train(processed_df: pd.DataFrame):
-    """Gets the training features (X_train) from the processed data."""
-    features = [col for col in processed_df.columns if '_lag1' in col]
-    X = processed_df[features]
-    train_size = int(0.75 * len(X))
-    return X[:train_size]
-
-def get_y_train(processed_df: pd.DataFrame):
-    """Gets the training target (y_train) from the processed data."""
-    y = processed_df['output']
-    train_size = int(0.75 * len(y))
-    return y[:train_size]
-
-def get_X_test(processed_df: pd.DataFrame):
-    """Gets the testing features (X_test) from the processed data."""
-    features = [col for col in processed_df.columns if '_lag1' in col]
-    X = processed_df[features]
-    train_size = int(0.75 * len(X))
-    return X[train_size:]
-
-def get_y_test(processed_df: pd.DataFrame):
-    """Gets the testing target (y_test) from the processed data."""
-    y = processed_df['output']
-    train_size = int(0.75 * len(y))
-    return y[train_size:]
-
-# Step 3: Model Training
-def train_model(X_train: pd.DataFrame, y_train: pd.Series):
-    """Initializes and trains the XGBoost model."""
-    model = xgb.XGBRegressor(
-        objective='reg:squarederror',
-        n_estimators=100,
-        learning_rate=0.1,
-        max_depth=3,
-        random_state=42
-    )
-    model.fit(X_train, y_train)
-    return model
-
-# Step 4: Prediction
-def make_predictions(model, X_test: pd.DataFrame):
-    """Makes predictions on the test set using the trained model."""
-    return model.predict(X_test)
-
-# Step 5: Format Final Results
-def format_results(y_test: pd.Series, predictions) -> pd.DataFrame:
-    """Combines test data and predictions into a final DataFrame for R."""
-    results_df = pd.DataFrame({
-        'period': y_test.index,
-        'actual_output': y_test.values,
-        'predicted_output': predictions
-    })
-    return results_df
-
-# Encoder Function (for saving the final output)
-def save_arrow(df: pd.DataFrame, path: str):
-    """Encoder function to save a pandas DataFrame to an Arrow file."""
-    feather.write_feather(df, path)
-
-\end{Code}
-```
-
-### The Visualization
-
-This R snippet uses [ggplot2]{.pkg} to create a plot comparing the model's output
-against the XGBoost predictions and is saved in `functions/functions.R`.
-
-```{=tex}
-\begin{CodeInput}
-# This script contains the helper functions for the R portion of the pipeline.
-# It defines the visualization logic using ggplot2.
-
-# 1. Load required R packages
-# This is only needed if you run this script by hand,
-# in which case, uncomment lines 8 and 9.
-# With rixpress, the packages get loaded automatically.
-#library(ggplot2)
-#library(dplyr)
-
-#-------------------------------------------------------------------------------
-# 2. Main Function: Create the Visualization
-#-------------------------------------------------------------------------------
-
-#' Create a plot comparing actual vs. predicted output
-#'
-#' This function takes a data frame with actual and predicted time series data
-#' and generates a ggplot visualization.
-#'
-#' @param predictions_df A data frame containing columns: 'period',
-#'   'actual_output', and 'predicted_output'. This data frame will be the
-#'   output of the Python XGBoost step.
-#' @return A ggplot object.
-#'
-plot_predictions <- function(predictions_df) {
-  # Create the plot object
-  p <- ggplot(predictions_df, aes(x = period)) +
-
-    # Add a line for the actual output from the RBC model simulation
-    geom_line(
-      aes(y = actual_output, color = "Actual (RBC Model)"),
-      linewidth = 1
-    ) +
-
-    # Add a dashed line for the XGBoost model's predictions
-    geom_line(
-      aes(y = predicted_output, color = "Predicted (XGBoost)"),
-      linetype = "dashed",
-      linewidth = 1
-    ) +
-
-    # Define custom colors and legend labels
-    scale_color_manual(
-      name = "Series",
-      values = c("Actual (RBC Model)" = "blue", "Predicted (XGBoost)" = "red")
-    ) +
-
-    # Add informative labels and a title
-    labs(
-      title = "XGBoost Prediction of RBC Model Output",
-      subtitle = "Forecasting next-quarter output based on current-quarter economic variables",
-      x = "Time (Quarters)",
-      y = "Output (Log-deviations from steady state)"
-    ) +
-
-    # Use a clean theme
-    theme_minimal() +
-
-    # Improve legend position
-    theme(legend.position = "bottom")
-
-  # Return the final ggplot object
-  return(p)
-}
-\end{CodeInput}
-```
-
-### Running the project
-
-The structure of the project should look like this:
-
-```bash
-.
-├── functions
-│   ├── functions.jl
-│   ├── functions.py
-│   └── functions.R
-├── gen-env.R
-└── gen-pipeline.R
-
-2 directories, 5 files
-```
-
-Assuming we are working on a system that has Nix installed, we can "drop" into
-a temporary Nix shell with R and `{rix}` available:
-
-```r
-nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
-```
-
-Once the shell is ready, start R (by simply typing `R`) and run `source("gen-env.R")`
-to generate the adequate `default.nix` file. This is the environment that we can use
-to work interactively with the code while developing, and in which the pipeline will
-be executed. We can then leave the R session and the shell entirely (by typing `exit`)
-and type `nix-shell` to drop into the environment as defined by the `default.nix`.
-The pipeline can now be built using `source("gen-pipeline.R")`.
+This appendix provides a conceptual walk-through of the polyglot pipeline example discussed in Section 5. 
+The pipeline simulates a Real Business Cycle (RBC) model in [Julia]{.proglang}, 
+trains an [XGBoost]{.pkg} model in [Python]{.proglang}, visualizes the results in [R]{.proglang}, 
+and compiles a final report with [Quarto]{.pkg}.
+
+The complete, runnable source code for this example is available in the paper's GitHub repository. 
+A shell script, `run_polyglot_example.sh`, is provided to automate the entire process from environment 
+creation to final output, as described at the end of this section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/run_polyglot_example.sh}].
+
+#### Project Structure and Components
+
+The project is organized into a set of scripts, each with a distinct responsibility. 
+The core logic is separated from the environment definition and pipeline orchestration.
+
+*   `functions/`: This directory contains the helper scripts with the core analytical code for each language.
+    *   `functions.jl`: The [Julia]{.proglang} script for the economic simulation.
+    *   `functions.py`: The [Python]{.proglang} script for the machine learning workflow.
+    *   `functions.R`: The [R]{.proglang} script for data visualization.
+*   `gen-env.R`: An [R]{.proglang} script that defines the reproducible computational environment.
+*   `gen-pipeline.R`: The main [R]{.proglang} script that defines and orchestrates the entire multi-language pipeline.
+
+#### Step 1: The Environment Definition
+
+The foundation of the project is the reproducible environment, defined in `gen-env.R`.
+This script uses the `rix()` function to programmatically generate a `default.nix` file. 
+This file serves as a complete blueprint for the computational environment, specifying:
+
+*   The exact versions of [R]{.proglang}, [Python]{.proglang}, and [Julia]{.proglang}.
+*   A list of required packages for each language (e.g., `ggplot2` for R, `xgboost` for Python, `DataFrames` for Julia).
+*   System-level dependencies like [Quarto]{.pkg}.
+
+Crucially, the entire environment is pinned to a specific date (`2025-10-14`), 
+ensuring that anyone who builds the environment, now or in the future, will get
+the exact same software versions, guaranteeing reproducibility.
+
+By dropping into a temporary shell using the following command:
+
+```{=tex}
+\begin{CodeInput}
+$> nix-shell -I \
++  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/tarball/2025-10-20 -p \
++  R rPackages.rix
+\end{CodeInput}
+```
+
+it is possible to generate the `default.nix` by sourcing `gen-env.R`. Then, one
+leaves this temporary shell, and builds the environment using the command `nix-shell`, 
+which also drop the user into the development shell.
+
+#### Step 2: The Core Analytical Logic
+
+The scientific logic for each stage of the pipeline is encapsulated in the separate 
+helper scripts within the `functions/` directory.
+
+*   **The RBC Model Simulation (`functions.jl`):** This [Julia]{.proglang} script contains a single pure
+  function that implements the state-space solution to the RBC model, based on the economic theory of 
+  @depaoli2009rbc. It takes the model's economic parameters as inputs and returns the simulated time-series
+  data as a DataFrame.
+*   **The XGBoost Forecasting Model (`functions.py`):** This [Python]{.proglang} script handles the 
+  complete machine learning workflow through a series of modular functions. Its responsibilities include
+  feature engineering (creating lagged variables), splitting the data into training and testing sets, 
+  training the [XGBoost]{.pkg} model, and generating predictions.
+*   **The Visualization (`functions.R`):** This [R]{.proglang} script contains a function that uses
+  `[ggplot2]{.pkg}` to create the final visualization. It is designed to take the data frame of 
+  actual and predicted values produced by the Python script and generate a plot comparing the two series.
+
+#### Step 3: Orchestrating the Pipeline
+
+The entire workflow is defined and orchestrated by `gen-pipeline.R`. This script acts as the master plan, 
+using functions from the `[rixpress]{.pkg}` package to define each computational step as a *derivation*.
+It declaratively outlines the dependencies between steps:
+
+1.  It begins by defining the RBC model parameters in [Julia]{.proglang}.
+2.  It specifies that the **RBC simulation** in `functions.jl` depends on these parameters.
+3.  It then defines the series of **Python steps** (feature preparation, training, prediction), 
+  making each one dependent on the output of the previous one. The first Python step is explicitly 
+  made dependent on the output from the Julia simulation. `rixpress` handles the passing of data 
+  between languages, in this case using the Arrow file format for efficiency.
+4.  Finally, it defines the **R visualization** step, which depends on the final predictions from 
+  the Python model, and a **Quarto rendering** step that depends on the generated plot.
+
+When this script is run in the development shell (by executing `source("gen-pipeline.R")`), 
+[rixpress]{.pkg} translates the declared pipeline into a master Nix expression
+that Nix can execute, automatically handling the caching of results and re-running only the necessary 
+steps if a piece of code or data changes.
+
+#### Running the Project
+
+To execute this entire example, a user with Nix installed can simply run the `run_polyglot_example.sh` 
+script available in the root of the paper's repository. This script automates the full process:
+
+1.  It first executes `gen-env.R` inside a temporary Nix shell to build the `default.nix` file.
+2.  It then executes `gen-pipeline.R` inside the newly defined environment. This triggers Nix to run the
+  entire polyglot pipeline in the correct order.
+
+Upon completion, the script will have generated all intermediate artifacts and the final `readme.html` 
+report containing the visualization.

### 2025-10-22 paper.qmd
Update link to run_polyglot_example.sh script

@@
 A shell script, `run_polyglot_example.sh`, is provided to automate the entire process from environment 
-creation to final output, as described at the end of this section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/run_polyglot_example.sh}].
+creation to final output, as described at the end of this section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/polyglot-example/run_polyglot_example.sh}].
 

### 2025-10-22 paper.qmd
Update paper.qmd

@@
 The package can then generate a visual representation of the pipeline's directed
-acyclic graph by running `rxp_dag()`, as can be seen in Figure @fig-dag.
+acyclic graph by running `rxp_dag()`, as can be seen in @fig-dag.
 
@@
 on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, I am
-grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of `{rix}`,
-and to William Landau and Anthony Martinez for their reviews of `{rixpress}`. I
+grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of [rix]{.pkg},
+and to William Landau and Anthony Martinez for their reviews of [rixpress]{.pkg}. I
 also acknowledge the contributions of Philipp Baumann, [rix]{.pkg}’s co-author,
 Richard J. Acton, Jordi Rosell, Elio Campitelli, László Kupcsik, and 
-Michael Heming for `{rix}`. Their expertise and feedback greatly 
+Michael Heming for [rix]{.pkg}. Their expertise and feedback greatly 
 improved the quality and usability of these packages.
@@
 *   The exact versions of [R]{.proglang}, [Python]{.proglang}, and [Julia]{.proglang}.
-*   A list of required packages for each language (e.g., `ggplot2` for R, `xgboost` for Python, `DataFrames` for Julia).
+*   A list of required packages for each language 
+  (e.g., [ggplot2]{.pkg} for [R]{.proglang}, [xgboost]{.pkg} for [Python]{.proglang}, [DataFrames]{.pkg} for [Julia]{.proglang}).
 *   System-level dependencies like [Quarto]{.pkg}.
@@
 
-*   **The RBC Model Simulation (`functions.jl`):** This [Julia]{.proglang} script contains a single pure
-  function that implements the state-space solution to the RBC model, based on the economic theory of 
-  @depaoli2009rbc. It takes the model's economic parameters as inputs and returns the simulated time-series
+*   The RBC Model Simulation (`functions.jl`): This [Julia]{.proglang} script contains a single pure
+  function that implements the state-space solution to the RBC model, based on @depaoli2009rbc.
+  It takes the model's economic parameters as inputs and returns the simulated time-series
   data as a DataFrame.
-*   **The XGBoost Forecasting Model (`functions.py`):** This [Python]{.proglang} script handles the 
+*   The XGBoost Forecasting Model (`functions.py`): This [Python]{.proglang} script handles the 
   complete machine learning workflow through a series of modular functions. Its responsibilities include
@@
   training the [XGBoost]{.pkg} model, and generating predictions.
-*   **The Visualization (`functions.R`):** This [R]{.proglang} script contains a function that uses
-  `[ggplot2]{.pkg}` to create the final visualization. It is designed to take the data frame of 
+*   The Visualization (`functions.R`): This [R]{.proglang} script contains a function that uses
+  [ggplot2]{.pkg} to create the final visualization. It is designed to take the data frame of 
   actual and predicted values produced by the Python script and generate a plot comparing the two series.
@@
 The entire workflow is defined and orchestrated by `gen-pipeline.R`. This script acts as the master plan, 
-using functions from the `[rixpress]{.pkg}` package to define each computational step as a *derivation*.
+using functions from the [rixpress]{.pkg} package to define each computational step as a *derivation*.
 It declaratively outlines the dependencies between steps:
@@
   making each one dependent on the output of the previous one. The first Python step is explicitly 
-  made dependent on the output from the Julia simulation. `rixpress` handles the passing of data 
+  made dependent on the output from the Julia simulation. [rixpress]{.pkg} handles the passing of data 
   between languages, in this case using the Arrow file format for efficiency.
@@
 
-To execute this entire example, a user with Nix installed can simply run the `run_polyglot_example.sh` 
-script available in the root of the paper's repository. This script automates the full process:
+To execute this entire example, a user with Nix installed can simply run the 
+`run_polyglot_example.sh` script available in the root of the paper's repository.
+This script automates the full process:
 

### 2025-10-22 paper.qmd
Update paper.qmd

@@
 
-This call to rix() generates a `default.nix` file for a development shell that encapsulates 
+This call to `rix()` generates a `default.nix` file for a development shell that encapsulates 
 a complete and reproducible research environment. It provides [R]{.proglang} and its packages 
@@
   It takes the model's economic parameters as inputs and returns the simulated time-series
-  data as a DataFrame.
-*   The XGBoost Forecasting Model (`functions.py`): This [Python]{.proglang} script handles the 
+  data as a [DataFrame]{.pkg}.
+*   The [XGBoost]{.pkg} Forecasting Model (`functions.py`): This [Python]{.proglang} script handles the 
   complete machine learning workflow through a series of modular functions. Its responsibilities include
@@
 1.  It begins by defining the RBC model parameters in [Julia]{.proglang}.
-2.  It specifies that the **RBC simulation** in `functions.jl` depends on these parameters.
-3.  It then defines the series of **Python steps** (feature preparation, training, prediction), 
-  making each one dependent on the output of the previous one. The first Python step is explicitly 
-  made dependent on the output from the Julia simulation. [rixpress]{.pkg} handles the passing of data 
-  between languages, in this case using the Arrow file format for efficiency.
+2.  It specifies that the RBC simulation in `functions.jl` depends on these parameters.
+3.  It then defines the series of [Python]{.proglang} steps (feature preparation, training, prediction), 
+  making each one dependent on the output of the previous one. The first [Python]{.proglang} step is explicitly 
+  made dependent on the output from the [Julia]{.proglang} simulation. [rixpress]{.pkg} handles the passing of data 
+  between languages, in this case using the [Arrow]{.pkg} file format for efficiency.
 4.  Finally, it defines the **R visualization** step, which depends on the final predictions from 
-  the Python model, and a **Quarto rendering** step that depends on the generated plot.
+  the [Python]{.proglang} model, and a **Quarto rendering** step that depends on the generated plot.
 
 When this script is run in the development shell (by executing `source("gen-pipeline.R")`), 
-[rixpress]{.pkg} translates the declared pipeline into a master Nix expression
-that Nix can execute, automatically handling the caching of results and re-running only the necessary 
+[rixpress]{.pkg} translates the declared pipeline into a master [Nix]{.proglang} expression
+that [Nix]{.pkg} can execute, automatically handling the caching of results and re-running only the necessary 
 steps if a piece of code or data changes.
@@
 
-To execute this entire example, a user with Nix installed can simply run the 
-`run_polyglot_example.sh` script available in the root of the paper's repository.
+This entire example can be executed by running the `run_polyglot_example.sh` 
+script available in the root of the paper's repository.
 This script automates the full process:
 
-1.  It first executes `gen-env.R` inside a temporary Nix shell to build the `default.nix` file.
-2.  It then executes `gen-pipeline.R` inside the newly defined environment. This triggers Nix to run the
+1.  It first executes `gen-env.R` inside a temporary [Nix]{.pkg} shell to build the `default.nix` file.
+2.  It then executes `gen-pipeline.R` inside the newly defined environment. This triggers [Nix]{.pkg} to run the
   entire polyglot pipeline in the correct order.
 
-Upon completion, the script will have generated all intermediate artifacts and the final `readme.html` 
-report containing the visualization.
+Upon completion, the script will have generated all intermediate artifacts and the final `readme.html` report found in
+the `pipeline-output` folder containing the visualization.

### 2025-10-22 paper.qmd
Update paper.qmd

@@
 
-- **[rixpress]{.pkg}** orchestrates polyglot analytical pipelines using
+- [rixpress]{.pkg} orchestrates polyglot analytical pipelines using
 [Nix]{.pkg} as the build engine. Each pipeline step runs in its own
@@
 This enables true polyglot workflows where [R]{.proglang}, [Python]{.proglang} and 
-[Julia]{.proglang} steps coexist with different dependencies.
+[Julia]{.proglang} steps coexist with different dependencies. [Python]{.proglang} 
+user can use [ryxpress]{.pkg}, a port of [rixpress]{.pkg}.
 
@@
 executing the analysis workflow within that environment, which is the role of
-the [rixpress]{.pkg} package.
+the [rixpress]{.pkg} package (or [ryxpress]{.pkg} for [Python]{.proglang} users).
 
@@
 is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
-`functions.R`), the full contents of which are detailed in the Appendix. The
-high-level orchestration script demonstrates how `{rixpress}` defines a
+`functions.R`). Further details are detailed step-by-step instructions are provided
+in the Appendix. The high-level orchestration script demonstrates how `{rixpress}` defines a
 granular, multi-step machine learning workflow that crosses language boundaries:
@@
 
+As a sidenote: [Python]{.proglang} users that wish to use [ryxpress]{.pkg} 
+would define the pipeline exactly the same way, as an [R]{.proglang} script.
+[ryxpress]{.pkg} calls [R]{.proglang} and [rixpress]{.pkg} under the
+hood to generate the `pipeline.nix` file and then build it.
+
 ![Graphical representation of the polyglot pipeline. Purple nodes are
@@
 
-While [rix]{.pkg} and [rixpress]{.pkg} provide significant steps forward, it is 
+While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide significant steps forward, it is 
 important to acknowledge current limitations. Both packages are relatively recent. 
-[rixpress]{.pkg}, in particular, requires further development, especially to enhance
+[rixpress]{.pkg} (and [ryxpress]{.pkg}), in particular, requires further development, especially to enhance
 the ease of debugging complex pipelines. Its visualization capabilities are also still
@@
 
+To build the pipeline from an interactive [Python]{.proglang} session, one would execute the following lines:
+
+```{=tex}
+\begin{CodeInput}
+Python> from ryxpress import rxp_make
+Python> rxp_make()
+\end{CodeInput}
+```
+
 #### Running the Project

### 2025-10-22 paper.qmd
Update paper.qmd

@@
   researchers must also manage software dependencies, computational environments, and workflow execution. 
-  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, but no single solution handles multi-language environments, 
-  system dependencies, and pipeline orchestration together. I present a two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
+  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, but no single solution handles polyglot environments, 
+  system dependencies, and pipeline orchestration together. I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
   environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments 
@@
 
-## Introduction: Reproducibility is also about software {#secntro}
+## Introduction: Reproducibility is also about software {#sec-intro}
 
@@
 By *linked and executable code and data*, Peng referred to compiled source code
-and runnable scripts. In this paper, we interpret this notion more broadly as
+and runnable scripts. In this paper, I interpret this notion more broadly as
 the *computational environment*: the complete set of software required to
@@
 researchers through two [R]{.proglang} packages: [rix]{.pkg} and
-[rixpress]{.pkg}. Before introducing our approach, I survey existing tools and
-their limitations to contextualize our contribution.
+[rixpress]{.pkg}. Before introducing these packages, I survey existing tools and
+their limitations to contextualize my contribution.
 
@@
 process differs across operating systems. Further details are discussed in the
-[[renv]{.pkg} documentation](https://rstudio.github.io/renv/articles/renv.html#caveats).
+[renv]{.pkg} documentation^[(https://rstudio.github.io/renv/articles/renv.html#caveats)].
 
@@
 These tools represent significant progress in reproducibility, but they share a
-common limitation: **none addresses system-level dependencies or supports
-multi-language environments**. A project using [R]{.proglang},
+common limitation: none addresses system-level dependencies or supports
+polyglot environments. A project using [R]{.proglang},
 [Python]{.proglang}, and system tools like [Quarto]{.pkg} or \LaTeX requires
@@
 which is a running instance of the image. Containers are typically run
-nonnteractively, allowing the complete computational environment to be
+non-interactively, allowing the complete computational environment to be
 instantiated and executed with a single command.
@@
   configuration that works reliably only on Linux. Instead, web-based IDEs (like RStudio
-Server) are used instead, but come with the inherent limitations of web-based solutions.
+  Server) are used, but come with the inherent limitations of web-based solutions.
 
@@
 
-4. Single-language focus: While [Docker]{.pkg} can support multi-language
+4. Single-language focus: While [Docker]{.pkg} can support polyglot
   environments, orchestrating them requires manual coordination. There is no
@@
 [Make]{.pkg} for workflow orchestration. Each tool has its own syntax, concepts,
-and failure modes. The complexity compounds for multi-language projects—a Python
-project would require learning virtual environments or conda, and coordinating R
-and Python in a single analysis requires additional tooling.
+and failure modes. The complexity compounds for polyglot projects; for instance,
+a [Python]{.proglang} project would require learning virtual environments, 
+and coordinating [R]{.proglang} and [Python]{.proglang} in a single analysis requires additional tooling.
 
@@
 Together, these packages provide what existing tools cannot: a single framework
-for managing reproducible, multi-language computational environments and
+for managing reproducible, polyglot computational environments and
 executing complex analytical workflows with step-level isolation. While
@@
 reproducibility. Section 3 presents [rix]{.pkg} and demonstrates environment
-definition for various use cases. Section 4 discusses our `rstats-on-nix` fork
+definition for various use cases. Section 4 discusses the `rstats-on-nix` fork
 of the official package repository and how it addresses practical limitations.
@@
 As of writing, the primary Nix package collection, `nixpkgs`, provides access to over 
-120,000 packages, including nearly all of CRAN and Bioconductor, making it the largest,
-and among the most up-to-date package repository according to Repology:
-
-![nixpkgs is the largest and among the most up-to-date packages repository](repology.png){#fig-repology}
+120,000 packages, including nearly all of CRAN and Bioconductor. According to Repology,
+a service that tracks package repositories across distributions, `nixpkgs` is both the 
+largest and among the most up-to-date package repositories available (see Figure @fig-repology).
+
+![Package repository size and freshness according to Repology. `nixpkgs` leads in both 
+total package count and update recency, making it well-suited for reproducible research 
+environments.](repology.png){#fig-repology}
 
@@
 
-For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, we refer readers to @dolstra2004nix.
+For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, I refer readers to @dolstra2004nix.
 
@@
 and virtual environments for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
-multi-language environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
+polyglot environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
 a [LaTeX]{.proglang} distribution, or any of the numerous other tools available in `nixpkgs`. 
 This enables the creation of a truly complete, project-specific, and deeply reproducible environment
-that can be used interactively for development or nonnteractively for automated analysis. 
+that can be used interactively for development or non-interactively for automated analysis. 
 As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
@@
 then be built on a system with [Nix]{.pkg} installed using the `nix-build`
-command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
+command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
 installer: \url{https://determinate.systems/posts/determinate-nixnstaller}].
@@
 highly appealing. To lower the barrier to adoption of [Nix]{.pkg} for
-reproducibility, we developed the [rix]{.pkg} package.
+reproducibility, I developed the [rix]{.pkg} package.
 
@@
 repository to build software. However, when generating expressions with
-[rix]{.pkg}, our fork `rstats-on-nix/nixpkgs` is used instead.
+[rix]{.pkg}, the fork `rstats-on-nix/nixpkgs` is used instead.
 
@@
 example, when RStudio was temporarily broken due to a dependency issue
-(`boost`), a pull request was submitted to the official `nixpkgs` repository. We
-backported the fix to our fork, making RStudio available to users of our fork
-earlier than upstream, as merging PRs in the official repository can take some
+(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
+backported the fix to the `rstats-on-nix` fork, making RStudio available to users of 
+[rix]{.pkg} earlier than upstream, as merging PRs in the official repository can take some
 time.
@@
 to build many packages from source, which can be time-consuming. To mitigate
-this, we provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
+this, I provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
 complementing the public Nix cache. Instructions for using Cachix are in
@@
 
-To illustrate these capabilities with a realistic research task, we present a
+To illustrate these capabilities with a realistic research task, I present a
 polyglot pipeline that simulates a Real Business Cycle (RBC) model in
@@
 in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
-report. The code to this example can be found on 
+report. The code to this example can be found in the  
 [following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
@@
 is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
-`functions.R`). Further details are detailed step-by-step instructions are provided
+`functions.R`). Further details are provided as well as step-by-step instructions 
 in the Appendix. The high-level orchestration script demonstrates how `{rixpress}` defines a
@@
 
-As a sidenote: [Python]{.proglang} users that wish to use [ryxpress]{.pkg} 
-would define the pipeline exactly the same way, as an [R]{.proglang} script.
-[ryxpress]{.pkg} calls [R]{.proglang} and [rixpress]{.pkg} under the
-hood to generate the `pipeline.nix` file and then build it.
+As a sidenote: [Python]{.proglang} users who wish to use [ryxpress]{.pkg} 
+define pipelines using the same [R]{.proglang}-based DSL shown above. This design 
+choice keeps the pipeline definition language consistent across both [R]{.proglang} 
+and [Python]{.proglang} ecosystems. [ryxpress]{.pkg} calls [R]{.proglang} and 
+[rixpress]{.pkg} under the hood to generate and build the `pipeline.nix` file.
 
@@
 recompute steps with changed inputs or code. This provides the efficiency of
-[targets]{.pkg} with multi-language support and bit-for-bit reproducibility.
+[targets]{.pkg} with polyglot support and bit-for-bit reproducibility.
 Artifacts can be inspected interactively in [R]{.proglang} using
@@
 However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
-[rix]{.pkg}, we aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
+[rix]{.pkg}, I aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
 by providing a familiar interface and workflow. By building reproducible
@@
 environment. This ensures deep reproducibility, efficient caching, and seamless
-orchestration of multi-language analyses, making it easier to manage complex
+orchestration of polyglot analyses, making it easier to manage complex
 projects with confidence that results can be reproduced exactly.
@@
 While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide significant steps forward, it is 
-important to acknowledge current limitations. Both packages are relatively recent. 
-[rixpress]{.pkg} (and [ryxpress]{.pkg}), in particular, requires further development, especially to enhance
+important to acknowledge current limitations. 
+[rixpress]{.pkg} (and [ryxpress]{.pkg}) in particular requires further development, especially to enhance
 the ease of debugging complex pipelines. Its visualization capabilities are also still
@@
 *   `gen-env.R`: An [R]{.proglang} script that defines the reproducible computational environment.
-*   `gen-pipeline.R`: The main [R]{.proglang} script that defines and orchestrates the entire multi-language pipeline.
+*   `gen-pipeline.R`: The main [R]{.proglang} script that defines and orchestrates the entire polyglot pipeline.
 

### 2025-10-22 paper.qmd
Update paper.qmd

@@
   researchers must also manage software dependencies, computational environments, and workflow execution. 
-  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, but no single solution handles polyglot environments, 
-  system dependencies, and pipeline orchestration together. I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
+  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, 
+  but no single solution handles polyglot environments, system dependencies, 
+  and workflow orchestration together.
+  I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
   environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments 
@@
 process differs across operating systems. Further details are discussed in the
-[renv]{.pkg} documentation^[(https://rstudio.github.io/renv/articles/renv.html#caveats)].
+[renv]{.pkg} documentation^[See \url{https://rstudio.github.io/renv/articles/renv.html}.].
 
@@
 
-For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, I refer readers to @dolstra2004nix.
+For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, see @dolstra2004nix.
 
@@
 command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
-installer: \url{https://determinate.systems/posts/determinate-nixnstaller}].
+installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
 Once the build completes, the user can enter the interactive shell with
@@
 
-The `rstats-on-nix` fork allows us to circumvent these limitations. For example, it provides
+The `rstats-on-nix` fork allows me to circumvent these limitations. For example, it provides
 daily snapshots of CRAN. Each day, the [R]{.proglang} package set is updated and
@@
 date added to the list of available dates in [rix]{.pkg}. This ensures users can
-reliably install packages, and allows us to backport fixes if needed. For
+reliably install packages, and allows me to backport fixes if needed. For
 example, when RStudio was temporarily broken due to a dependency issue
@@
 in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
-report. The code to this example can be found in the  
+report. The code for this example can be found in the  
 [following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
@@
 environment. This ensures deep reproducibility, efficient caching, and seamless
-orchestration of polyglot analyses, making it easier to manage complex
-projects with confidence that results can be reproduced exactly.
+orchestration of polyglot analyses. Most importantly, it provides
+native polyglot support without the manual coordination required by 
+containerization approaches, making complex multi-language workflows 
+accessible to researchers without systems administration expertise.
 
@@
 The source code of this paper is hosted on GitHub and can be found
-on following 
+at following 
 [link](https://github.com/b-rodrigues/rix_paper)^[https://github.com/b-rodrigues/rix_paper].

### 2025-10-22 paper.qmd
chore: remove end of line extre whitespaces
chore: fix typo
fix: reword duplicated sentence

@@
 snapshot of October 20, 2025, while all other system tools are fixed to the version of
-`nixpkgs` from that same date. Typing `nix-shell` in a terminal in the folder that containts
-this `default.nix`, will drop the user into the shell. It should be noted that it is 
-also possible to configure IDEs to dynamically load this shell to provide the right
-tooling for developing.
+`nixpkgs` from that same date. Typing `nix-shell` in a terminal within the folder that contains
+this `default.nix` will drop the user into a new shell. It should be noted that it uses the
+`nixpkgs` from that same date. It is also possible to configure IDEs to dynamically load this
+new shell to provide the proper development tooling.
 

### 2025-10-22 paper.qmd
feat: add new point about "Incomplete Reproducibility"

@@
   another runs in R.
+
+5. Incomplete Reproducibility: Reproducibility should be understood as a spectrum
+  rather than a binary property, a notion emphasised in "Reproducibility in Software
+  Engineering" by @dellaieraMasterThesis2024. Docker illustrates this spectrum well:
+  it provides strong run-time reproducibility—that is, consistency in space
+  (where the same image behaves identically across different systems) but weak
+  build-time reproducibility, which concerns consistency over time. As defined in
+  the thesis, reproducibility at build time means being able to produce an identical
+  build artefact regardless of when or where it is built. However, Docker inherently
+  fails to guarantee this property. Its reliance on mutable base images and non-deterministic
+  instructions such as `apt-get update` leads to temporal drift: rebuilding the same
+  `Dockerfile` at different points in time can produce distinct outputs. In the
+  experimental evaluation conducted in the thesis, even with pinned base images,
+  checksums of the resulting containers varied across builds and machines. Thus, while
+  Docker offers partial reproducibility: strong at run time (space) and weak at build time (time),
+  it cannot on its own ensure verifiable, deterministic software artefacts. Achieving
+  true reproducibility requires pinning image digests, freezing dependency versions,
+  and employing deterministic build systems such as Nix or Guix for complete control
+  over both space and time [@Malka_2024].
 

### 2025-10-22 paper.qmd
chore: update `Acknowledgments`
chore: replace `buildInputs` with `packages`

@@
   pkgs.mkShell {
-    buildInputs = [ system_packages ];
+    packages = [ system_packages ];
     shellHook = "R --vanilla";
@@
 also acknowledge the contributions of Philipp Baumann, [rix]{.pkg}’s co-author,
-Richard J. Acton, Jordi Rosell, Elio Campitelli, László Kupcsik, and
+Richard J. Acton, Jordi Rosell, Elio Campitelli, László Kupcsik, Pol Dellaiera, and
 Michael Heming for [rix]{.pkg}. Their expertise and feedback greatly

### 2025-10-23 paper.qmd
Update paper.qmd
formatting

--- 2025-10-22 paper.qmd
@@
   Reproducible analysis requires more than clean, well-documented code:
-  researchers must also manage software dependencies, computational environments, and workflow execution.
-  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge,
-  but no single solution handles polyglot environments, system dependencies,
-  and workflow orchestration together.
-  I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot
-  environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments
-  spanning [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system dependencies. Building on these environments, [rixpress]{.pkg} orchestrates polyglot pipelines where each
-  computational step runs in its own hermetically sealed environment, with automatic caching and dependency tracking.
-  This approach enables researchers to develop analyses interactively while maintaining bit-for-bit reproducibility, supports
-  collaboration across heterogeneous computational environments, and ensures that analyses remain executable years into the future.
+  researchers must also manage software dependencies, computational
+  environments, and workflow execution. Existing tools like [Docker]{.pkg} and
+  [renv]{.pkg} address parts of this challenge, but no single solution handles
+  polyglot environments, system dependencies, and workflow orchestration
+  together. I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to
+  provide both reproducible polyglot environments and declarative workflow
+  orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define
+  reproducible environments spanning [R]{.proglang}, [Python]{.proglang},
+  [Julia]{.proglang}, and system dependencies. Building on these environments,
+  [rixpress]{.pkg} orchestrates polyglot pipelines where each computational step
+  runs in its own hermetically sealed environment, with automatic caching and
+  dependency tracking. This approach enables researchers to develop analyses
+  interactively while maintaining bit-for-bit reproducibility, supports
+  collaboration across heterogeneous computational environments, and ensures
+  that analyses remain executable years into the future.
 
@@
 By *linked and executable code and data*, Peng referred to compiled source code
-and runnable scripts. In this paper, I interpret this notion more broadly as
-the *computational environment*: the complete set of software required to
-execute an analysis. Here too, a continuum exists. At the minimal end, authors
-might only name the main software used—say, the [R]{.proglang} programming
-language. More careful authors might also specify the version of [R]{.proglang},
-or list the additional packages and their versions. Rarely, however, do authors
-specify the operating system on which the analysis was performed, even though
-differences in operating systems can lead to divergent results when using the
-same code and software versions, as shown by @neupane2019. It is even less
-common for authors to provide step-by-step installation instructions for the
-required software stack.
+and runnable scripts. In this paper, I interpret this notion more broadly as the
+*computational environment*: the complete set of software required to execute an
+analysis. Here too, a continuum exists. At the minimal end, authors might only
+name the main software used—say, the [R]{.proglang} programming language. More
+careful authors might also specify the version of [R]{.proglang}, or list the
+additional packages and their versions. Rarely, however, do authors specify the
+operating system on which the analysis was performed, even though differences in
+operating systems can lead to divergent results when using the same code and
+software versions, as shown by @neupane2019. It is even less common for authors
+to provide step-by-step installation instructions for the required software
+stack.
 
@@
 When an author includes this information, others attempting to reproduce the
-study (future readers, collaborators, or even the author at a later
-time) can easily see which version of [R]{.proglang} and which packages (with
-their versions) were used. However, reproducing the environment still requires
+study (future readers, collaborators, or even the author at a later time) can
+easily see which version of [R]{.proglang} and which packages (with their
+versions) were used. However, reproducing the environment still requires
 manually installing the correct package versions—a process that becomes
 particularly difficult when packages depend on system-level libraries. For
-example, the [sf]{.pkg} package requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} system libraries.
-Installing and configuring these dependencies varies substantially across
-operating systems and can be prohibitively difficult on some platforms.
+example, the [sf]{.pkg} package requires [GDAL]{.pkg}, [GEOS]{.pkg}, and
+[PROJ]{.pkg} system libraries. Installing and configuring these dependencies
+varies substantially across operating systems and can be prohibitively difficult
+on some platforms.
 
@@
 [renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
-[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
-resolve this conflict. Users must manually install system libraries, and the
-process differs across operating systems. Further details are discussed in the
-[renv]{.pkg} documentation^[See \url{https://rstudio.github.io/renv/articles/renv.html}.].
+[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg}
+cannot resolve this conflict. Users must manually install system libraries, and
+the process differs across operating systems. Further details are discussed in
+the [renv]{.pkg} documentation^[See
+\url{https://rstudio.github.io/renv/articles/renv.html}.].
 
@@
 existed on CRAN at a given date:
+
 ```{=tex}
@@
 These tools represent significant progress in reproducibility, but they share a
-common limitation: none addresses system-level dependencies or supports
-polyglot environments. A project using [R]{.proglang},
-[Python]{.proglang}, and system tools like [Quarto]{.pkg} or \LaTeX requires
-coordinating multiple package managers ([renv]{.pkg} for R, virtual environments
-for Python, manual installation for system tools), each with its own
-configuration and potential for conflicts.
+common limitation: none addresses system-level dependencies or supports polyglot
+environments. A project using [R]{.proglang}, [Python]{.proglang}, and system
+tools like [Quarto]{.pkg} or \LaTeX requires coordinating multiple package
+managers ([renv]{.pkg} for R, virtual environments for Python, manual
+installation for system tools), each with its own configuration and potential
+for conflicts.
 
@@
   changes are lost when the container stops unless explicitly saved. Running
-  graphical applications like RStudio Desktop from containers requires complex X11
-  configuration that works reliably only on Linux. Instead, web-based IDEs (like RStudio
-  Server) are used, but come with the inherent limitations of web-based solutions.
+  graphical applications like RStudio Desktop from containers requires complex
+  X11 configuration that works reliably only on Linux. Instead, web-based IDEs
+  (like RStudio Server) are used, but come with the inherent limitations of
+  web-based solutions.
 
@@
   interactively using standard R installations, then create a `Dockerfile`
-  afterward to enable reproduction. While this achieves post-hoc reproducibility,
-  it fails to address the challenge of deploying consistent environments across
-  multiple machines *during* development, a critical need for collaborative
-  projects.
+  afterward to enable reproduction. While this achieves post-hoc
+  reproducibility, it fails to address the challenge of deploying consistent
+  environments across multiple machines *during* development, a critical need
+  for collaborative projects.
 
@@
 
-5. Incomplete Reproducibility: Reproducibility should be understood as a spectrum
-  rather than a binary property, a notion emphasised in "Reproducibility in Software
-  Engineering" by @dellaieraMasterThesis2024. Docker illustrates this spectrum well:
-  it provides strong run-time reproducibility—that is, consistency in space
-  (where the same image behaves identically across different systems) but weak
-  build-time reproducibility, which concerns consistency over time. As defined in
-  the thesis, reproducibility at build time means being able to produce an identical
-  build artefact regardless of when or where it is built. However, Docker inherently
-  fails to guarantee this property. Its reliance on mutable base images and non-deterministic
-  instructions such as `apt-get update` leads to temporal drift: rebuilding the same
-  `Dockerfile` at different points in time can produce distinct outputs. In the
-  experimental evaluation conducted in the thesis, even with pinned base images,
-  checksums of the resulting containers varied across builds and machines. Thus, while
-  Docker offers partial reproducibility: strong at run time (space) and weak at build time (time),
-  it cannot on its own ensure verifiable, deterministic software artefacts. Achieving
-  true reproducibility requires pinning image digests, freezing dependency versions,
-  and employing deterministic build systems such as Nix or Guix for complete control
-  over both space and time [@Malka_2024].
+5. Incomplete Reproducibility: Reproducibility should be understood as a
+  spectrum rather than a binary property, a notion emphasised in
+  "Reproducibility in Software Engineering" by @dellaieraMasterThesis2024.
+  Docker illustrates this spectrum well: it provides strong run-time
+  reproducibility—that is, consistency in space (where the same image behaves
+  identically across different systems) but weak build-time reproducibility,
+  which concerns consistency over time. As defined in the thesis,
+  reproducibility at build time means being able to produce an identical build
+  artefact regardless of when or where it is built. However, Docker inherently
+  fails to guarantee this property. Its reliance on mutable base images and
+  non-deterministic instructions such as `apt-get update` leads to temporal
+  drift: rebuilding the same `Dockerfile` at different points in time can
+  produce distinct outputs. In the experimental evaluation conducted in the
+  thesis, even with pinned base images, checksums of the resulting containers
+  varied across builds and machines. Thus, while Docker offers partial
+  reproducibility: strong at run time (space) and weak at build time (time), it
+  cannot on its own ensure verifiable, deterministic software artifacts.
+  Achieving true reproducibility requires pinning image digests, freezing
+  dependency versions, and employing deterministic build systems such as Nix or
+  Guix for complete control over both space and time [@Malka_2024].
 
@@
 and failure modes. The complexity compounds for polyglot projects; for instance,
-a [Python]{.proglang} project would require learning virtual environments,
-and coordinating [R]{.proglang} and [Python]{.proglang} in a single analysis requires additional tooling.
+a [Python]{.proglang} project would require learning virtual environments, and
+coordinating [R]{.proglang} and [Python]{.proglang} in a single analysis
+requires additional tooling.
 
@@
 
-1. Manages complete environments including programming languages, packages,
-and system dependencies;
+1. Manages complete environments including programming languages, packages, and
+   system dependencies;
 2. Supports multiple languages natively, not through workarounds;
@@
 reproducible software installation by deploying *component closures*—packages
-bundled with all their dependencies and transitive dependencies [@dolstra2004nix].
-This creates self-contained software environments that work identically across
-machines and over time. [Nix]{.pkg} can replace [Docker]{.pkg} for environment
-isolation, [renv]{.pkg} for package management, and even [Make]{.pkg} for
-workflow orchestration—all within a single, unified framework.
+bundled with all their dependencies and transitive dependencies
+[@dolstra2004nix]. This creates self-contained software environments that work
+identically across machines and over time. [Nix]{.pkg} can replace
+[Docker]{.pkg} for environment isolation, [renv]{.pkg} for package management,
+and even [Make]{.pkg} for workflow orchestration—all within a single, unified
+framework.
 
@@
 
-- [rix]{.pkg} generates [Nix]{.pkg} expressions from intuitive R function
-calls, eliminating the need to learn the [Nix]{.proglang} language. It handles
+- [rix]{.pkg} generates [Nix]{.pkg} expressions from intuitive R function calls,
+eliminating the need to learn the [Nix]{.proglang} language. It handles
 environment definition including [R]{.proglang}, [Python]{.proglang},
@@
 
-- [rixpress]{.pkg} orchestrates polyglot analytical pipelines using
-[Nix]{.pkg} as the build engine. Each pipeline step runs in its own
-hermetically sealed environment with automatic caching and dependency tracking.
-This enables true polyglot workflows where [R]{.proglang}, [Python]{.proglang} and
-[Julia]{.proglang} steps coexist with different dependencies. [Python]{.proglang}
-user can use [ryxpress]{.pkg}, a port of [rixpress]{.pkg}.
+- [rixpress]{.pkg} orchestrates polyglot analytical pipelines using [Nix]{.pkg}
+as the build engine. Each pipeline step runs in its own hermetically sealed
+environment with automatic caching and dependency tracking. This enables true
+polyglot workflows where [R]{.proglang}, [Python]{.proglang} and
+[Julia]{.proglang} steps coexist with different dependencies.
+[Python]{.proglang} user can use [ryxpress]{.pkg}, a port of [rixpress]{.pkg}.
 
 Together, these packages provide what existing tools cannot: a single framework
-for managing reproducible, polyglot computational environments and
-executing complex analytical workflows with step-level isolation. While
-[Nix]{.pkg} remains complex for advanced use cases, [rix]{.pkg} and
-[rixpress]{.pkg} abstract this complexity for common research workflows, making
-deep reproducibility accessible without requiring systems administration
-expertise.
+for managing reproducible, polyglot computational environments and executing
+complex analytical workflows with step-level isolation. While [Nix]{.pkg}
+remains complex for advanced use cases, [rix]{.pkg} and [rixpress]{.pkg}
+abstract this complexity for common research workflows, making deep
+reproducibility accessible without requiring systems administration expertise.
 
@@
 
-[Nix]{.pkg} is a powerful, cross-platform package manager designed for installing
-and building software in a fully reproducible and reliable manner. Unlike traditional
-package managers, [Nix]{.pkg} emphasizes immutable infrastructure and a functional
-approach, which significantly enhances the consistency of computational environments.
-As of writing, the primary Nix package collection, `nixpkgs`, provides access to over
-120,000 packages, including nearly all of CRAN and Bioconductor. According to Repology,
-a service that tracks package repositories across distributions, `nixpkgs` is both the
-largest and among the most up-to-date package repositories available (see Figure @fig-repology).
-
-![Package repository size and freshness according to Repology. `nixpkgs` leads in both
-total package count and update recency, making it well-suited for reproducible research
-environments.](repology.png){#fig-repology}
-
-This extensive coverage allows users to install not only [R]{.proglang} itself but
-also all its required packages and system-level dependencies for any given project.
-
-While [Nix]{.pkg} is the default package manager for the NixOS
-Linux distribution, it can be installed as a standalone tool on other Linux
+[Nix]{.pkg} is a powerful, cross-platform package manager designed for
+installing and building software in a fully reproducible and reliable manner.
+Unlike traditional package managers, [Nix]{.pkg} emphasizes immutable
+infrastructure and a functional approach, which significantly enhances the
+consistency of computational environments. As of writing, the primary Nix
+package collection, `nixpkgs`, provides access to over 120,000 packages,
+including nearly all of CRAN and Bioconductor. According to Repology, a service
+that tracks package repositories across distributions, `nixpkgs` is both the
+largest and among the most up-to-date package repositories available (see Figure
+@fig-repology).
+
+![Package repository size and freshness according to Repology. `nixpkgs` leads
+in both total package count and update recency, making it well-suited for
+reproducible research environments.](repology.png){#fig-repology}
+
+This extensive coverage allows users to install not only [R]{.proglang} itself
+but also all its required packages and system-level dependencies for any given
+project.
+
+While [Nix]{.pkg} is the default package manager for the NixOS Linux
+distribution, it can be installed as a standalone tool on other Linux
 distributions, macOS, and effectively on Windows via the Windows Subsystem for
@@
 opposed to the standard `install.packages()` function, lies in its comprehensive
-dependency management. [Nix]{.pkg} ensures that all dependencies of each package are
-installed, irrespective of whether they are other [R]{.proglang} packages or
+dependency management. [Nix]{.pkg} ensures that all dependencies of each package
+are installed, irrespective of whether they are other [R]{.proglang} packages or
 underlying system libraries.
 
-Consider, for example, the [sf]{.pkg} package, which is fundamental for spatial data analysis in [R]{.proglang}.
-[sf]{.pkg} relies on several complex external system libraries such as [GDAL]{.proglang},
-[GEOS]{.proglang}, and [PROJ]{.proglang}. Manually installing and configuring these libraries
-to be compatible with [sf]{.pkg} can be challenging and platform-specific. With [Nix]{.pkg},
-however, the user merely needs to declare [sf]{.pkg} as a project requirement.
-[Nix]{.pkg} then automatically installs and configures all necessary system libraries
-as transitive dependencies. This seamless process is possible because the maintainers of the
-[R]{.proglang} programming language for [Nix]{.pkg}
-have declaratively specified these dependencies for [sf]{.pkg} in `nixpkgs`, allowing the entire software
-stack to be provisioned effortlessly from an end-user perspective.
-
-This concept is central to [Nix]{.pkg}'s design, where packages are referred to as *component closures*.
-As @dolstra2004nix explains:
+Consider, for example, the [sf]{.pkg} package, which is fundamental for spatial
+data analysis in [R]{.proglang}. [sf]{.pkg} relies on several complex external
+system libraries such as [GDAL]{.proglang}, [GEOS]{.proglang}, and
+[PROJ]{.proglang}. Manually installing and configuring these libraries to be
+compatible with [sf]{.pkg} can be challenging and platform-specific. With
+[Nix]{.pkg}, however, the user merely needs to declare [sf]{.pkg} as a project
+requirement. [Nix]{.pkg} then automatically installs and configures all
+necessary system libraries as transitive dependencies. This seamless process is
+possible because the maintainers of the [R]{.proglang} programming language for
+[Nix]{.pkg} have declaratively specified these dependencies for [sf]{.pkg} in
+`nixpkgs`, allowing the entire software stack to be provisioned effortlessly
+from an end-user perspective.
+
+This concept is central to [Nix]{.pkg}'s design, where packages are referred to
+as *component closures*. As @dolstra2004nix explains:
 
@@
 
-The core of [Nix]{.pkg}'s reproducibility lies in its functional package management paradigm
-and its unique approach to builds. When installing software with [Nix]{.pkg}, it evaluates
-an *expression* written in the [Nix]{.proglang} language. These expressions define *derivations*,
-which are declarative blueprints describing how to build a package. A derivation specifies all inputs:
-source code, build commands, and crucially, all direct and transitive dependencies. This declarative
-nature ensures that every build process is fully specified and independent of the environment where
-it's executed. Furthermore, all [Nix]{.pkg} expressions for official packages are hosted in the `nixpkgs`
-GitHub repository. By referencing a specific commit of `nixpkgs` (a process known as *pinning a revision*) users
-guarantee that all packages installed are derived from the exact same set of build instructions.
-This means that if an environment is built today using a pinned revision, it will produce the
-identical set of software versions and configurations if rebuilt tomorrow, next year, or on a
-different machine, as long as the `nixpkgs` repository remains accessible.
-
-Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports reproducibility.
-[Nix]{.pkg} is a *functional* package manager in the programming sense: it treats software builds as pure functions.
-Given the same inputs (source code, dependencies, build scripts), a [Nix]{.pkg} build will always produce the
-identical output, regardless of when or where it's executed. This is achieved by disallowing side effects and
-ensuring that builds operate within hermetic (isolated) environments, with no hidden dependencies on global system
-state. While this functional approach greatly enhances reproducibility, it can sometimes introduce complexity for
-package maintainers, especially for software that needs to download external assets during installation (e.g.,
-certain Bioconductor packages). For end-users, however, this complexity is largely abstracted away. Additionally,
-[Nix]{.pkg} supports multiple versions or *variants* of a package on the same system. Each package installed into
-the "Nix store" (typically in `/nix/store`) has a unique identifier derived from a hash of all its inputs. This
-isolation ensures that projects always use their intended software versions, allowing multiple versions of
-[R]{.proglang} to coexist and preventing "dependency hell" and global environment pollution.
-
-For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, see @dolstra2004nix.
-
-By leveraging these principles, [Nix]{.pkg} can effectively replace and integrate the functionalities of
-tools like [renv]{.pkg} and [Docker]{.pkg} for [R]{.proglang} projects, or `requirements.txt` files
-and virtual environments for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
-polyglot environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
-a [LaTeX]{.proglang} distribution, or any of the numerous other tools available in `nixpkgs`.
-This enables the creation of a truly complete, project-specific, and deeply reproducible environment
-that can be used interactively for development or non-interactively for automated analysis.
-As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
-the environment can be rebuilt reliably in the future.
-
-[Nix]{.pkg} provides strong cross-platform support across major operating systems. It runs natively on Linux
-(x86_64) with full functionality and optimal integration. On macOS (Intel and Apple Silicon), however,
-reproducibility is less reliable due to dependencies on impure system frameworks and Xcode toolchains outside
-the Nix store. As a result, macOS users may need to update project pins after system or Xcode upgrades to
-restore builds. On Windows, [Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though graphical
-applications may need extra setup. [Nix]{.pkg} environments on WSL can be used interactively via VS Code or Positron
-for a smooth development experience. It is also important to note that [Nix]{.pkg} can be installed
-within a [Docker]{.pkg} image to deploy and reproduce environments reliably in containerized workflows.
-
-While [Nix]{.pkg} is a powerful system, it comes with notable challenges. The steepest is its learning curve:
-[Nix]{.pkg} uses its own declarative language (also called [Nix]{.proglang}), which can be difficult for
-those new to functional or declarative paradigms. Complexity becomes most apparent when writing custom
-derivations or troubleshooting intricate build issues. Build times can also be long, especially for large
-environments, when pre-built binaries aren’t available. Additionally, because [Nix]{.pkg} stores all package
-outputs immutably in the Nix store, disk usage tends to be higher than with traditional package managers.
-Finally, while `nixpkgs` is extensive, some new or niche packages may be missing or fail to build across platforms.
-
-To make these capabilities more approachable for [R]{.proglang} users seeking reproducible,
-project-specific environments without learning the full [Nix]{.proglang} language, I created the
-[rix]{.pkg} and [rixpress]{.pkg} packages.
+The core of [Nix]{.pkg}'s reproducibility lies in its functional package
+management paradigm and its unique approach to builds. When installing software
+with [Nix]{.pkg}, it evaluates an *expression* written in the [Nix]{.proglang}
+language. These expressions define *derivations*, which are declarative
+blueprints describing how to build a package. A derivation specifies all inputs:
+source code, build commands, and crucially, all direct and transitive
+dependencies. This declarative nature ensures that every build process is fully
+specified and independent of the environment where it's executed. Furthermore,
+all [Nix]{.pkg} expressions for official packages are hosted in the `nixpkgs`
+GitHub repository. By referencing a specific commit of `nixpkgs` (a process
+known as *pinning a revision*) users guarantee that all packages installed are
+derived from the exact same set of build instructions. This means that if an
+environment is built today using a pinned revision, it will produce the
+identical set of software versions and configurations if rebuilt tomorrow, next
+year, or on a different machine, as long as the `nixpkgs` repository remains
+accessible.
+
+Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports
+reproducibility. [Nix]{.pkg} is a *functional* package manager in the
+programming sense: it treats software builds as pure functions. Given the same
+inputs (source code, dependencies, build scripts), a [Nix]{.pkg} build will
+always produce the identical output, regardless of when or where it's executed.
+This is achieved by disallowing side effects and ensuring that builds operate
+within hermetic (isolated) environments, with no hidden dependencies on global
+system state. While this functional approach greatly enhances reproducibility,
+it can sometimes introduce complexity for package maintainers, especially for
+software that needs to download external assets during installation (e.g.,
+certain Bioconductor packages). For end-users, however, this complexity is
+largely abstracted away. Additionally, [Nix]{.pkg} supports multiple versions or
+*variants* of a package on the same system. Each package installed into the "Nix
+store" (typically in `/nix/store`) has a unique identifier derived from a hash
+of all its inputs. This isolation ensures that projects always use their
+intended software versions, allowing multiple versions of [R]{.proglang} to
+coexist and preventing "dependency hell" and global environment pollution.
+
+For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, see
+@dolstra2004nix.
+
+By leveraging these principles, [Nix]{.pkg} can effectively replace and
+integrate the functionalities of tools like [renv]{.pkg} and [Docker]{.pkg} for
+[R]{.proglang} projects, or `requirements.txt` files and virtual environments
+for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
+polyglot environments, which can seamlessly combine [R]{.proglang},
+[Python]{.proglang}, [Julia]{.proglang}, a [LaTeX]{.proglang} distribution, or
+any of the numerous other tools available in `nixpkgs`. This enables the
+creation of a truly complete, project-specific, and deeply reproducible
+environment that can be used interactively for development or non-interactively
+for automated analysis. As long as the `nixpkgs` repository (or a suitable fork,
+as discussed in Section @sec-fork) remains accessible, the environment can be
+rebuilt reliably in the future.
+
+[Nix]{.pkg} provides strong cross-platform support across major operating
+systems. It runs natively on Linux (x86_64) with full functionality and optimal
+integration. On macOS (Intel and Apple Silicon), however, reproducibility is
+less reliable due to dependencies on impure system frameworks and Xcode
+toolchains outside the Nix store. As a result, macOS users may need to update
+project pins after system or Xcode upgrades to restore builds. On Windows,
+[Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though
+graphical applications may need extra setup. [Nix]{.pkg} environments on WSL can
+be used interactively via VS Code or Positron for a smooth development
+experience. It is also important to note that [Nix]{.pkg} can be installed
+within a [Docker]{.pkg} image to deploy and reproduce environments reliably in
+containerized workflows.
+
+While [Nix]{.pkg} is a powerful system, it comes with notable challenges. The
+steepest is its learning curve: [Nix]{.pkg} uses its own declarative language
+(also called [Nix]{.proglang}), which can be difficult for those new to
+functional or declarative paradigms. Complexity becomes most apparent when
+writing custom derivations or troubleshooting intricate build issues. Build
+times can also be long, especially for large environments, when pre-built
+binaries aren’t available. Additionally, because [Nix]{.pkg} stores all package
+outputs immutably in the Nix store, disk usage tends to be higher than with
+traditional package managers. Finally, while `nixpkgs` is extensive, some new or
+niche packages may be missing or fail to build across platforms.
+
+To make these capabilities more approachable for [R]{.proglang} users seeking
+reproducible, project-specific environments without learning the full
+[Nix]{.proglang} language, I created the [rix]{.pkg} and [rixpress]{.pkg}
+packages.
 
@@
 
-This call to `rix()` generates a `default.nix` file for a development shell that encapsulates
-a complete and reproducible research environment. It provides [R]{.proglang} and its packages
-(including [AER]{.pkg} at version 1.2-8), several TeXLive packages for \LaTeX document authoring,
-development versions of [rix]{.pkg} and [fusen]{.pkg} pulled directly from GitHub at a specific commit,
-and the Positron editor. The reproducibility of this environment is guaranteed by pinning all
-components to a single point in time: the [R]{.proglang} packages are resolved from the CRAN
-snapshot of October 20, 2025, while all other system tools are fixed to the version of
-`nixpkgs` from that same date. Typing `nix-shell` in a terminal within the folder that contains
-this `default.nix` will drop the user into a new shell. It should be noted that it uses the
-`nixpkgs` from that same date. It is also possible to configure IDEs to dynamically load this
-new shell to provide the proper development tooling.
+This call to `rix()` generates a `default.nix` file for a development shell that
+encapsulates a complete and reproducible research environment. It provides
+[R]{.proglang} and its packages (including [AER]{.pkg} at version 1.2-8),
+several TeXLive packages for \LaTeX document authoring, development versions of
+[rix]{.pkg} and [fusen]{.pkg} pulled directly from GitHub at a specific commit,
+and the Positron editor. The reproducibility of this environment is guaranteed
+by pinning all components to a single point in time: the [R]{.proglang} packages
+are resolved from the CRAN snapshot of October 20, 2025, while all other system
+tools are fixed to the version of `nixpkgs` from that same date. Typing
+`nix-shell` in a terminal within the folder that contains this `default.nix`
+will drop the user into a new shell. It should be noted that it uses the
+`nixpkgs` from that same date. It is also possible to configure IDEs to
+dynamically load this new shell to provide the proper development tooling.
 
@@
 
-However, installing from source might fail, especially if the package needs
-to be compiled.
+However, installing from source might fail, especially if the package needs to
+be compiled.
 
@@
 is impractical. While CRAN and Bioconductor update daily, the [R]{.proglang}
-packages in `nixpkgs` are only updated with new [R]{.proglang} releases. This limitation is
-due to [Nix]{.pkg}’s governance as a Linux distribution package manager.
-
-The `rstats-on-nix` fork allows me to circumvent these limitations. For example, it provides
-daily snapshots of CRAN. Each day, the [R]{.proglang} package set is updated and
-committed to a dated branch using GitHub Actions. Users can select a specific
-date with:
+packages in `nixpkgs` are only updated with new [R]{.proglang} releases. This
+limitation is due to [Nix]{.pkg}’s governance as a Linux distribution package
+manager.
+
+The `rstats-on-nix` fork allows me to circumvent these limitations. For example,
+it provides daily snapshots of CRAN. Each day, the [R]{.proglang} package set is
+updated and committed to a dated branch using GitHub Actions. Users can select a
+specific date with:
 
@@
 (`boost`), a pull request was submitted to the official `nixpkgs` repository. I
-backported the fix to the `rstats-on-nix` fork, making RStudio available to users of
-[rix]{.pkg} earlier than upstream, as merging PRs in the official repository can take some
-time.
-
-I have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as March 2019. The
-process involves checking out a `nixpkgs` commit on the selected date, updating
-the [R]{.proglang} package set using Posit CRAN and Bioconductor snapshots,
-backporting fixes, and ensuring popular packages work on both x86-linux
-(including WSL2) and aarch64-darwin (Apple Silicon). These changes are committed
-to a dated branch in `rstats-on-nix/nixpkgs`. Users can see all available dates
-with `rix::available_dates()`.
+backported the fix to the `rstats-on-nix` fork, making RStudio available to
+users of [rix]{.pkg} earlier than upstream, as merging PRs in the official
+repository can take some time.
+
+I have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as
+March 2019. The process involves checking out a `nixpkgs` commit on the selected
+date, updating the [R]{.proglang} package set using Posit CRAN and Bioconductor
+snapshots, backporting fixes, and ensuring popular packages work on both
+x86-linux (including WSL2) and aarch64-darwin (Apple Silicon). These changes are
+committed to a dated branch in `rstats-on-nix/nixpkgs`. Users can see all
+available dates with `rix::available_dates()`.
 
@@
 executing the analysis workflow within that environment, which is the role of
-the [rixpress]{.pkg} package (or [ryxpress]{.pkg} for [Python]{.proglang} users).
+the [rixpress]{.pkg} package (or [ryxpress]{.pkg} for [Python]{.proglang}
+users).
 
@@
 in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
-report. The code for this example can be found in the
-[following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
+report. The code for this example can be found in the [following
+repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
 with additional details provided in the Appendix of this manuscript.
@@
 is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
-`functions.R`). Further details are provided as well as step-by-step instructions
-in the Appendix. The high-level orchestration script demonstrates how `{rixpress}` defines a
-granular, multi-step machine learning workflow that crosses language boundaries:
+`functions.R`). Further details are provided as well as step-by-step
+instructions in the Appendix. The high-level orchestration script demonstrates
+how `{rixpress}` defines a granular, multi-step machine learning workflow that
+crosses language boundaries:
 
@@
 The `rxp_populate()` function translates this [R]{.proglang} list into a
-`pipeline.nix` file, which declaratively defines the entire workflow.
-Data flows from derivation to derivation by being serialized into the
-efficient and language-agnostic  Arrow format using the pair of `encoder/decoder`
-functions. Other universal formats, such as `csv` or `json` could have been used.
-
-As a sidenote: [Python]{.proglang} users who wish to use [ryxpress]{.pkg}
-define pipelines using the same [R]{.proglang}-based DSL shown above. This design
-choice keeps the pipeline definition language consistent across both [R]{.proglang}
-and [Python]{.proglang} ecosystems. [ryxpress]{.pkg} calls [R]{.proglang} and
-[rixpress]{.pkg} under the hood to generate and build the `pipeline.nix` file.
+`pipeline.nix` file, which declaratively defines the entire workflow. Data flows
+from derivation to derivation by being serialized into the efficient and
+language-agnostic Arrow format using the pair of `encoder/decoder` functions.
+Other universal formats, such as `csv` or `json` could have been used.
+
+As a sidenote: [Python]{.proglang} users who wish to use [ryxpress]{.pkg} define
+pipelines using the same [R]{.proglang}-based DSL shown above. This design
+choice keeps the pipeline definition language consistent across both
+[R]{.proglang} and [Python]{.proglang} ecosystems. [ryxpress]{.pkg} calls
+[R]{.proglang} and [rixpress]{.pkg} under the hood to generate and build the
+`pipeline.nix` file.
 
@@
 recompute steps with changed inputs or code. This provides the efficiency of
-[targets]{.pkg} with polyglot support and bit-for-bit reproducibility.
-Artifacts can be inspected interactively in [R]{.proglang} using
+[targets]{.pkg} with polyglot support and bit-for-bit reproducibility. Artifacts
+can be inspected interactively in [R]{.proglang} using
 `rxp_read("artifact_name")` or `rxp_load("artifact_name")`.
@@
 includes several additional features not covered here for brevity. As mentioned
-previously, it is also possible to configure popular IDEs to work interactively and seamlessly with
-both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth, reproducible workflow
-from within the development environment. Detailed setup instructions are
-provided in the vignettes of both packages.
+previously, it is also possible to configure popular IDEs to work interactively
+and seamlessly with both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth,
+reproducible workflow from within the development environment. Detailed setup
+instructions are provided in the vignettes of both packages.
 
@@
 environment. This ensures deep reproducibility, efficient caching, and seamless
-orchestration of polyglot analyses. Most importantly, it provides
-native polyglot support without the manual coordination required by
-containerization approaches, making complex multi-language workflows
-accessible to researchers without systems administration expertise.
-
-While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide significant steps forward, it is
-important to acknowledge current limitations.
-[rixpress]{.pkg} (and [ryxpress]{.pkg}) in particular requires further development, especially to enhance
-the ease of debugging complex pipelines. Its visualization capabilities are also still
-fairly limited; for instance, they do not currently indicate outdated derivations when
-a user modifies code that would necessitate a rebuild. These areas represent important
-avenues for future work to further mature the framework and improve the user experience.
+orchestration of polyglot analyses. Most importantly, it provides native
+polyglot support without the manual coordination required by containerization
+approaches, making complex multi-language workflows accessible to researchers
+without systems administration expertise.
+
+While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide
+significant steps forward, it is important to acknowledge current limitations.
+[rixpress]{.pkg} (and [ryxpress]{.pkg}) in particular requires further
+development, especially to enhance the ease of debugging complex pipelines. Its
+visualization capabilities are also still fairly limited; for instance, they do
+not currently indicate outdated derivations when a user modifies code that would
+necessitate a rebuild. These areas represent important avenues for future work
+to further mature the framework and improve the user experience.
 
@@
 on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, I am
-grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of [rix]{.pkg},
-and to William Landau and Anthony Martinez for their reviews of [rixpress]{.pkg}. I
-also acknowledge the contributions of Philipp Baumann, [rix]{.pkg}’s co-author,
-Richard J. Acton, Jordi Rosell, Elio Campitelli, László Kupcsik, and
-Michael Heming for [rix]{.pkg}. Their expertise and feedback greatly
-improved the quality and usability of these packages.
+grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of
+[rix]{.pkg}, and to William Landau and Anthony Martinez for their reviews of
+[rixpress]{.pkg}. I also acknowledge the contributions of Philipp Baumann,
+[rix]{.pkg}’s co-author, Richard J. Acton, Jordi Rosell, Elio Campitelli, László
+Kupcsik, and Michael Heming for [rix]{.pkg}. Their expertise and feedback
+greatly improved the quality and usability of these packages. I would also like
+to thank Pol Dellaiera for providing feedback on the manuscript.
 
@@
 This command utilizes the `default.nix` file to build the exact computational
-environment required to compile the manuscript. To ensure continuous reproducibility,
-the paper is also automatically re-compiled via GitHub Actions on every push to the
-repository, using the same Nix environment.
+environment required to compile the manuscript. To ensure continuous
+reproducibility, the paper is also automatically re-compiled via GitHub Actions
+on every push to the repository, using the same Nix environment.
 
@@
 
-This appendix provides a conceptual walk-through of the polyglot pipeline example discussed in Section 5.
-The pipeline simulates a Real Business Cycle (RBC) model in [Julia]{.proglang},
-trains an [XGBoost]{.pkg} model in [Python]{.proglang}, visualizes the results in [R]{.proglang},
-and compiles a final report with [Quarto]{.pkg}.
-
-The complete, runnable source code for this example is available in the paper's GitHub repository.
-A shell script, `run_polyglot_example.sh`, is provided to automate the entire process from environment
-creation to final output, as described at the end of this section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/polyglot-example/run_polyglot_example.sh}].
+This appendix provides a conceptual walk-through of the polyglot pipeline
+example discussed in Section 5. The pipeline simulates a Real Business Cycle
+(RBC) model in [Julia]{.proglang}, trains an [XGBoost]{.pkg} model in
+[Python]{.proglang}, visualizes the results in [R]{.proglang}, and compiles a
+final report with [Quarto]{.pkg}.
+
+The complete, runnable source code for this example is available in the paper's
+GitHub repository. A shell script, `run_polyglot_example.sh`, is provided to
+automate the entire process from environment creation to final output, as
+described at the end of this
+section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/polyglot-example/run_polyglot_example.sh}].
 
@@
 
-The project is organized into a set of scripts, each with a distinct responsibility.
-The core logic is separated from the environment definition and pipeline orchestration.
+The project is organized into a set of scripts, each with a distinct
+responsibility. The core logic is separated from the environment definition and
+pipeline orchestration.
 
@@
 
-The foundation of the project is the reproducible environment, defined in `gen-env.R`.
-This script uses the `rix()` function to programmatically generate a `default.nix` file.
-This file serves as a complete blueprint for the computational environment, specifying:
-
-*   The exact versions of [R]{.proglang}, [Python]{.proglang}, and [Julia]{.proglang}.
-*   A list of required packages for each language
-  (e.g., [ggplot2]{.pkg} for [R]{.proglang}, [xgboost]{.pkg} for [Python]{.proglang}, [DataFrames]{.pkg} for [Julia]{.proglang}).
+The foundation of the project is the reproducible environment, defined in
+`gen-env.R`. This script uses the `rix()` function to programmatically generate
+a `default.nix` file. This file serves as a complete blueprint for the
+computational environment, specifying:
+
+*   The exact versions of [R]{.proglang}, [Python]{.proglang}, and
+    [Julia]{.proglang}.
+*   A list of required packages for each language (e.g., [ggplot2]{.pkg} for
+  [R]{.proglang}, [xgboost]{.pkg} for [Python]{.proglang}, [DataFrames]{.pkg}
+  for [Julia]{.proglang}).
 *   System-level dependencies like [Quarto]{.pkg}.
@@
 it is possible to generate the `default.nix` by sourcing `gen-env.R`. Then, one
-leaves this temporary shell, and builds the environment using the command `nix-shell`,
-which also drop the user into the development shell.
+leaves this temporary shell, and builds the environment using the command
+`nix-shell`, which also drop the user into the development shell.
 
@@
 
-The scientific logic for each stage of the pipeline is encapsulated in the separate
-helper scripts within the `functions/` directory.
-
-*   The RBC Model Simulation (`functions.jl`): This [Julia]{.proglang} script contains a single pure
-  function that implements the state-space solution to the RBC model, based on @depaoli2009rbc.
-  It takes the model's economic parameters as inputs and returns the simulated time-series
-  data as a [DataFrame]{.pkg}.
-*   The [XGBoost]{.pkg} Forecasting Model (`functions.py`): This [Python]{.proglang} script handles the
-  complete machine learning workflow through a series of modular functions. Its responsibilities include
-  feature engineering (creating lagged variables), splitting the data into training and testing sets,
-  training the [XGBoost]{.pkg} model, and generating predictions.
-*   The Visualization (`functions.R`): This [R]{.proglang} script contains a function that uses
-  [ggplot2]{.pkg} to create the final visualization. It is designed to take the data frame of
-  actual and predicted values produced by the Python script and generate a plot comparing the two series.
+The scientific logic for each stage of the pipeline is encapsulated in the
+separate helper scripts within the `functions/` directory.
+
+*   The RBC Model Simulation (`functions.jl`): This [Julia]{.proglang} script
+  contains a single pure function that implements the state-space solution to
+  the RBC model, based on @depaoli2009rbc. It takes the model's economic
+  parameters as inputs and returns the simulated time-series data as a
+  [DataFrame]{.pkg}.
+*   The [XGBoost]{.pkg} Forecasting Model (`functions.py`): This
+  [Python]{.proglang} script handles the complete machine learning workflow
+  through a series of modular functions. Its responsibilities include feature
+  engineering (creating lagged variables), splitting the data into training and
+  testing sets, training the [XGBoost]{.pkg} model, and generating predictions.
+*   The Visualization (`functions.R`): This [R]{.proglang} script contains a
+  function that uses [ggplot2]{.pkg} to create the final visualization. It is
+  designed to take the data frame of actual and predicted values produced by the
+  Python script and generate a plot comparing the two series.
 
@@
 
-The entire workflow is defined and orchestrated by `gen-pipeline.R`. This script acts as the master plan,
-using functions from the [rixpress]{.pkg} package to define each computational step as a *derivation*.
-It declaratively outlines the dependencies between steps:
+The entire workflow is defined and orchestrated by `gen-pipeline.R`. This script
+acts as the master plan, using functions from the [rixpress]{.pkg} package to
+define each computational step as a *derivation*. It declaratively outlines the
+dependencies between steps:
 
 1.  It begins by defining the RBC model parameters in [Julia]{.proglang}.
-2.  It specifies that the RBC simulation in `functions.jl` depends on these parameters.
-3.  It then defines the series of [Python]{.proglang} steps (feature preparation, training, prediction),
-  making each one dependent on the output of the previous one. The first [Python]{.proglang} step is explicitly
-  made dependent on the output from the [Julia]{.proglang} simulation. [rixpress]{.pkg} handles the passing of data
-  between languages, in this case using the [Arrow]{.pkg} file format for efficiency.
-4.  Finally, it defines the **R visualization** step, which depends on the final predictions from
-  the [Python]{.proglang} model, and a **Quarto rendering** step that depends on the generated plot.
-
-When this script is run in the development shell (by executing `source("gen-pipeline.R")`),
-[rixpress]{.pkg} translates the declared pipeline into a master [Nix]{.proglang} expression
-that [Nix]{.pkg} can execute, automatically handling the caching of results and re-running only the necessary
+2.  It specifies that the RBC simulation in `functions.jl` depends on these
+    parameters.
+3.  It then defines the series of [Python]{.proglang} steps (feature
+  preparation, training, prediction), making each one dependent on the output of
+  the previous one. The first [Python]{.proglang} step is explicitly made
+  dependent on the output from the [Julia]{.proglang} simulation.
+  [rixpress]{.pkg} handles the passing of data between languages, in this case
+  using the [Arrow]{.pkg} file format for efficiency.
+4.  Finally, it defines the **R visualization** step, which depends on the final
+  predictions from the [Python]{.proglang} model, and a **Quarto rendering**
+  step that depends on the generated plot.
+
+When this script is run in the development shell (by executing
+`source("gen-pipeline.R")`), [rixpress]{.pkg} translates the declared pipeline
+into a master [Nix]{.proglang} expression that [Nix]{.pkg} can execute,
+automatically handling the caching of results and re-running only the necessary
 steps if a piece of code or data changes.
 
-To build the pipeline from an interactive [Python]{.proglang} session, one would execute the following lines:
+To build the pipeline from an interactive [Python]{.proglang} session, one would
+execute the following lines:
 
@@
 This script automates the full process:
-
-1.  It first executes `gen-env.R` inside a temporary [Nix]{.pkg} shell to build the `default.nix` file.
-2.  It then executes `gen-pipeline.R` inside the newly defined environment. This triggers [Nix]{.pkg} to run the
-  entire polyglot pipeline in the correct order.
-
-Upon completion, the script will have generated all intermediate artifacts and the final `readme.html` report found in
-the `pipeline-output` folder containing the visualization.
+1.  It first executes `gen-env.R` inside a temporary [Nix]{.pkg} shell to build
+    the `default.nix` file.
+2.  It then executes `gen-pipeline.R` inside the newly defined environment. This
+  triggers [Nix]{.pkg} to run the entire polyglot pipeline in the correct order.
+Upon completion, the script will have generated all intermediate artifacts and
+the final `readme.html` report found in the `pipeline-output` folder containing
+the visualization.

### 2025-10-23 paper.qmd
fixes #3

@@
 to provide step-by-step installation instructions for the required software
-stack.
+stack, an omission often driven by institutional constraints. Journals, for
+instance, can inadvertently work against reproducibility by imposing strict page
+or word limits that leave no room for the necessary technical documentation,
+effectively discouraging thoroughness in the name of brevity.
 
@@
   triggers [Nix]{.pkg} to run the entire polyglot pipeline in the correct order.
+
 Upon completion, the script will have generated all intermediate artifacts and

### 2025-10-23 paper.qmd
thank Edvin Syk
github pages

@@
         keep-tex: true
-    jss-html: default
+    jss-html:
+        embed-resources: true
+        toc: true
 author:
@@
 greatly improved the quality and usability of these packages. I would also like
-to thank Pol Dellaiera for providing feedback on the manuscript.
+to thank Pol Dellaiera and Edvin Syk for providing feedback on the manuscript.
 

### 2025-10-23 paper.qmd
sharing the link

@@
 
-This command utilizes the `default.nix` file to build the exact computational
-environment required to compile the manuscript. To ensure continuous
-reproducibility, the paper is also automatically re-compiled via GitHub Actions
-on every push to the repository, using the same Nix environment.
+The `default.nix` file defines the exact computational environment required to
+compile the manuscript, ensuring full reproducibility of the build process. To
+guarantee the reproducibility of the manuscript, it is automatically
+recompiled via GitHub Actions upon each commit, using the same Nix environment.
+An HTML version is additionally deployed to GitHub Pages, providing an
+accessible format for viewing on smaller devices at this
+[link](https://b-rodrigues.github.io/rix_paper/)^[[https://b-rodrigues.github.io/rix_paper/](https://b-rodrigues.github.io/rix_paper/)].
 

### 2025-10-23 paper.qmd
code formatting for html

@@
 
+::: {.content-hidden when-format="pdf"}
+```R
+sessionInfo()
+```
+:::
+
+::: {.content-hidden when-format="pdf"}
+```R
+R version 4.3.2 (2023-10-31)
+Platform: aarch64-unknown-linux-gnu (64-bit)
+Running under: Ubuntu 22.04.3 LTS
+
+Matrix products: default
+BLAS:   /usr/lib/aarch64-linux-gnu/openblas-pthread/libblas.so.3
+LAPACK: /usr/lib/aarch64-linux-gnu/openblas-pthread/libopenblasp[...]
+
+locale:
+ LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C
+ LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8
+ ...
+
+attached base packages:
+ stats     graphics  grDevices utils     datasets  methods   base
+
+other attached packages:
+ nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163
+
+loaded via a namespace (and not attached):
+ compiler_4.3.2  Matrix_1.6-1.1  tools_4.3.2     splines_4.3.2
+ grid_4.3.2      lattice_0.21-9
+```
+:::
+
 ```{=tex}
@@
 
+::: {.content-hidden when-format="pdf"}
+```R
+groundhog.library("
+ library('purrr')
+ library('ggplot2')",
+ "2017-10-04",
+ tolerate.R.version = "4.2.2")
+```
+:::
+
 ```{=tex}
@@
 CRAN snapshots, configured via the `.Rprofile` file:
+
+::: {.content-hidden when-format="pdf"}
+```R
+options(repos =
+  c(REPO_NAME =
+    "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
+  )
+)
+```
+:::
 
@@
 
+::: {.content-hidden when-format="pdf"}
+```nix
+let
+  pkgs = import (fetchTarball
+    "https://github.com/NixOS/nixpkgs/archive/976fa336.tar.gz"
+  ) {};
+  system_packages = builtins.attrValues {
+    inherit (pkgs) R;
+  };
+in
+  pkgs.mkShell {
+    buildInputs = [ system_packages ];
+    shellHook = "R --vanilla";
+  }
+```
+:::
+
 ```{=tex}
@@
 
+::: {.content-hidden when-format="pdf"}
+```bash
+nix-shell -I \
+  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/tarball/2025-10-20 -p \
+  R rPackages.rix
+```
+:::
+
 ```{=tex}
@@
 
+::: {.content-hidden when-format="pdf"}
+```r
+library('rix')
+
+rix(r_ver = "4.3.1",
+  r_pkgs = c("dplyr", "chronicler"),
+  project_path = ".",
+  overwrite = TRUE)
+```
+:::
+
 ```{=tex}
@@
 instead of a specific [R]{.proglang} version:
+
+::: {.content-hidden when-format="pdf"}
+```r
+rix(date = "2025-10-20",
+  r_pkgs = c("rix", "dplyr", "chronicler", "AER@1.2-8"),
+  system_pkgs = c("quarto", "git"),
+  tex_pkgs = c(
+        "amsmath",
+        "framed",
+        "fvextra",
+        "environ",
+        "fontawesome5",
+        "orcidlink",
+        "pdfcol",
+        "tcolorbox",
+        "tikzfill"
+  ),
+  git_pkgs = list(
+      package_name = "fusen",
+      repo_url = "https://github.com/ThinkR-open/fusen",
+      commit = "60346860111be79fc2beb33c53e195f97504a667"
+  ),
+  ide = "positron",
+  project_path = ".",
+  overwrite = TRUE)
+```
+:::
 
@@
 
+::: {.content-hidden when-format="pdf"}
+```bash
+cd /path/to/project/ && nix-shell default.nix --run "Rscript analysis.R"
+```
+:::
+
 ```{=tex}
@@
 
+::: {.content-hidden when-format="pdf"}
+```r
+rix(..., r_pkgs = "dplyr@1.0.7", ...)
+```
+:::
+
 ```{=tex}
@@
 specific date with:
+
+::: {.content-hidden when-format="pdf"}
+```r
+rix(date = "2024-12-14", ...)
+```
+:::
 
@@
 crosses language boundaries:
+
+::: {.content-hidden when-format="pdf"}
+```r
+library('rixpress')
+
+pipeline_steps <- list(
+  # STEP 0: Define RBC Model Parameters in Julia
+  rxp_jl(alpha, 0.3),
+  rxp_jl(beta, 1 / 1.01),
+  # ... (other parameters omitted for brevity) ...
+
+  # STEP 1: Julia - Simulate the RBC model
+  rxp_jl(
+    name = simulated_rbc_data,
+    expr = "simulate_rbc_model(alpha, beta, delta, rho, sigma, sigma_z)",
+    user_functions = "functions/functions.jl",
+    encoder = "arrow_write"
+  ),
+
+  # STEP 2.1: Python - Prepare features
+  rxp_py(
+    name = processed_data,
+    expr = "prepare_features(simulated_rbc_data)",
+    user_functions = "functions/functions.py",
+    decoder = "pyarrow.feather.read_feather"
+  ),
+
+  # STEP 2.2: Python - Split data (X_train, y_train, etc.)
+  rxp_py(name = X_train, expr = "get_X_train(processed_data)", ...),
+  # ... (other data splits omitted for brevity) ...
+
+  # STEP 2.3: Python - Train the XGBoost model
+  rxp_py(
+    name = trained_model,
+    expr = "train_model(X_train, y_train)",
+    user_functions = "functions/functions.py"
+  ),
+
+  # STEP 2.4: Python - Make predictions and format results
+  # ... (prediction and formatting steps omitted for brevity) ...
+  rxp_py(
+    name = final_predictions_df,
+    expr = "format_results(y_test, model_predictions)",
+    user_functions = "functions/functions.py",
+    encoder = "save_arrow"
+  ),
+
+  # STEP 3: R - Visualize the predictions
+  rxp_r(
+    name = output_plot,
+    expr = plot_predictions(final_predictions_df),
+    user_functions = "functions/functions.R",
+    decoder = arrow::read_feather
+  ),
+
+  # STEP 4: Quarto - Compile the final report
+  rxp_qmd(
+    name = final_report,
+    qmd_file = "readme.qmd"
+  )
+)
+
+# Generate the 'pipeline.nix' file from the R list
+rxp_populate(pipeline_steps, build = TRUE)
+```
+:::
 
@@
 
+::: {.content-hidden when-format="pdf"}
+```bash
+nix-shell --run "quarto render paper.qmd --to jss-pdf"
+```
+:::
+
 ```{=tex}
@@
 By dropping into a temporary shell using the following command:
+
+::: {.content-hidden when-format="pdf"}
+```bash
+nix-shell -I \
+  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/tarball/2025-10-20 -p \
+  R rPackages.rix
+```
+:::
 
@@
 
+::: {.content-hidden when-format="pdf"}
+```python
+from ryxpress import rxp_make
+rxp_make()
+```
+:::
+
 ```{=tex}

### 2025-10-23 paper.qmd
link to pdf
container -> docker container

@@
 Once the image is built, the analysis can be executed inside a *container*,
-which is a running instance of the image. Containers are typically run
+which is a running instance of the image. [Docker]{.pkg} containers are typically run
 non-interactively, allowing the complete computational environment to be
@@
 
-1. Poor interactive support: While containers can be modified at runtime,
+1. Poor interactive support: While [Docker]{.pkg} containers can be modified at runtime,
   changes are lost when the container stops unless explicitly saved. Running
@@
   produce distinct outputs. In the experimental evaluation conducted in the
-  thesis, even with pinned base images, checksums of the resulting containers
-  varied across builds and machines. Thus, while Docker offers partial
-  reproducibility: strong at run time (space) and weak at build time (time), it
-  cannot on its own ensure verifiable, deterministic software artifacts.
-  Achieving true reproducibility requires pinning image digests, freezing
-  dependency versions, and employing deterministic build systems such as Nix or
-  Guix for complete control over both space and time [@Malka_2024].
+  thesis, even with pinned base images, checksums of the resulting [Docker]{.pkg}
+  containers varied across builds and machines. Thus, while Docker offers
+  partial reproducibility: strong at run time (space) and weak at build time
+  (time), it cannot on its own ensure verifiable, deterministic software
+  artifacts. Achieving true reproducibility requires pinning image digests,
+  freezing dependency versions, and employing deterministic build systems such
+  as Nix or Guix for complete control over both space and time [@Malka_2024].
 
@@
 [link](https://b-rodrigues.github.io/rix_paper/)^[[https://b-rodrigues.github.io/rix_paper/](https://b-rodrigues.github.io/rix_paper/)].
+The PDF version of the manuscript can be found at
+[link](https://b-rodrigues.github.io/rix_paper/paper.pdf)^[[https://b-rodrigues.github.io/rix_paper/paper.pdf](https://b-rodrigues.github.io/rix_paper/paper.pdf)].
 

### 2025-10-23 paper.qmd
more intro to nix

@@
 bundled with all their dependencies and transitive dependencies
-[@dolstra2004nix]. This creates self-contained software environments that work
+[@dolstra2004nix]. Think of installing a package like packing for a trip.
+Traditional package managers are like packing just your clothes, assuming you'll
+find a toothbrush and soap at your destination. [Nix]{.pkg} packs everything you
+might need—the entire self-contained kit. This is what [Nix]{.pkg} calls a
+*component closure*.
+
+This creates self-contained software environments that work
 identically across machines and over time. [Nix]{.pkg} can replace
@@
 
-Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports
-reproducibility. [Nix]{.pkg} is a *functional* package manager in the
-programming sense: it treats software builds as pure functions. Given the same
-inputs (source code, dependencies, build scripts), a [Nix]{.pkg} build will
-always produce the identical output, regardless of when or where it's executed.
+Pinning is essential, but it is not the only mechanism through which [Nix]{.pkg}
+achieves reproducibility. [Nix]{.pkg} adopts principles from functional
+programming, conceptualizing software builds as *pure functions*. In
+mathematics, a pure function is defined as one that always produces the same
+output for a given input—for instance, if \( f(x) = x^2 \), then \( f(2) = 4 \)
+invariably holds. Analogously, [Nix]{.pkg} ensures that, given identical inputs
+(such as source code, dependencies, and build instructions), the resulting
+output (the built package) remains identical, irrespective of the environment or
+time of execution. In this sense, [Nix]{.pkg} functions as a *functional package
+manager*, applying the deterministic properties of pure functions to software
+construction.
+
 This is achieved by disallowing side effects and ensuring that builds operate
@@
 largely abstracted away. Additionally, [Nix]{.pkg} supports multiple versions or
-*variants* of a package on the same system. Each package installed into the "Nix
-store" (typically in `/nix/store`) has a unique identifier derived from a hash
-of all its inputs. This isolation ensures that projects always use their
-intended software versions, allowing multiple versions of [R]{.proglang} to
-coexist and preventing "dependency hell" and global environment pollution.
+*variants* of a package on the same system.
+
+[Nix]{.pkg} installs all packages into a centralized, content-addressed
+directory called the [Nix]{.pkg} store (typically `/nix/store`). Each package
+gets a unique subdirectory whose name includes a hash of all its inputs: source
+code, dependencies, build instructions. This means
+`/nix/store/abc123-dplyr-1.0.0` and `/nix/store/xyz789-dplyr-1.1.0` can coexist
+without conflicts.
+
+This isolation ensures that projects always use their intended software
+versions, allowing multiple versions of [R]{.proglang} to coexist and preventing
+"dependency hell" and global environment pollution.
 
@@
 Once the build completes, the user can enter the interactive shell with
-`nix-shell`. This shell contains all the packages specified in `default.nix` and
+`nix-shell`, which executes following steps:
+
+- Nix reads the expression and identifies what needs to be installed;
+- It checks if these packages already exist in `/nix/store`;
+- If not, it downloads or builds them (with all dependencies);
+- It creates a new shell environment and updates the `$PATH` to include the
+  right binaries from `/nix/store`.
+- The user is now in an interactive [Nix]{.pkg} shell session.
+
+This shell contains all the packages specified in `default.nix` and
 can be used for development, similar to activating a virtual environment in the

### 2025-10-23 paper.qmd
more on python

@@
   [Julia]{.proglang}, and system dependencies. Building on these environments,
-  [rixpress]{.pkg} orchestrates polyglot pipelines where each computational step
-  runs in its own hermetically sealed environment, with automatic caching and
-  dependency tracking. This approach enables researchers to develop analyses
-  interactively while maintaining bit-for-bit reproducibility, supports
-  collaboration across heterogeneous computational environments, and ensures
-  that analyses remain executable years into the future.
+  [rixpress]{.pkg} (and its [Python]{.proglang} port, [ryxpress]{.pkg})
+  orchestrates polyglot pipelines where each computational step runs in its own
+  hermetically sealed environment, with automatic caching and dependency
+  tracking. This approach enables researchers to develop analyses interactively
+  while maintaining bit-for-bit reproducibility, supports collaboration across
+  heterogeneous computational environments, and ensures that analyses remain
+  executable years into the future.
 
@@
 data analysis in [R]{.proglang}. [sf]{.pkg} relies on several complex external
-system libraries such as [GDAL]{.proglang}, [GEOS]{.proglang}, and
-[PROJ]{.proglang}. Manually installing and configuring these libraries to be
+system libraries such as [GDAL]{.pkg}, [GEOS]{.pkg}, and
+[PROJ]{.pkg}. Manually installing and configuring these libraries to be
 compatible with [sf]{.pkg} can be challenging and platform-specific. With
@@
 
+It is also possible to include [Python]{.proglang} and [Julia]{.proglang}
+packages using the `py_conf` and `jl_conf` arguments respectively.
+[Julia]{.proglang} packages work the same way as [R]{.proglang} packages: they
+are pinned to their versions as of the specified date.
+
+[Python]{.proglang} packages, however, require special handling. Unlike CRAN or
+the Julia package registry, PyPI (the [Python]{.proglang} Package Index)
+contains over 500,000 packages with highly variable quality—many lack proper
+build specifications, have broken dependencies, or are abandoned. Because of
+this heterogeneity, [Nix]{.pkg} maintainers cannot automatically package all of
+PyPI. Instead, they manually curate and package individual [Python]{.proglang}
+packages that meet quality standards. This means some [Python]{.proglang}
+packages, or specific versions, may not be available through [Nix]{.pkg}.
+
+For projects requiring [Python]{.proglang} packages not yet available in
+[Nix]{.pkg}, [rix]{.pkg} supports integration with [uv]{.pkg}, a modern
+[Python]{.proglang} package manager. This allows users to combine Nix's
+system-level reproducibility (Python interpreter, system libraries) with
+[uv]{.pkg}'s comprehensive package availability. While this hybrid approach
+sacrifices some of Nix's deterministic build guarantees for Python packages, it
+provides a pragmatic solution when the needed packages are unavailable in
+`nixpkgs`.
+
+Users can include [uv]{.pkg} as a system package:
+
+::: {.content-hidden when-format="pdf"}
+```r
+rix(date = "2025-10-20",
+  r_pkgs = c("rix", "dplyr", "chronicler"),
+  system_pkgs = c("uv"),
+  project_path = ".",
+  overwrite = TRUE)
+```
+:::
+```{=tex}
+\begin{CodeInput}
+R> rix(date = "2025-10-20",
++    r_pkgs = c("rix", "dplyr", "chronicler"),
++    system_pkgs = c("uv"),
++    project_path = ".",
++    overwrite = TRUE)
+\end{CodeInput}
+```
+
+This hybrid approach offers complementary strengths: [Nix]{.pkg} manages the
+[Python]{.proglang} interpreter, system libraries (e.g., [GDAL]{.pkg},
+[HDF5]{.pkg}), and [R]{.proglang} packages with full determinism, while
+[uv]{.pkg} handles [Python]{.proglang} packages through its own lockfile
+mechanism (`uv.lock`). Within the Nix environment, users run standard [uv]{.pkg}
+commands (e.g., `uv pip install pandas==2.0.0`) to install [Python]{.proglang}
+packages. The [uv]{.pkg} lockfile records exact package versions and their
+hashes, providing reproducibility for the [Python]{.proglang} package layer.
+
+While this approach doesn't achieve Nix's level of build-time determinism for
+[Python]{.proglang} packages ([uv]{.pkg} still downloads wheels or builds
+packages at install time rather than using pre-built, content-addressed
+artifacts) it provides a pragmatic balance: comprehensive [Python]{.proglang}
+package availability with reasonable reproducibility guarantees. For polyglot
+projects, this means [R]{.proglang} and system-level dependencies remain fully
+reproducible through [Nix]{.pkg}, while [Python]{.proglang} packages gain the
+reproducibility that [uv]{.pkg} can provide.
+
 [rix]{.pkg} can generate [Nix]{.pkg} expressions even if [Nix]{.pkg} is not
@@
 
-The user defines the pipeline in an [R]{.proglang} script (`gen-pipeline.R`)
-using functions inspired by [targets]{.pkg}. The underlying logic for each step
-is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
-`functions.R`). Further details are provided as well as step-by-step
-instructions in the Appendix. The high-level orchestration script demonstrates
-how `{rixpress}` defines a granular, multi-step machine learning workflow that
-crosses language boundaries:
+For new projects, [rixpress]{.pkg} provides `rxp_init()`, which generates the
+necessary boilerplate structure including a pipeline definition file and
+template helper scripts. Users then define their pipeline in the script
+`gen-pipeline.R` using functions inspired by [targets]{.pkg}, and the
+computational environment the pipeline will be executed in the `gen-env.R`
+script. If functions are needed for pipeline steps, these are to be specified, in
+language-specific helper scripts: `functions.jl` for [Julia]{.proglang},
+`functions.py` for [Python]{.proglang}, and `functions.R` for [R]{.proglang}.
+This separation keeps the pipeline definition clean and readable while
+encapsulating the scientific logic in appropriate language-native files.
+Importantly, these helper scripts are standard language files—they contain no
+[rixpress]{.pkg}-specific code and can be reused in other projects or executed
+independently outside of [Nix]{.pkg}.
+
+The following example demonstrates how [rixpress]{.pkg} orchestrates a granular,
+multi-step data science workflow that crosses language boundaries. (Complete
+setup instructions and a detailed walkthrough are provided in Appendix Section
+@sec-polyglot-example.)
 
@@
 
+Computation steps, referred to as *derivations*, are defined using the functions
+`rxp_r()`, `rxp_jl()`, and `rxp_py()`, corresponding to [R]{.proglang},
+[Julia]{.proglang}, and [Python]{.proglang} environments, respectively. Variants
+of these functions suffixed with `_file()` designate *import derivations*, which
+introduce external data into the pipeline. It is important to note that any data
+imported in this way is stored within the Nix store, ensuring reproducibility
+but also persisting the data in the build environment. The function `rxp_qmd()`
+is used to compile a [Quarto]{.pkg} document (there is also `rxp_rmd()` which
+serves the same purpose for [R Markdown]{.pkg} documents).
+
 The `rxp_populate()` function translates this [R]{.proglang} list into a
@@
 [R]{.proglang} and [rixpress]{.pkg} under the hood to generate and build the
-`pipeline.nix` file.
+`pipeline.nix` file. To use [ryxpress]{.pkg}, one should simply add this
+package to the `py_conf` argument of `rix()`.
 
@@
 
-For Python users, a port called [ryxpress]{.pkg} allows building the same
-pipelines and inspecting outputs from Python sessions. [rixpress]{.pkg} also
-includes several additional features not covered here for brevity. As mentioned
-previously, it is also possible to configure popular IDEs to work interactively
-and seamlessly with both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth,
-reproducible workflow from within the development environment. Detailed setup
-instructions are provided in the vignettes of both packages.
+[rixpress]{.pkg} also includes several additional features not covered here for
+brevity. As mentioned previously, it is also possible to configure popular IDEs
+to work interactively and seamlessly with both [rix]{.pkg} and [rixpress]{.pkg},
+enabling a smooth, reproducible workflow from within the development
+environment. Detailed setup instructions are provided in the vignettes of both
+packages.
 

### 2025-10-23 paper.qmd
formatting

@@
 common limitation: none addresses system-level dependencies or supports polyglot
-environments. A project using [R]{.proglang}, [Python]{.proglang}, and system
-tools like [Quarto]{.pkg} or \LaTeX requires coordinating multiple package
-managers ([renv]{.pkg} for R, virtual environments for Python, manual
-installation for system tools), each with its own configuration and potential
-for conflicts.
-
-The most comprehensive approach to reproducibility is containerization with
+environments. A project using [R]{.proglang} [@r-core-team], [Python]{.proglang}
+[@python3], and system tools like [Quarto]{.pkg} or \LaTeX requires coordinating
+multiple package managers ([renv]{.pkg} for R, virtual environments for
+[Python]{.proglang}, manual installation for system tools), each with its own
+configuration and potential for conflicts.
+
+The most comprehensive approach to reproducibility is containerisation with
 [Docker]{.pkg}. [Docker]{.pkg} packages a *data product* together with all its
@@
 
-1. Poor interactive support: While [Docker]{.pkg} containers can be modified at runtime,
-  changes are lost when the container stops unless explicitly saved. Running
-  graphical applications like RStudio Desktop from containers requires complex
-  X11 configuration that works reliably only on Linux. Instead, web-based IDEs
-  (like RStudio Server) are used, but come with the inherent limitations of
-  web-based solutions.
+1. Poor interactive support: While [Docker]{.pkg} containers can be modified at
+  runtime, changes are lost when the container stops unless explicitly saved.
+  Running graphical applications like RStudio Desktop from containers requires
+  complex X11 configuration that works reliably only on Linux. Instead,
+  web-based IDEs (like RStudio Server) are used, but come with the inherent
+  limitations of web-based solutions.
 
@@
   environments, orchestrating them requires manual coordination. There is no
-  standard way to specify that one step should run in a Python environment while
-  another runs in R.
-
-5. Incomplete Reproducibility: Reproducibility should be understood as a
-  spectrum rather than a binary property, a notion emphasised in
-  "Reproducibility in Software Engineering" by @dellaieraMasterThesis2024.
-  Docker illustrates this spectrum well: it provides strong run-time
-  reproducibility—that is, consistency in space (where the same image behaves
-  identically across different systems) but weak build-time reproducibility,
-  which concerns consistency over time. As defined in the thesis,
-  reproducibility at build time means being able to produce an identical build
-  artefact regardless of when or where it is built. However, Docker inherently
-  fails to guarantee this property. Its reliance on mutable base images and
-  non-deterministic instructions such as `apt-get update` leads to temporal
-  drift: rebuilding the same `Dockerfile` at different points in time can
-  produce distinct outputs. In the experimental evaluation conducted in the
-  thesis, even with pinned base images, checksums of the resulting [Docker]{.pkg}
-  containers varied across builds and machines. Thus, while Docker offers
-  partial reproducibility: strong at run time (space) and weak at build time
-  (time), it cannot on its own ensure verifiable, deterministic software
-  artifacts. Achieving true reproducibility requires pinning image digests,
-  freezing dependency versions, and employing deterministic build systems such
-  as Nix or Guix for complete control over both space and time [@Malka_2024].
+  standard way to specify that one step should run in a [Python]{.pkg}
+  environment while another runs in R.
+
+5. Incomplete Reproducibility: Reproducibility should be viewed as a spectrum
+   rather than a binary property, as emphasised in *Reproducibility in Software
+   Engineering* by @dellaieraMasterThesis2024. [Docker]{.pkg} illustrates this
+   spectrum well: it achieves strong run-time reproducibility—consistency across
+   systems (space)—but weak build-time reproducibility, which concerns
+   consistency across time. As defined in the thesis, build-time reproducibility
+   means producing identical artefacts regardless of when or where they are
+   built. [Docker]{.pkg} cannot guarantee this, as its reliance on mutable base
+   images and non-deterministic commands such as `apt-get update` (or similar
+   commands, depending on which base image is used) introduces temporal drift:
+   rebuilding the same `Dockerfile` at different times can yield different
+   results. Even with pinned base images, the thesis found that container
+   checksums varied across builds and machines. Thus, while [Docker]{.pkg}
+   provides partial reproducibility, strong in space but weak in time, verifiable,
+   deterministic artefacts require pinning image digests, freezing dependencies,
+   and using deterministic build systems such as [Nix]{.pkg} for full control
+   over both dimensions [@Malka_2024].
 
@@
 However, [targets]{.pkg} operates within a single R session. While it can call
-Python via [reticulate]{.pkg} or execute system commands, all steps share the
-same computational environment. For truly polyglot pipelines (where different
-steps may require incompatible dependencies or different language versions) this
-limitation becomes problematic. One might run [targets]{.pkg} inside a
-[Docker]{.pkg} container to ensure reproducibility, but this requires the entire
-pipeline to use a single environment.
+[Python]{.proglang} via [reticulate]{.pkg} or execute system commands, all steps
+share the same computational environment. For truly polyglot pipelines (where
+different steps may require incompatible dependencies or different language
+versions) this limitation becomes problematic. One might run [targets]{.pkg}
+inside a [Docker]{.pkg} container to ensure reproducibility, but this requires
+the entire pipeline to use a single environment.
 
@@
 environment definition including [R]{.proglang}, [Python]{.proglang},
-[Julia]{.proglang}, system tools, and their dependencies.
+[Julia]{.proglang} [@bezanson2017julia], system tools, and their dependencies.
 
@@
 infrastructure and a functional approach, which significantly enhances the
-consistency of computational environments. As of writing, the primary Nix
+consistency of computational environments. As of writing, the primary [Nix]{.pkg}
 package collection, `nixpkgs`, provides access to over 120,000 packages,
@@
 [Python]{.proglang} package manager. This allows users to combine Nix's
-system-level reproducibility (Python interpreter, system libraries) with
-[uv]{.pkg}'s comprehensive package availability. While this hybrid approach
-sacrifices some of Nix's deterministic build guarantees for Python packages, it
-provides a pragmatic solution when the needed packages are unavailable in
-`nixpkgs`.
+system-level reproducibility ([Python]{.proglang} interpreter, system libraries)
+with [uv]{.pkg}'s comprehensive package availability. While this hybrid approach
+sacrifices some of Nix's deterministic build guarantees for [Python]{.proglang}
+packages, it provides a pragmatic solution when the needed packages are
+unavailable in `nixpkgs`.
 
@@
 
-[rixpress]{.pkg} overcomes these limitations by using Nix not just as a package
-manager, but as the build automation engine itself. In a [rixpress]{.pkg}
-pipeline, each step is defined as a Nix derivation, providing two key benefits:
-
-  1. True Polyglot Pipelines: Each step can have its own Nix environment. A
-     Python step can run in a pure Python environment, an [R]{.proglang} step in
-     an [R]{.proglang} environment, and a Quarto rendering step in yet another,
-     all within the same pipeline.
-  2. Deep Reproducibility: Each step is a hermetically sealed Nix derivation
-     whose output is cached in the Nix store based on the hash of all its
+[rixpress]{.pkg} overcomes these limitations by using [Nix]{.pkg} not just as a
+package manager, but as the build automation engine itself. In a
+[rixpress]{.pkg} pipeline, each step is defined as a [Nix]{.pkg} derivation,
+providing two key benefits:
+
+  1. True Polyglot Pipelines: Each step can have its own [Nix]{.pkg} environment. A
+     [Python]{.proglang} step can run in a pure [Python]{.proglang} environment,
+     an [R]{.proglang} step in an [R]{.proglang} environment, and a Quarto
+     rendering step in yet another, all within the same pipeline.
+  2. Deep Reproducibility: Each step is a hermetically sealed [Nix]{.pkg} derivation
+     whose output is cached in the [Nix]{.pkg} store based on the hash of all its
      inputs. All artifacts are direct children of the computational environment
@@
 polyglot pipeline that simulates a Real Business Cycle (RBC) model in
-[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
+[Julia]{.proglang}, uses the resulting data to train an
 [XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
@@
 multi-step data science workflow that crosses language boundaries. (Complete
-setup instructions and a detailed walkthrough are provided in Appendix Section
-@sec-polyglot-example.)
+setup instructions and a detailed walkthrough are provided in Appendix)
 
@@
 introduce external data into the pipeline. It is important to note that any data
-imported in this way is stored within the Nix store, ensuring reproducibility
-but also persisting the data in the build environment. The function `rxp_qmd()`
-is used to compile a [Quarto]{.pkg} document (there is also `rxp_rmd()` which
-serves the same purpose for [R Markdown]{.pkg} documents).
+imported in this way is stored within the [Nix]{.pkg} store, ensuring
+reproducibility but also persisting the data in the build environment. The
+function `rxp_qmd()` is used to compile a [Quarto]{.pkg} document (there is also
+`rxp_rmd()` which serves the same purpose for [R Markdown]{.pkg} documents).
 
@@
 Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
-pipelines. By leveraging Nix as a build automation engine, [rixpress]{.pkg}
+pipelines. By leveraging [Nix]{.pkg} as a build automation engine, [rixpress]{.pkg}
 allows polyglot workflows where each step runs in its own hermetically sealed
@@
 compile the manuscript, ensuring full reproducibility of the build process. To
-guarantee the reproducibility of the manuscript, it is automatically
-recompiled via GitHub Actions upon each commit, using the same Nix environment.
-An HTML version is additionally deployed to GitHub Pages, providing an
-accessible format for viewing on smaller devices at this
+guarantee the reproducibility of the manuscript, it is automatically recompiled
+via GitHub Actions upon each commit, using the same [Nix]{.pkg} environment. An
+HTML version is additionally deployed to GitHub Pages, providing an accessible
+format for viewing on smaller devices at this
 [link](https://b-rodrigues.github.io/rix_paper/)^[[https://b-rodrigues.github.io/rix_paper/](https://b-rodrigues.github.io/rix_paper/)].

### 2025-10-23 paper.qmd
british english

@@
 analysis. Here too, a continuum exists. At the minimal end, authors might only
-name the main software used—say, the [R]{.proglang} programming language. More
-careful authors might also specify the version of [R]{.proglang}, or list the
-additional packages and their versions. Rarely, however, do authors specify the
-operating system on which the analysis was performed, even though differences in
-operating systems can lead to divergent results when using the same code and
-software versions, as shown by @neupane2019. It is even less common for authors
-to provide step-by-step installation instructions for the required software
-stack, an omission often driven by institutional constraints. Journals, for
-instance, can inadvertently work against reproducibility by imposing strict page
-or word limits that leave no room for the necessary technical documentation,
-effectively discouraging thoroughness in the name of brevity.
+name the main software used—say, the [R]{.proglang} programming language
+[@r-core-team]. More careful authors might also specify the version of
+[R]{.proglang}, or list the additional packages and their versions. Rarely,
+however, do authors specify the operating system on which the analysis was
+performed, even though differences in operating systems can lead to divergent
+results when using the same code and software versions, as shown by
+@neupane2019. It is even less common for authors to provide step-by-step
+installation instructions for the required software stack, an omission often
+driven by institutional constraints. Journals, for instance, can inadvertently
+work against reproducibility by imposing strict page or word limits that leave
+no room for the necessary technical documentation, effectively discouraging
+thoroughness in the name of brevity.
 
@@
 orchestration*. I present a comprehensive framework that addresses both
-challenges using the [Nix]{.pkg} package manager, making it accessible to
-researchers through two [R]{.proglang} packages: [rix]{.pkg} and
+challenges using the [Nix]{.pkg} package manager [@dolstra2004nix], making it
+accessible to researchers through two [R]{.proglang} packages: [rix]{.pkg} and
 [rixpress]{.pkg}. Before introducing these packages, I survey existing tools and
-their limitations to contextualize my contribution.
+their limitations to contextualise my contribution.
 
@@
 common limitation: none addresses system-level dependencies or supports polyglot
-environments. A project using [R]{.proglang} [@r-core-team], [Python]{.proglang}
+environments. A project using [R]{.proglang}, [Python]{.proglang}
 [@python3], and system tools like [Quarto]{.pkg} or \LaTeX requires coordinating
@@
 However, achieving this required mastering multiple complex tools: [renv]{.pkg}
-for package management, [Docker]{.pkg} for environment containerization, and
+for package management, [Docker]{.pkg} for environment containerisation, and
 [Make]{.pkg} for workflow orchestration. Each tool has its own syntax, concepts,
@@
 installing and building software in a fully reproducible and reliable manner.
-Unlike traditional package managers, [Nix]{.pkg} emphasizes immutable
+Unlike traditional package managers, [Nix]{.pkg} emphasises immutable
 infrastructure and a functional approach, which significantly enhances the
@@
 achieves reproducibility. [Nix]{.pkg} adopts principles from functional
-programming, conceptualizing software builds as *pure functions*. In
+programming, conceptualising software builds as *pure functions*. In
 mathematics, a pure function is defined as one that always produces the same
@@
 
-[Nix]{.pkg} installs all packages into a centralized, content-addressed
+[Nix]{.pkg} installs all packages into a centralised, content-addressed
 directory called the [Nix]{.pkg} store (typically `/nix/store`). Each package
@@
 packages at install time rather than using pre-built, content-addressed
-artifacts) it provides a pragmatic balance: comprehensive [Python]{.proglang}
+artefacts) it provides a pragmatic balance: comprehensive [Python]{.proglang}
 package availability with reasonable reproducibility guarantees. For polyglot
@@
      whose output is cached in the [Nix]{.pkg} store based on the hash of all its
-     inputs. All artifacts are direct children of the computational environment
+     inputs. All artefacts are direct children of the computational environment
      (because the computational environment is actually a dependency of the
-     artifacts), ensuring they are rebuilt if the environment changes, keeping
+     artefacts), ensuring they are rebuilt if the environment changes, keeping
      environment and outputs always in sync.
@@
 [Julia]{.proglang}, uses the resulting data to train an
-[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
+[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualises the results
 in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
@@
 
-  # STEP 3: R - Visualize the predictions
+  # STEP 3: R - Visualise the predictions
   rxp_r(
@@
 +
-+  # STEP 3: R - Visualize the predictions
++  # STEP 3: R - Visualise the predictions
 +  rxp_r(
@@
 `pipeline.nix` file, which declaratively defines the entire workflow. Data flows
-from derivation to derivation by being serialized into the efficient and
+from derivation to derivation by being serialised into the efficient and
 language-agnostic Arrow format using the pair of `encoder/decoder` functions.
@@
 recompute steps with changed inputs or code. This provides the efficiency of
-[targets]{.pkg} with polyglot support and bit-for-bit reproducibility. Artifacts
+[targets]{.pkg} with polyglot support and bit-for-bit reproducibility. Artefacts
 can be inspected interactively in [R]{.proglang} using
-`rxp_read("artifact_name")` or `rxp_load("artifact_name")`.
+`rxp_read("artefact_name")` or `rxp_load("artefact_name")`.
 
@@
 development, especially to enhance the ease of debugging complex pipelines. Its
-visualization capabilities are also still fairly limited; for instance, they do
+visualisation capabilities are also still fairly limited; for instance, they do
 not currently indicate outdated derivations when a user modifies code that would
@@
 (RBC) model in [Julia]{.proglang}, trains an [XGBoost]{.pkg} model in
-[Python]{.proglang}, visualizes the results in [R]{.proglang}, and compiles a
+[Python]{.proglang}, visualises the results in [R]{.proglang}, and compiles a
 final report with [Quarto]{.pkg}.
@@
     *   `functions.py`: The [Python]{.proglang} script for the machine learning workflow.
-    *   `functions.R`: The [R]{.proglang} script for data visualization.
+    *   `functions.R`: The [R]{.proglang} script for data visualisation.
 *   `gen-env.R`: An [R]{.proglang} script that defines the reproducible computational environment.
@@
   testing sets, training the [XGBoost]{.pkg} model, and generating predictions.
-*   The Visualization (`functions.R`): This [R]{.proglang} script contains a
-  function that uses [ggplot2]{.pkg} to create the final visualization. It is
+*   The Visualisation (`functions.R`): This [R]{.proglang} script contains a
+  function that uses [ggplot2]{.pkg} to create the final visualisation. It is
   designed to take the data frame of actual and predicted values produced by the
@@
   using the [Arrow]{.pkg} file format for efficiency.
-4.  Finally, it defines the **R visualization** step, which depends on the final
+4.  Finally, it defines the **R visualisation** step, which depends on the final
   predictions from the [Python]{.proglang} model, and a **Quarto rendering**
@@
 
-Upon completion, the script will have generated all intermediate artifacts and
+Upon completion, the script will have generated all intermediate artefacts and
 the final `readme.html` report found in the `pipeline-output` folder containing
-the visualization.
+the visualisation.

### 2025-10-23 paper.qmd
Add Python and Julia to keywords-formatted

@@
 keywords: [reproducibility, R, Nix]
-keywords-formatted: [reproducibility, "[R]{.proglang}", "[Nix]{.pkg}"]
+keywords-formatted: [reproducibility, "[R]{.proglang}", "[Python]{.proglang}", "[Julia]{.proglang}", "[Nix]{.pkg}"]
 number-sections: true

### 2025-10-24 paper.qmd
revision of intro

@@
 
-A more robust approach than simply listing package versions is to use the
-[renv]{.pkg} package. [renv]{.pkg} captures the project's software state and
-writes it to a `renv.lock` file, which includes the exact versions of
-[R]{.proglang} and all required packages. This lockfile serves as a blueprint
-for restoring the environment automatically, ensuring that others can recreate
-the same setup with minimal effort. Below is an abbreviated example of an
-`renv.lock` file:
+A more robust approach than simply listing package versions is to use tools that
+help data scientists define per-project environments. This is common practice
+for [Python]{.proglang} users, who have many tools at their disposal, such as
+the built-in *virtual environments* module, but also sophisticated third-party
+tools. One particularly popular tool is [uv]{.pkg}, which manages both
+project-specific [Python]{.proglang} interpreters (enabling different Python
+versions per project) and package libraries using a lock file mechanism.
+
+In the [R]{.proglang} ecosystem, virtual environments are not a built-in
+language feature as they are in [Python]{.proglang}, but tools like [renv]{.pkg}
+provide equivalent functionality. [renv]{.pkg} captures the project's software
+state and writes it to a lock file (called `renv.lock`), which includes the
+exact versions of [R]{.proglang} and all required packages. This lockfile serves
+as a blueprint for restoring the environment automatically, ensuring that others
+can recreate the same setup with minimal effort. Below is an abbreviated example
+of an `renv.lock` file:
+
 ```{=tex}
@@
 
-However, [renv]{.pkg} has important limitations. It does not restore the version
-of [R]{.proglang} itself—installing the correct R version must be done
-separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
-[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
-[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg}
-cannot resolve this conflict. Users must manually install system libraries, and
-the process differs across operating systems. Further details are discussed in
-the [renv]{.pkg} documentation^[See
-\url{https://rstudio.github.io/renv/articles/renv.html}.].
-
-Other packages provide similar functionality with different trade-offs.
-[groundhog]{.pkg} by @simonsohn2023 makes it easy to install packages as they
-existed on CRAN at a given date:
-
-::: {.content-hidden when-format="pdf"}
-```R
-groundhog.library("
- library('purrr')
- library('ggplot2')",
- "2017-10-04",
- tolerate.R.version = "4.2.2")
-```
-:::
-
-```{=tex}
-\begin{CodeInput}
-R> groundhog.library("
-+   library('purrr')
-+   library('ggplot2')",
-+   "2017-10-04",
-+   tolerate.R.version = "4.2.2")
-\end{CodeInput}
-```
-
-[groundhog]{.pkg} places packages in project-specific libraries but does not
-install [R]{.proglang} itself. Without the `tolerate.R.version` argument, it
-will refuse to proceed if the R version does not match expectations.
-[rang]{.pkg}, developed by @chan2023, offers similar date-based package
-installation. The Posit Package Manager provides another approach through dated
-CRAN snapshots, configured via the `.Rprofile` file:
-
-::: {.content-hidden when-format="pdf"}
-```R
-options(repos =
-  c(REPO_NAME =
-    "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
-  )
-)
-```
-:::
-
-```{=tex}
-\begin{CodeInput}
-R> options(repos =
-+    c(REPO_NAME =
-+      "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
-+    )
-+  )
-\end{CodeInput}
-```
-
-This approach installs all packages from the specified snapshot date, but unless
-users explicitly manage separate libraries for different projects, all projects
-will share the same package versions.
-
-These tools represent significant progress in reproducibility, but they share a
-common limitation: none addresses system-level dependencies or supports polyglot
-environments. A project using [R]{.proglang}, [Python]{.proglang}
-[@python3], and system tools like [Quarto]{.pkg} or \LaTeX requires coordinating
-multiple package managers ([renv]{.pkg} for R, virtual environments for
-[Python]{.proglang}, manual installation for system tools), each with its own
-configuration and potential for conflicts.
+However, both [renv]{.pkg} and [uv]{.pkg} have important limitations. While
+[uv]{.pkg} manages Python interpreter versions directly, [renv]{.pkg} does not
+restore the version of [R]{.proglang} itself—installing the correct R version
+must be done separately using tools like [rig]{.pkg} [@rlib2023]. More
+critically, neither [uv]{.pkg} nor [renv]{.pkg} handle system-level
+dependencies. If [sf]{.pkg} requires [GDAL]{.pkg} version 3.0 but the system has
+version 2.4 installed, neither tool can resolve this conflict. Users must
+manually install system libraries, and the process differs across operating
+systems. Further details are discussed in the [renv]{.pkg} documentation^[See
+https://rstudio.github.io/renv/articles/renv.html.].
+
+Pre-compiled binary packages (Python wheels or R binaries from Posit Package
+Manager) can simplify installation by avoiding compilation, but they do not
+eliminate the need for system libraries that would be required at runtime—these
+must still be installed separately.
+
+Other [R]{.proglang} packages provide similar functionality with different
+trade-offs. [groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023]
+enable date-based package installation, retrieving packages as they existed on
+CRAN at a specified date. The Posit Package Manager offers another approach
+through dated CRAN snapshots. However, like [renv]{.pkg}, these tools focus on
+[R]{.proglang} packages and do not address system-level dependencies or
+[R]{.proglang} version management.
+
+These tools represent significant progress in reproducibility within their
+respective ecosystems, but they share a common limitation: none addresses
+system-level dependencies or supports polyglot environments. A project using
+[R]{.proglang}, [Python]{.proglang} [@python3], and system tools like
+[Quarto]{.pkg} or \LaTeX requires coordinating multiple package managers
+([renv]{.pkg} for R, [uv]{.pkg} for [Python]{.proglang}, manual installation for
+system tools), each with its own configuration and potential for conflicts.
 
@@
   environments, orchestrating them requires manual coordination. There is no
-  standard way to specify that one step should run in a [Python]{.pkg}
+  standard way to specify that one step should run in a [Python]{.proglang}
   environment while another runs in R.
@@
    checksums varied across builds and machines. Thus, while [Docker]{.pkg}
-   provides partial reproducibility, strong in space but weak in time, verifiable,
-   deterministic artefacts require pinning image digests, freezing dependencies,
-   and using deterministic build systems such as [Nix]{.pkg} for full control
-   over both dimensions [@Malka_2024].
+   provides strong run-time reproducibility, achieving verifiable, deterministic
+   artefacts requires pinning image digests, freezing dependencies, and using
+   deterministic build systems such as [Nix]{.pkg} for full control over both
+   temporal and spatial dimensions [@Malka_2024].
 

### 2025-10-24 paper.qmd
tighter intro

@@
 driven by institutional constraints. Journals, for instance, can inadvertently
-work against reproducibility by imposing strict page or word limits that leave
-no room for the necessary technical documentation, effectively discouraging
-thoroughness in the name of brevity.
+hinder reproducibility by imposing strict page or word limits that leave no room
+for the necessary technical documentation, discouraging thoroughness in the name
+of brevity.
 
@@
 hinders reproducibility but also complicates everyday research workflows.
-Researchers working across multiple machines must be able to recreate the same
-environment consistently, and collaborators must share identical computational
-setups to avoid inconsistencies.
+Researchers using multiple machines must recreate environments consistently, and
+collaborators must share identical computational setups to avoid
+inconsistencies.
 
@@
 
-This paper focuses specifically on two critical but often overlooked aspects of
+This paper focuses on two critical but often overlooked aspects of
 reproducibility: *computational environment management* and *workflow
-orchestration*. I present a comprehensive framework that addresses both
-challenges using the [Nix]{.pkg} package manager [@dolstra2004nix], making it
-accessible to researchers through two [R]{.proglang} packages: [rix]{.pkg} and
+orchestration*. I present a comprehensive framework addressing both challenges
+using the [Nix]{.pkg} package manager [@dolstra2004nix], making it accessible to
+researchers through two [R]{.proglang} packages: [rix]{.pkg} and
 [rixpress]{.pkg}. Before introducing these packages, I survey existing tools and
@@
 version, platform details, and all loaded packages. Its output can be saved to a
-file and included as part of a study's reproducibility record. Below is an
+file and included as part of a study’s reproducibility record. Below is an
 example output from `sessionInfo()`:
@@
 study (future readers, collaborators, or even the author at a later time) can
-easily see which version of [R]{.proglang} and which packages (with their
-versions) were used. However, reproducing the environment still requires
-manually installing the correct package versions—a process that becomes
-particularly difficult when packages depend on system-level libraries. For
-example, the [sf]{.pkg} package requires [GDAL]{.pkg}, [GEOS]{.pkg}, and
-[PROJ]{.pkg} system libraries. Installing and configuring these dependencies
-varies substantially across operating systems and can be prohibitively difficult
-on some platforms.
+see which version of [R]{.proglang} and which packages (with their versions)
+were used. However, reproducing the environment still requires manually
+installing the correct package versions—a process that becomes difficult when
+packages depend on system-level libraries. For example, the [sf]{.pkg} package
+requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} libraries. Installing and
+configuring these dependencies varies substantially across operating systems and
+can be prohibitively difficult on some platforms.
 
 A more robust approach than simply listing package versions is to use tools that
-help data scientists define per-project environments. This is common practice
-for [Python]{.proglang} users, who have many tools at their disposal, such as
-the built-in *virtual environments* module, but also sophisticated third-party
-tools. One particularly popular tool is [uv]{.pkg}, which manages both
-project-specific [Python]{.proglang} interpreters (enabling different Python
-versions per project) and package libraries using a lock file mechanism.
-
-In the [R]{.proglang} ecosystem, virtual environments are not a built-in
-language feature as they are in [Python]{.proglang}, but tools like [renv]{.pkg}
-provide equivalent functionality. [renv]{.pkg} captures the project's software
-state and writes it to a lock file (called `renv.lock`), which includes the
-exact versions of [R]{.proglang} and all required packages. This lockfile serves
-as a blueprint for restoring the environment automatically, ensuring that others
-can recreate the same setup with minimal effort. Below is an abbreviated example
-of an `renv.lock` file:
-
-```{=tex}
-\begin{Code}
-{
-  "R": {
-    "Version": "4.2.2",
-    "Repositories": [
-      {
-        "Name": "CRAN",
-        "URL": "https://packagemanager.rstudio.com/all/latest"
-      }
-    ]
-  },
-  "Packages": {
-    "MASS": {
-      "Package": "MASS",
-      "Version": "7.3-58.1",
-      "Source": "Repository",
-      "Repository": "CRAN",
-      "Hash": "762e1804143a332333c054759f89a706",
-      "Requirements": []
-    },
-    "Matrix": {
-      "Package": "Matrix",
-      "Version": "1.5-1",
-      "Source": "Repository",
-      "Repository": "CRAN",
-      "Hash": "539dc0c0c05636812f1080f473d2c177",
-      "Requirements": ["lattice"]
-    }
-    ... additional packages omitted ...
-  }
-}
-\end{Code}
-```
-
-This lockfile lists each package alongside its version and the repository from
-which it was downloaded. Creating it requires only a single command:
-`renv::init()`. Others can then reproduce the same package library by running
-`renv::restore()`, which installs the exact package versions in an isolated,
-project-specific library that does not interfere with the user's global R
-library.
-
-However, both [renv]{.pkg} and [uv]{.pkg} have important limitations. While
-[uv]{.pkg} manages Python interpreter versions directly, [renv]{.pkg} does not
-restore the version of [R]{.proglang} itself—installing the correct R version
-must be done separately using tools like [rig]{.pkg} [@rlib2023]. More
-critically, neither [uv]{.pkg} nor [renv]{.pkg} handle system-level
-dependencies. If [sf]{.pkg} requires [GDAL]{.pkg} version 3.0 but the system has
-version 2.4 installed, neither tool can resolve this conflict. Users must
-manually install system libraries, and the process differs across operating
-systems. Further details are discussed in the [renv]{.pkg} documentation^[See
-https://rstudio.github.io/renv/articles/renv.html.].
-
-Pre-compiled binary packages (Python wheels or R binaries from Posit Package
-Manager) can simplify installation by avoiding compilation, but they do not
-eliminate the need for system libraries that would be required at runtime—these
-must still be installed separately.
-
-Other [R]{.proglang} packages provide similar functionality with different
-trade-offs. [groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023]
-enable date-based package installation, retrieving packages as they existed on
-CRAN at a specified date. The Posit Package Manager offers another approach
-through dated CRAN snapshots. However, like [renv]{.pkg}, these tools focus on
-[R]{.proglang} packages and do not address system-level dependencies or
-[R]{.proglang} version management.
+define per-project environments. This is common for [Python]{.proglang} users,
+who have many tools at their disposal, such as the built-in *virtual
+environments* module, or third-party tools like [uv]{.pkg}, which manages both
+project-specific interpreters and package libraries using a lock file mechanism.
+
+In the [R]{.proglang} ecosystem, virtual environments are not built-in as they
+are in [Python]{.proglang}, but tools like [renv]{.pkg} provide equivalent
+functionality. [renv]{.pkg} captures the project’s software state and writes it
+to a lock file (`renv.lock`), which includes the exact versions of
+[R]{.proglang} and all required packages. This lockfile serves as a blueprint
+for restoring the environment automatically, ensuring others can recreate the
+same setup with minimal effort.
+
+However, both [renv]{.pkg} and [uv]{.pkg} have limitations. While [uv]{.pkg}
+manages Python interpreter versions directly, [renv]{.pkg} does not restore the
+version of [R]{.proglang} itself—installing the correct version must be done
+separately using tools like [rig]{.pkg} [@rlib2023]. More critically, neither
+handles system-level dependencies. If [sf]{.pkg} requires [GDAL]{.pkg} version
+3.0 but the system has version 2.4 installed, neither tool can resolve this
+conflict. Users must manually install system libraries, and the process differs
+across operating systems.
+
+Pre-compiled binaries simplify installation but do not eliminate the need for
+system libraries required at runtime. Other [R]{.proglang} packages such as
+[groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023] enable
+date-based package installation, while the Posit Package Manager provides dated
+CRAN snapshots. However, like [renv]{.pkg}, these focus on R packages and do not
+address system-level dependencies or R version management.
 
 These tools represent significant progress in reproducibility within their
-respective ecosystems, but they share a common limitation: none addresses
-system-level dependencies or supports polyglot environments. A project using
-[R]{.proglang}, [Python]{.proglang} [@python3], and system tools like
-[Quarto]{.pkg} or \LaTeX requires coordinating multiple package managers
-([renv]{.pkg} for R, [uv]{.pkg} for [Python]{.proglang}, manual installation for
-system tools), each with its own configuration and potential for conflicts.
-
-The most comprehensive approach to reproducibility is containerisation with
-[Docker]{.pkg}. [Docker]{.pkg} packages a *data product* together with all its
+ecosystems, but they share a common limitation: none handles system-level
+dependencies or polyglot environments. A project using [R]{.proglang},
+[Python]{.proglang} [@python3], and system tools like [Quarto]{.pkg} or \LaTeX
+requires coordinating multiple package managers ([renv]{.pkg} for R, [uv]{.pkg}
+for Python, manual installation for system tools), each with its own
+configuration and potential conflicts.
+
+The most comprehensive approach to date is containerisation with [Docker]{.pkg}.
+[Docker]{.pkg} packages a *data product* together with all its
 dependencies—including the operating system, system libraries, programming
-languages, and packages—into a self-contained image. A statistical analysis can
-be viewed as such a data product. The steps to create an image are specified in
-a `Dockerfile`, which defines the structure and content of the [Docker]{.pkg}
-image.
-
-Once the image is built, the analysis can be executed inside a *container*,
-which is a running instance of the image. [Docker]{.pkg} containers are typically run
-non-interactively, allowing the complete computational environment to be
-instantiated and executed with a single command.
+languages, and packages—into a self-contained image. Once built, the analysis
+can be executed inside a *container*, a running instance of the image.
 
@@
 correct versions of [R]{.proglang} and its packages but also system-level
-dependencies. Since [Docker]{.pkg} images are effectively minimal Linux systems,
-they naturally include libraries required by packages like [sf]{.pkg}. This
-ensures that future replicators have access to the same software environment,
-including all necessary system components. Moreover, these images can be easily
-shared via registries like Docker Hub, enabling seamless reproducibility across
-machines and collaborators.
-
-The Rocker project, introduced by @boettiger2017, provides a collection of
-pre-built [Docker]{.pkg} images for the [R]{.proglang} community. These images
-come with specific R versions and, in some cases, preinstalled packages, making
-them convenient base images for building reproducible analysis environments
-without starting from scratch.
-
-Despite its strengths, [Docker]{.pkg} presents several challenges for
-statistical computing workflows:
-
-1. Poor interactive support: While [Docker]{.pkg} containers can be modified at
-  runtime, changes are lost when the container stops unless explicitly saved.
-  Running graphical applications like RStudio Desktop from containers requires
-  complex X11 configuration that works reliably only on Linux. Instead,
-  web-based IDEs (like RStudio Server) are used, but come with the inherent
-  limitations of web-based solutions.
-
-2. Steep learning curve: [Docker]{.pkg} images are minimal Linux systems, so
-  writing reliable `Dockerfile`s requires familiarity with Linux system
-  administration. To ensure reproducibility, `Dockerfile`s should reference
-  specific image *digests* rather than *tags*, but this practice is rarely
-  followed. Packages like [dockerfiler]{.pkg} [@dockerfiler] help automate
-  `Dockerfile` creation but cannot eliminate the need for basic Linux knowledge.
-
-3. Post-hoc reproducibility: A common workflow is to develop analyses
-  interactively using standard R installations, then create a `Dockerfile`
-  afterward to enable reproduction. While this achieves post-hoc
-  reproducibility, it fails to address the challenge of deploying consistent
-  environments across multiple machines *during* development, a critical need
-  for collaborative projects.
-
-4. Single-language focus: While [Docker]{.pkg} can support polyglot
-  environments, orchestrating them requires manual coordination. There is no
-  standard way to specify that one step should run in a [Python]{.proglang}
-  environment while another runs in R.
-
-5. Incomplete Reproducibility: Reproducibility should be viewed as a spectrum
-   rather than a binary property, as emphasised in *Reproducibility in Software
-   Engineering* by @dellaieraMasterThesis2024. [Docker]{.pkg} illustrates this
-   spectrum well: it achieves strong run-time reproducibility—consistency across
-   systems (space)—but weak build-time reproducibility, which concerns
-   consistency across time. As defined in the thesis, build-time reproducibility
-   means producing identical artefacts regardless of when or where they are
-   built. [Docker]{.pkg} cannot guarantee this, as its reliance on mutable base
-   images and non-deterministic commands such as `apt-get update` (or similar
-   commands, depending on which base image is used) introduces temporal drift:
-   rebuilding the same `Dockerfile` at different times can yield different
-   results. Even with pinned base images, the thesis found that container
-   checksums varied across builds and machines. Thus, while [Docker]{.pkg}
-   provides strong run-time reproducibility, achieving verifiable, deterministic
-   artefacts requires pinning image digests, freezing dependencies, and using
+dependencies. Since [Docker]{.pkg} images are minimal Linux systems, they
+include the libraries required by packages such as [sf]{.pkg}, ensuring that
+future replicators have access to the same environment. Moreover, these images
+can be shared via registries like Docker Hub, enabling reproducibility across
+systems.
+
+The Rocker project [@boettiger2017] provides pre-built [Docker]{.pkg} images for
+the [R]{.proglang} community, offering specific R versions and, in some cases,
+preinstalled packages. These serve as convenient bases for reproducible analysis
+environments.
+
+Despite its strengths, [Docker]{.pkg} presents several challenges:
+
+1. Poor interactive support: While containers can be modified at runtime,
+   changes are lost unless explicitly saved. Running graphical applications like
+   RStudio Desktop requires complex X11 configuration that works reliably only
+   on Linux. Web-based IDEs such as RStudio Server are alternatives but have
+   their own limitations.
+
+2. Steep learning curve: Writing reliable `Dockerfile`s requires familiarity
+   with Linux system administration. To ensure reproducibility, `Dockerfile`s
+   should reference image *digests* rather than *tags*, but this is rarely done.
+   Packages like [dockerfiler]{.pkg} [@dockerfiler] help automate this but do
+   not remove the need for basic Linux knowledge.
+
+3. Post-hoc reproducibility: Analyses are often developed interactively and
+   containerised only afterward, achieving post-hoc reproducibility rather than
+   consistent environments during development.
+
+4. Single-language focus: While [Docker]{.pkg} can support polyglot setups,
+   orchestrating them requires manual coordination. There is no standard way to
+   specify that one step runs in [Python]{.proglang} and another in [R]{.proglang}.
+
+5. Incomplete reproducibility: Reproducibility is best viewed as a spectrum
+   rather than a binary property [@dellaieraMasterThesis2024]. [Docker]{.pkg}
+   achieves strong run-time reproducibility—consistency across systems
+   (space)—but weak build-time reproducibility, or consistency over time.
+   Mutable base images and non-deterministic commands such as `apt-get update`
+   introduce temporal drift: rebuilding the same `Dockerfile` at different times
+   can yield different results. Even with pinned base images, container
+   checksums can vary across builds and machines. Thus, while [Docker]{.pkg}
+   provides strong run-time reproducibility, achieving deterministic artefacts
+   requires pinning image digests, freezing dependencies, and using
    deterministic build systems such as [Nix]{.pkg} for full control over both
-   temporal and spatial dimensions [@Malka_2024].
+   time and space [@Malka_2024].
 
@@
 
-Within [R]{.proglang}, the [targets]{.pkg} package by @landau2021 provides a
-modern, declarative approach to workflow management. [targets]{.pkg} tracks
-dependencies between code and data, caches intermediate results, and only
-recomputes steps affected by changes. This dramatically improves efficiency for
-complex analyses.
-
-However, [targets]{.pkg} operates within a single R session. While it can call
-[Python]{.proglang} via [reticulate]{.pkg} or execute system commands, all steps
-share the same computational environment. For truly polyglot pipelines (where
-different steps may require incompatible dependencies or different language
-versions) this limitation becomes problematic. One might run [targets]{.pkg}
-inside a [Docker]{.pkg} container to ensure reproducibility, but this requires
-the entire pipeline to use a single environment.
-
-@mcdermott2021 provides an excellent example of achieving the gold standard of
-reproducibility. The accompanying GitHub
-repository^[https://github.com/grantmcdermott/skeptic-priors] demonstrates the
-state of the art:
-
-- Package versions recorded in an `renv.lock` file
-- A `Makefile` automating the full analysis
-- A `Dockerfile` providing a complete computational environment
-
-However, achieving this required mastering multiple complex tools: [renv]{.pkg}
-for package management, [Docker]{.pkg} for environment containerisation, and
-[Make]{.pkg} for workflow orchestration. Each tool has its own syntax, concepts,
-and failure modes. The complexity compounds for polyglot projects; for instance,
-a [Python]{.proglang} project would require learning virtual environments, and
-coordinating [R]{.proglang} and [Python]{.proglang} in a single analysis
-requires additional tooling.
-
-What researchers need is a tool that:
-
-1. Manages complete environments including programming languages, packages, and
-   system dependencies;
-2. Supports multiple languages natively, not through workarounds;
-3. Works interactively without the complexity of containers;
-4. Provides step-level isolation so different pipeline steps can use
-  different environments;
+Within [R]{.proglang}, the [targets]{.pkg} package [@landau2021] provides a
+modern, declarative approach to workflow management. It tracks dependencies,
+caches results, and only recomputes affected steps, improving efficiency for
+complex analyses. However, [targets]{.pkg} operates within a single
+[R]{.proglang} session. For truly polyglot pipelines—where steps may require
+incompatible dependencies or different language versions—this becomes
+problematic. Running [targets]{.pkg} inside a [Docker]{.pkg} container ensures
+reproducibility but forces the entire pipeline to share one environment.
+
+@mcdermott2021 provides an excellent example of best practice. The accompanying
+repository^[See
+[https://github.com/grantmcdermott/skeptic-priors](https://github.com/grantmcdermott/skeptic-priors)]
+demonstrates:
+
+- Package versions recorded in an `renv.lock` file;
+- A `Makefile` automating the full analysis;
+- A `Dockerfile` providing the computational environment.
+
+Achieving this required mastering multiple complex tools: [renv]{.pkg} for
+package management, [Docker]{.pkg} for containerisation, and [Make]{.pkg} for
+orchestration. The complexity increases for polyglot projects, which must manage
+multiple ecosystems simultaneously.
+
+Researchers therefore need a tool that:
+
+1. Manages complete environments including languages, packages, and system
+   dependencies;
+2. Supports multiple languages natively;
+3. Works interactively without container complexity;
+4. Provides step-level isolation for different environments;
 5. Ensures bit-for-bit reproducibility through deterministic builds;
@@
 
-The [Nix]{.pkg} package manager provides these capabilities. [Nix]{.pkg} ensures
-reproducible software installation by deploying *component closures*—packages
-bundled with all their dependencies and transitive dependencies
-[@dolstra2004nix]. Think of installing a package like packing for a trip.
-Traditional package managers are like packing just your clothes, assuming you'll
-find a toothbrush and soap at your destination. [Nix]{.pkg} packs everything you
-might need—the entire self-contained kit. This is what [Nix]{.pkg} calls a
-*component closure*.
-
-This creates self-contained software environments that work
-identically across machines and over time. [Nix]{.pkg} can replace
-[Docker]{.pkg} for environment isolation, [renv]{.pkg} for package management,
-and even [Make]{.pkg} for workflow orchestration—all within a single, unified
-framework.
-
-However, [Nix]{.pkg} has a steep learning curve. It is a complex system with its
-own purely functional programming language designed to declaratively define how
-software is built and configured. While this functional approach ensures
-reproducibility, it creates a significant barrier to adoption.
+The [Nix]{.pkg} package manager provides these capabilities. It ensures
+reproducible installation by deploying *component closures*—packages bundled
+with all their direct and transitive dependencies [@dolstra2004nix]. Think of
+installing a package as packing for a trip: traditional package managers assume
+you’ll find essentials at your destination, while [Nix]{.pkg} packs everything
+you need. This produces self-contained software environments that behave
+identically across machines and over time.
+
+[Nix]{.pkg} can replace [Docker]{.pkg} for isolation, [renv]{.pkg} for package
+management, and even [Make]{.pkg} for orchestration—all within a single, unified
+framework. However, it has a steep learning curve, with its own functional
+language for declaratively defining software builds and configurations. While
+this ensures reproducibility, it also creates a barrier to adoption.
 
@@
   eliminating the need to learn the [Nix]{.proglang} language. It handles
-environment definition including [R]{.proglang}, [Python]{.proglang},
-[Julia]{.proglang} [@bezanson2017julia], system tools, and their dependencies.
+  environment definitions including [R]{.proglang}, [Python]{.proglang},
+  [Julia]{.proglang} [@bezanson2017julia], system tools, and dependencies.
 
 - [rixpress]{.pkg} orchestrates polyglot analytical pipelines using [Nix]{.pkg}
-as the build engine. Each pipeline step runs in its own hermetically sealed
-environment with automatic caching and dependency tracking. This enables true
-polyglot workflows where [R]{.proglang}, [Python]{.proglang} and
-[Julia]{.proglang} steps coexist with different dependencies.
-[Python]{.proglang} user can use [ryxpress]{.pkg}, a port of [rixpress]{.pkg}.
-
-Together, these packages provide what existing tools cannot: a single framework
-for managing reproducible, polyglot computational environments and executing
-complex analytical workflows with step-level isolation. While [Nix]{.pkg}
-remains complex for advanced use cases, [rix]{.pkg} and [rixpress]{.pkg}
-abstract this complexity for common research workflows, making deep
-reproducibility accessible without requiring systems administration expertise.
-
-The remainder of this paper proceeds as follows. Section 2 introduces the
-[Nix]{.pkg} package manager and explains why its functional approach enables
-reproducibility. Section 3 presents [rix]{.pkg} and demonstrates environment
-definition for various use cases. Section 4 discusses the `rstats-on-nix` fork
-of the official package repository and how it addresses practical limitations.
-Section 5 introduces [rixpress]{.pkg} and demonstrates polyglot pipeline
-orchestration through a complete example. Section 6 concludes with discussion of
-appropriate use cases and future directions.
+  as the build engine. Each step runs in its own hermetic environment with
+  automatic caching and dependency tracking, enabling true polyglot workflows
+  across R, Python, and Julia. [Python]{.proglang} users can use
+  [ryxpress]{.pkg}, a direct port.
+
+Together, these packages provide a single framework for managing reproducible,
+polyglot environments and executing complex workflows with step-level isolation.
+While [Nix]{.pkg} remains complex for advanced use, [rix]{.pkg} and
+[rixpress]{.pkg} abstract this complexity, making deep reproducibility
+accessible without systems administration expertise.
+
+The remainder of this paper proceeds as follows. Section 2 introduces
+[Nix]{.pkg} and explains how its functional model enables reproducibility.
+Section 3 presents [rix]{.pkg} and demonstrates environment definition. Section
+4 discusses the `rstats-on-nix` fork of the package repository. Section 5
+introduces [rixpress]{.pkg} and demonstrates polyglot pipeline orchestration.
+Section 6 concludes with discussion of future directions.
 

### 2025-10-24 paper.qmd
repo cleanup

@@
 However, both [renv]{.pkg} and [uv]{.pkg} have limitations. While [uv]{.pkg}
-manages Python interpreter versions directly, [renv]{.pkg} does not restore the
-version of [R]{.proglang} itself—installing the correct version must be done
-separately using tools like [rig]{.pkg} [@rlib2023]. More critically, neither
-handles system-level dependencies. If [sf]{.pkg} requires [GDAL]{.pkg} version
-3.0 but the system has version 2.4 installed, neither tool can resolve this
-conflict. Users must manually install system libraries, and the process differs
-across operating systems.
+manages [Python]{.proglang} interpreter versions directly, [renv]{.pkg} does not
+restore the version of [R]{.proglang} itself—installing the correct version must
+be done separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
+neither handles system-level dependencies. If [sf]{.pkg} requires [GDAL]{.pkg}
+version 3.0 but the system has version 2.4 installed, neither tool can resolve
+this conflict. Users must manually install system libraries, and the process
+differs across operating systems.
 
@@
 requires coordinating multiple package managers ([renv]{.pkg} for R, [uv]{.pkg}
-for Python, manual installation for system tools), each with its own
+for [Python]{.proglang}, manual installation for system tools), each with its own
 configuration and potential conflicts.
@@
   automatic caching and dependency tracking, enabling true polyglot workflows
-  across R, Python, and Julia. [Python]{.proglang} users can use
-  [ryxpress]{.pkg}, a direct port.
+  across [R]{.proglang}, [Python]{.proglang}, and [Julia]{.proglang}.
+  [Python]{.proglang} users can use [ryxpress]{.pkg}, a direct port.
 
@@
 in both total package count and update recency, making it well-suited for
-reproducible research environments.](repology.png){#fig-repology}
+reproducible research environments.](images/repology.png){#fig-repology}
 
@@
 [Julia]{.proglang}, yellow nodes are [Python]{.pkg}, light blue nodes are
-[R]{.proglang}, and dark blue nodes are Quarto derivations.](dag.png){#fig-dag}
+[R]{.proglang}, and dark blue nodes are Quarto derivations.](images/dag.png){#fig-dag}
 
@@
   designed to take the data frame of actual and predicted values produced by the
-  Python script and generate a plot comparing the two series.
+  [Python]{.proglang} script and generate a plot comparing the two series.
 

### 2025-10-24 paper.qmd
better conclusion

@@
 on, and prepare fixes that can then be contributed upstream.
+
+While the reliance on the `rstats-on-nix` fork might initially raise
+sustainability concerns, this risk should be contextualized. The fork serves a
+focused purpose: it updates CRAN and [Julia]{.proglang} package registries daily
+and backports fixes when builds break. Importantly, users are never locked into
+this fork—they retain the ability to point directly to any commit in the
+upstream `nixpkgs` repository, maintaining full access to the official
+[Nix]{.pkg} ecosystem. The fork introduces no fundamental architectural changes
+or proprietary extensions; it functions as a convenience layer providing tested
+package sets and dated snapshots. Should maintenance cease (which is unlikely,
+as I directly use it myself), users could seamlessly transition to upstream
+`nixpkgs` commits, though with trade-offs: less frequent [R]{.proglang} and
+[Julia]{.proglang} package updates. The fork's value lies not in creating
+dependency, but in reducing friction through testing infrastructure and curated
+"known-good" dates. Thus, the fork represents an enhancement to user experience
+rather than a critical single point of failure—the underlying [Nix]{.pkg}
+infrastructure and official `nixpkgs` repository remain the foundation, ensuring
+long-term viability of environments created with [rix]{.pkg}.
 
@@
 
+Adopting [Nix]{.pkg}-based workflows represents a significant paradigm shift, and
+researchers should approach it incrementally. A reasonable pathway begins with
+using [rix]{.pkg} to generate reproducible environments for new projects, while
+maintaining existing workflows for ongoing work. This allows learning [Nix]{.pkg}
+concepts without risking active research. As comfort grows, researchers can
+experiment with [rixpress]{.pkg} for smaller analyses before tackling complex
+polyglot pipelines. Institutions can support adoption by providing workshops,
+maintaining institutional binary caches with private packages, and designating
+local experts. The investment pays increasing returns: early projects require
+substantial learning, but subsequent projects leverage accumulated knowledge and
+reusable environment definitions. For research groups, having one member develop
+[Nix]{.pkg} expertise can enable the entire team to benefit from reproducible
+environments without everyone climbing the learning curve. Ultimately, the
+decision to adopt these tools depends on project timescales, reproducibility
+requirements, and institutional context—but the framework exists for those who
+need it.
+
 While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide
 significant steps forward, it is important to acknowledge current limitations.
-[rixpress]{.pkg} (and [ryxpress]{.pkg}) in particular requires further
-development, especially to enhance the ease of debugging complex pipelines. Its
-visualisation capabilities are also still fairly limited; for instance, they do
-not currently indicate outdated derivations when a user modifies code that would
-necessitate a rebuild. These areas represent important avenues for future work
-to further mature the framework and improve the user experience.
+Debugging complex [rixpress]{.pkg} pipelines remains more difficult than
+debugging [targets]{.pkg} workflows, particularly when failures occur deep in
+the [Nix]{.pkg} build process. The visualization capabilities, while useful, lack
+features like automatic detection of outdated derivations that make tools like
+[targets]{.pkg} particularly developer-friendly. Cross-platform reproducibility,
+while generally strong, has known limitations on macOS due to system framework
+dependencies. These are not insurmountable problems—they represent areas for
+continued development and community contribution. Researchers adopting these
+tools should do so with realistic expectations: they provide powerful
+capabilities for deep reproducibility and polyglot workflows, but require
+patience during the learning phase and may occasionally present challenges that
+simpler tools avoid.
 

### 2025-10-25 paper.qmd
feat: small suggestions

@@
     url: https://www.brodrigues.co
+  - name: Philipp Baumann
+    affiliations:
+      - name: Alliance SwissPass, Switzerland
+        department: Data and Analytics
+        address: 7, Länggassstrasse
+        city: Bern
+        country: Switzerland
+        postal-code: 3012
+    orcid: 0000-0002-3194-8975
+    email: baumann-philipp@protonmail.com
 abstract: |
@@
   polyglot environments, system dependencies, and workflow orchestration
-  together. I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to
+  together. We present two [R]{.proglang} packages that leverage [Nix]{.pkg} to
   provide both reproducible polyglot environments and declarative workflow
@@
   [rixpress]{.pkg} (and its [Python]{.proglang} port, [ryxpress]{.pkg})
-  orchestrates polyglot pipelines where each computational step runs in its own
+  orchestrate polyglot pipelines. Each computational step runs in its own
   hermetically sealed environment, with automatic caching and dependency
   tracking. This approach enables researchers to develop analyses interactively
-  while maintaining bit-for-bit reproducibility, supports collaboration across
-  heterogeneous computational environments, and ensures that analyses remain
-  executable years into the future.
+  while maintaining bit-for-bit reproducibility. Beyond, it supports
+  collaboration across heterogeneous computational environments, and ensures
+  that analyses can be reproduced years into the future.
 
@@
 By *linked and executable code and data*, Peng referred to compiled source code
-and runnable scripts. In this paper, I interpret this notion more broadly as the
+and runnable scripts. In this paper, we interpret this notion more broadly as the
 *computational environment*: the complete set of software required to execute an
@@
 reproducibility: *computational environment management* and *workflow
-orchestration*. I present a comprehensive framework addressing both challenges
+orchestration*. We present a comprehensive framework addressing both challenges
 using the [Nix]{.pkg} package manager [@dolstra2004nix], making it accessible to
 researchers through two [R]{.proglang} packages: [rix]{.pkg} and
-[rixpress]{.pkg}. Before introducing these packages, I survey existing tools and
-their limitations to contextualise my contribution.
+[rixpress]{.pkg}. Before introducing these packages, we survey existing tools and
+their limitations to contextualise our contributions.
 
@@
 were used. However, reproducing the environment still requires manually
-installing the correct package versions—a process that becomes difficult when
+installing the correct package versions --- a process that becomes difficult when
 packages depend on system-level libraries. For example, the [sf]{.pkg} package
@@
 In the [R]{.proglang} ecosystem, virtual environments are not built-in as they
-are in [Python]{.proglang}, but tools like [renv]{.pkg} provide equivalent
+are in [Python]{.proglang}, however tools like [renv]{.pkg} provide equivalent
 functionality. [renv]{.pkg} captures the project’s software state and writes it
@@
 development environment in a single text file and then rebuild it anywhere is
-highly appealing. To lower the barrier to adoption of [Nix]{.pkg} for
-reproducibility, I developed the [rix]{.pkg} package.
+highly appealing. [rix]{.pkg} aims to lower the barrier to adoption of [Nix]{.pkg} for
+reproducibility.
 

### 2025-10-26 paper.qmd
action

--- 2025-10-24 paper.qmd
@@
 abstract: |
-  Reproducible analysis requires more than clean, well-documented code:
-  researchers must also manage software dependencies, computational
-  environments, and workflow execution. Existing tools like [Docker]{.pkg} and
-  [renv]{.pkg} address parts of this challenge, but no single solution handles
-  polyglot environments, system dependencies, and workflow orchestration
-  together. I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to
-  provide both reproducible polyglot environments and declarative workflow
-  orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define
-  reproducible environments spanning [R]{.proglang}, [Python]{.proglang},
-  [Julia]{.proglang}, and system dependencies. Building on these environments,
-  [rixpress]{.pkg} (and its [Python]{.proglang} port, [ryxpress]{.pkg})
-  orchestrates polyglot pipelines where each computational step runs in its own
-  hermetically sealed environment, with automatic caching and dependency
-  tracking. This approach enables researchers to develop analyses interactively
-  while maintaining bit-for-bit reproducibility, supports collaboration across
-  heterogeneous computational environments, and ensures that analyses remain
-  executable years into the future.
+  Reproducible analysis requires more than clean, well-documented code; it
+  demands the precise management of software dependencies, computational
+  environments, and workflow execution. While researchers can combine tools like
+  [Docker]{.pkg}, [renv]{.pkg} for [R]{.proglang} packages (or [uv]{.pkg} for
+  [Python]{.proglang} packages) and [Make]{.pkg} to address these needs, the
+  resulting toolchain is often complex and fragile, requiring the coordination
+  of multiple disparate systems. This paper introduces a unified framework built
+  on the [Nix]{.pkg} package manager, made accessible through two [R]{.proglang}
+  packages. The first, [rix]{.pkg}, generates declarative [Nix]{.pkg}
+  expressions to define version-pinned, reproducible environments that encompass
+  [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system-level
+  dependencies. The second, [rixpress]{.pkg} (and its [Python]{.proglang} port,
+  [ryxpress]{.pkg}), leverages these environments to orchestrate polyglot
+  pipelines where each computational step runs in a hermetically sealed,
+  language-specific environment with automatic caching. By treating the entire
+  computational environment as code, this integrated approach ensures a high
+  degree of reproducibility, supports collaboration across heterogeneous
+  systems, and helps guarantee that analyses remain executable long into the
+  future.
 

### 2025-10-26 docker-polyglot-example/rbc/functions/functions.R
setup docker example

@@ -1,0 +1,53 @@
+library(ggplot2)
+library(dplyr)
+
+#' Create a plot comparing actual vs. predicted output
+#'
+#' This function takes a data frame with actual and predicted time series data
+#' and generates a ggplot visualization.
+#'
+#' @param predictions_df A data frame containing columns: 'period',
+#'   'actual_output', and 'predicted_output'. This data frame will be the
+#'   output of the Python XGBoost step.
+#' @return A ggplot object.
+#'
+plot_predictions <- function(predictions_df) {
+  # Create the plot object
+  p <- ggplot(predictions_df, aes(x = period)) +
+
+    # Add a line for the actual output from the RBC model simulation
+    geom_line(
+      aes(y = actual_output, color = "Actual (RBC Model)"),
+      linewidth = 1
+    ) +
+
+    # Add a dashed line for the XGBoost model's predictions
+    geom_line(
+      aes(y = predicted_output, color = "Predicted (XGBoost)"),
+      linetype = "dashed",
+      linewidth = 1
+    ) +
+
+    # Define custom colors and legend labels
+    scale_color_manual(
+      name = "Series",
+      values = c("Actual (RBC Model)" = "blue", "Predicted (XGBoost)" = "red")
+    ) +
+
+    # Add informative labels and a title
+    labs(
+      title = "XGBoost Prediction of RBC Model Output",
+      subtitle = "Forecasting next-quarter output based on current-quarter economic variables",
+      x = "Time (Quarters)",
+      y = "Output (Log-deviations from steady state)"
+    ) +
+
+    # Use a clean theme
+    theme_minimal() +
+
+    # Improve legend position
+    theme(legend.position = "bottom")
+
+  # Return the final ggplot object
+  return(p)
+}

### 2025-10-26 docker-polyglot-example/rbc/functions/functions.jl

@@ -1,0 +1,103 @@
+using LinearAlgebra, Distributions, DataFrames, Arrow, Random
+
+#-------------------------------------------------------------------------------
+# 1. Main Simulation Function
+#-------------------------------------------------------------------------------
+
+"""
+    simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
+
+    Takes RBC model parameters as input, solves for the state-space representation
+    using the method of undetermined coefficients (as per the slides), simulates
+    the model for 250 quarters, and returns a DataFrame.
+"""
+function simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
+
+    # --- STEADY-STATE VALUES AND CONSTRUCTED PARAMETERS (Slides p. 16, 31) ---
+    y_k = ((1/β) - 1 + δ) / α
+
+    # --- SOLVE FOR POLICY FUNCTION COEFFICIENTS (Slides p. 19-22) ---
+    # We solve for the coefficients of the policy functions:
+    # c_hat_t = c_ck * k_hat_{t-1} + c_cz * z_hat_t
+    # k_hat_t = c_kk * k_hat_{t-1} + c_kz * z_hat_t
+
+    # Solve the quadratic equation for c_ck (coefficient of capital on consumption)
+    # Based on Campbell (1994) and slide 22, we take the stable root.
+    # a*x^2 + b*x + c = 0
+    a_quad = y_k - δ
+    b_quad = -( (1-β*(1-δ))*(y_k - δ) + y_k*(1+α) + 1 )
+    c_quad = y_k * α
+
+    # The stable solution is the smaller positive root
+    c_ck = (-b_quad - sqrt(b_quad^2 - 4*a_quad*c_quad)) / (2*a_quad)
+
+    # Now solve for the other coefficients based on c_ck
+    c_kk = (y_k * α) / (y_k - δ - c_ck)
+    c_cz = (y_k * (1 - c_kk * (1 - α) * β * ρ)) /
+           ( (y_k - δ - c_ck) * (1 - β * ρ) + σ * (1 - c_ck) * (1 - β * ρ) )
+    c_kz = (c_cz * (1 - c_ck)) / (y_k * (1 - α))
+
+    # --- BUILD STATE-SPACE MATRICES FROM SOLVED COEFFICIENTS ---
+    # State vector: s_t = [k_{t-1}, z_t]'
+    # Transition:   s_t = T*s_{t-1} + R*ε_t
+
+    T = [c_kk  c_kz * ρ
[…]

### 2025-10-26 docker-polyglot-example/rbc/functions/functions.py

@@ -1,0 +1,90 @@
+import pandas as pd
+import pyarrow.feather as feather
+from sklearn.model_selection import train_test_split
+import xgboost as xgb
+
+#-------------------------------------------------------------------------------
+# Step 1: Feature Engineering
+#-------------------------------------------------------------------------------
+def prepare_features(simulated_df: pd.DataFrame) -> pd.DataFrame:
+    """Takes the raw simulated data and creates lagged features and the target variable."""
+    df = simulated_df.copy()
+
+    # Create lagged features
+    for col in ['output', 'consumption', 'investment', 'capital', 'technology']:
+        df[f'{col}_lag1'] = df[col].shift(1)
+
+    # Drop the first row which now has NaNs
+    df.dropna(inplace=True)
+
+    return df
+
+#-------------------------------------------------------------------------------
+# Step 2: Data Splitting Functions
+#-------------------------------------------------------------------------------
+def get_X_train(processed_df: pd.DataFrame):
+    """Gets the training features (X_train) from the processed data."""
+    features = [col for col in processed_df.columns if '_lag1' in col]
+    X = processed_df[features]
+    train_size = int(0.75 * len(X))
+    return X[:train_size]
+
+def get_y_train(processed_df: pd.DataFrame):
+    """Gets the training target (y_train) from the processed data."""
+    y = processed_df['output']
+    train_size = int(0.75 * len(y))
+    return y[:train_size]
+
+def get_X_test(processed_df: pd.DataFrame):
+    """Gets the testing features (X_test) from the processed data."""
+    features = [col for col in processed_df.columns if '_lag1' in col]
+    X = processed_df[features]
+    train_size = int(0.75 * len(X))
+    return X[train_size:]
+
+def get_y_test(processed_df: pd.DataFrame):
+    """Gets the testing target (y_test) from the processed data."""
+    y = processed_df['output']
+    train_size = int(0.75 * len(y))
+    return y[train_size:]
+
+#-------------------------------------------------------------------------------
+# Step 3: Model Training
+#-------------------------------------------------------------------------------
+def train_model(X_train: pd.DataFrame, y_train: pd.Series):
+    """Initializes and trains the XGBoost model."""
+    model = xgb.XGBRegressor(
+        objective='reg:squarederror',
+        n_estimators=100,
+        learning_rate=0.1,
+        max_depth=3,
+        random_state=42
+    )
+    model.fit(X_train, y_train)
+    return model
+
+#-------------------------------------------------------------------------------
+# Step 4: Prediction
[…]

### 2025-10-26 docker-polyglot-example/rbc/readme.qmd
docker example

---
title: "RBC Model Analysis Report"
format: 
  html:
    embed-resources: true
    toc: true
---

## Model Performance

The plot below compares the actual output from the RBC model simulation against the predictions generated by the XGBoost model.

```{r}
#| echo: false
#| label: rbc-plot
#| fig-cap: "Comparison of Actual RBC Model Output vs. XGBoost Predictions"

# This chunk now simply includes the plot that was generated by the 
# 03_visualize_results.R script in the pipeline.
knitr::include_graphics("plots/output_plot.png")
```

### 2025-10-26 docker-polyglot-example/rbc/01_simulate_rbc.jl

@@ -1,0 +1,41 @@
+# ==============================================================================
+# 01_simulate_rbc.jl
+#
+# Simulates data from a Real Business Cycle (RBC) model and saves it.
+#
+# Usage: julia 01_simulate_rbc.jl <output_path>
+# Example: julia 01_simulate_rbc.jl data/simulated_rbc_data.arrow
+# ==============================================================================
+
+# Load required packages
+using Distributions, DataFrames, Arrow, Random
+
+# Include the helper functions from the separate file.
+include("functions/functions.jl")
+
+# Check if the output path is provided as a command-line argument.
+if length(ARGS) != 1
+    error("Usage: julia 01_simulate_rbc.jl <output_path>")
+end
+
+# Get the output path from the first command-line argument.
+output_path = ARGS[1]
+
+# --- Define RBC Model Parameters ---
+# These were previously defined as individual steps in the rixpress script.
+α = 0.3         # Capital's share of income
+β = 1 / 1.01    # Discount factor
+δ = 0.025       # Depreciation rate
+ρ = 0.95        # Technology shock persistence
+σ = 1.0         # Risk aversion (log-utility)
+σ_z = 0.01      # Technology shock standard deviation
+
+# --- Run the Simulation ---
+println("Simulating RBC model...")
+simulated_df = simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
+
+# --- Save the Output ---
+println("Saving simulated data to: ", output_path)
+arrow_write(simulated_df, output_path)
+
+println("Julia script finished successfully.")

### 2025-10-26 docker-polyglot-example/rbc/02_train_model.py

@@ -1,0 +1,56 @@
+# ==============================================================================
+# 02_train_model.py
+#
+# Loads simulated data, trains an XGBoost model, and saves predictions.
+#
+# Usage: python 02_train_model.py <input_path> <output_path>
+# Example: python 02_train_model.py data/simulated_rbc_data.arrow data/predictions.arrow
+# ==============================================================================
+
+import sys
+import pandas as pd
+import pyarrow.feather as feather
+import xgboost as xgb
+from functions.functions import * # Import all helper functions
+
+def main():
+    # Check for correct command-line arguments.
+    if len(sys.argv) != 3:
+        print("Usage: python 02_train_model.py <input_path> <output_path>")
+        sys.exit(1)
+
+    input_path = sys.argv[1]
+    output_path = sys.argv[2]
+
+    # --- Load Data ---
+    print(f"Loading simulated data from: {input_path}")
+    simulated_df = feather.read_feather(input_path)
+
+    # --- Feature Engineering ---
+    print("Preparing features...")
+    processed_df = prepare_features(simulated_df)
+
+    # --- Data Splitting ---
+    print("Splitting data into training and testing sets...")
+    X_train = get_X_train(processed_df)
+    y_train = get_y_train(processed_df)
+    X_test = get_X_test(processed_df)
+    y_test = get_y_test(processed_df)
+
+    # --- Model Training ---
+    print("Training XGBoost model...")
+    model = train_model(X_train, y_train)
+
+    # --- Prediction ---
+    print("Making predictions on the test set...")
+    predictions = make_predictions(model, X_test)
+
+    # --- Format and Save Results ---
+    print("Formatting and saving results...")
+    results_df = format_results(y_test, predictions)
+    save_arrow(results_df, output_path)
+
+    print(f"Python script finished. Predictions saved to: {output_path}")
+
+if __name__ == "__main__":
+    main()

### 2025-10-26 docker-polyglot-example/rbc/03_visualize_results.R

@@ -1,0 +1,48 @@
+# ==============================================================================
+# 03_visualize_results.R
+#
+# Loads model predictions and creates a visualization.
+#
+# Usage: Rscript 03_visualize_results.R <input_path> <output_path>
+# Example: Rscript 03_visualize_results.R data/predictions.arrow plots/output_plot.png
+# ==============================================================================
+
+# Load required packages
+library(arrow)
+library(ggplot2)
+library(dplyr)
+
+# Source the helper function containing the plotting logic.
+source("functions/functions.R")
+
+# Get command-line arguments.
+args <- commandArgs(trailingOnly = TRUE)
+if (length(args) != 2) {
+  stop(
+    "Usage: Rscript 03_visualize_results.R <input_path> <output_path>",
+    call. = FALSE
+  )
+}
+
+input_path <- args[1]
+output_path <- args[2]
+
+# --- Load Data ---
+cat("Loading prediction data from:", input_path, "\n")
+predictions_df <- read_feather(input_path)
+
+# --- Generate Plot ---
+cat("Generating plot...\n")
+output_plot <- plot_predictions(predictions_df)
+
+# --- Save Plot ---
+cat("Saving plot to:", output_path, "\n")
+ggsave(
+  output_path,
+  plot = output_plot,
+  width = 10,
+  height = 6,
+  dpi = 300
+)
+
+cat("R script finished successfully.\n")

### 2025-10-26 docker-polyglot-example/rbc/functions/functions.R

@@ -1,5 +1,5 @@
-library(ggplot2)
-library(dplyr)
+#library(ggplot2)
+#library(dplyr)
 
 #' Create a plot comparing actual vs. predicted output
 #'

### 2025-10-26 docker-polyglot-example/rbc/functions/functions.jl

@@ -1,4 +1,4 @@
-using LinearAlgebra, Distributions, DataFrames, Arrow, Random
+#using LinearAlgebra, Distributions, DataFrames, Arrow, Random
 
 #-------------------------------------------------------------------------------
 # 1. Main Simulation Function

### 2025-10-26 case-study/nix/readme.qmd
cleanup

---
title: "A Polyglot Data Science Pipeline with rixpress"
format:
  html:
    embed-resources: true
    toc: true
---

```{r, include = FALSE}
library(rixpress)
```

*Disclaimer: The code and this text were written with the assistance of LLMs.*

## Introduction: Showcasing Reproducible Pipelines

This document demonstrates the power of the `{rixpress}` R package to build,
orchestrate, and execute a fully reproducible, polyglot data science pipeline.
The primary goal is to showcase `{rixpress}`'s capabilities, not to conduct a
deep economic analysis.

To provide a compelling example, our pipeline will perform a task common in
modern quantitative research:

1.  **Simulate** synthetic economic data using a macroeconomic model in **Julia**.
2.  **Train** a machine learning model on this data to make forecasts in **Python**.
3.  **Visualize** the model's performance in **R**.

I want to insist that this is merely a toy example project to illustrate the
functionality of `{rixpress}`. I make no claims as to the accuracy of the LLM-generated
code for model simulation.

### The Economic Model: A Means to an End

We will use a foundational model from modern macroeconomics—the **Real Business
Cycle (RBC) model**—to generate our data. The theory and log-linearized
equations used in our Julia script are taken directly from the excellent lecture
slides by Bianca De Paoli.

::: {.callout-note}
### Source Attribution

The economic theory and equations for the RBC model are based on the following
source. This document serves as a guide for our implementation.

**Source:** De Paoli, B. (2009). *Slides 1: The RBC Model, Analytical and
Numerical solutions*.
[https://personal.lse.ac.uk/depaoli/RBC_slides1.pdf](https://personal.lse.ac.uk/depaoli/RBC_slides1.pdf)
:::

Our Julia script implements the state-space solution to this system.

## The Pipeline Components

Our project consists of a set of scripts, each with a dedicated purpose, all
orchestrated by `{rixpress}`.

### 1. The Environment (`gen-env.R`)

Reproducibility starts with the environment. We use the `{rix}` package to
declaratively define a `default.nix` file. This file locks down the exact
versions of R, Julia, Python, and all their respective packages, ensuring the
pipeline runs identically today, tomorrow, or on any machine with Nix installed.

```r
#| eval: false
#| code-summary: "Environment Definition with {rix}"
s
# This script defines the polyglot environment our pipeline will run in.
library(rix)

# Define the complete execution environment
rix(
  # Pin the environment to a specific date to ensure that all package
  # versions are resolved as they were on this day.
  date = "2025-10-14",

  # 1. R Packages
  # We need packages for plotting, data manipulation, and reading arrow files.
  r_pkgs = c(
    "ggplot2",
    "dplyr",
    "arrow",
    "rix",
    "rixpress"
  ),

  # 2. Julia Configuration
  # We specify the Julia version and the list of packages needed
  # for our manual RBC model simulation.
  jl_conf = list(
    jl_version = "lts",
    jl_pkgs = c(
      "Distributions", # For creating random shocks
      "DataFrames", # For structuring the output
      "Arrow", # For saving the data in a cross-language format
      "Random"
    )
  ),

  # 3. Python Configuration
  # We specify the Python version and the packages needed for the
  # machine learning step.
  py_conf = list(
    py_version = "3.13",
    py_pkgs = c(
      "pandas",
      "scikit-learn",
      "xgboost",
      "pyarrow"
    )
  ),

  # We set the IDE to 'none' for a minimal environment. You could change
  # this to "rstudio" if you prefer to work interactively in RStudio.
  ide = "none",

  # Define the project path and allow overwriting the default.nix file.
  project_path = ".",
  overwrite = TRUE
)

```

Assuming we are working on a system that has Nix installed, we can "drop" into
a temporary Nix shell with R and `{rix}` available:

```r
nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
```

Once the shell is ready, start R (by simply typing `R`) and run `source("gen-env.R")`
to generate the adequate `default.nix` file. This is the environment that we can use
to work interactively with the code while developing, and in which the pipeline will
be executed.

### 2. The Simulation (`functions.jl`)

This Julia script contains a pure function that simulates the RBC model and
returns a DataFrame. The code is a direct implementation of the state-space
solution derived from the equations above and is saved under
`functions/functions.jl`.

```julia
#| eval: false
#| code-summary: "Julia: RBC Model Simulation"

# This script contains the helper functions for the Julia portion of the pipeline.
# It implements the correct, stable, and canonical log-linearized solution to
# a standard Real Business Cycle (RBC) model, based on the provided lecture slides.

#using LinearAlgebra, Distributions, DataFrames, Arrow, Random

#-------------------------------------------------------------------------------
# 1. Main Simulation Function
#-------------------------------------------------------------------------------

"""
    simulate_rbc_model(α, β, δ, ρ, σ, σ_z)

    Takes RBC model parameters as input, solves for the state-space representation
    using the method of undetermined coefficients (as per the slides), simulates
    the model for 250 quarters, and returns a DataFrame.
"""
function simulate_rbc_model(α, β, δ, ρ, σ, σ_z)

    # --- STEADY-STATE VALUES AND CONSTRUCTED PARAMETERS (Slides p. 16, 31) ---
    y_k = ((1/β) - 1 + δ) / α

    # --- SOLVE FOR POLICY FUNCTION COEFFICIENTS (Slides p. 19-22) ---
    # We solve for the coefficients of the policy functions:
    # c_hat_t = c_ck * k_hat_{t-1} + c_cz * z_hat_t
    # k_hat_t = c_kk * k_hat_{t-1} + c_kz * z_hat_t

    # Solve the quadratic equation for c_ck (coefficient of capital on consumption)
    # Based on Campbell (1994) and slide 22, we take the stable root.
    # a*x^2 + b*x + c = 0
    a_quad = y_k - δ
    b_quad = -( (1-β*(1-δ))*(y_k - δ) + y_k*(1+α) + 1 )
    c_quad = y_k * α

    # The stable solution is the smaller positive root
    c_ck = (-b_quad - sqrt(b_quad^2 - 4*a_quad*c_quad)) / (2*a_quad)

    # Now solve for the other coefficients based on c_ck
    c_kk = (y_k * α) / (y_k - δ - c_ck)
    c_cz = (y_k * (1 - c_kk * (1 - α) * β * ρ)) /
           ( (y_k - δ - c_ck) * (1 - β * ρ) + σ * (1 - c_ck) * (1 - β * ρ) )
    c_kz = (c_cz * (1 - c_ck)) / (y_k * (1 - α))

    # --- BUILD STATE-SPACE MATRICES FROM SOLVED COEFFICIENTS ---
    # State vector: s_t = [k_{t-1}, z_t]'
    # Transition:   s_t = T*s_{t-1} + R*ε_t

    T = [c_kk  c_kz * ρ
         0     ρ      ]

    R = [c_kz ; 1]

    # Observation Matrix for [output, consumption, investment]
    # y_hat_t = α*k_{t-1} + z_t
    # c_hat_t = c_ck*k_{t-1} + c_cz*z_t
    # i_hat_t = (i_y_ss)^-1 * (k_t - (1-δ)k_{t-1})
    #         = (i_y_ss)^-1 * ( (c_kk - (1-δ))k_{t-1} + c_kz*z_t )

    i_y_ss = δ * (α / ((1/β) - 1 + δ))

    C = [α      1
         c_ck   c_cz
         (c_kk - (1-δ))/i_y_ss   c_kz/i_y_ss]

    # --- SIMULATE THE MODEL ---
    Random.seed!(1234)
    n_periods = 250
    shocks = randn(n_periods) * σ_z

    states = zeros(2, n_periods)
    for t in 2:n_periods
        # The state is [k_{t-1}, z_t]'
        # To get the next state [k_t, z_{t+1}]', we first find k_t
        k_t = T[1,1]*states[1, t] + T[1,2]*states[2, t] + R[1]*shocks[t]
        z_tp1 = T[2,1]*states[1, t] + T[2,2]*states[2, t] + R[2]*shocks[t]

        states[1, t] = k_t # This is now k_t, which is "k_lag" for t+1
        states[2, t] = z_tp1 # This is now z_{t+1}
    end
    # Re-aligning states to be [k_{t-1}, z_t]
    k_lag = [0; states[1, 1:end-1]]
    z = [0; states[2, 1:end-1]]

    observables = C * [k_lag'; z']

    df = DataFrame(
        period = 1:n_periods,
        output = observables[1, :],
        consumption = observables[2, :],
        investment = observables[3, :],
        capital = k_lag,
        technology = z
    )

    return df
end

#-------------------------------------------------------------------------------
# 2. Encoder Function (Unchanged)
#-------------------------------------------------------------------------------
"""
    arrow_write(df::DataFrame, path::String)
    Encoder function to save a Julia DataFrame to an Arrow file.
"""
function arrow_write(df::DataFrame, path::String)
    Arrow.write(path, df)
end

```

At the end of the script, we create a wrapper around `Arrow.write()` called
`arrow_write()` to serialize the generated data frame into an Arrow file.

While developing, it is possible to start the Julia interpreter and test things
and see if they work. The generated data frame looks like:

```{r}
# This is an Arrow data file, so rxp_read(),
# not knowing how to read it, will simply return
# the path to the file in the Nix store.
# We can pass this file to the adequate reader.
rxp_read("simulated_rbc_data") |>
  arrow::read_feather() |>
  head()
```

### 3. The Prediction Model (`functions.py`)

This Python script defines a function to perform our machine learning task. It
takes the DataFrame from the Julia step, trains an XGBoost model to predict
next-quarter's output, and returns a new DataFrame containing the predictions.
Of course, we could have simply used the RBC model itself for the predictions,
but again, this is a toy example.

The Python script contains several little functions to make the code as modular
as possible.

```python
#| eval: false
#| code-summary: "Python: XGBoost Forecasting"

# This script contains modular helper functions for the Python portion of the pipeline.
# Each function performs a single, distinct task in the ML workflow.

# 1. Load required Python packages
# This is only needed if you run this script by hand,
# in which case, uncomment lines 8 to 11.
# With rixpress, the packages get loaded automatically.
#import pandas as pd
#import pyarrow.feather as feather
#from sklearn.model_selection import train_test_split
#import xgboost as xgb
# This script contains the helper functions for the Python portion of the pipeline.
# It defines the ML model training and prediction logic as a pure function.
# This script contains the helper functions for the Python portion of the pipeline.
# It defines the ML model training and prediction logic as a pure function.

# Step 1: Feature Engineering

def prepare_features(simulated_df: pd.DataFrame) -> pd.DataFrame:
    """Takes the raw simulated data and creates lagged features and the target variable."""
    df = simulated_df.copy()

    # Create lagged features
    for col in ['output', 'consumption', 'investment', 'capital', 'technology']:
        df[f'{col}_lag1'] = df[col].shift(1)

    # Drop the first row which now has NaNs
    df.dropna(inplace=True)

    return df

# Step 2: Data Splitting Functions
def get_X_train(processed_df: pd.DataFrame):
    """Gets the training features (X_train) from the processed data."""
    features = [col for col in processed_df.columns if '_lag1' in col]
    X = processed_df[features]
    train_size = int(0.75 * len(X))
    return X[:train_size]

def get_y_train(processed_df: pd.DataFrame):
    """Gets the training target (y_train) from the processed data."""
    y = processed_df['output']
    train_size = int(0.75 * len(y))
    return y[:train_size]

def get_X_test(processed_df: pd.DataFrame):
    """Gets the testing features (X_test) from the processed data."""
    features = [col for col in processed_df.columns if '_lag1' in col]
    X = processed_df[features]
    train_size = int(0.75 * len(X))
    return X[train_size:]

def get_y_test(processed_df: pd.DataFrame):
    """Gets the testing target (y_test) from the processed data."""
    y = processed_df['output']
    train_size = int(0.75 * len(y))
    return y[train_size:]

# Step 3: Model Training
def train_model(X_train: pd.DataFrame, y_train: pd.Series):
    """Initializes and trains the XGBoost model."""
    model = xgb.XGBRegressor(
        objective='reg:squarederror',
        n_estimators=100,
        learning_rate=0.1,
        max_depth=3,
        random_state=42
    )
    model.fit(X_train, y_train)
    return model

# Step 4: Prediction
def make_predictions(model, X_test: pd.DataFrame):
    """Makes predictions on the test set using the trained model."""
    return model.predict(X_test)

# Step 5: Format Final Results
def format_results(y_test: pd.Series, predictions) -> pd.DataFrame:
    """Combines test data and predictions into a final DataFrame for R."""
    results_df = pd.DataFrame({
        'period': y_test.index,
        'actual_output': y_test.values,
        'predicted_output': predictions
    })
    return results_df

# Encoder Function (for saving the final output)
def save_arrow(df: pd.DataFrame, path: str):
    """Encoder function to save a pandas DataFrame to an Arrow file."""
    feather.write_feather(df, path)
```

Just as for the Julia script, I write a wrapper around `feather.write_feather()`
to serialize the data into an Arrow file and pass it finally to R. Just as before,
it is possible to start a Python interpreter as defined in the `default.nix` file
and use to test and debug.

### 4. The Visualization (`functions.R`)

Finally, this R script contains a function that takes the predictions from the
Python step and creates a publication-quality visualization using `{ggplot2}`.

```r
#| eval: false
#| code-summary: "R: ggplot2 Visualization"

# This script contains the helper functions for the R portion of the pipeline.
# It defines the visualization logic using ggplot2.

# 1. Load required R packages
# This is only needed if you run this script by hand,
# in which case, uncomment lines 8 and 9.
# With rixpress, the packages get loaded automatically.

#library(ggplot2)
#library(dplyr)

#-------------------------------------------------------------------------------
# 2. Main Function: Create the Visualization
#-------------------------------------------------------------------------------

#' Create a plot comparing actual vs. predicted output
#'
#' This function takes a data frame with actual and predicted time series data
#' and generates a ggplot visualization.
#'
#' @param predictions_df A data frame containing columns: 'period',
#'   'actual_output', and 'predicted_output'. This data frame will be the
#'   output of the Python XGBoost step.
#' @return A ggplot object.
#'
plot_predictions <- function(predictions_df) {
  # Create the plot object
  p <- ggplot(predictions_df, aes(x = period)) +

    # Add a line for the actual output from the RBC model simulation
    geom_line(
      aes(y = actual_output, color = "Actual (RBC Model)"),
      linewidth = 1
    ) +

    # Add a dashed line for the XGBoost model's predictions
    geom_line(
      aes(y = predicted_output, color = "Predicted (XGBoost)"),
      linetype = "dashed",
      linewidth = 1
    ) +

    # Define custom colors and legend labels
    scale_color_manual(
      name = "Series",
      values = c("Actual (RBC Model)" = "blue", "Predicted (XGBoost)" = "red")
    ) +

    # Add informative labels and a title
    labs(
      title = "XGBoost Prediction of RBC Model Output",
      subtitle = "Forecasting next-quarter output based on current-quarter economic variables",
      x = "Time (Quarters)",
      y = "Output (Log-deviations from steady state)"
    ) +

    # Use a clean theme
    theme_minimal() +

    # Improve legend position
    theme(legend.position = "bottom")

  # Return the final ggplot object
  return(p)
}

```

This simply returns the following plot:

```{r}
rxp_read("output_plot")
```

## Orchestration with `{rixpress}`

The `gen-pipeline.R` script is the "conductor" of our orchestra. It uses
`{rixpress}`'s intuitive, R-based syntax to define the dependencies and data
flow between our multi-language components.

Notice how the output of one step becomes the input to the next, and how
`{rixpress}` uses `encoder` and `decoder` arguments to seamlessly handle the
serialization (saving to disk) and deserialization (reading from disk) of data
between languages using the wrappers we wrote.

```r
#| eval: false
#| code-summary: "The {rixpress} Pipeline Definition"

# This script defines and orchestrates the entire reproducible analytical pipeline
# using the {rixpress} package.

library(rixpress)

# Define the full pipeline as a list of derivations and pipe it to rxp_populate()
# to generate the Nix build instructions and execute the pipeline.
list(
  # STEP 0: Define RBC Model Parameters as Derivations
  # This makes the parameters an explicit part of the pipeline.
  # Changing a parameter will cause downstream steps to rebuild.
  rxp_jl(alpha, 0.3), # Capital's share of income
  rxp_jl(beta, 1 / 1.01), # Discount factor
  rxp_jl(delta, 0.025), # Depreciation rate
  rxp_jl(rho, 0.95), # Technology shock persistence
  rxp_jl(sigma, 1.0), # Risk aversion (log-utility)
  rxp_jl(sigma_z, 0.01), # Technology shock standard deviation

  # STEP 1: Julia - Simulate a Real Business Cycle (RBC) model.
  # This derivation runs our Julia script to generate the source data.
  rxp_jl(
    name = simulated_rbc_data,
    expr = "simulate_rbc_model(alpha, beta, delta, rho, sigma, sigma_z)",
    user_functions = "functions/functions.jl", # The file containing the function
    encoder = "arrow_write" # The function to use for saving the output
  ),

 # STEP 2.1: Python - Prepare features (lagging data)
  rxp_py(
    name = processed_data,
    expr = "prepare_features(simulated_rbc_data)",
    user_functions = "functions/functions.py",
    # Decode the Arrow file from Julia into a pandas DataFrame
    decoder = "feather.read_feather"
    # Note: No encoder needed here. {rixpress} will use pickle by default
    # to pass the DataFrame between Python steps.
  ),

  # STEP 2.2: Python - Split data into training and testing sets
  rxp_py(
    name = X_train,
    expr = "get_X_train(processed_data)",
    user_functions = "functions/functions.py"
  ),

  rxp_py(
    name = y_train,
    expr = "get_y_train(processed_data)",
    user_functions = "functions/functions.py"
  ),

  rxp_py(
    name = X_test,
    expr = "get_X_test(processed_data)",
    user_functions = "functions/functions.py"
  ),

  rxp_py(
    name = y_test,
    expr = "get_y_test(processed_data)",
    user_functions = "functions/functions.py"
  ),

  # STEP 2.3: Python - Train the model
  rxp_py(
    name = trained_model,
    expr = "train_model(X_train, y_train)",
    user_functions = "functions/functions.py"
  ),

  # STEP 2.4: Python - Make predictions
  rxp_py(
    name = model_predictions,
    expr = "make_predictions(trained_model, X_test)",
    user_functions = "functions/functions.py"
  ),

  # STEP 2.5: Python - Format final results for R
  rxp_py(
    name = predictions,
    expr = "format_results(y_test, model_predictions)",
    user_functions = "functions/functions.py",
    # We need an encoder here to save the final DataFrame as an Arrow file
    # so the R step can read it.
    encoder = "save_arrow"
  ),

  # STEP 3: R - Visualize the predictions from the Python model.
  # This final derivation depends on the output of the Python step.
  rxp_r(
    name = output_plot,
    expr = plot_predictions(predictions), # The function to call from functions.R
    user_functions = "functions/functions.R",
    # Specify how to load the upstream data (from Python) into R.
    decoder = arrow::read_feather
  ),

  # STEP 4: Quarto - Compile the final report.
  rxp_qmd(
    name = final_report,
    qmd_file = "readme.qmd"
  )
) |>
  rxp_populate(
    py_imports = c(
      pandas = "import pandas as pd",
      pyarrow = "import pyarrow.feather as feather",
      sklearn = "from sklearn.model_selection import train_test_split",
      xgboost = "import xgboost as xgb"
    ),
    project_path = ".", # The root of our project
    build = TRUE, # Set to TRUE to execute the pipeline immediately
    verbose = 0
  )

# Plot DAG for CI
rxp_dag_for_ci()

```

To run the pipeline, start the environment by simply typing `nix-shell` in the terminal,
which will use the environment as defined in the `default.nix` and run `source("gen-pipeline.R")`.
If there are issues, setting the `verbose` argument of `rxp_populate()` to `1` is helpful
to see the full error logs.

Calling `rxp_ggdag()` shows the following plot:

```{r}
rxp_ggdag()
```

In an interactive session, one can use `rxp_read()`, `rxp_load()` to read or
load artifacts into the R session respectively and `rxp_trace()` to check the
lineage of an artifact. The pipeline could also be built from a Python session,
using `ryxpress`, the Python port of `{rixpress}`.

## The Final Result

After running the entire pipeline, the final artifact is a `{ggplot2}` object
named `output_plot`, which is saved in the Nix store. We can load and display it
directly in this document using the `rixpress::rxp_read()` function.

## Conclusion

This project successfully demonstrates how `{rixpress}` leverages the Nix
package manager to create a robust, end-to-end reproducible pipeline that
seamlessly integrates Julia, Python, and R. By defining each component as a pure
function and orchestrating the data flow with `{rixpress}`, we ensure that the
entire workflow—from the environment to the final plot—can be reproduced
perfectly on any machine.

---
::: {.callout-tip}
### How to Render This Document

To render this Quarto file and display the plot, you must first run the entire
pipeline as described in the previous steps. Then, from your terminal, execute
the following command:

`nix-shell --run "quarto render index.qmd"`
:::

### 2025-10-26 case-study/nix/functions/functions.jl

@@ -1,0 +1,107 @@
+# This script contains the helper functions for the Julia portion of the pipeline.
+# It implements the correct, stable, and canonical log-linearized solution to
+# a standard Real Business Cycle (RBC) model, based on the provided lecture slides.
+
+#using LinearAlgebra, Distributions, DataFrames, Arrow, Random
+
+#-------------------------------------------------------------------------------
+# 1. Main Simulation Function
+#-------------------------------------------------------------------------------
+
+"""
+    simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
+
+    Takes RBC model parameters as input, solves for the state-space representation
+    using the method of undetermined coefficients (as per the slides), simulates
+    the model for 250 quarters, and returns a DataFrame.
+"""
+function simulate_rbc_model(α, β, δ, ρ, σ, σ_z)
+
+    # --- STEADY-STATE VALUES AND CONSTRUCTED PARAMETERS (Slides p. 16, 31) ---
+    y_k = ((1/β) - 1 + δ) / α
+
+    # --- SOLVE FOR POLICY FUNCTION COEFFICIENTS (Slides p. 19-22) ---
+    # We solve for the coefficients of the policy functions:
+    # c_hat_t = c_ck * k_hat_{t-1} + c_cz * z_hat_t
+    # k_hat_t = c_kk * k_hat_{t-1} + c_kz * z_hat_t
+
+    # Solve the quadratic equation for c_ck (coefficient of capital on consumption)
+    # Based on Campbell (1994) and slide 22, we take the stable root.
+    # a*x^2 + b*x + c = 0
+    a_quad = y_k - δ
+    b_quad = -( (1-β*(1-δ))*(y_k - δ) + y_k*(1+α) + 1 )
+    c_quad = y_k * α
+
+    # The stable solution is the smaller positive root
+    c_ck = (-b_quad - sqrt(b_quad^2 - 4*a_quad*c_quad)) / (2*a_quad)
+
+    # Now solve for the other coefficients based on c_ck
+    c_kk = (y_k * α) / (y_k - δ - c_ck)
+    c_cz = (y_k * (1 - c_kk * (1 - α) * β * ρ)) /
+           ( (y_k - δ - c_ck) * (1 - β * ρ) + σ * (1 - c_ck) * (1 - β * ρ) )
[…]

### 2025-10-26 case-study/nix/functions/functions.py

@@ -1,0 +1,101 @@
+# This script contains modular helper functions for the Python portion of the pipeline.
+# Each function performs a single, distinct task in the ML workflow.
+
+# 1. Load required Python packages
+# This is only needed if you run this script by hand,
+# in which case, uncomment lines 8 to 11.
+# With rixpress, the packages get loaded automatically.
+#import pandas as pd
+#import pyarrow.feather as feather
+#from sklearn.model_selection import train_test_split
+#import xgboost as xgb
+# This script contains the helper functions for the Python portion of the pipeline.
+# It defines the ML model training and prediction logic as a pure function.
+# This script contains the helper functions for the Python portion of the pipeline.
+# It defines the ML model training and prediction logic as a pure function.
+
+#-------------------------------------------------------------------------------
+# Step 1: Feature Engineering
+#-------------------------------------------------------------------------------
+def prepare_features(simulated_df: pd.DataFrame) -> pd.DataFrame:
+    """Takes the raw simulated data and creates lagged features and the target variable."""
+    df = simulated_df.copy()
+
+    # Create lagged features
+    for col in ['output', 'consumption', 'investment', 'capital', 'technology']:
+        df[f'{col}_lag1'] = df[col].shift(1)
+
+    # Drop the first row which now has NaNs
+    df.dropna(inplace=True)
+
+    return df
+
+#-------------------------------------------------------------------------------
+# Step 2: Data Splitting Functions
+#-------------------------------------------------------------------------------
+def get_X_train(processed_df: pd.DataFrame):
+    """Gets the training features (X_train) from the processed data."""
+    features = [col for col in processed_df.columns if '_lag1' in col]
+    X = processed_df[features]
+    train_size = int(0.75 * len(X))
+    return X[:train_size]
+
+def get_y_train(processed_df: pd.DataFrame):
+    """Gets the training target (y_train) from the processed data."""
+    y = processed_df['output']
+    train_size = int(0.75 * len(y))
+    return y[:train_size]
+
+def get_X_test(processed_df: pd.DataFrame):
+    """Gets the testing features (X_test) from the processed data."""
+    features = [col for col in processed_df.columns if '_lag1' in col]
+    X = processed_df[features]
+    train_size = int(0.75 * len(X))
+    return X[train_size:]
+
+def get_y_test(processed_df: pd.DataFrame):
+    """Gets the testing target (y_test) from the processed data."""
+    y = processed_df['output']
+    train_size = int(0.75 * len(y))
+    return y[train_size:]
+
[…]

### 2025-10-26 case-study/nix/gen-env.R

@@ -1,0 +1,63 @@
+# This script defines the polyglot environment our pipeline will run in.
+# It uses the {rix} package to create a 'default.nix' file, which serves
+# as a blueprint for a fully reproducible environment.
+
+library(rix)
+
+# Define the complete execution environment
+rix(
+  # Pin the environment to a specific date to ensure that all package
+  # versions are resolved as they were on this day.
+  date = "2025-10-14",
+
+  # 1. R Packages
+  # We need packages for plotting, data manipulation, and reading arrow files.
+  # We also include reticulate as it can be useful for rixpress internals.
+  r_pkgs = c(
+    "ggplot2",
+    "ggdag",
+    "dplyr",
+    "arrow",
+    "rix",
+    "rixpress",
+    "quarto"
+  ),
+
+  # 2. Julia Configuration
+  # We specify the Julia version and the list of packages needed
+  # for our manual RBC model simulation.
+  jl_conf = list(
+    jl_version = "lts",
+    jl_pkgs = c(
+      "Distributions", # For creating random shocks
+      "DataFrames", # For structuring the output
+      "Arrow", # For saving the data in a cross-language format
+      "Random"
+    )
+  ),
+
+  # 3. Python Configuration
+  # We specify the Python version and the packages needed for the
+  # machine learning step.
+  py_conf = list(
+    py_version = "3.13",
+    py_pkgs = c(
+      "pandas",
+      "scikit-learn",
+      "xgboost",
+      "pyarrow",
+      "ryxpress"
+    )
+  ),
+
+  # We set the IDE to 'none' for a minimal environment. You could change
+  # this to "rstudio" if you prefer to work interactively in RStudio.
+  ide = "none",
+
+  # Define the project path and allow overwriting the default.nix file.
+  project_path = ".",
+  overwrite = TRUE
+)
+
+# A message to confirm the script has run successfully.
+message("Successfully generated 'default.nix' environment file.")

### 2025-10-26 case-study/nix/gen-pipeline.R

@@ -1,0 +1,114 @@
+# This script defines and orchestrates the entire reproducible analytical pipeline
+# using the {rixpress} package.
+
+library(rixpress)
+
+# Define the full pipeline as a list of derivations and pipe it to rxp_populate()
+# to generate the Nix build instructions and execute the pipeline.
+list(
+  # STEP 0: Define RBC Model Parameters as Derivations
+  # This makes the parameters an explicit part of the pipeline.
+  # Changing a parameter will cause downstream steps to rebuild.
+  rxp_jl(alpha, 0.3), # Capital's share of income
+  rxp_jl(beta, 1 / 1.01), # Discount factor
+  rxp_jl(delta, 0.025), # Depreciation rate
+  rxp_jl(rho, 0.95), # Technology shock persistence
+  rxp_jl(sigma, 1.0), # Risk aversion (log-utility)
+  rxp_jl(sigma_z, 0.01), # Technology shock standard deviation
+
+  # STEP 1: Julia - Simulate a Real Business Cycle (RBC) model.
+  # This derivation runs our Julia script to generate the source data.
+  rxp_jl(
+    name = simulated_rbc_data,
+    expr = "simulate_rbc_model(alpha, beta, delta, rho, sigma, sigma_z)",
+    user_functions = "functions/functions.jl", # The file containing the function
+    encoder = "arrow_write" # The function to use for saving the output
+  ),
+
+  # STEP 2.1: Python - Prepare features (lagging data)
+  rxp_py(
+    name = processed_data,
+    expr = "prepare_features(simulated_rbc_data)",
+    user_functions = "functions/functions.py",
+    # Decode the Arrow file from Julia into a pandas DataFrame
+    decoder = "feather.read_feather"
+    # Note: No encoder needed here. {rixpress} will use pickle by default
+    # to pass the DataFrame between Python steps.
+  ),
+
+  # STEP 2.2: Python - Split data into training and testing sets
+  rxp_py(
+    name = X_train,
+    expr = "get_X_train(processed_data)",
+    user_functions = "functions/functions.py"
+  ),
+
+  rxp_py(
+    name = y_train,
+    expr = "get_y_train(processed_data)",
+    user_functions = "functions/functions.py"
+  ),
+
+  rxp_py(
+    name = X_test,
+    expr = "get_X_test(processed_data)",
+    user_functions = "functions/functions.py"
+  ),
+
+  rxp_py(
+    name = y_test,
+    expr = "get_y_test(processed_data)",
[…]

### 2025-10-27 paper.qmd
I -> We

@@
 
-To make [Nix]{.pkg} accessible to researchers, I developed two [R]{.proglang}
+To make [Nix]{.pkg} accessible to researchers, we developed two [R]{.proglang}
 packages:
@@
 reproducible, project-specific environments without learning the full
-[Nix]{.proglang} language, I created the [rix]{.pkg} and [rixpress]{.pkg}
+[Nix]{.proglang} language, we created the [rix]{.pkg} and [rixpress]{.pkg}
 packages.
@@
 then be built on a system with [Nix]{.pkg} installed using the `nix-build`
-command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
+command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
 installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
@@
 
-I strive to provide an available date per week: each Monday, a GitHub Action
+We strive to provide an available date per week: each Monday, a GitHub Action
 tests popular packages on Linux and macOS, and only if all tests succeed is the
@@
 
-I have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as
+We have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as
 March 2019. The process involves checking out a `nixpkgs` commit on the selected
@@
 to build many packages from source, which can be time-consuming. To mitigate
-this, I provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
+this, we provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
 complementing the public Nix cache. Instructions for using Cachix are in
@@
 or proprietary extensions; it functions as a convenience layer providing tested
-package sets and dated snapshots. Should maintenance cease (which is unlikely,
-as I directly use it myself), users could seamlessly transition to upstream
-`nixpkgs` commits, though with trade-offs: less frequent [R]{.proglang} and
-[Julia]{.proglang} package updates. The fork's value lies not in creating
-dependency, but in reducing friction through testing infrastructure and curated
-"known-good" dates. Thus, the fork represents an enhancement to user experience
-rather than a critical single point of failure—the underlying [Nix]{.pkg}
-infrastructure and official `nixpkgs` repository remain the foundation, ensuring
-long-term viability of environments created with [rix]{.pkg}.
+package sets and dated snapshots. Should maintenance cease, users could
+seamlessly transition to upstream `nixpkgs` commits, though with trade-offs:
+less frequent [R]{.proglang} and [Julia]{.proglang} package updates. The fork's
+value lies not in creating dependency, but in reducing friction through testing
+infrastructure and curated "known-good" dates. Thus, the fork represents an
+enhancement to user experience rather than a critical single point of
+failure—the underlying [Nix]{.pkg} infrastructure and official `nixpkgs`
+repository remain the foundation, ensuring long-term viability of environments
+created with [rix]{.pkg}.
 
@@
 
-To illustrate these capabilities with a realistic research task, I present a
+To illustrate these capabilities with a realistic research task, we present a
 polyglot pipeline that simulates a Real Business Cycle (RBC) model in
@@
 However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
-[rix]{.pkg}, I aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
+[rix]{.pkg}, we aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
 by providing a familiar interface and workflow. By building reproducible
@@
 
-I thank the rOpenSci reviewers and contributors who provided valuable feedback
-on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, I am
+We thank the rOpenSci reviewers and contributors who provided valuable feedback
+on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, we are
 grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of
 [rix]{.pkg}, and to William Landau and Anthony Martinez for their reviews of
-[rixpress]{.pkg}. I also acknowledge the contributions of Philipp Baumann,
-[rix]{.pkg}’s co-author, Richard J. Acton, Jordi Rosell, Elio Campitelli, László
-Kupcsik, and Michael Heming for [rix]{.pkg}. Their expertise and feedback
-greatly improved the quality and usability of these packages. I would also like
-to thank Pol Dellaiera and Edvin Syk for providing feedback on the manuscript.
+[rixpress]{.pkg}. We also acknowledge the contributions of Richard J. Acton,
+Jordi Rosell, Elio Campitelli, László Kupcsik, and Michael Heming for
+[rix]{.pkg}. Their expertise and feedback greatly improved the quality and
+usability of these packages. We would also like to thank Pol Dellaiera and Edvin
+Syk for providing feedback on the manuscript.
 

### 2025-10-27 paper.qmd
typo

@@
 
+Despite these shortcomings, [Docker]{.pkg} remains valuable for specific use
+cases: production deployment, CI/CD integration, and cross-platform
+distribution. The two approaches are not mutually exclusive. [Nix]{.pkg} can be
+installed and used inside [Docker]{.pkg}, combining [Nix]{.pkg}'s deterministic
+builds with [Docker]{.pkg}'s deployment infrastructure. In this workflow,
+[Nix]{.pkg} handles environment reproducibility during the build of the image,
+while [Docker]{.pkg} serves as the distribution mechanism. This addresses
+[Docker]{.pkg}'s build-time reproducibility issues while maintaining its deployment
+advantages. However, for research workflows where interactive development is
+primary and [Docker]{.pkg}'s complexity is a barrier, [Nix]{.pkg} alone may suffice.
+
+Thankfully, it is entirely possible to combine both [Docker]{.pkg} and [Nix]{.pkg}:
+[Nix]{.pkg} to setup reproducible environments and [Docker]{.pkg} for deployment
+across the many different already existing infrastructures.
+
 Even with a perfectly reproducible environment, researchers face another
@@
 within hermetic (isolated) environments, with no hidden dependencies on global
-system state. While this functional approach greatly enhances reproducibility,
-it can sometimes introduce complexity for package maintainers, especially for
-software that needs to download external assets during installation (e.g.,
-certain Bioconductor packages). For end-users, however, this complexity is
-largely abstracted away. Additionally, [Nix]{.pkg} supports multiple versions or
-*variants* of a package on the same system.
-
-[Nix]{.pkg} installs all packages into a centralised, content-addressed
-directory called the [Nix]{.pkg} store (typically `/nix/store`). Each package
-gets a unique subdirectory whose name includes a hash of all its inputs: source
-code, dependencies, build instructions. This means
-`/nix/store/abc123-dplyr-1.0.0` and `/nix/store/xyz789-dplyr-1.1.0` can coexist
-without conflicts.
+system state. The principles of functional package management (hermetic builds,
+immutable artifacts, and declarative dependencies) are designed to produce
+deterministic outputs. The effectiveness of this model was recently validated in
+the most extensive study of build reproducibility to date. The effectiveness of
+this model was recently validated at unprecedented scale by @malka2025, who
+rebuilt 709,816 packages from historical nixpkgs snapshots (2017–2023). They
+found rebuildability rates exceeding 99% and bitwise reproducibility rates
+between 69% and 91%, with an upward trend over time. Importantly, they
+identified that about 15% of unreproducible builds failed due to embedded build
+timestamps, which a solvable issue rather than a fundamental limitation.
+
+While this functional approach greatly enhances reproducibility, it can
+sometimes introduce complexity for package maintainers, especially for software
+that needs to download external assets during installation (e.g., certain
+Bioconductor packages). For end-users, however, this complexity is largely
+abstracted away thanks to both [rix]{.pkg} and [rixpress]{.pkg}.
+
+Additionally, [Nix]{.pkg} supports multiple versions or *variants* of a package
+on the same system. [Nix]{.pkg} installs all packages into a centralised,
+content-addressed directory called the [Nix]{.pkg} store (typically
+`/nix/store`). Each package gets a unique subdirectory whose name includes a
+hash of all its inputs: source code, dependencies, build instructions. This
+means `/nix/store/abc123-dplyr-1.0.0` and `/nix/store/xyz789-dplyr-1.1.0` can
+coexist without conflicts.
 
@@
 
-To illustrate these capabilities with a realistic research task, we present a
-polyglot pipeline that simulates a Real Business Cycle (RBC) model in
-[Julia]{.proglang}, uses the resulting data to train an
-[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualises the results
-in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
-report. The code for this example can be found in the [following
-repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
+To illustrate these capabilities, we present a polyglot pipeline that simulates
+a Real Business Cycle (RBC) model in [Julia]{.proglang}, uses the resulting data
+to train an [XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualises
+the results in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a
+[Quarto]{.pkg} report. The code for this example can be found in the [following
+repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/case-study)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/case-study],
 with additional details provided in the Appendix of this manuscript.
@@
 [Julia]{.proglang}, yellow nodes are [Python]{.pkg}, light blue nodes are
-[R]{.proglang}, and dark blue nodes are Quarto derivations.](images/dag.png){#fig-dag}
+[R]{.proglang}, and dark blue nodes are Quarto
+derivations.](images/dag.png){#fig-dag}
 
@@
 packages.
+
+## A Comparison of Pipeline Paradigms: Imperative vs. Declarative {#sec-comparison}
+
+To illustrate the practical benefits of our framework, we implemented the RBC
+pipeline example from Section @sec-rixpress using two distinct paradigms. The
+first is a traditional, imperative stack combining Docker for environment
+isolation, Make for workflow orchestration, and language-specific package
+managers. The second is our proposed declarative framework using `rix` and
+`rixpress`. While both approaches successfully execute the same polyglot
+analysis to produce identical outputs, their comparison illuminates fundamental
+differences in conceptual models, system complexity, and the nature of the
+reproducibility guarantees they provide.
+
+### The Imperative Composition of Specialized Tools
+
+The traditional approach represents a composition of specialized tools, each
+solving part of the reproducibility puzzle. This paradigm is entirely
+imperative; the researcher must explicitly script the procedural steps for both
+environment construction and workflow execution.
+
+The codification of the environment is a procedural build process specified in a
+`Dockerfile`. Successfully authoring such a file necessitates that the
+researcher act as a system administrator, manually scripting a sequence of shell
+commands to install each component. This includes managing `apt` repositories
+for R, invoking external installers like `uv` and `rig` for Python and R
+runtimes, and using `curl` and `tar` for Julia. Subsequently, separate commands
+must be run to install packages within each language's ecosystem. The procedural
+verbosity and intertwined concerns of this approach are reflected in the
+resulting `Dockerfile`, which spans 108 lines to define the complete
+environment.
+
+Here is a sketch of this `Dockerfile`:
+
+```dockerfile
+# Add R repository and install specific version
+RUN apt-get update && apt-get install -y software-properties-common
+RUN add-apt-repository ppa:...
+RUN curl -L https://rig.r-pkg.org/... | sh
+RUN rig add 4.5.1
+
+# Install Python with uv
+RUN curl -LsSf https://astral.sh/uv/install.sh | sh
+RUN uv python install 3.13
+
+# Download and extract Julia
+RUN curl -fsSL https://julialang-s3.julialang.org/... -o julia.tar.gz
+RUN tar -xzf julia.tar.gz -C /opt/
+
+# Install packages for each language separately
+RUN echo 'options(repos = c(CRAN = ...))' > /root/.Rprofile
+RUN Rscript -e 'install.packages(...)'
+
+# Install Python packages using uv with specific versions for reproducibility.
+RUN echo "pandas==2.3.3" > /tmp/requirements.txt && \
+    echo "scikit-learn==1.7.2" >> /tmp/requirements.txt && \
+    # ... more packages ...
+
+RUN uv pip install --no-cache -r /tmp/requirements.txt && \
+    rm /tmp/requirements.txt
+
+# Install specific versions of Julia packages for reproducibility
+RUN julia -e 'using Pkg; \
+    Pkg.add(name="Arrow", version="2.8.0"); \
+    # ... more packages ...'
+```
+
+This procedural philosophy extends from environment definition to workflow
+orchestration. A `Makefile` defines the pipeline as a directed acyclic graph of
+file dependencies. While robust, this model requires the researcher to manually
+encode the dependency graph, tightly coupling the analytical logic to specific
+file paths. Furthermore, since `make` is agnostic to the content of these files,
+it provides no mechanism to enforce type or schema consistency between steps.
+
+A direct consequence of this file-based orchestration model is the proliferation
+of wrapper scripts. Each step in the `Makefile` invokes a script containing a
+significant amount of procedural boilerplate unrelated to the scientific logic.
+This includes parsing command-line arguments, handling data serialization and
+deserialization, and managing file paths. The core analytical functions are thus
+obscured by this necessary but distracting scaffolding.
+
+The cumulative effect is a system of high structural complexity. The traditional
+implementation requires nine separate files to fully specify the environment,
+workflow, and analysis. This includes the `Dockerfile`, `Makefile`, three
+wrapper scripts for orchestration, and three `functions` files containing the
+core logic, in addition to the final report. The cognitive overhead is
+substantial, as a user must comprehend the conventions and interactions of this
+entire toolchain.
+
+### A Unified Declarative Framework
+
+The Nix-based framework inverts this imperative model. Instead of scripting the
+procedural construction of the environment and workflow, the researcher declares
+the desired end-state.
+
+Environment specification is reduced to a single declarative statement in the
+`gen-env.R` script. The researcher specifies *what* the environment should
+contain, leaving the procedural implementation of *how* to build it entirely to
+the Nix toolchain. The `gen-env.R` script replaces the 108-line `Dockerfile`,
+abstracting away all system administration details. `rix` translates this
+high-level specification into a formal Nix expression that deterministically
+resolves all transitive dependencies for the complete polyglot environment.
+
+```r
+library(rix)
+rix(
+  date = "2025-10-14",
+  r_pkgs = c("ggplot2", "dplyr", "arrow"),
+  jl_conf = list(jl_version = "lts", ...),
+  py_conf = list(py_version = "3.13", ...),
+  ...
+)
+```
+
+Workflow orchestration undergoes a similar conceptual shift from procedural to
+declarative. The `gen-pipeline.R` script defines the pipeline not as a set of
+file rules, but as a graph of data objects. Dependencies are not manually
+specified but are automatically inferred by `rixpress` from the functional
+relationships in the code. This object-centric model renders the procedural
+wrapper scripts unnecessary. The scientific logic is expressed as a set of pure
+functions that accept data structures as inputs and return them as outputs, free
+of any boilerplate for file I/O or command-line parsing.
+
+```python
+# functions/functions.py
+def train_and_predict_output(simulated_df: pd.DataFrame) -> pd.DataFrame:
+    # ... scientific logic ...
+    return format_results(predictions)
+```
+
+This declarative approach significantly reduces system complexity. The entire
+implementation requires only six files: two for the high-level environment and
+pipeline declarations, three containing the pure scientific functions, and the
+final report. The reduction is not merely quantitative; the remaining files are
+conceptually simpler, containing only analytical logic and specifications, with
+all procedural scaffolding abstracted away by the framework.
+
+### Reproducibility Guarantees: Space and Time
+
+The critical distinction between the two paradigms lies in the nature of their
+reproducibility guarantees. The traditional stack, anchored by Docker, provides
+strong **spatial reproducibility**—the ability to execute the analysis
+identically across different machines. However, it is vulnerable to **temporal
+drift**, a concept that distinguishes reproducibility across space from
+consistency over time [@malka2025space]. This temporal vulnerability arises from
+several sources, including the resolution of mutable base image tags (e.g.,
+`ubuntu:24.04`) and the potential for language-level package managers to resolve
+dependencies differently over time.
+
+The Nix-based approach is architected to solve this problem of temporal drift.
+By pinning the entire software ecosystem to a single, immutable revision of the
+`nixpkgs` repository, the framework provides strong guarantees of **temporal
+reproducibility**. This claim is supported by large-scale empirical evidence; a
+study by @malka2024yes, which rebuilt over 700,000 historical packages,
+confirmed that Nix's functional model achieves a bit-for-bit reproducibility
+rate exceeding 90% and a rebuildability rate over 99% across a multi-year
+timeframe.
+
+### Summary of Differences
+
+| Dimension | Traditional Stack | Nix-Based Framework |
+|-----------|------------------|---------------------|
+| **Paradigm** | Imperative composition of specialized tools | Declarative specification in unified framework |
+| **Environment setup** | Step-by-step shell commands | High-level specification |
+| **Workflow definition** | File-based rules in Make syntax | Object-based graph in R syntax |
+| **Scientific code** | Wrapped in CLI parsing and file I/O boilerplate | Pure functions with no plumbing |
+| **File count** | 9 files (4 for orchestration and plumbing) | 6 files (2 for orchestration) |
+| **Expertise required** | Linux system administration, Make syntax | R programming |
+| **Spatial reproducibility** | Strong (via Docker isolation) | Strong (via Nix isolation) |
+| **Temporal reproducibility** | Moderate (vulnerable to drift) | Strong (validated empirically [@malka2025]) |
+
+This declarative rigor, however, is not without pragmatic costs. Our validation
+on GitHub Actions shows that the imperative Docker-based runner completes the
+entire pipeline in approximately three minutes, whereas the declarative
+Nix-based runner takes around five minutes. This difference reflects the
+trade-off between the upfront computational cost of Nix's rigorous dependency
+resolution and the faster, but more brittle, imperative builds of the
+traditional approach.
 
@@
 described at the end of this
-section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/polyglot-example/run_polyglot_example.sh}].
+section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/repro-script/run_polyglot_example.sh}].
 

### 2025-10-27 paper.qmd
rewrite, missing conclusion

@@
 
-The most comprehensive approach to date is containerisation with [Docker]{.pkg}.
-[Docker]{.pkg} packages a *data product* together with all its
-dependencies—including the operating system, system libraries, programming
-languages, and packages—into a self-contained image. Once built, the analysis
-can be executed inside a *container*, a running instance of the image.
-
-The strength of [Docker]{.pkg} lies in its ability to bundle not only the
-correct versions of [R]{.proglang} and its packages but also system-level
-dependencies. Since [Docker]{.pkg} images are minimal Linux systems, they
-include the libraries required by packages such as [sf]{.pkg}, ensuring that
-future replicators have access to the same environment. Moreover, these images
-can be shared via registries like Docker Hub, enabling reproducibility across
-systems.
-
-The Rocker project [@boettiger2017] provides pre-built [Docker]{.pkg} images for
-the [R]{.proglang} community, offering specific R versions and, in some cases,
-preinstalled packages. These serve as convenient bases for reproducible analysis
-environments.
-
-Despite its strengths, [Docker]{.pkg} presents several challenges:
-
-1. Poor interactive support: While containers can be modified at runtime,
-   changes are lost unless explicitly saved. Running graphical applications like
-   RStudio Desktop requires complex X11 configuration that works reliably only
-   on Linux. Web-based IDEs such as RStudio Server are alternatives but have
-   their own limitations.
-
-2. Steep learning curve: Writing reliable `Dockerfile`s requires familiarity
-   with Linux system administration. To ensure reproducibility, `Dockerfile`s
-   should reference image *digests* rather than *tags*, but this is rarely done.
-   Packages like [dockerfiler]{.pkg} [@dockerfiler] help automate this but do
-   not remove the need for basic Linux knowledge.
-
-3. Post-hoc reproducibility: Analyses are often developed interactively and
-   containerised only afterward, achieving post-hoc reproducibility rather than
-   consistent environments during development.
-
-4. Single-language focus: While [Docker]{.pkg} can support polyglot setups,
-   orchestrating them requires manual coordination. There is no standard way to
-   specify that one step runs in [Python]{.proglang} and another in [R]{.proglang}.
-
-5. Incomplete reproducibility: Reproducibility is best viewed as a spectrum
-   rather than a binary property [@dellaieraMasterThesis2024]. [Docker]{.pkg}
-   achieves strong run-time reproducibility—consistency across systems
-   (space)—but weak build-time reproducibility, or consistency over time.
-   Mutable base images and non-deterministic commands such as `apt-get update`
-   introduce temporal drift: rebuilding the same `Dockerfile` at different times
-   can yield different results. Even with pinned base images, container
-   checksums can vary across builds and machines. Thus, while [Docker]{.pkg}
-   provides strong run-time reproducibility, achieving deterministic artefacts
-   requires pinning image digests, freezing dependencies, and using
-   deterministic build systems such as [Nix]{.pkg} for full control over both
-   time and space [@Malka_2024].
-
-Despite these shortcomings, [Docker]{.pkg} remains valuable for specific use
-cases: production deployment, CI/CD integration, and cross-platform
-distribution. The two approaches are not mutually exclusive. [Nix]{.pkg} can be
-installed and used inside [Docker]{.pkg}, combining [Nix]{.pkg}'s deterministic
-builds with [Docker]{.pkg}'s deployment infrastructure. In this workflow,
-[Nix]{.pkg} handles environment reproducibility during the build of the image,
-while [Docker]{.pkg} serves as the distribution mechanism. This addresses
-[Docker]{.pkg}'s build-time reproducibility issues while maintaining its deployment
-advantages. However, for research workflows where interactive development is
-primary and [Docker]{.pkg}'s complexity is a barrier, [Nix]{.pkg} alone may suffice.
-
-Thankfully, it is entirely possible to combine both [Docker]{.pkg} and [Nix]{.pkg}:
-[Nix]{.pkg} to setup reproducible environments and [Docker]{.pkg} for deployment
-across the many different already existing infrastructures.
-
-Even with a perfectly reproducible environment, researchers face another
-challenge: reliably executing the analysis workflow. Which scripts run first?
-What are the dependencies between steps? Manual execution is error-prone and
-poorly documented. Build automation tools like [Make]{.pkg} address this by
+The most comprehensive approach to managing a complete computational environment
+to date is containerization, for which [Docker]{.pkg} has become the de facto
+standard. Docker's principal strength lies in its capacity to package an entire
+data product—including the operating system, all system-level libraries,
+programming language runtimes, and packages—into a single, self-contained
+artifact known as an image. Once this image is built, the analysis can be
+executed within a container, which is a running instance of that image. This
+methodology effectively solves the problem of missing system-level dependencies;
+because Docker images are typically minimal Linux systems, they can bundle the
+specific versions of libraries like [GDAL]{.pkg} or [GEOS]{.pkg} required by
+packages such as [sf]{.pkg}. These images can then be shared via public
+registries, enabling a high degree of reproducibility across different host
+systems. The Rocker project, for instance, provides invaluable pre-built Docker
+images tailored for the [R]{.proglang} community, offering a convenient
+foundation for reproducible research environments [@boettiger2017].
+
+However, while Docker excels at providing this form of spatial reproducibility,
+its application in interactive research workflows reveals several significant
+limitations. First, the container model is often poorly aligned with the
+iterative, interactive nature of data analysis. Running graphical applications
+like RStudio Desktop requires complex and platform-specific configuration, while
+ephemeral container filesystems mean that changes made during a session are lost
+by default. Consequently, a common workflow involves developing an analysis
+interactively on a host system and only containerizing it post-hoc, which
+achieves reproducibility as an afterthought rather than as a continuous
+guarantee during the research process. Packages like @dockerfiler are specifically
+made to enable this type of post-hoc reproducibility.
+
+Second, the effective use of [Docker]{.pkg} introduces a considerable learning
+curve. Authoring a reliable and truly reproducible `Dockerfile` requires a
+degree of familiarity with Linux system administration. Best practices, such as
+referencing base images by their immutable content digests rather than mutable
+tags, are necessary for long-term stability but are not widely adopted.
+Furthermore, while [Docker]{.pkg} can house polyglot environments, it offers no native
+mechanism for orchestrating a pipeline where distinct steps might require
+different, and potentially conflicting, sets of dependencies. The entire
+analysis is typically confined to a single, monolithic environment.
+
+The most critical limitation, however, lies in its guarantees of temporal
+reproducibility. As noted by @dellaieraMasterThesis2024, reproducibility is best
+viewed as a spectrum. Docker provides strong run-time reproducibility
+(consistency across space), but offers weaker build-time reproducibility
+(consistency over time). The imperative commands within a `Dockerfile`, such as
+`apt-get update`, are non-deterministic by nature and can introduce temporal
+drift, where rebuilding the same file at different times yields different binary
+artifacts. As empirically demonstrated by @malka2025, achieving deterministic
+builds requires a system designed for this purpose, a role for which functional
+package managers like [Nix]{.pkg} are architecturally better suited.^[This
+contrast between the procedural, imperative commands of a `Dockerfile` and the
+declarative, functional model of [Nix]{.pkg} highlights a fundamental dichotomy
+in approaches to reproducibility, a concept we will examine in detail as we
+introduce the [Nix]{.pkg} package manager in Section @sec-nix.]
+
+This is not to say that [Docker]{.pkg} has no place in a reproducible research
+workflow. Its utility for production deployment, continuous integration, and
+cross-platform distribution is well-established. Indeed, the two approaches are
+not mutually exclusive. As shall be explained in @sec-nix, it is possible
+to use [Nix]{.pkg} within [Docker]{.pkg} to benefit from the advantages of
+both tools.
+
+Making sure the development environment is reproducible is but one piece of the
+puzzle, though. Even with a perfectly reproducible environment, researchers face
+another challenge: reliably executing the analysis workflow. Which scripts run
+first? What are the dependencies between steps? Manual execution is error-prone
+and poorly documented. Build automation tools like [Make]{.pkg} address this by
 defining analyses as ordered, reproducible steps.
@@
 
-@mcdermott2021 provides an excellent example of best practice. The accompanying
-repository^[See
+The current state-of-the-art in reproducible research exemplifies the complexity
+of navigating this fragmented landscape. The work of @mcdermott2021 provides an
+excellent case study in best practices, with an accompanying repository that is
+fully reproducible.^[See
 [https://github.com/grantmcdermott/skeptic-priors](https://github.com/grantmcdermott/skeptic-priors)]
-demonstrates:
-
-- Package versions recorded in an `renv.lock` file;
-- A `Makefile` automating the full analysis;
-- A `Dockerfile` providing the computational environment.
-
-Achieving this required mastering multiple complex tools: [renv]{.pkg} for
-package management, [Docker]{.pkg} for containerisation, and [Make]{.pkg} for
-orchestration. The complexity increases for polyglot projects, which must manage
+Achieving this gold standard, however, required the
+masterful integration of multiple complex and disparate tools: [renv]{.pkg} for
+managing [R]{.proglang} packages, [Docker]{.pkg} for containerizing the system
+environment, and [Make]{.pkg} for orchestrating the analysis. This demonstrates
+that while a high degree of reproducibility is attainable, it requires a heroic
+effort, forcing researchers to become experts in a brittle and complex
+toolchain.
+
+Furthermore, the complexity increases for polyglot projects, which must manage
 multiple ecosystems simultaneously.
@@
 
-## The Nix package manager {#sec-nix}
-
-[Nix]{.pkg} is a powerful, cross-platform package manager designed for
-installing and building software in a fully reproducible and reliable manner.
-Unlike traditional package managers, [Nix]{.pkg} emphasises immutable
-infrastructure and a functional approach, which significantly enhances the
-consistency of computational environments. As of writing, the primary [Nix]{.pkg}
-package collection, `nixpkgs`, provides access to over 120,000 packages,
-including nearly all of CRAN and Bioconductor. According to Repology, a service
-that tracks package repositories across distributions, `nixpkgs` is both the
-largest and among the most up-to-date package repositories available (see Figure
+## The Nix Package Manager {#sec-nix}
+
+[Nix]{.pkg} is a cross-platform package manager designed for reproducible
+software installation and deployment. Unlike traditional package managers,
+[Nix]{.pkg} emphasizes immutable infrastructure and a functional programming
+paradigm to ensure computational environment consistency. The primary
+[Nix]{.pkg} package collection, `nixpkgs`, provides access to over 120,000
+packages, including nearly all of CRAN and Bioconductor. According to Repology,
+a service tracking package repositories across distributions, `nixpkgs` is both
+the largest and among the most up-to-date repositories available (Figure
 @fig-repology).
@@
 
-This extensive coverage allows users to install not only [R]{.proglang} itself
-but also all its required packages and system-level dependencies for any given
+This extensive coverage allows users to install not only [R]{.proglang}
+but also all required packages and system-level dependencies for any given
 project.
 
-While [Nix]{.pkg} is the default package manager for the NixOS Linux
-distribution, it can be installed as a standalone tool on other Linux
-distributions, macOS, and effectively on Windows via the Windows Subsystem for
-Linux (WSL2), treating these as equivalent platforms in practice.
-
-A key advantage of using [Nix]{.pkg} to install [R]{.proglang} packages, as
-opposed to the standard `install.packages()` function, lies in its comprehensive
-dependency management. [Nix]{.pkg} ensures that all dependencies of each package
-are installed, irrespective of whether they are other [R]{.proglang} packages or
-underlying system libraries.
-
-Consider, for example, the [sf]{.pkg} package, which is fundamental for spatial
-data analysis in [R]{.proglang}. [sf]{.pkg} relies on several complex external
-system libraries such as [GDAL]{.pkg}, [GEOS]{.pkg}, and
-[PROJ]{.pkg}. Manually installing and configuring these libraries to be
-compatible with [sf]{.pkg} can be challenging and platform-specific. With
-[Nix]{.pkg}, however, the user merely needs to declare [sf]{.pkg} as a project
-requirement. [Nix]{.pkg} then automatically installs and configures all
-necessary system libraries as transitive dependencies. This seamless process is
-possible because the maintainers of the [R]{.proglang} programming language for
-[Nix]{.pkg} have declaratively specified these dependencies for [sf]{.pkg} in
-`nixpkgs`, allowing the entire software stack to be provisioned effortlessly
-from an end-user perspective.
-
-This concept is central to [Nix]{.pkg}'s design, where packages are referred to
-as *component closures*. As @dolstra2004nix explains:
+A key advantage of [Nix]{.pkg} over traditional installation methods lies in its
+comprehensive dependency management. Consider the [sf]{.pkg} package for spatial
+data analysis in [R]{.proglang}. It requires several complex system libraries:
+[GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg}. Manually installing and
+configuring these libraries is challenging and platform-specific. With
+[Nix]{.pkg}, users simply declare [sf]{.pkg} as a project requirement, and
+[Nix]{.pkg} automatically installs and configures all necessary system libraries
+as transitive dependencies.
+
+This seamless process stems from [Nix]{.pkg}'s concept of *component closures*.
+As @dolstra2004nix explains:
 
 > The idea is to always deploy component closures: if we deploy a component,
-> then we must also deploy its dependencies, their dependencies, and so on. That
-> is, we must always deploy a set of components that is closed under the
-> ''depends on'' relation. Since closures are selfcontained, they are the units
-> of complete software deployment. After all, if a set of components is not
-> closed, it is not safe to deploy, since using them might cause other
-> components to be referenced that are missing on the target system.
-
-The core of [Nix]{.pkg}'s reproducibility lies in its functional package
-management paradigm and its unique approach to builds. When installing software
-with [Nix]{.pkg}, it evaluates an *expression* written in the [Nix]{.proglang}
-language. These expressions define *derivations*, which are declarative
-blueprints describing how to build a package. A derivation specifies all inputs:
-source code, build commands, and crucially, all direct and transitive
-dependencies. This declarative nature ensures that every build process is fully
-specified and independent of the environment where it's executed. Furthermore,
-all [Nix]{.pkg} expressions for official packages are hosted in the `nixpkgs`
-GitHub repository. By referencing a specific commit of `nixpkgs` (a process
-known as *pinning a revision*) users guarantee that all packages installed are
-derived from the exact same set of build instructions. This means that if an
-environment is built today using a pinned revision, it will produce the
-identical set of software versions and configurations if rebuilt tomorrow, next
-year, or on a different machine, as long as the `nixpkgs` repository remains
-accessible.
-
-Pinning is essential, but it is not the only mechanism through which [Nix]{.pkg}
-achieves reproducibility. [Nix]{.pkg} adopts principles from functional
-programming, conceptualising software builds as *pure functions*. In
-mathematics, a pure function is defined as one that always produces the same
-output for a given input—for instance, if \( f(x) = x^2 \), then \( f(2) = 4 \)
-invariably holds. Analogously, [Nix]{.pkg} ensures that, given identical inputs
-(such as source code, dependencies, and build instructions), the resulting
-output (the built package) remains identical, irrespective of the environment or
-time of execution. In this sense, [Nix]{.pkg} functions as a *functional package
-manager*, applying the deterministic properties of pure functions to software
-construction.
-
-This is achieved by disallowing side effects and ensuring that builds operate
-within hermetic (isolated) environments, with no hidden dependencies on global
-system state. The principles of functional package management (hermetic builds,
-immutable artifacts, and declarative dependencies) are designed to produce
-deterministic outputs. The effectiveness of this model was recently validated in
-the most extensive study of build reproducibility to date. The effectiveness of
-this model was recently validated at unprecedented scale by @malka2025, who
-rebuilt 709,816 packages from historical nixpkgs snapshots (2017–2023). They
-found rebuildability rates exceeding 99% and bitwise reproducibility rates
-between 69% and 91%, with an upward trend over time. Importantly, they
-identified that about 15% of unreproducible builds failed due to embedded build
-timestamps, which a solvable issue rather than a fundamental limitation.
-
-While this functional approach greatly enhances reproducibility, it can
-sometimes introduce complexity for package maintainers, especially for software
-that needs to download external assets during installation (e.g., certain
-Bioconductor packages). For end-users, however, this complexity is largely
-abstracted away thanks to both [rix]{.pkg} and [rixpress]{.pkg}.
-
-Additionally, [Nix]{.pkg} supports multiple versions or *variants* of a package
-on the same system. [Nix]{.pkg} installs all packages into a centralised,
-content-addressed directory called the [Nix]{.pkg} store (typically
-`/nix/store`). Each package gets a unique subdirectory whose name includes a
-hash of all its inputs: source code, dependencies, build instructions. This
-means `/nix/store/abc123-dplyr-1.0.0` and `/nix/store/xyz789-dplyr-1.1.0` can
-coexist without conflicts.
-
-This isolation ensures that projects always use their intended software
-versions, allowing multiple versions of [R]{.proglang} to coexist and preventing
-"dependency hell" and global environment pollution.
+> then we must also deploy its dependencies, their dependencies, and so on.
+> [...] Since closures are self-contained, they are the units of complete
+> software deployment.
+
+In other words, every package deployment includes everything it needs to run,
+with no hidden dependencies on the host system.
+
+While comprehensive dependency tracking could theoretically be implemented in
+any package manager, what distinguishes [Nix]{.pkg} (alongside its cousin
+[Guix]{.pkg}) is its *functional package management* paradigm. Traditional
+package managers require imperative, step-by-step installation instructions: *do
+this, then this, then that*. [Nix]{.pkg} inverts this model; users declare the
+desired end state (*I require an environment with these packages*), and
+[Nix]{.pkg} determines how to achieve it.
+
+This is possible because [Nix]{.pkg} packages are defined as *derivations*:
+declarative blueprints specifying all inputs (source code, build commands,
+dependencies) needed to build a package. When installing software, [Nix]{.pkg}
+evaluates expressions written in the [Nix]{.proglang} language to produce these
+derivations. The declarative nature ensures every build process is fully
+specified and independent of the execution environment.
+
+For example, the [rJava]{.pkg} package definition includes the correct
+`JAVA_HOME` configuration pointing to the [jdk]{.pkg} dependency. Outside
+[Nix]{.pkg}, users must manually set this variable if automatic detection fails.
+With [Nix]{.pkg}, the package maintainers handle this complexity once (by
+writing the correct [Nix]{.pkg} expressions that perform the installation
+purely), and all users benefit.
+
+[Nix]{.pkg} conceptualizes software builds as *pure functions* from functional
+programming. In mathematics, a pure function always produces the same output for
+a given input: if $f(x) = x^2$, then $f(2) = 4$ invariably holds. Analogously,
+[Nix]{.pkg} ensures that identical inputs (source code, dependencies, build
+instructions) produce identical outputs (built packages), regardless of when or
+where the build occurs.
+
+This determinism is achieved through hermetic (isolated) builds with no hidden
+dependencies on global system state nor undeclared side effects. The
+effectiveness of this model was validated at unprecedented scale by @malka2025,
+who rebuilt 709,816 packages from historical `nixpkgs` snapshots (2017–2023),
+finding rebuildability rates exceeding 99% and bitwise reproducibility rates
+between 69% and 91%, with an upward trend. Importantly, about 15% of
+unreproducible builds failed due to embedded timestamps—a solvable
+implementation issue rather than a fundamental model limitation.
+
+All [Nix]{.pkg} package definitions are hosted in the `nixpkgs` GitHub
+repository. By *pinning* to a specific commit (or equivalently, a specific
+date), users guarantee that all installed packages derive from identical build
+instructions. An environment built today using a pinned revision will produce
+the same software versions and configurations if rebuilt years later, provided
+the `nixpkgs` repository remains accessible (it is archived and distributed
+across multiple mirrors).
+
+[Nix]{.pkg} installs all packages into a content-addressed directory called the
+Nix store (typically `/nix/store`). Each package occupies a unique subdirectory
+whose name includes a cryptographic hash of all its inputs: source code,
+dependencies, and build instructions. This means `/nix/store/abc123-dplyr-1.0.0`
+and `/nix/store/xyz789-dplyr-1.1.0` can coexist without conflicts, allowing
+multiple [R]{.proglang} versions on the same system and eliminating "dependency
+hell."
+
+While this functional approach enhances reproducibility, it can introduce
+complexity for package maintainers, particularly for software downloading
+external assets during installation (e.g., certain Bioconductor packages). For
+end-users, however, this complexity is largely abstracted away by [rix]{.pkg}
+and [rixpress]{.pkg}, which we introduce in the following sections.
 
@@
 
-By leveraging these principles, [Nix]{.pkg} can effectively replace and
-integrate the functionalities of tools like [renv]{.pkg} and [Docker]{.pkg} for
-[R]{.proglang} projects, or `requirements.txt` files and virtual environments
-for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
-polyglot environments, which can seamlessly combine [R]{.proglang},
-[Python]{.proglang}, [Julia]{.proglang}, a [LaTeX]{.proglang} distribution, or
-any of the numerous other tools available in `nixpkgs`. This enables the
-creation of a truly complete, project-specific, and deeply reproducible
-environment that can be used interactively for development or non-interactively
-for automated analysis. As long as the `nixpkgs` repository (or a suitable fork,
-as discussed in Section @sec-fork) remains accessible, the environment can be
-rebuilt reliably in the future.
-
-[Nix]{.pkg} provides strong cross-platform support across major operating
-systems. It runs natively on Linux (x86_64) with full functionality and optimal
-integration. On macOS (Intel and Apple Silicon), however, reproducibility is
-less reliable due to dependencies on impure system frameworks and Xcode
-toolchains outside the Nix store. As a result, macOS users may need to update
-project pins after system or Xcode upgrades to restore builds. On Windows,
-[Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though
-graphical applications may need extra setup. [Nix]{.pkg} environments on WSL can
-be used interactively via VS Code or Positron for a smooth development
-experience. It is also important to note that [Nix]{.pkg} can be installed
-within a [Docker]{.pkg} image to deploy and reproduce environments reliably in
-containerized workflows.
-
-While [Nix]{.pkg} is a powerful system, it comes with notable challenges. The
-steepest is its learning curve: [Nix]{.pkg} uses its own declarative language
-(also called [Nix]{.proglang}), which can be difficult for those new to
-functional or declarative paradigms. Complexity becomes most apparent when
-writing custom derivations or troubleshooting intricate build issues. Build
-times can also be long, especially for large environments, when pre-built
-binaries aren’t available. Additionally, because [Nix]{.pkg} stores all package
-outputs immutably in the Nix store, disk usage tends to be higher than with
-traditional package managers. Finally, while `nixpkgs` is extensive, some new or
-niche packages may be missing or fail to build across platforms.
+By leveraging these principles, [Nix]{.pkg} provides a unified framework that
+can subsume the functionalities of disparate tools. For [R]{.proglang} projects,
+it can replace the combination of [renv]{.pkg} for package management and
+[Docker]{.pkg} for system-level isolation. Similarly, for [Python]{.proglang}
+projects, it offers a more comprehensive alternative to `requirements.txt` files
+and virtual environments. Furthermore, [Nix]{.pkg} excels at composing polyglot
+environments, seamlessly integrating [R]{.proglang}, [Python]{.proglang},
+[Julia]{.proglang}, a [LaTeX]{.proglang} distribution, and any of the numerous
+other tools available in `nixpkgs`. This enables the creation of a truly
+complete, project-specific, and deeply reproducible environment suitable for
+both interactive development and non-interactive, automated analysis. The
+long-term viability of this reproducibility is contingent only on the continued
+accessibility of the `nixpkgs` repository or a suitable fork, as will be
+discussed in Section @sec-fork.
+
+[Nix]{.pkg} offers robust cross-platform support, running natively on Linux
+(x86_64) with optimal functionality. On macOS (Intel and Apple Silicon),
+however, its reproducibility guarantees are somewhat compromised by dependencies
+on proprietary system frameworks and Xcode toolchains that reside outside
+of the [Nix]{.pkg} store; consequently, macOS users may need to update project
+pins following system upgrades. For Windows users, [Nix]{.pkg} operates
+effectively through the Windows Subsystem for Linux (WSL2), where environments
+can be integrated into development workflows using editors like VS Code or
+Positron.
+
+While [Nix]{.pkg} is a powerful system, its adoption is not without notable
+challenges. The most significant is its steep learning curve; it uses its own
+programming language, also called [Nix]{.proglang}, which is purely functional, a
+paradigm that may be unfamiliar to many researchers. This complexity is most
+apparent when authoring custom build instructions or troubleshooting intricate
+dependency issues. Additionally, build times can be substantial for large
+environments when pre-built binary artifacts are not available from a cache. The
+immutable, content-addressed nature of the [Nix]{.pkg} store also leads to
+higher disk usage compared to traditional package managers. Finally, while the
+`nixpkgs` repository is extensive, certain niche packages may be unavailable or
+fail to build on all platforms.
+
+Beyond these technical aspects, a practical hurdle to adoption is the ubiquity
+of [Docker]{.pkg}. Given the extensive infrastructure and community knowledge
+built around containerization, [Nix]{.pkg} can be perceived as a competing
+rather than a complementary technology. This perception, however, is a false
+dichotomy. [Docker]{.pkg} and [Nix]{.pkg} allow researchers to solve similar
+issues but these tools are not alternatives of one another. Thus, a powerful
+hybrid model involves using [Nix]{.pkg} to deterministically build the
+environment *during the construction of a [Docker]{.pkg} image*, which is then
+deployed using standard container infrastructure. This approach leverages the
+strengths of both tools: [Nix]{.pkg} provides robust build-time reproducibility,
+while [Docker]{.pkg} serves as a universal distribution and deployment
+mechanism. For many research workflows, however, where interactive development
+is paramount and the overhead of containerization is a barrier, the unified
+model provided by [Nix]{.pkg} alone can provide a more direct and streamlined
+path to reproducibility.
 
@@
 Once the build completes, the user can enter the interactive shell with
-`nix-shell`, which executes following steps:
-
-- Nix reads the expression and identifies what needs to be installed;
+`nix-shell`, which executes the following steps:
+
+- [Nix]{.pkg} reads the expression and identifies what needs to be installed;
 - It checks if these packages already exist in `/nix/store`;
@@
 development environment in a single text file and then rebuild it anywhere is
-highly appealing. [rix]{.pkg} aims to lower the barrier to adoption of [Nix]{.pkg} for
-reproducibility.
-
-[rix]{.pkg} provides the `rix()` function, which simplifies generating
-[Nix]{.pkg} expressions. It is available on CRAN and can be installed like any
-other R package. Additionally, it can bootstrap an [R]{.proglang} development
-environment on a system where [R]{.proglang} is not yet installed but
-[Nix]{.pkg} is available. This can be done by running (inside of a terminal):
+highly appealing. [rix]{.pkg} aims to lower the barrier to adoption of
+[Nix]{.pkg} for reproducibility.
+
+[rix]{.pkg}, an [R]{.proglang} package, provides the `rix()` function, which
+simplifies generating [Nix]{.pkg} expressions. It is available on CRAN and can
+be installed like any other [R]{.proglang} package. Additionally, it can
+bootstrap an [R]{.proglang} development environment on a system where
+[R]{.proglang} is not yet installed but [Nix]{.pkg} is available. This can be
+done by running (inside of a terminal):
 
@@
 by pinning all components to a single point in time: the [R]{.proglang} packages
-are resolved from the CRAN snapshot of October 20, 2025, while all other system
+are resolved from the CRAN snapshot of October 20th 2025, while all other system
 tools are fixed to the version of `nixpkgs` from that same date. Typing
 `nix-shell` in a terminal within the folder that contains this `default.nix`
-will drop the user into a new shell. It should be noted that it uses the
-`nixpkgs` from that same date. It is also possible to configure IDEs to
-dynamically load this new shell to provide the proper development tooling.
+will drop the user into a new shell. It is also possible to configure IDEs to
+dynamically load this new shell to provide the proper development tooling for
+better interactive use.
 
@@
 this heterogeneity, [Nix]{.pkg} maintainers cannot automatically package all of
-PyPI. Instead, they manually curate and package individual [Python]{.proglang}
-packages that meet quality standards. This means some [Python]{.proglang}
-packages, or specific versions, may not be available through [Nix]{.pkg}.
-
-For projects requiring [Python]{.proglang} packages not yet available in
+PyPI as they do for CRAN or the [Julia]{.proglang} registry. Instead, they
+manually curate and package individual [Python]{.proglang} packages that meet
+quality standards. This means some [Python]{.proglang} packages, or specific
+versions, may not be available through [Nix]{.pkg}.
+
+For projects requiring [Python]{.proglang} packages not available in
 [Nix]{.pkg}, [rix]{.pkg} supports integration with [uv]{.pkg}, a modern
@@
 reproducible through [Nix]{.pkg}, while [Python]{.proglang} packages gain the
-reproducibility that [uv]{.pkg} can provide.
-
-[rix]{.pkg} can generate [Nix]{.pkg} expressions even if [Nix]{.pkg} is not
-installed on the system. This is useful for continuous integration and
-continuous deployment (CI/CD) workflows on platforms such as GitHub Actions. For
+reproducibility that [uv]{.pkg} can provide. We recommend of course to use
+[Python]{.pkg} through [Nix]{.pkg} whenever possible and only resort to the
+hybrid approach with [uv]{.pkg} as a last resort.^[It is also possible to package
+the missing [Python]{.pkg} for [Nix]{.pkg} and submit a pull request to `nixpkgs`,
+but we recognise that this option may not be beginner-friendly.]
+
+Anywhere [Nix]{.pkg} can be installed, users can benefit from its features. For
 instance, the repository containing the source code for this
@@
 file, and compiles the paper using [Quarto]{.pkg} within the reproducible
-environment. This ensure that *exactly* the same environment is used on the
-author's computer and on the CI/CD without any additional, platform-specific,
-configuration.
+environment. This ensures that *exactly* the same environment is used on the
+authors' machines and on the CI/CD platform without any additional,
+platform-specific, configuration.
 
@@
 Using a fork offers several advantages. First, it provides flexibility that the
-official `nixpkgs` repository cannot always accommodate. [Nix]{.pkg} is
-primarily the package manager for the NixOS Linux distribution, and governance
-and technical choices made upstream can limit what [rix]{.pkg} aims to provide.
+official `nixpkgs` repository cannot always accommodate. Because [Nix]{.pkg}
+serves as the package manager for the NixOS Linux distribution, upstream
+priorities center on system-level stability rather than rapid updates to
+language ecosystems like [R]{.proglang} or [Julia]{.proglang}.
 
@@
 
-The `rstats-on-nix` fork allows me to circumvent these limitations. For example,
-it provides daily snapshots of CRAN. Each day, the [R]{.proglang} package set is
-updated and committed to a dated branch using GitHub Actions. Users can select a
-specific date with:
+The `rstats-on-nix` fork allows us to circumvent these limitations. For example,
+it provides daily snapshots of CRAN and the [Julia]{.proglang}. Each day, the
+[R]{.proglang} package set is updated and committed to a dated branch using
+GitHub Actions. Users can select a specific date with:
 
@@
 We strive to provide an available date per week: each Monday, a GitHub Action
-tests popular packages on Linux and macOS, and only if all tests succeed is the
-date added to the list of available dates in [rix]{.pkg}. This ensures users can
-reliably install packages, and allows me to backport fixes if needed. For
+tests popular [R]{.proglang} packages on Linux and macOS, and only if all tests
+succeed is the date added to the list of available dates in [rix]{.pkg}. Users
+can see all available dates with `rix::available_dates()`. This ensures users
+can reliably install packages, and allows us to backport fixes if needed. For
 example, when RStudio was temporarily broken due to a dependency issue
@@
 x86-linux (including WSL2) and aarch64-darwin (Apple Silicon). These changes are
-committed to a dated branch in `rstats-on-nix/nixpkgs`. Users can see all
-available dates with `rix::available_dates()`.
+committed to a dated branch in `rstats-on-nix/nixpkgs`.
 
@@
 
-The fork also allows me to catch issues (such as packages’ builds breaking) early
+The fork also allows us to catch issues (such as packages’ builds breaking) early
 on, and prepare fixes that can then be contributed upstream.
@@
 Defining a reproducible environment with [rix]{.pkg} addresses the first major
-challenge of reproducibility. The second challenge is reliably and efficiently
-executing the analysis workflow within that environment, which is the role of
-the [rixpress]{.pkg} package (or [ryxpress]{.pkg} for [Python]{.proglang}
+challenge of reproducibility: environment capture. The second challenge, reliably
+and efficiently executing the analysis workflow within that environment, is
+addressed by [rixpress]{.pkg} (or [ryxpress]{.pkg} for [Python]{.proglang}
 users).
@@
      rendering step in yet another, all within the same pipeline.
-  2. Deep Reproducibility: Each step is a hermetically sealed [Nix]{.pkg} derivation
-     whose output is cached in the [Nix]{.pkg} store based on the hash of all its
-     inputs. All artefacts are direct children of the computational environment
-     (because the computational environment is actually a dependency of the
-     artefacts), ensuring they are rebuilt if the environment changes, keeping
-     environment and outputs always in sync.
-
-To illustrate these capabilities, we present a polyglot pipeline that simulates
-a Real Business Cycle (RBC) model in [Julia]{.proglang}, uses the resulting data
-to train an [XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualises
-the results in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a
-[Quarto]{.pkg} report. The code for this example can be found in the [following
+  2. Deep Reproducibility: Each step is a hermetically sealed [Nix]{.pkg}
+     derivation whose output is cached in the [Nix]{.pkg} store based on the
+     hash of all its inputs. All artefacts depend directly on their
+     computational environment, ensuring that any change in dependencies
+     triggers a rebuild.
+
+As an example, we demonstrate a polyglot pipeline that simulates a Real Business
+Cycle (RBC) model in [Julia]{.proglang}, trains an [XGBoost]{.pkg} forecasting
+model in [Python]{.proglang} on the simulated data, visualises results in
+[R]{.proglang}, and compiles a [Quarto]{.pkg} report. The code for this example
+can be found in the [following
 repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/case-study)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/case-study],
-with additional details provided in the Appendix of this manuscript.
+with additional details provided in the Appendix of this manuscript. Although
+this example is purely synthetic and not of scientific significance, it
+effectively demonstrates the capabilities of [rixpress]{.pkg}.
 
@@
 computational environment the pipeline will be executed in the `gen-env.R`
-script. If functions are needed for pipeline steps, these are to be specified, in
-language-specific helper scripts: `functions.jl` for [Julia]{.proglang},
+script. If functions are needed for pipeline steps, these are to be specified,
+in language-specific helper scripts: `functions.jl` for [Julia]{.proglang},
 `functions.py` for [Python]{.proglang}, and `functions.R` for [R]{.proglang}.
@@
 [rixpress]{.pkg}-specific code and can be reused in other projects or executed
-independently outside of [Nix]{.pkg}.
+independently outside of [Nix]{.pkg}. This design allows users to preserve
+standard development practices while still benefiting from declarative
+orchestration.
 
@@
 
+In summary, [rixpress]{.pkg} extends the guarantees of [rix]{.pkg}
+from static environments to dynamic workflows, enabling end-to-end, polyglot
+reproducibility.
+
 ## A Comparison of Pipeline Paradigms: Imperative vs. Declarative {#sec-comparison}
@@
 pipeline example from Section @sec-rixpress using two distinct paradigms. The
-first is a traditional, imperative stack combining Docker for environment
+first is a traditional, imperative, stack combining Docker for environment
 isolation, Make for workflow orchestration, and language-specific package
@@
 The traditional approach represents a composition of specialized tools, each
-solving part of the reproducibility puzzle. This paradigm is entirely
+solving part of the reproducibility puzzle. This paradigm is purely
 imperative; the researcher must explicitly script the procedural steps for both
@@
 must be run to install packages within each language's ecosystem. The procedural
-verbosity and intertwined concerns of this approach are reflected in the
-resulting `Dockerfile`, which spans 108 lines to define the complete
-environment.
+verbosity and intertwined concerns of this approach are reflected in a
+`Dockerfile` spanning 108 lines to define the complete environment.
 
@@
 
-```dockerfile
+::: {.content-hidden when-format="pdf"}
+```r
 # Add R repository and install specific version
@@
     # ... more packages ...'
+```
+:::
+
+```{=tex}
+\begin{CodeInput}
+# Add R repository and install specific version
+RUN apt-get update && apt-get install -y software-properties-common
+RUN add-apt-repository ppa:...
+RUN curl -L https://rig.r-pkg.org/... | sh
+RUN rig add 4.5.1
+
+# Install Python with uv
+RUN curl -LsSf https://astral.sh/uv/install.sh | sh
+RUN uv python install 3.13
+
+# Download and extract Julia
+RUN curl -fsSL https://julialang-s3.julialang.org/... -o julia.tar.gz
+RUN tar -xzf julia.tar.gz -C /opt/
+
+# Install packages for each language separately
+RUN echo 'options(repos = c(CRAN = ...))' > /root/.Rprofile
+RUN Rscript -e 'install.packages(...)'
+
+# Install Python packages using uv with specific versions for reproducibility.
+RUN echo "pandas==2.3.3" > /tmp/requirements.txt && \
+    echo "scikit-learn==1.7.2" >> /tmp/requirements.txt && \
+    # ... more packages ...
+
+RUN uv pip install --no-cache -r /tmp/requirements.txt && \
+    rm /tmp/requirements.txt
+
+# Install specific versions of Julia packages for reproducibility
+RUN julia -e 'using Pkg; \
+    Pkg.add(name="Arrow", version="2.8.0"); \
+\end{CodeInput}
 ```
@@
 This includes parsing command-line arguments, handling data serialization and
-deserialization, and managing file paths. The core analytical functions are thus
-obscured by this necessary but distracting scaffolding.
+deserialization, and managing file paths. The core analytical logic is thus
+obscured by necessary but extraneous scaffolding.
 
@@
 
+::: {.content-hidden when-format="pdf"}
 ```r
@@
 ```
+:::
+
+```{=tex}
+\begin{CodeInput}
+R> library(rix)
+R> rix(
++   date = "2025-10-14",
++   r_pkgs = c("ggplot2", "dplyr", "arrow"),
++   jl_conf = list(jl_version = "lts", ...),
++   py_conf = list(py_version = "3.13", ...),
++   ...
++ )
+\end{CodeInput}
+```
 
@@
 
+::: {.content-hidden when-format="pdf"}
 ```python
@@
   return format_results(predictions)
+```
+:::
+
+```{=tex}
+# functions/functions.py
+Py> def train_and_predict_output(simulated_df: pd.DataFrame) -> pd.DataFrame:
++     # ... scientific logic ...
++     return format_results(predictions)
+end{CodeInput}
 ```
@@
 drift**, a concept that distinguishes reproducibility across space from
-consistency over time [@malka2025space]. This temporal vulnerability arises from
+consistency over time [@malka2024]. This temporal vulnerability arises from
 several sources, including the resolution of mutable base image tags (e.g.,
@@
 reproducibility**. This claim is supported by large-scale empirical evidence; a
-study by @malka2024yes, which rebuilt over 700,000 historical packages,
+study by @malka2025, which rebuilt over 700,000 historical packages,
 confirmed that Nix's functional model achieves a bit-for-bit reproducibility
@@
 | **Expertise required** | Linux system administration, Make syntax | R programming |
-| **Spatial reproducibility** | Strong (via Docker isolation) | Strong (via Nix isolation) |
-| **Temporal reproducibility** | Moderate (vulnerable to drift) | Strong (validated empirically [@malka2025]) |
+| **Spatial reproducibility** | Strong | Strong |
+| **Temporal reproducibility** | Moderate (vulnerable to drift) | Strong |
 
@@
 entire pipeline in approximately three minutes, whereas the declarative
-Nix-based runner takes around five minutes. This difference reflects the
-trade-off between the upfront computational cost of Nix's rigorous dependency
-resolution and the faster, but more brittle, imperative builds of the
-traditional approach.
+Nix-based runner takes around five minutes. Nix’s deterministic dependency
+resolution guarantees bit-for-bit reproducibility at the expense of initial
+computation time, whereas the imperative system prioritizes immediacy over
+long-term stability.
 
@@
 This appendix provides a conceptual walk-through of the polyglot pipeline
-example discussed in Section 5. The pipeline simulates a Real Business Cycle
-(RBC) model in [Julia]{.proglang}, trains an [XGBoost]{.pkg} model in
-[Python]{.proglang}, visualises the results in [R]{.proglang}, and compiles a
-final report with [Quarto]{.pkg}.
+example discussed in Section 5. The pipeline simulates a RBC model in
+[Julia]{.proglang}, trains an [XGBoost]{.pkg} model in [Python]{.proglang},
+visualises the results in [R]{.proglang}, and compiles a final report with
+[Quarto]{.pkg}.
 

### 2025-10-27 paper.qmd
typo in python code block

@@
 ```{=tex}
+\begin{CodeInput}
 # functions/functions.py
@@
 +     return format_results(predictions)
-end{CodeInput}
+\end{CodeInput}
 ```

### 2025-10-28 paper.qmd
Refine language on Nix and reproducibility tools
Updated language for clarity and precision in the discussion of Nix and its tools, emphasizing reproducibility and user accessibility.

@@
 [Nix]{.pkg} package collection, `nixpkgs`, provides access to over 120,000
-packages, including nearly all of CRAN and Bioconductor. According to Repology,
+packages, including nearly all of CRAN and Bioconductor.^[Only about 80 packages 
+currently fail to build on the NixOS build farm, plus roughly 30 packages marked as broken
+by the maintainers of the [R]{.proglang} ecosystem. These could, in principle, be fixed, 
+but doing so would require significant effort. Marking them as broken signals to users 
+that their build failures are known and intentional.] According to Repology,
 a service tracking package repositories across distributions, `nixpkgs` is both
@@
 Many tools exist to improve reproducibility, but [Nix]{.pkg} stands out because 
-it deploys complete software environments *closed under the "depends on"
-relation*: it installs not only a package, but all its dependencies and their
-dependencies. This makes [Nix]{.pkg} particularly powerful for reproducible
-research.
-
-However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
-[rix]{.pkg}, we aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
-by providing a familiar interface and workflow. By building reproducible
-development shells with [Nix]{.pkg}, researchers can accommodate a wide range of
-use cases: running scripts and pipelines, developing interactive [shiny]{.pkg}
-applications, or serving [plumber]{.pkg} APIs.
-
-Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
-pipelines. By leveraging [Nix]{.pkg} as a build automation engine, [rixpress]{.pkg}
-allows polyglot workflows where each step runs in its own hermetically sealed
-environment. This ensures deep reproducibility, efficient caching, and seamless
-orchestration of polyglot analyses. Most importantly, it provides native
-polyglot support without the manual coordination required by containerization
-approaches, making complex multi-language workflows accessible to researchers
-without systems administration expertise.
-
-Adopting [Nix]{.pkg}-based workflows represents a significant paradigm shift, and
-researchers should approach it incrementally. A reasonable pathway begins with
-using [rix]{.pkg} to generate reproducible environments for new projects, while
-maintaining existing workflows for ongoing work. This allows learning [Nix]{.pkg}
-concepts without risking active research. As comfort grows, researchers can
-experiment with [rixpress]{.pkg} for smaller analyses before tackling complex
-polyglot pipelines. Institutions can support adoption by providing workshops,
-maintaining institutional binary caches with private packages, and designating
-local experts. The investment pays increasing returns: early projects require
-substantial learning, but subsequent projects leverage accumulated knowledge and
-reusable environment definitions. For research groups, having one member develop
-[Nix]{.pkg} expertise can enable the entire team to benefit from reproducible
-environments without everyone climbing the learning curve. Ultimately, the
-decision to adopt these tools depends on project timescales, reproducibility
-requirements, and institutional context—but the framework exists for those who
-need it.
-
-While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide
-significant steps forward, it is important to acknowledge current limitations.
-Debugging complex [rixpress]{.pkg} pipelines remains more difficult than
-debugging [targets]{.pkg} workflows, particularly when failures occur deep in
-the [Nix]{.pkg} build process. The visualization capabilities, while useful, lack
-features like automatic detection of outdated derivations that make tools like
-[targets]{.pkg} particularly developer-friendly. Cross-platform reproducibility,
-while generally strong, has known limitations on macOS due to system framework
-dependencies. These are not insurmountable problems—they represent areas for
-continued development and community contribution. Researchers adopting these
-tools should do so with realistic expectations: they provide powerful
-capabilities for deep reproducibility and polyglot workflows, but require
-patience during the learning phase and may occasionally present challenges that
-simpler tools avoid.
+it deploys complete software environments closed under the “depends on” relation:
+it installs not only a package but all of its transitive dependencies. 
+This makes [Nix]{.pkg} uniquely suited for reproducible research.
+
+Yet solving such a complex problem makes [Nix]{.pkg} itself complex, as it relies on a 
+functional paradigm still likely unfamiliar to most researchers.
+[rix]{.pkg} lowers this barrier for [R]{.proglang} users by providing a familiar interface and workflow. 
+By declaratively specifying reproducible development shells with [Nix]{.pkg}, 
+researchers can accommodate diverse use cases—from running analytical pipelines 
+to developing interactive [shiny]{.pkg} applications or serving [plumber]{.pkg} APIs.
+
+Furthermore, [rixpress]{.pkg} extends this declarative model to entire analysis pipelines. 
+By leveraging [Nix]{.pkg} as a unified build automation system, [rixpress]{.pkg} enables polyglot 
+workflows where each step runs in a hermetically sealed environment. 
+This ensures both spatial and temporal reproducibility, efficient caching, and seamless 
+orchestration of multi-language analyses. Crucially, it provides native polyglot support 
+without the manual orchestration required by container-based approaches, 
+making complex workflows accessible without systems administration expertise.
+
+Adopting [Nix]{.pkg}-based workflows represents a conceptual shift from procedural 
+scripting to declarative specification. A practical adoption pathway begins with using 
+[rix]{.pkg} to generate reproducible environments for new projects while maintaining existing workflows. 
+This allows gradual learning of [Nix]{.pkg} concepts without disrupting active research. 
+As confidence grows, researchers can extend to [rixpress]{.pkg} for smaller analyses before tackling complex 
+pipelines. Institutions can support adoption by offering workshops, maintaining binary caches, 
+and designating local experts. While early projects require investment, later projects benefit from
+reusable environment definitions and cumulative expertise. Ultimately, adoption depends on 
+project scope, reproducibility needs, and institutional context—but the framework now exists 
+for those who require robust, long-term reproducibility.
+
+While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) represent significant progress,
+current limitations remain. Debugging complex [rixpress]{.pkg} pipelines is more challenging than
+troubleshooting [targets]{.pkg} workflows, especially when failures occur deep in the [Nix]{.pkg} build process. 
+Visualization capabilities, though useful, currently lack features such as automatic detection 
+of outdated derivations. Cross-platform reproducibility remains generally strong but is constrained on macOS by 
+proprietary framework dependencies. These challenges are active areas for development and community contribution. 
+Researchers adopting these tools should do so with realistic expectations: they provide strong guarantees of spatial 
+and temporal reproducibility and scalable polyglot workflows, but they require patience during the learning phase 
+and may occasionally present challenges that simpler tools avoid.
 

### 2025-10-28 paper.qmd
Update paper.qmd

@@
 who have many tools at their disposal, such as the built-in *virtual
-environments* module, or third-party tools like [uv]{.pkg}, which manages both
+environments* module, or third-party tools like [uv]{.pkg} [@uv], which manages both
 project-specific interpreters and package libraries using a lock file mechanism.
@@
 but doing so would require significant effort. Marking them as broken signals to users 
-that their build failures are known and intentional.] According to Repology,
+that their build failures are known and intentional. It should be further noted that some
+packages may build successfully, but then fail at runtime as they try to further install
+runtime dependencies on first run. One such example is [torch]{.pkg}.] According to Repology,
 a service tracking package repositories across distributions, `nixpkgs` is both
@@
 to build many packages from source, which can be time-consuming. To mitigate
-this, we provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
+this, we provide a binary cache sponsored by 
+[Cachix](https://www.cachix.org/)^[[https://www.cachix.org/](https://www.cachix.org/)],
 complementing the public Nix cache. Instructions for using Cachix are in
@@
 isolation, Make for workflow orchestration, and language-specific package
-managers. The second is our proposed declarative framework using `rix` and
+managers.^[This choice of tools is illustrative: we could have used, for instance,
+[Snakemake]{.pkg} [@moelder2021snakemake] for workflow orchestration or [conda]{.pkg} [@anaconda] 
+instead of [uv]{.pkg}, 
+but the broader point remains—these are all imperative approaches] 
+The second is our proposed declarative framework using `rix` and
 `rixpress`. While both approaches successfully execute the same polyglot
@@
 entire pipeline in approximately three minutes, whereas the declarative
-Nix-based runner takes around five minutes. Nix’s deterministic dependency
+Nix-based runner takes around five minutes.^[To run the benchmark yourself,
+fork the paper’s repository, enable GitHub Actions, and push a single change to
+trigger the runs.] Nix’s deterministic dependency
 resolution guarantees bit-for-bit reproducibility at the expense of initial

### 2025-10-28 paper.qmd
Update paper.qmd

@@
 manually curate and package individual [Python]{.proglang} packages that meet
-quality standards. This means some [Python]{.proglang} packages, or specific
+quality standards.^[On October 2025, this amounts to roughly 10000 packages, 
+according to the search engine for [Nix]{.pkg} packages.] 
+This means some [Python]{.proglang} packages, or specific
 versions, may not be available through [Nix]{.pkg}.
@@
 
-| Dimension | Traditional Stack | Nix-Based Framework |
+| Dimension | Traditional Stack | Nix-Based Framework with rix and rixpress |
 |-----------|------------------|---------------------|

### 2025-10-28 paper.qmd
Revise content on reproducibility and polyglot workflows
Enhanced discussion on polyglot environments and reproducibility challenges in data science. Updated references to tools and trends in managing dependencies across multiple programming languages.

@@
 A more robust approach than simply listing package versions is to use tools that
-define per-project environments. This is common for [Python]{.proglang} users,
+define per-project environments. This is common for [Python]{.proglang} [@python3] users,
 who have many tools at their disposal, such as the built-in *virtual
@@
 
-Pre-compiled binaries simplify installation but do not eliminate the need for
-system libraries required at runtime. Other [R]{.proglang} packages such as
+Other [R]{.proglang} packages such as
 [groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023] enable
@@
 address system-level dependencies or R version management. 
+Furthermore, using pre-compiled binaries of packages simplifies installation but do not eliminate the need for
+system libraries required at runtime.
 
@@
 ecosystems, but they share a common limitation: none handles system-level
-dependencies or polyglot environments. A project using [R]{.proglang},
-[Python]{.proglang} [@python3], and system tools like [Quarto]{.pkg} or \LaTeX
-requires coordinating multiple package managers ([renv]{.pkg} for R, [uv]{.pkg}
-for [Python]{.proglang}, manual installation for system tools), each with its own
-configuration and potential conflicts.
+dependencies or the increasingly polyglot nature of modern data science. 
+The era of a single tool dominating an entire analytical pipeline is giving 
+way to a more pragmatic approach where researchers combine the complementary 
+strengths of multiple languages [@peng2018]. This trend is particularly evident
+in the synergistic use of [R]{.proglang}, with its rich ecosystem for statistical modeling, 
+[Python]{.proglang}, with its extensive libraries for machine learning, and [Julia]{.proglang} 
+for complex simulations through the use of interoperability libraries such as [PythonCall]{.pkg} 
+(calling [Python]{.proglang} from [Julia]{.proglang}), [JuliaCall]{.pkg} 
+(calling [Julia]{.proglang} from [Python]{.proglang} or [R]{.proglang}), [reticulate]{.pkg} 
+(calling [Python]{.proglang} from [R]{.proglang}), [rpy2]{.pkg} and [ryp]{.pkg} (calling [R]{.proglang} from [Python]{.proglang}) [@osborne2024, @navarro2024].
+
+Recent survey data confirms this trend, showing that research scientists use, on average, 
+nearly two programming languages in their work, with the [R]{.proglang} and [Python]{.proglang} combination being 
+the most prevalent [@chen2025]. This creates a significant challenge for reproducibility:
+a project using [R]{.proglang}, [Python]{.proglang}, and system tools like
+[Quarto]{.pkg} or \LaTeX requires coordinating multiple, disparate package managers, 
+each with its own configuration and potential for conflict and using interoperability
+libraries.
+
+The demand for solutions that can manage this complexity is not merely academic; 
+it is reflected in major strategic shifts within the industry. A prominent example is the 
+evolution of Posit (formerly RStudio), which has pivoted from its origins as a premier R-centric 
+development environment to a multi-language scientific data platform that treats Python as a 
+first-class citizen. This industry trend, coupled with the proliferation of academic research
+into polyglot tooling, underscores a critical reality: the central challenge for modern reproducible
+science is no longer *if* polyglot workflows should be used, but *how* they can be managed in a unified, 
+efficient, and reproducible manner.
 

### 2025-10-28 paper.qmd
Update paper.qmd

@@
 language-agnostic Arrow format using the pair of `encoder/decoder` functions.
-Other universal formats, such as `csv` or `json` could have been used.
+Other universal formats, such as `csv` or `json` could have been used. It should
+also be noted that if interoperability libraries such as [reticulate]{.pkg} or 
+[rds2py]{.pkg} are available, it becomes possible to seamlessly transfer
+supported objects to and from [R]{.proglang} and [Python]{.proglang} (for now,
+such a feature is not yet integrated for conversion of arbitrary [Julia]{.proglang} objects).
 

### 2025-10-28 paper.qmd
Update paper.qmd

@@
 Furthermore, using pre-compiled binaries of packages simplifies installation but do not eliminate the need for
-system libraries required at runtime.
-
-These tools represent significant progress in reproducibility within their
+system libraries required at runtime. These tools represent significant progress in reproducibility within their
 ecosystems, but they share a common limitation: none handles system-level
 dependencies or the increasingly polyglot nature of modern data science.
-The era of a single tool dominating an entire analytical pipeline is giving 
+
+Indeed, there is evidence that the era of a single tool dominating an entire analytical pipeline is giving 
 way to a more pragmatic approach where researchers combine the complementary 
@@
 (calling [Python]{.proglang} from [R]{.proglang}), [rpy2]{.pkg} and [ryp]{.pkg} (calling [R]{.proglang} from [Python]{.proglang}) [@osborne2024, @navarro2024].
-
 Recent survey data confirms this trend, showing that research scientists use, on average, 
@@
 evolution of Posit (formerly RStudio), which has pivoted from its origins as a premier R-centric 
-development environment to a multi-language scientific data platform that treats Python as a 
-first-class citizen. This industry trend, coupled with the proliferation of academic research
+service and software supplier to a company that treats [Python]{.proglang} as a 
+first-class citizen alongside [R]{.proglang}. This industry trend, coupled with the proliferation of academic research
 into polyglot tooling, underscores a critical reality: the central challenge for modern reproducible

### 2025-10-28 paper.qmd
Update paper.qmd

@@
 (calling [Julia]{.proglang} from [Python]{.proglang} or [R]{.proglang}), [reticulate]{.pkg} 
-(calling [Python]{.proglang} from [R]{.proglang}), [rpy2]{.pkg} and [ryp]{.pkg} (calling [R]{.proglang} from [Python]{.proglang}) [@osborne2024, @navarro2024].
+(calling [Python]{.proglang} from [R]{.proglang}), [rpy2]{.pkg} and [ryp]{.pkg} (calling [R]{.proglang} from [Python]{.proglang}) [@osborne2024; @navarro2024].
 Recent survey data confirms this trend, showing that research scientists use, on average, 

### 2025-10-28 paper.qmd
Update paper.qmd

@@
 The complete, runnable source code for this example is available in the paper's
-GitHub repository. A shell script, `run_polyglot_example.sh`, is provided to
+GitHub repository. A shell script, `repro_script.sh`, is provided to
 automate the entire process from environment creation to final output, as
 described at the end of this
-section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/repro-script/run_polyglot_example.sh}].
+section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/repro-script/repro_script.sh}].
 
@@
 
-This entire example can be executed by running the `run_polyglot_example.sh`
+This entire example can be executed by running the `repro_script.sh`
 script available in the root of the paper's repository.

### 2025-10-28 paper.qmd
Update paper.qmd

@@
 # functions/functions.py
-Py> def train_and_predict_output(simulated_df: pd.DataFrame) -> pd.DataFrame:
+>>> def train_and_predict_output(simulated_df: pd.DataFrame) -> pd.DataFrame:
 +     # ... scientific logic ...
@@
 \begin{CodeInput}
-Python> from ryxpress import rxp_make
-Python> rxp_make()
+>>> from ryxpress import rxp_make
+>>> rxp_make()
 \end{CodeInput}

### 2025-10-28 paper.qmd
Update paper.qmd

@@
 achieves reproducibility as an afterthought rather than as a continuous
-guarantee during the research process. Packages like @dockerfiler are specifically
-made to enable this type of post-hoc reproducibility.
+guarantee during the research process. Packages like [dockerfiler]{.pkg} 
+[@dockerfiler] are specifically made to enable this type of post-hoc reproducibility.
 
@@
 example, when RStudio was temporarily broken due to a dependency issue
-(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
+([boost]{.pkg}), a pull request was submitted to the official `nixpkgs` repository. I
 backported the fix to the `rstats-on-nix` fork, making RStudio available to
@@
 from derivation to derivation by being serialised into the efficient and
-language-agnostic Arrow format using the pair of `encoder/decoder` functions.
+language-agnostic [Arrow]{.pkg} format using the pair of `encoder/decoder` functions.
 Other universal formats, such as `csv` or `json` could have been used. It should

### 2025-10-28 paper.qmd
Update paper.qmd
formatting

@@
 By *linked and executable code and data*, Peng referred to compiled source code
-and runnable scripts. In this paper, we interpret this notion more broadly as the
-*computational environment*: the complete set of software required to execute an
-analysis. Here too, a continuum exists. At the minimal end, authors might only
-name the main software used—say, the [R]{.proglang} programming language
-[@r-core-team]. More careful authors might also specify the version of
+and runnable scripts. In this paper, we interpret this notion more broadly as
+the *computational environment*: the complete set of software required to
+execute an analysis. Here too, a continuum exists. At the minimal end, authors
+might only name the main software used—say, the [R]{.proglang} programming
+language [@r-core-team]. More careful authors might also specify the version of
 [R]{.proglang}, or list the additional packages and their versions. Rarely,
@@
 researchers through two [R]{.proglang} packages: [rix]{.pkg} and
-[rixpress]{.pkg}. Before introducing these packages, we survey existing tools and
-their limitations to contextualise our contributions.
+[rixpress]{.pkg}. Before introducing these packages, we survey existing tools
+and their limitations to contextualise our contributions.
 
@@
 were used. However, reproducing the environment still requires manually
-installing the correct package versions --- a process that becomes difficult when
-packages depend on system-level libraries. For example, the [sf]{.pkg} package
-requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} libraries. Installing and
-configuring these dependencies varies substantially across operating systems and
-can be prohibitively difficult on some platforms.
+installing the correct package versions --- a process that becomes difficult
+when packages depend on system-level libraries. For example, the [sf]{.pkg}
+package requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} libraries.
+Installing and configuring these dependencies varies substantially across
+operating systems and can be prohibitively difficult on some platforms.
 
 A more robust approach than simply listing package versions is to use tools that
-define per-project environments. This is common for [Python]{.proglang} [@python3] users,
-who have many tools at their disposal, such as the built-in *virtual
-environments* module, or third-party tools like [uv]{.pkg} [@uv], which manages both
-project-specific interpreters and package libraries using a lock file mechanism.
+define per-project environments. This is common for [Python]{.proglang}
+[@python3] users, who have many tools at their disposal, such as the built-in
+*virtual environments* module, or third-party tools like [uv]{.pkg} [@uv], which
+manages both project-specific interpreters and package libraries using a lock
+file mechanism.
 
@@
 
-Other [R]{.proglang} packages such as
-[groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023] enable
-date-based package installation, while the Posit Package Manager provides dated
-CRAN snapshots. However, like [renv]{.pkg}, these focus on R packages and do not
-address system-level dependencies or R version management. 
-Furthermore, using pre-compiled binaries of packages simplifies installation but do not eliminate the need for
-system libraries required at runtime. These tools represent significant progress in reproducibility within their
-ecosystems, but they share a common limitation: none handles system-level
+Other [R]{.proglang} packages such as [groundhog]{.pkg} [@simonsohn2023] and
+[rang]{.pkg} [@chan2023] enable date-based package installation, while the Posit
+Package Manager provides dated CRAN snapshots. However, like [renv]{.pkg}, these
+focus on R packages and do not address system-level dependencies or R version
+management. Furthermore, using pre-compiled binaries of packages simplifies
+installation but do not eliminate the need for system libraries required at
+runtime. These tools represent significant progress in reproducibility within
+their ecosystems, but they share a common limitation: none handles system-level
 dependencies or the increasingly polyglot nature of modern data science.
 
-Indeed, there is evidence that the era of a single tool dominating an entire analytical pipeline is giving 
-way to a more pragmatic approach where researchers combine the complementary 
-strengths of multiple languages [@peng2018]. This trend is particularly evident
-in the synergistic use of [R]{.proglang}, with its rich ecosystem for statistical modeling, 
-[Python]{.proglang}, with its extensive libraries for machine learning, and [Julia]{.proglang} 
-for complex simulations through the use of interoperability libraries such as [PythonCall]{.pkg} 
+Indeed, there is evidence that the era of a single tool dominating an entire
+analytical pipeline is giving way to a more pragmatic approach where researchers
+combine the complementary strengths of multiple languages [@peng2018]. This
+trend is particularly evident in the synergistic use of [R]{.proglang}, with its
+rich ecosystem for statistical modeling, [Python]{.proglang}, with its extensive
+libraries for machine learning, and [Julia]{.proglang} for complex simulations
+through the use of interoperability libraries such as [PythonCall]{.pkg}
 (calling [Python]{.proglang} from [Julia]{.proglang}), [JuliaCall]{.pkg}
-(calling [Julia]{.proglang} from [Python]{.proglang} or [R]{.proglang}), [reticulate]{.pkg} 
-(calling [Python]{.proglang} from [R]{.proglang}), [rpy2]{.pkg} and [ryp]{.pkg} (calling [R]{.proglang} from [Python]{.proglang}) [@osborne2024; @navarro2024].
-Recent survey data confirms this trend, showing that research scientists use, on average, 
-nearly two programming languages in their work, with the [R]{.proglang} and [Python]{.proglang} combination being 
-the most prevalent [@chen2025]. This creates a significant challenge for reproducibility:
-a project using [R]{.proglang}, [Python]{.proglang}, and system tools like
-[Quarto]{.pkg} or \LaTeX requires coordinating multiple, disparate package managers, 
-each with its own configuration and potential for conflict and using interoperability
-libraries.
+(calling [Julia]{.proglang} from [Python]{.proglang} or [R]{.proglang}),
+[reticulate]{.pkg} (calling [Python]{.proglang} from [R]{.proglang}),
+[rpy2]{.pkg} and [ryp]{.pkg} (calling [R]{.proglang} from [Python]{.proglang})
+[@osborne2024; @navarro2024]. Recent survey data confirms this trend, showing
+that research scientists use, on average, nearly two programming languages in
+their work, with the [R]{.proglang} and [Python]{.proglang} combination being
+the most prevalent [@chen2025]. This creates a significant challenge for
+reproducibility: a project using [R]{.proglang}, [Python]{.proglang}, and system
+tools like [Quarto]{.pkg} or \LaTeX requires coordinating multiple, disparate
+package managers, each with its own configuration and potential for conflict and
+using interoperability libraries.
 
 The demand for solutions that can manage this complexity is not merely academic;
-it is reflected in major strategic shifts within the industry. A prominent example is the 
-evolution of Posit (formerly RStudio), which has pivoted from its origins as a premier R-centric 
-service and software supplier to a company that treats [Python]{.proglang} as a 
-first-class citizen alongside [R]{.proglang}. This industry trend, coupled with the proliferation of academic research
-into polyglot tooling, underscores a critical reality: the central challenge for modern reproducible
-science is no longer *if* polyglot workflows should be used, but *how* they can be managed in a unified, 
-efficient, and reproducible manner.
+it is reflected in major strategic shifts within the industry. A prominent
+example is the evolution of Posit (formerly RStudio), which has pivoted from its
+origins as a premier R-centric service and software supplier to a company that
+treats [Python]{.proglang} as a first-class citizen alongside [R]{.proglang}.
+This industry trend, coupled with the proliferation of academic research into
+polyglot tooling, underscores a critical reality: the central challenge for
+modern reproducible science is no longer *if* polyglot workflows should be used,
+but *how* they can be managed in a unified, efficient, and reproducible manner.
 
@@
 guarantee during the research process. Packages like [dockerfiler]{.pkg}
-[@dockerfiler] are specifically made to enable this type of post-hoc reproducibility.
+[@dockerfiler] are specifically made to enable this type of post-hoc
+reproducibility.
 
@@
 tags, are necessary for long-term stability but are not widely adopted.
-Furthermore, while [Docker]{.pkg} can house polyglot environments, it offers no native
-mechanism for orchestrating a pipeline where distinct steps might require
+Furthermore, while [Docker]{.pkg} can house polyglot environments, it offers no
+native mechanism for orchestrating a pipeline where distinct steps might require
 different, and potentially conflicting, sets of dependencies. The entire
@@
 cross-platform distribution is well-established. Indeed, the two approaches are
-not mutually exclusive. As shall be explained in @sec-nix, it is possible
-to use [Nix]{.pkg} within [Docker]{.pkg} to benefit from the advantages of
-both tools.
+not mutually exclusive. As shall be explained in @sec-nix, it is possible to use
+[Nix]{.pkg} within [Docker]{.pkg} to benefit from the advantages of both tools.
 
@@
 [https://github.com/grantmcdermott/skeptic-priors](https://github.com/grantmcdermott/skeptic-priors)]
-Achieving this gold standard, however, required the
-masterful integration of multiple complex and disparate tools: [renv]{.pkg} for
-managing [R]{.proglang} packages, [Docker]{.pkg} for containerizing the system
-environment, and [Make]{.pkg} for orchestrating the analysis. This demonstrates
-that while a high degree of reproducibility is attainable, it requires a heroic
-effort, forcing researchers to become experts in a brittle and complex
-toolchain.
+Achieving this gold standard, however, required the masterful integration of
+multiple complex and disparate tools: [renv]{.pkg} for managing [R]{.proglang}
+packages, [Docker]{.pkg} for containerizing the system environment, and
+[Make]{.pkg} for orchestrating the analysis. This demonstrates that while a high
+degree of reproducibility is attainable, it requires a heroic effort, forcing
+researchers to become experts in a brittle and complex toolchain.
 
@@
 packages, including nearly all of CRAN and Bioconductor.^[Only about 80 packages
-currently fail to build on the NixOS build farm, plus roughly 30 packages marked as broken
-by the maintainers of the [R]{.proglang} ecosystem. These could, in principle, be fixed, 
-but doing so would require significant effort. Marking them as broken signals to users 
-that their build failures are known and intentional. It should be further noted that some
-packages may build successfully, but then fail at runtime as they try to further install
-runtime dependencies on first run. One such example is [torch]{.pkg}.] According to Repology,
-a service tracking package repositories across distributions, `nixpkgs` is both
-the largest and among the most up-to-date repositories available (Figure
+currently fail to build on the NixOS build farm, plus roughly 30 packages marked
+as broken by the maintainers of the [R]{.proglang} ecosystem. These could, in
+principle, be fixed, but doing so would require significant effort. Marking them
+as broken signals to users that their build failures are known and intentional.
+It should be further noted that some packages may build successfully, but then
+fail at runtime as they try to further install runtime dependencies on first
+run. One such example is [torch]{.pkg}.] According to Repology, a service
+tracking package repositories across distributions, `nixpkgs` is both the
+largest and among the most up-to-date repositories available (Figure
 @fig-repology).
@@
 
-This extensive coverage allows users to install not only [R]{.proglang}
-but also all required packages and system-level dependencies for any given
-project.
+This extensive coverage allows users to install not only [R]{.proglang} but also
+all required packages and system-level dependencies for any given project.
 
@@
 
-This shell contains all the packages specified in `default.nix` and
-can be used for development, similar to activating a virtual environment in the
+This shell contains all the packages specified in `default.nix` and can be used
+for development, similar to activating a virtual environment in the
 [Python]{.proglang} ecosystem.
@@
 quality standards.^[On October 2025, this amounts to roughly 10000 packages,
-according to the search engine for [Nix]{.pkg} packages.] 
-This means some [Python]{.proglang} packages, or specific
-versions, may not be available through [Nix]{.pkg}.
+according to the search engine for [Nix]{.pkg} packages.] This means some
+[Python]{.proglang} packages, or specific versions, may not be available through
+[Nix]{.pkg}.
 
@@
 [Python]{.pkg} through [Nix]{.pkg} whenever possible and only resort to the
-hybrid approach with [uv]{.pkg} as a last resort.^[It is also possible to package
-the missing [Python]{.pkg} for [Nix]{.pkg} and submit a pull request to `nixpkgs`,
-but we recognise that this option may not be beginner-friendly.]
+hybrid approach with [uv]{.pkg} as a last resort.^[It is also possible to
+package the missing [Python]{.pkg} for [Nix]{.pkg} and submit a pull request to
+`nixpkgs`, but we recognise that this option may not be beginner-friendly.]
 
@@
 example, when RStudio was temporarily broken due to a dependency issue
-([boost]{.pkg}), a pull request was submitted to the official `nixpkgs` repository. I
-backported the fix to the `rstats-on-nix` fork, making RStudio available to
-users of [rix]{.pkg} earlier than upstream, as merging PRs in the official
-repository can take some time.
+([boost]{.pkg}), a pull request was submitted to the official `nixpkgs`
+repository. I backported the fix to the `rstats-on-nix` fork, making RStudio
+available to users of [rix]{.pkg} earlier than upstream, as merging PRs in the
+official repository can take some time.
 
@@
 this example is purely synthetic and not of scientific significance, it
-effectively demonstrates the capabilities of [rixpress]{.pkg}.
+effectively demonstrates the capabilities of [rixpress]{.pkg}. It should also
+be noted that the code [Julia]{.proglang}, [Python]{.proglang} and [R]{.proglang}
+was generated by a large language model.
 
@@
 from derivation to derivation by being serialised into the efficient and
-language-agnostic [Arrow]{.pkg} format using the pair of `encoder/decoder` functions.
-Other universal formats, such as `csv` or `json` could have been used. It should
-also be noted that if interoperability libraries such as [reticulate]{.pkg} or 
-[rds2py]{.pkg} are available, it becomes possible to seamlessly transfer
-supported objects to and from [R]{.proglang} and [Python]{.proglang} (for now,
-such a feature is not yet integrated for conversion of arbitrary [Julia]{.proglang} objects).
+language-agnostic [Arrow]{.pkg} format using the pair of `encoder/decoder`
+functions. Other universal formats, such as `csv` or `json` could have been
+used. It should also be noted that if interoperability libraries such as
+[reticulate]{.pkg} or [rds2py]{.pkg} are available, it becomes possible to
+seamlessly transfer supported objects to and from [R]{.proglang} and
+[Python]{.proglang} (for now, such a feature is not yet integrated for
+conversion of arbitrary [Julia]{.proglang} objects).
 
@@
 isolation, Make for workflow orchestration, and language-specific package
-managers.^[This choice of tools is illustrative: we could have used, for instance,
-[Snakemake]{.pkg} [@moelder2021snakemake] for workflow orchestration or [conda]{.pkg} [@anaconda] 
-instead of [uv]{.pkg}, 
-but the broader point remains—these are all imperative approaches] 
-The second is our proposed declarative framework using `rix` and
-`rixpress`. While both approaches successfully execute the same polyglot
-analysis to produce identical outputs, their comparison illuminates fundamental
-differences in conceptual models, system complexity, and the nature of the
-reproducibility guarantees they provide.
+managers.^[This choice of tools is illustrative: we could have used, for
+instance, [Snakemake]{.pkg} [@moelder2021snakemake] for workflow orchestration
+or [conda]{.pkg} [@anaconda] instead of [uv]{.pkg}, but the broader point
+remains—these are all imperative approaches] The second is our proposed
+declarative framework using `rix` and `rixpress`. While both approaches
+successfully execute the same polyglot analysis to produce identical outputs,
+their comparison illuminates fundamental differences in conceptual models,
+system complexity, and the nature of the reproducibility guarantees they
+provide.
 
@@
 The traditional approach represents a composition of specialized tools, each
-solving part of the reproducibility puzzle. This paradigm is purely
-imperative; the researcher must explicitly script the procedural steps for both
-environment construction and workflow execution.
+solving part of the reproducibility puzzle. This paradigm is purely imperative;
+the researcher must explicitly script the procedural steps for both environment
+construction and workflow execution.
 
@@
 reproducibility**. This claim is supported by large-scale empirical evidence; a
-study by @malka2025, which rebuilt over 700,000 historical packages,
-confirmed that Nix's functional model achieves a bit-for-bit reproducibility
-rate exceeding 90% and a rebuildability rate over 99% across a multi-year
-timeframe.
+study by @malka2025, which rebuilt over 700,000 historical packages, confirmed
+that Nix's functional model achieves a bit-for-bit reproducibility rate
+exceeding 90% and a rebuildability rate over 99% across a multi-year time-frame.
 
@@
 entire pipeline in approximately three minutes, whereas the declarative
-Nix-based runner takes around five minutes.^[To run the benchmark yourself,
-fork the paper’s repository, enable GitHub Actions, and push a single change to
-trigger the runs.] Nix’s deterministic dependency
-resolution guarantees bit-for-bit reproducibility at the expense of initial
-computation time, whereas the imperative system prioritizes immediacy over
-long-term stability.
+Nix-based runner takes around five minutes.^[To run the benchmark yourself, fork
+the paper’s repository, enable GitHub Actions, and push a single change to
+trigger the runs.] Nix’s deterministic dependency resolution guarantees
+bit-for-bit reproducibility at the expense of initial computation time, whereas
+the imperative system prioritizes immediacy over long-term stability.
 
@@
 Many tools exist to improve reproducibility, but [Nix]{.pkg} stands out because
-it deploys complete software environments closed under the “depends on” relation:
-it installs not only a package but all of its transitive dependencies. 
+it deploys complete software environments closed under the “depends on”
+relation: it installs not only a package but all of its transitive dependencies.
 This makes [Nix]{.pkg} uniquely suited for reproducible research.
 
-Yet solving such a complex problem makes [Nix]{.pkg} itself complex, as it relies on a 
-functional paradigm still likely unfamiliar to most researchers.
-[rix]{.pkg} lowers this barrier for [R]{.proglang} users by providing a familiar interface and workflow. 
-By declaratively specifying reproducible development shells with [Nix]{.pkg}, 
-researchers can accommodate diverse use cases—from running analytical pipelines 
-to developing interactive [shiny]{.pkg} applications or serving [plumber]{.pkg} APIs.
-
-Furthermore, [rixpress]{.pkg} extends this declarative model to entire analysis pipelines. 
-By leveraging [Nix]{.pkg} as a unified build automation system, [rixpress]{.pkg} enables polyglot 
-workflows where each step runs in a hermetically sealed environment. 
-This ensures both spatial and temporal reproducibility, efficient caching, and seamless 
-orchestration of multi-language analyses. Crucially, it provides native polyglot support 
-without the manual orchestration required by container-based approaches, 
-making complex workflows accessible without systems administration expertise.
-
-Adopting [Nix]{.pkg}-based workflows represents a conceptual shift from procedural 
-scripting to declarative specification. A practical adoption pathway begins with using 
-[rix]{.pkg} to generate reproducible environments for new projects while maintaining existing workflows. 
-This allows gradual learning of [Nix]{.pkg} concepts without disrupting active research. 
-As confidence grows, researchers can extend to [rixpress]{.pkg} for smaller analyses before tackling complex 
-pipelines. Institutions can support adoption by offering workshops, maintaining binary caches, 
-and designating local experts. While early projects require investment, later projects benefit from
-reusable environment definitions and cumulative expertise. Ultimately, adoption depends on 
-project scope, reproducibility needs, and institutional context—but the framework now exists 
+Yet solving such a complex problem makes [Nix]{.pkg} itself complex, as it
+relies on a functional paradigm still likely unfamiliar to most researchers.
+[rix]{.pkg} lowers this barrier for [R]{.proglang} users by providing a familiar
+interface and workflow. By declaratively specifying reproducible development
+shells with [Nix]{.pkg}, researchers can accommodate diverse use cases—from
+running analytical pipelines to developing interactive [shiny]{.pkg}
+applications or serving [plumber]{.pkg} APIs.
+
+Furthermore, [rixpress]{.pkg} extends this declarative model to entire analysis
+pipelines. By leveraging [Nix]{.pkg} as a unified build automation system,
+[rixpress]{.pkg} enables polyglot workflows where each step runs in a
+hermetically sealed environment. This ensures both spatial and temporal
+reproducibility, efficient caching, and seamless orchestration of multi-language
+analyses. Crucially, it provides native polyglot support without the manual
+orchestration required by container-based approaches, making complex workflows
+accessible without systems administration expertise.
+
+Adopting [Nix]{.pkg}-based workflows represents a conceptual shift from
+procedural scripting to declarative specification. A practical adoption pathway
+begins with using [rix]{.pkg} to generate reproducible environments for new
+projects while maintaining existing workflows. This allows gradual learning of
+[Nix]{.pkg} concepts without disrupting active research. As confidence grows,
+researchers can extend to [rixpress]{.pkg} for smaller analyses before tackling
+complex pipelines. Institutions can support adoption by offering workshops,
+maintaining binary caches, and designating local experts. While early projects
+require investment, later projects benefit from reusable environment definitions
+and cumulative expertise. Ultimately, adoption depends on project scope,
+reproducibility needs, and institutional context—but the framework now exists
 for those who require robust, long-term reproducibility.
 
-While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) represent significant progress,
-current limitations remain. Debugging complex [rixpress]{.pkg} pipelines is more challenging than
-troubleshooting [targets]{.pkg} workflows, especially when failures occur deep in the [Nix]{.pkg} build process. 
-Visualization capabilities, though useful, currently lack features such as automatic detection 
-of outdated derivations. Cross-platform reproducibility remains generally strong but is constrained on macOS by 
-proprietary framework dependencies. These challenges are active areas for development and community contribution. 
-Researchers adopting these tools should do so with realistic expectations: they provide strong guarantees of spatial 
-and temporal reproducibility and scalable polyglot workflows, but they require patience during the learning phase 
-and may occasionally present challenges that simpler tools avoid.
+While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) represent
+significant progress, current limitations remain. Debugging complex
+[rixpress]{.pkg} pipelines is more challenging than troubleshooting
+[targets]{.pkg} workflows, especially when failures occur deep in the
+[Nix]{.pkg} build process. Visualization capabilities, though useful, currently
+lack features such as automatic detection of outdated derivations.
+Cross-platform reproducibility remains generally strong but is constrained on
+macOS by proprietary framework dependencies. These challenges are active areas
+for development and community contribution. Researchers adopting these tools
+should do so with realistic expectations: they provide strong guarantees of
+spatial and temporal reproducibility and scalable polyglot workflows, but they
+require patience during the learning phase and may occasionally present
+challenges that simpler tools avoid.
 

### 2025-10-28 paper.qmd
r2u and portable binaries

@@
 
-While comprehensive dependency tracking could theoretically be implemented in
-any package manager, what distinguishes [Nix]{.pkg} (alongside its cousin
-[Guix]{.pkg}) is its *functional package management* paradigm. Traditional
-package managers require imperative, step-by-step installation instructions: *do
-this, then this, then that*. [Nix]{.pkg} inverts this model; users declare the
-desired end state (*I require an environment with these packages*), and
-[Nix]{.pkg} determines how to achieve it.
-
-This is possible because [Nix]{.pkg} packages are defined as *derivations*:
-declarative blueprints specifying all inputs (source code, build commands,
-dependencies) needed to build a package. When installing software, [Nix]{.pkg}
-evaluates expressions written in the [Nix]{.proglang} language to produce these
-derivations. The declarative nature ensures every build process is fully
-specified and independent of the execution environment.
+Comprehensive dependency tracking, however, is not unique to [Nix]{.pkg}. The
+[r2u]{.pkg} project by @r2u achieves something similar for Ubuntu: once the r2u
+repository is configured, every [R]{.pkg} package is installed as a binary with
+all its dependencies automatically resolved. In principle, any package manager
+could implement such tracking.
+
+What distinguishes [Nix]{.pkg} (and its sibling [Guix]{.pkg}, the only other
+functional package manager) is its *functional package* management model. This
+model makes full dependency specification intrinsic to the system, mandatory by
+design, and replaces the imperative, step-by-step nature of traditional package
+managers (*install this, then that*) with a declarative one: users describe the
+desired end state (*an environment with these packages*), and [Nix]{.pkg}
+determines how to realize it.
+
+A further advantage is that packages built by [Nix]{.pkg} act as universal
+binaries across any Linux system where [Nix]{.pkg} is installed. This “build
+once, run anywhere” property has long been solved by [Nix]{.pkg}, and is now
+being revisited by Posit through its initiative to provide portable CRAN
+binaries [@posit_portable_r_binaries].
+
+This universality stems from [Nix]{.pkg}’s use of derivations, declarative
+blueprints that specify all inputs (source code, build commands, dependencies)
+required to build a package. When software is installed, [Nix]{.pkg} evaluates
+expressions written in the [Nix]{.proglang} language to produce these
+derivations. Because every input is explicitly declared, builds are fully
+reproducible and independent of the surrounding environment.
 
 For example, the [rJava]{.pkg} package definition includes the correct
-`JAVA_HOME` configuration pointing to the [jdk]{.pkg} dependency. Outside
+`JAVA_HOME` configuration pointing to its [jdk]{.pkg} dependency. Outside
 [Nix]{.pkg}, users must manually set this variable if automatic detection fails.
-With [Nix]{.pkg}, the package maintainers handle this complexity once (by
-writing the correct [Nix]{.pkg} expressions that perform the installation
-purely), and all users benefit.
-
-[Nix]{.pkg} conceptualizes software builds as *pure functions* from functional
-programming. In mathematics, a pure function always produces the same output for
-a given input: if $f(x) = x^2$, then $f(2) = 4$ invariably holds. Analogously,
-[Nix]{.pkg} ensures that identical inputs (source code, dependencies, build
-instructions) produce identical outputs (built packages), regardless of when or
-where the build occurs.
+Within [Nix]{.pkg}, maintainers encode such details once, ensuring every user
+benefits from a consistent, reliable installation.
+
+At its core, [Nix]{.pkg} treats software builds as pure functions, a concept
+borrowed from functional programming. A pure function always produces the same
+output for a given input—for example, if $f(x) = x^2$, then $f(2) = 4$
+invariably holds. Likewise, [Nix]{.pkg} guarantees that identical inputs (source
+code, dependencies, build instructions) yield identical outputs (built
+packages), regardless of when or where the build occurs.
 

### 2025-10-28 paper.qmd
precision

@@
 A further advantage is that packages built by [Nix]{.pkg} act as universal
-binaries across any Linux system where [Nix]{.pkg} is installed. This “build
-once, run anywhere” property has long been solved by [Nix]{.pkg}, and is now
-being revisited by Posit through its initiative to provide portable CRAN
-binaries [@posit_portable_r_binaries].
+binaries across any Linux system (on the same architecture) where [Nix]{.pkg} is
+installed. This “build once, run anywhere” property has long been solved by
+[Nix]{.pkg}, and is now being revisited by Posit through its initiative to
+provide portable CRAN binaries [@posit_portable_r_binaries].
 

### 2025-10-29 paper.qmd
Update paper.qmd

@@
         postal-code: 2327
-      - Journal of Statistical Software
     orcid: 0000-0002-3211-3689

### 2025-10-31 paper.qmd
chore: add new section on nix containers
chore: add rix and rixpress citations; start docker-nix combo

@@
 using the [Nix]{.pkg} package manager [@dolstra2004nix], making it accessible to
-researchers through two [R]{.proglang} packages: [rix]{.pkg} and
-[rixpress]{.pkg}. Before introducing these packages, we survey existing tools
-and their limitations to contextualise our contributions.
+researchers through two [R]{.proglang} packages: [rix]{.pkg} [@R-rix] and
+[rixpress]{.pkg} [@R-rixpress]. Before introducing these packages, we survey
+existing tools and their limitations to contextualise our contributions.
 
@@
 [Nix]{.pkg} within [Docker]{.pkg} to benefit from the advantages of both tools.
+
+Provisioning [Nix]{.pkg} in container images lowers the burden to install it on
+the operating system of the host.
 

### 2025-10-31 paper.qmd
Update paper.qmd

--- 2025-10-29 paper.qmd
@@
 can be found in the [following
-repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/case-study)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/case-study],
+repository](https://github.com/b-rodrigues/rix_paper/tree/master/case-study)^[https://github.com/b-rodrigues/rix_paper/tree/master/case-study],
 with additional details provided in the Appendix of this manuscript. Although

### 2025-11-04 paper.qmd
chore: rephrase Nix and container technologies

--- 2025-10-31 paper.qmd
@@
 
-Provisioning [Nix]{.pkg} in container images lowers the burden to install it on
-the operating system of the host.
+By combining the strengths of Nix and container technologies, it is possible to
+enhance both build-time and runtime isolation. For example, installing [Nix].pkg
+directly into container images reduces the setup overhead by avoiding the need
+to install derivations into the host operating system's `/nix` directory.
 
